### 2024-07-28 paper/aamas/aamas25-submission.tex
tmp

%%%% For anonymized submission, use this
%\documentclass[sigconf,anonymous]{aamas} 
%%%% For camera-ready, use this
\documentclass[sigconf]{aamas}

\usepackage{balance} % for balancing columns on the final page

\setcopyright{ifaamas}
\acmConference[AAMAS '25]{Proc.\@ of the 24th International Conference
on Autonomous Agents and Multiagent Systems (AAMAS 2025)}{May 19 -- 23, 2025}
{Detroit, Michigan, USA}{A.~El~Fallah~Seghrouchni, Y.~Vorobeychik, S.~Das, A.~Nowe (eds.)}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{}
\acmPrice{}
\acmISBN{}

%%% Use this command to specify your EasyChair submission number.
%%% In anonymous mode, it will be printed on the first page.
\acmSubmissionID{<<EasyChair submission id>>}

\title{Title}

%%% Provide names, affiliations, and email addresses for all authors.

\author{Francesco Bertolotti}
\affiliation{
  \institution{Università degli Studi di Milano}
  \city{Milan}
  \country{Italy}}
\email{francesco.bertolotti@di.unimi.it}

\author{Walter Cazzola}
\affiliation{
  \institution{Università degli Studi di Milano}
  \city{Milan}
  \country{Italy}}
\email{walter.cazzola@di.unimi.it}

\author{Gianluca Aguzzi}
\affiliation{
  \institution{Università di Bologna}
  \city{Bologna}
  \country{Italy}}
\email{gianluca.aguzzi@unibo.it}

\author{Mirko Viroli}
\affiliation{
  \institution{Università di Bologna}
  \city{Bologna}
  \country{Italy}}
\email{mirko.viroli@unibo.it}

\keywords{Multi-Agent, Differetiable Environment, World Model}

\begin{document}

\pagestyle{fancy}
\fancyhead{}
\input{sects/abstract.tex}
\maketitle 
\input{sects/acks.tex}

\bibliographystyle{ACM-Reference-Format} 
\bibliography{ml+nn,foundations,strings,aosd,my_work,grammars,software_engineering,reflection,dsu,dsl,pl,splc,programming,roles,oolanguages,testing,./bibs/local}

\end{document}

### 2024-07-28 src/agents/MLPOFA.py

@@ -20,11 +20,13 @@ class MLPOFA(Agent, torch.nn.Module):
 
         # actor layers
         self.actor_obs2emb = utils.layer_init(torch.nn.Linear(observation_size*agents, hidden_size, device=device))
+        self.actor_pos     = torch.nn.Parameter(torch.normal(torch.zeros(agents, hidden_size, device=device), 0.1 * torch.ones(agents, hidden_size, device=device)))
         self.actor_hidden  = torch.nn.ModuleList([l for _ in range(layers) for l in [utils.layer_init(torch.nn.Linear(hidden_size, hidden_size, device=device)), torch.nn.Tanh()]])
         self.actor_hid2act = utils.layer_init(torch.nn.Linear(hidden_size, action_size, device=device), std=actor_init_gain)
 
         # critic layers
         self.critic_obs2emb = self.actor_obs2emb if shared else copy.deepcopy(self.actor_obs2emb)
+        self.critic_pos     = self.actor_pos     if shared else copy.deepcopy(self.actor_pos)
         self.critic_hidden  = self.actor_hidden  if shared else copy.deepcopy(self.actor_hidden)
         self.critic_hid2val = utils.layer_init(torch.nn.Linear(hidden_size, 1, bias=False, device=device), std=critic_init_gain)
 
@@ -35,6 +37,7 @@ class MLPOFA(Agent, torch.nn.Module):
                 utils.Lambda(lambda x:x.unsqueeze(1).repeat(1,agents,1)),
 
                 self.critic_obs2emb,
+                utils.Lambda(lambda x:x + self.critic_pos),
                 torch.nn.Tanh(),
                 *self.critic_hidden,
                 
@@ -46,6 +49,7 @@ class MLPOFA(Agent, torch.nn.Module):
                 utils.Lambda(lambda x:x.unsqueeze(1).repeat(1,agents,1)),
 
                 self.actor_obs2emb,
+                utils.Lambda(lambda x:x + self.actor_pos),
                 torch.nn.Tanh(),
                 *self.actor_hidden,
                 

### 2024-07-28 src/algorithms/shac/Trajectory.py

@@ -4,7 +4,6 @@ from optimizers import add_adam_command
 from schedulers import add_constant_command, add_cosine_command
 import click, torch
 
-from utils import hash_tensors
 
 class Trajectory:
 

### 2024-07-28 src/callbacks/proxied/ProxyLog.py

@@ -1,7 +1,7 @@
 from callbacks.proxied import proxied
 from callbacks import Callback
 from loggers import File
-import click
+import torch, click
 
 class ProxyLog(Callback):
     def __init__(self, trainer, path="train.log"):
@@ -10,12 +10,15 @@ class ProxyLog(Callback):
         self.path    =    path
 
     def end_epoch_proxy(self, data):
+        nonempty = data["self"].mask.sum().item()
 
         self.logger.log({
             "update"   : data["self"].updates ,
+            "filled"   : nonempty,
             "epoch"    : data["epoch"]        ,
             "step"     : data["step"]         ,
             "loss"     : data["loss"].item()  ,
+            "dist"     : torch.nn.functional.avg_pool1d(data["self"].rewards[data["self"].mask].unsqueeze(0), kernel_size=nonempty//10, stride=nonempty//10).tolist()[0] ,
             "accuracy" : data["accuracy"]
         })
 

### 2024-12-27 paper/icml25/first/sects/background.tex
intro,back, and shac++

\section{Background}

To provide a context for our work, we first introduce the reinforcement learning (RL) framework and the SHAC algorithm. We model each task as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, where $\mathcal{S}$ is the state space, $\mathcal{A}$ is the action space, $F:\mathcal{S}\times\mathcal{A}\rightarrow\mathcal{S}$ is the transition function, $R:\mathcal{S}\times\mathcal{A}\times\mathcal{S}\rightarrow\mathbb{R}$ is the reward function, $\gamma\in\mathcal{R}$ is the discount factor, and $T\in\mathbb{N}$ is the time horizon.

The goal is to find a policy $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maximizes the expected return $J(\pi) = \mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r(s_t, a_t, s_{t+1})\right]$, where $\tau = (s_0, a_0, s_1, a_1, \ldots)$ is a trajectory generated by the policy $\pi$.

In particular a SHAC policy is trained according the loss $\mathcal{L}_\theta^{\pi}$: 
$$ \frac{1}{NH}\sum_{i=1}^N \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^i, a_t^i) + \gamma^H V(s^i_{t_0+T})$$
Where $N$ is the number of sampled trajectories, $H << T$ is a time horizon, $s_t^i$ is the state at time $t$ of trajectory $i$, $a_t^i$ is the action at time $t$ of trajectory $i$, and $V(s):\mathcal{S}\rightarrow\mathbb{R}$ is the value function that estimates the expected return starting from state $s$. Further, SHAC assumes reward and transition functions to be differentiable, $\hat{F},\hat{R} \in \mathcal{C}^1$.

The value function is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98}, $\hat{V}$. In practice, the value function is trained to minimize the loss $\mathcal{L}_\theta^{V}$.

$$ \frac{1}{N}\sum_i^N\left|\left| V(s^i) - \hat{V}(s^i) \right|\right|^2 $$
$$ \hat{V}(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
$$ G_t^k = \left(\sum_{l=0}^{k-1}\gamma^l r_{t+l}\right) + \gamma^k V(s_{t+k})$$

$G^t_k$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$. The estimates $\hat{V}$ is computed by discounting more longer (higher $k$) discouunted sums, $G^k_t$.

### 2024-12-27 paper/icml25/first/sects/introduction.tex
intro,back, and shac++

\section{Introduction}

Learnign control policies has important applications in robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, swarm intelligence \cite{Tang21} and many other fields. Recently, a reinforcement learning algorithm has gained popularity, SHAC\cite{Xu22}. SHAC is a simple and yet effective algorithm that exploits differentiable simulators to learn control policies effectively.

Despite its effectiveness, the main limitations of SHAC are two fold. Firstly, it requires a differentiable simulator and differentiable reward function to work properly. Both of which may be not feasible to obtain in practice, especially with inherently sparse rewards. Secondly, even when a differentiable simulator is available, SHAC has been shown to be sensitive to complex dynamics such as object collisions~\cite{Georgiev24}.

In this work, we aim to address both these limitation by approximating the gradients of the differentiable simulator. In particular, we will replace both the simulator and the reward function with a neural network trained in tandem with the policy network and value network. Overall, this will provide an algorithm that is more robust to complex dynamics, sparse rewards, and can work in a wider range of environments. On the other hand, however, this will also introduce additional complexity and computational cost tied to training additional networks.

We test our framework---dubbed \fname{}---against PPO\cite{Schulman17} and SHAC\cite{Xu22}, the former a state-of-the-art reinforcement learning algorithm for non-differentiable environments, and the latter, a state-of-the-art reinforcement learning algorithm for differentiable environments. We compare these algorithm on a set of environments with different numbers of agents. This allows us to evaluate the performance of these algorithm as the search space grows with the number of agents.

With this work, we aim to answer the following research questions:
\begin{compactitem}
    \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
    \item \RQ{2} How does our algorithm compare to PPO and SHAC in differentiable and non-differentiable environments?
    \item \RQ{3} How does the performance of these algorithms change as the search space increases?
\end{compactitem}

We summarize our contributions as follows:
\begin{compactitem}
    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC.
    \item We evaluate our algorithm against PPO and SHAC on a set of environments with increasing numbers of agents.
    \item We show that our algorithm achieve SHAC-like performance in differentiable environments and outperform PPO in non-differentiable environments.
\end{compactitem}

### 2024-12-27 paper/icml25/first/sects/shacpp.tex
intro,back, and shac++

\section{\fname{}}

As mentioned previously, we aim to address the limitations of SHAC by approximating the gradients of a differentiable simulator with neural networks. In particular, we aim to train a neural network to approximate the reward function by minimizing: 
$$ \mathbb{E}_{s,a}\left[\left|\left| R_{\theta}(s, a) - \hat{R}(s, a) \right|\right|^2\right] $$

In practice, this is achieved by caching the state-action-reward triplets obtained from the environment and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{R}$:
$$ \frac{1}{\left|B\right|}\sum_{s,a \in B}\left|\left| R_\theta(s, a) - \hat{R}(s, a) \right|\right|^2 $$
Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.

Similarly, we aim to train a neural network to approximate the transition function by minimizing. However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model\cite{Ma24}. This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $N$ actions from $\mathcal{A}^N$, predicts the next $N$ observations from $\mathcal{S}^N$. In practice, we aim to minimize:

$$ \mathbb{E}_{\tau}\left[\sum_{i=1}^N \left|\left| F_{\theta}(s^0, a^0, \dots, a^N)^i - \hat{F}(s^i, a^i) \right|\right|^2\right] $$

Where $\tau$ corresponds to a trajectory $(s^0, a^0, s^1, a^1, \dots, s^N, a^N)$. In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss. That is, we minimize the loss $\mathcal{L}_\theta^{F}$:
$$ \frac{1}{\left|B\right|}\sum_{\tau \in B}\sum_{i=1}^N\left|\left| F_{\theta}(s^0, a^0, \dots, a^N)^i - \hat{F}(s^i, a^i) \right|\right|^2 $$
Where $B$ represents a batch of trajectories sampled uniformly from the cache.

While SHAC is aims to minimize $\mathcal{L}\theta^\pi$ and $\mathcal{L}_\theta^V$, we also aim to minimize also $\mathcal{L}_\theta^R$ and $\mathcal{L}_\theta^F$. Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space. That is, suppose the action space is $(a,b)$ (with $a<b$), we minimize the loss $\mathcal{L'}_\theta^\pi:$

$$\mathcal{L}_\theta^\pi + \alpha \frac{1}{NH}\sum_{i=0}^N\sum_{t=t_0}^{t_0+H-1} d(a_t^i,(a,b))$$
$$d(x,(a,b)) = \begin{cases}(a-x)^2 & \text{if } x < a \\ (x-b)^2 & \text{if } x > b \\ 0 & \text{otherwise} \end{cases}$$

Where $\alpha$ is a hyper-parameter that controls the strength of the penalty. In our experiments, we set $\alpha=1$. In practice, actions are usually either clipped to the allowed action space or, an activation function is such as hyperbolic tangent is used to ensure that the actions are within the action space. This is acceptable for PPO and SHAC. However, in our scenario, this penalty is crucial. To understand why, consider that we are training the policy to optimize two neural networks that approximate the environment. Further, these networks are not trained on data generated from action inside the action space. As a result, it is very likely that these networks achieve their maximum for actions outside the allowed range. Leading to higher than normal pre-activation actions that disrupt the training process.

To summarize, our algorithm, aims to minimize the 4 presented losses, $\mathcal{L'}_\theta^{\pi}$, $\mathcal{L}_\theta^V$, $\mathcal{L}_\theta^R$, and $\mathcal{L}_\theta^F$. A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.

However, consider that, to ensure fast convergence of the reward and transition networks, we employ a cache to store seen trajectories. To ensure that the cache is balanced, and not filled completely with $0$-reward trajectories, we divide the cache $K$ equally sized bins. Each bin is filled with trajectories that have a similar reward. For the reward network, this corresponds to the immediate reward. For the transition network, this corresponds to the sum of reward accumulated during the trajectory.

And, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.

The full implementation is openly available at 
\begin{center}
\url{https://redacted/for/anonymity}
\end{center}

\begin{algorithm}[t]
    \begin{algorithmic}[1]
    \STATE Initialize $\pi_\theta,V_\theta,R_\theta,W_\theta$ networks
    \STATE Initialize learning rates $\eta_\pi, \eta_R, \eta_V, \eta_W$
    \STATE Initialize environment $\mathcal{E}$
    \STATE Initialize $\gamma, \alpha$
    \FOR{episode in $1, \ldots, N_{\text{episodes}}$}

        \STATE $s,a,r,v \sim \mathcal{E}(\pi)$
        \STATE $\hat{s} = W(s[0], a)$
        \STATE $\hat{r} = R(\hat{s})$
        \STATE $\hat{v} = V(\hat{s})$

        \STATE $\pi_\theta \gets \pi_\theta - \eta_\pi \nabla \mathcal{L'}^\pi_\theta(\hat{r},\hat{v},\gamma,\alpha)$
        \STATE $V_\theta \gets V_\theta - \eta_V \nabla \mathcal{L}^V_\theta(\hat{v},s,v)$
        \STATE $R_\theta \gets R_\theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,r)$
        \STATE $W_\theta \gets W_\theta - \eta_W \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$

    \ENDFOR

    \end{algorithmic}
    \label{alg:shacpp}
    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode}
\end{algorithm}

### 2025-01-02 paper/icml25/first/sects/introduction.tex
minors on intro

@@
 
-Learnign control policies has important applications in robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, swarm intelligence \cite{Tang21} and many other fields.
-Recently, a reinforcement learning algorithm has gained popularity, SHAC\cite{Xu22}.
+Learnign control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields.
+Recently, a reinforcement learning algorithm has gained popularity, SHAC~\cite{Xu22}.
 SHAC is a simple and yet effective algorithm that exploits differentiable simulators to learn control policies effectively.
 
-Despite its effectiveness, the main limitations of SHAC are two fold.
+Despite its effectiveness, the main limitations of SHAC are twofold.
 Firstly, it requires a differentiable simulator and differentiable reward function to work properly.
@@
 
-In this work, we aim to address both these limitation by approximating the gradients of the differentiable simulator.
+In this work, we aim to address both these limitations by approximating the gradients of the differentiable simulator.
 In particular, we will replace both the simulator and the reward function with a neural network trained in tandem with the policy network and value network.
@@
 
-We test our framework---dubbed \fname{}---against PPO\cite{Schulman17} and SHAC\cite{Xu22}, the former a state-of-the-art reinforcement learning algorithm for non-differentiable environments, and the latter, a state-of-the-art reinforcement learning algorithm for differentiable environments.
-We compare these algorithm on a set of environments with different numbers of agents.
-This allows us to evaluate the performance of these algorithm as the search space grows with the number of agents.
+We evaluate our framework---dubbed \fname{}---against PPO~\cite{Schulman17}, a state-of-the-art algorithm for non-differentiable environments, and SHAC~\cite{Xu22}, designed for differentiable environments.
+We compare these algorithms on environments with varying numbers of agents, allowing us to assess their performance as the search space grows.
 

### 2025-01-02 paper/icml25/first/sects/introduction.tex
minor on introduction

@@
 Learnign control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields. 
-Recently, a reinforcement learning algorithm has gained popularity, SHAC~\cite{Xu22}.
-SHAC is a simple and yet effective algorithm that exploits differentiable simulators to learn control policies effectively.
+These scenarios are varied, with can consider both single and multi-agent systems, and can have different dynamics and reward functions.
+Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings.
+Moreover, reinforcement learning algorithms tend to be sample inefficient due to the high variance of the gradient estimates and the exploration required.
+A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.
 
@@
 Both of which may be not feasible to obtain in practice, especially with inherently sparse rewards.
-Secondly, even when a differentiable simulator is available, SHAC has been shown to be sensitive to complex dynamics such as object collisions~\cite{Georgiev24}.
+Secondly, even when a differentiable simulator is available, SHAC has been shown to be sensitive to complex dynamics such as object collisions~\cite{Georgiev24} or multi-agent interactions, which is currently an unsplore area in the literature.
 
@@
 In particular, we will replace both the simulator and the reward function with a neural network trained in tandem with the policy network and value network.
-Overall, this will provide an algorithm that is more robust to complex dynamics, sparse rewards, and can work in a wider range of environments.
+Overall, this will provide an algorithm that is more robust to complex dynamics, sparse rewards, and can work in a wider range of environments, from sigle-agent to multi-agent scenarios.
 On the other hand, however, this will also introduce additional complexity and computational cost tied to training additional networks.
 
-We evaluate our framework---dubbed \fname{}---against PPO~\cite{Schulman17}, a state-of-the-art algorithm for non-differentiable environments, and SHAC~\cite{Xu22}, designed for differentiable environments.
-We compare these algorithms on environments with varying numbers of agents, allowing us to assess their performance as the search space grows.
+We evaluate our framework---dubbed \fname{}---against PPO~\cite{Schulman17}, a state-of-the-art algorithm for non-differentiable environments (used also in multi-agent scenarios with the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, designed for differentiable environments.
+We compare these algorithms on VMAS---a state-of-the-art simulator for complex physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}---across cooperative tasks with varying numbers of agents, enabling a systematic assessment of how their performance scales as the search space grows.
 
@@
     \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
-    \item \RQ{2} How does our algorithm compare to PPO and SHAC in differentiable and non-differentiable environments?
+    \item \RQ{2} How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?
     \item \RQ{3} How does the performance of these algorithms change as the search space increases?
@@
     \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC.
+    \item We assess the SHAC algorithm in multi-agent scenarios, where it has not been previously evaluated.
     \item We evaluate our algorithm against PPO and SHAC on a set of environments with increasing numbers of agents.

### 2025-01-07 paper/icml25/first/figs/shac.tex
minors + shac/shacpp figs

\documentclass[crop,dvipsnames,tikz]{standalone}

\usetikzlibrary{calc}

\begin{document}
\begin{tikzpicture}

    \def\nodes{3}
    \def\nnodes{4}

    % Loop to place numbers
    \foreach \i in {0,...,\nodes} {
        \pgfmathsetmacro{\sx}{\i*2}
        \pgfmathsetmacro{\ax}{\i*2+1}

        % Draw number
        \node[circle, fill=Orange,draw=Orange] (s\i) at (\sx, 0) {\color{white}$s_\i$};
        \node[circle, fill=RoyalBlue,draw=RoyalBlue] (a\i) at (\ax, -1.4) {\color{white}$a_\i$};
        \node[circle, fill=DarkOrchid,draw=DarkOrchid] (r\i) at (\ax, +1.4) {\color{white}$r_\i$};

    }
    \pgfmathsetmacro{\sx}{\nnodes*2}
    \node[circle, draw=Orange, fill=Orange] (s\nnodes) at (\sx, 0) {\color{white}$s_\nnodes$};

    % reward function
    \foreach \i in {0, ..., \nodes} {
        \pgfmathtruncatemacro{\j}{\i+1}
        \draw[-latex,line width=1,color=DarkOrchid] (a\i) -- (r\i);
        \draw[-latex,line width=1,color=DarkOrchid] (s\i) -- ($(s\i) + (.6,0)$) to[out=0,in=270]   ($(r\i) - (0,.8)$) -- (r\i);
        \draw[-latex,line width=1,color=DarkOrchid] (s\j) -- ($(s\j) - (.6,0)$) to[out=180,in=270] ($(r\i) - (0,.8)$) -- (r\i);
        \node[circle, fill=DarkOrchid,draw=white, inner sep=1.5pt] (tmp) at ($(r\i) - (0,.8)$) {};
        \node at ($(tmp) - (.25,0)$) {\color{DarkOrchid}$\hat{R}$};
    }

    % policy function
    \foreach \i in {0, ..., \nodes} {
        \draw[-latex,line width=1,color=RoyalBlue] (s\i) -- ($(s\i) - (0,1.2)$) to[out=270,in=180] ($(a\i) - (.9,0)$) -- (a\i);
        \node[circle, fill=RoyalBlue,draw=white, inner sep=1.5pt] (tmp) at ($(s\i) - (0,1.2)$) {};
        \node at ($(tmp) - (.2,0)$) {\color{RoyalBlue}$\pi$};
    }

    % value function
    \foreach \i in {1, ..., \nnodes} {
        \node[circle, fill=BlueGreen,draw=BlueGreen] (v\i) at ($(s\i) + (0,1.4)$) {\color{white}$v_\i$};
        \draw[-latex,line width=1, color=BlueGreen] (s\i) --  (v\i);
        \node[circle, fill=BlueGreen,draw=white, inner sep=1.5pt] (tmp) at ($(s\i) + (0,.6)$) {};
        \node at ($(tmp) - (.25,0)$) {\color{BlueGreen}$V_\theta$};
    }

    % transition function
    \foreach \i in {0, ..., \nodes} {
        \pgfmathtruncatemacro{\j}{\i+1}
        \draw[-latex,line width=1, color=orange] (s\i) -- (s\j) ;
        \draw[-latex,line width=1, color=orange] (a\i) -- ($(a\i) + (0,1.1)$) to[out=90,in=180] ($(s\j) - (.9,0)$) -- (s\j);
        \node[circle, fill=orange,draw=white, inner sep=1.5pt] (tmp) at ($(s\j) - (.8,0)$) {};
        \node at ($(tmp) - (0,.3)$) {\color{orange}$\hat{F}$};
    }

    % loss policy function
    \draw[draw=white,fill=LimeGreen,rounded corners=.5ex] ($(r0) + (-.5, 1)$) rectangle ($(v4) + (.5,1.2)$); 
    \node at ($(r0) + (-.75,1.05)$) {\color{LimeGreen}$\mathcal{L}^\pi_\theta$};
    \foreach \i in {0, ..., \nodes} {
        \pgfmathtruncatemacro{\j}{\i+1}
        \draw[-latex,line width=1, color=LimeGreen] (r\i) -- ($(r\i) + (0,1)$);
        \draw[-latex,line width=1, color=LimeGreen] (v\j) -- ($(v\j) + (0,1)$);
    }

    \node[circle, fill=gray] (l) at ($(v\nnodes) + (1.5,1.1)$) {\color{white}$l$};
    \draw[-latex,line width=1, color=gray] ($(v\nnodes) + (.5,1.1)$) -- (l);

\end{tikzpicture}
\end{document}

### 2025-01-07 paper/icml25/first/figs/shacpp.tex
minors + shac/shacpp figs

--- 2025-01-07 paper/icml25/first/figs/shac.tex
@@
         \node[circle, fill=Orange,draw=Orange] (s\i) at (\sx, 0) {\color{white}$s_\i$};
+        \node[circle, fill=Dandelion,draw=Dandelion, inner sep=2.2pt] (sa\i) at (\sx, 2.2) {\color{white}$s'_\i$};
         \node[circle, fill=RoyalBlue,draw=RoyalBlue] (a\i) at (\ax, -1.4) {\color{white}$a_\i$};
-        \node[circle, fill=DarkOrchid,draw=DarkOrchid] (r\i) at (\ax, +1.4) {\color{white}$r_\i$};
+        \node[circle, fill=DarkOrchid,draw=DarkOrchid, inner sep=2.2pt] (r\i) at (\ax, +3.7) {\color{white}$r'_\i$};
+        \node[circle, fill=DarkOrchid,draw=white,inner sep=1.5pt] (R\i) at ($(r\i) - (0,.8)$) {};
+        \draw[-latex,color=DarkOrchid,line width=1] (R\i) -- (r\i);
+        \node at ($(R\i) - (.3,0)$) {\color{DarkOrchid}$R_\theta$};
 
@@
     \node[circle, draw=Orange, fill=Orange] (s\nnodes) at (\sx, 0) {\color{white}$s_\nnodes$};
+    \node[circle, draw=Dandelion, fill=Dandelion, inner sep=2.2pt] (sa\nnodes) at (\sx, 2.2) {\color{white}$s'_\nnodes$};
 
@@
         \pgfmathtruncatemacro{\j}{\i+1}
-        \draw[-latex,line width=1,color=DarkOrchid] (a\i) -- (r\i);
-        \draw[-latex,line width=1,color=DarkOrchid] (s\i) -- ($(s\i) + (.6,0)$) to[out=0,in=270]   ($(r\i) - (0,.8)$) -- (r\i);
-        \draw[-latex,line width=1,color=DarkOrchid] (s\j) -- ($(s\j) - (.6,0)$) to[out=180,in=270] ($(r\i) - (0,.8)$) -- (r\i);
-        \node[circle, fill=DarkOrchid,draw=white, inner sep=1.5pt] (tmp) at ($(r\i) - (0,.8)$) {};
-        \node at ($(tmp) - (.25,0)$) {\color{DarkOrchid}$\hat{R}$};
+        \draw[-,line width=1,color=DarkOrchid] (a\i) -- (R\i);
+        \draw[-,line width=1,color=DarkOrchid] (sa\i) -- ($(sa\i) + (.6,0)$) to[out=0,in=270]   ($(R\i) - (0,.2)$) -- (R\i);
+        \draw[-,line width=1,color=DarkOrchid] (sa\j) -- ($(sa\j) - (.6,0)$) to[out=180,in=270] ($(R\i) - (0,.2)$) -- (R\i);
     }
@@
     \foreach \i in {1, ..., \nnodes} {
-        \node[circle, fill=BlueGreen,draw=BlueGreen] (v\i) at ($(s\i) + (0,1.4)$) {\color{white}$v_\i$};
-        \draw[-latex,line width=1, color=BlueGreen] (s\i) --  (v\i);
-        \node[circle, fill=BlueGreen,draw=white, inner sep=1.5pt] (tmp) at ($(s\i) + (0,.6)$) {};
+        \node[circle, fill=BlueGreen,draw=BlueGreen,inner sep=2.2pt] (v\i) at ($(sa\i) + (0,1.5)$) {\color{white}$v'_\i$};
+        \draw[-latex,line width=1, color=BlueGreen] (sa\i) --  (v\i);
+        \node[circle, fill=BlueGreen,draw=white, inner sep=1.5pt] (tmp) at ($(sa\i) + (0,.7)$) {};
         \node at ($(tmp) - (.25,0)$) {\color{BlueGreen}$V_\theta$};
+    }
+
+    % world model
+    \draw[draw=white,fill=Dandelion,rounded corners=.5ex] ($(s0) + (-.5, 1)$) rectangle ($(s\nnodes) + (.5,1.2)$); 
+    \node at ($(s0) + (-.75,1.05)$) {\color{Dandelion}$F_\theta$};
+    \draw[-latex,line width=1, color=Dandelion] (s0) -- ($(s0) + (0,1)$);
+    \foreach \i in {0, ..., \nodes} {
+        \draw[-latex,line width=1, color=Dandelion] (a\i) -- ($(a\i) + (0,2.4)$);
+    }
+    \foreach \i in {0, ..., \nnodes} {
+        \draw[-latex,line width=1, color=Dandelion] ($(sa\i) - (0,1)$) -- (sa\i);
     }
@@
         \pgfmathtruncatemacro{\j}{\i+1}
-        \draw[-latex,line width=1, color=orange] (s\i) -- (s\j) ;
-        \draw[-latex,line width=1, color=orange] (a\i) -- ($(a\i) + (0,1.1)$) to[out=90,in=180] ($(s\j) - (.9,0)$) -- (s\j);
+        \draw[-latex,line width=1, color=orange, dashed] (s\i) -- (s\j) ;
+        \draw[-latex,line width=1, color=orange, dashed] (a\i) -- ($(a\i) + (0,1.1)$) to[out=90,in=180] ($(s\j) - (.9,0)$) -- (s\j);
         \node[circle, fill=orange,draw=white, inner sep=1.5pt] (tmp) at ($(s\j) - (.8,0)$) {};

### 2025-01-07 paper/icml25/first/sects/background.tex
minor on background
minors + shac/shacpp figs

@@
 
-To provide a context for our work, we first introduce the reinforcement learning (RL) framework and the SHAC algorithm.
-We model each task as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, where $\mathcal{S}$ is the state space, $\mathcal{A}$ is the action space, $F:\mathcal{S}\times\mathcal{A}\rightarrow\mathcal{S}$ is the transition function, $R:\mathcal{S}\times\mathcal{A}\times\mathcal{S}\rightarrow\mathbb{R}$ is the reward function, $\gamma\in\mathcal{R}$ is the discount factor, and $T\in\mathbb{N}$ is the time horizon.
+\subsection{Markov Decision Processes}
+To provide context for our work, we first introduce the reinforcement learning (RL) framework.
+Each task is modeled as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, where $\mathcal{S}$ represents the state space, $\mathcal{A}$ denotes the action space, $\hat{F}:\mathcal{S} \times \mathcal{A} \rightarrow \mathcal{S}$ is the transition function, $\hat{R}:\mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, $\gamma \in [0, 1]$ is the discount factor, and $T \in \mathbb{N}$ is the time horizon.
 
-The goal is to find a policy $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maximizes the expected return $J(\pi) = \mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r(s_t, a_t, s_{t+1})\right]$, where $\tau = (s_0, a_0, s_1, a_1, \ldots)$ is a trajectory generated by the policy $\pi$.
+Further, let $\mathcal{T}$ denote the space of trajectories, defined as $\mathcal{T} = (\mathcal{S} \times \mathcal{A} \times \mathcal{S} \times \mathbb{R})^*$, where each trajectory consists of a sequence of starting states, actions, subsequent states, and rewards.
+We will use the symbols $((s_0, a_0, s_1, r_0), (s_1, a_1, s_2, r_1), \ldots)$ to denote a trajectory in $\mathcal{T}$.
 
-In particular a SHAC policy is trained according the loss $\mathcal{L}_\theta^{\pi}$: 
-$$ \frac{1}{NH}\sum_{i=1}^N \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^i, a_t^i) + \gamma^H V(s^i_{t_0+T})$$
-Where $N$ is the number of sampled trajectories, $H << T$ is a time horizon, $s_t^i$ is the state at time $t$ of trajectory $i$, $a_t^i$ is the action at time $t$ of trajectory $i$, and $V(s):\mathcal{S}\rightarrow\mathbb{R}$ is the value function that estimates the expected return starting from state $s$.
-Further, SHAC assumes reward and transition functions to be differentiable, $\hat{F},\hat{R} \in \mathcal{C}^1$.
+The goal of the MDP is to find a policy $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maximizes the expected return 
+$$J(\pi) = \mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r(s_t, a_t, s_{t+1})\right]$$
+Where $\tau \in \mathcal{T}$ is a trajectory generated by the policy $\pi$ with the transition function $\hat{F}$.
 
-The value function is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98}, $\hat{V}$.
-In practice, the value function is trained to minimize the loss $\mathcal{L}_\theta^{V}$.
+\subsection{SHAC}
+
+To solve an MDP $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, SHAC trains a policy $\pi$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+$$\mathcal{L}_\theta^{\pi} = \frac{1}{NH}\sum_{i=1}^N \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^i, a_t^i) + \gamma^H V(s^i_{t_0+T}),$$  
+where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^i$ is the state at time $t$ for trajectory $i$, $a_t^i$ is the action taken at time $t$ for trajectory $i$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$.
+Additionally, SHAC assumes that the policy, reward, and transition functions are differentiable, i.e., $\pi, \hat{F}, \hat{R} \in \mathcal{C}^1$.
+
+The value function is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $\hat{V}$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{V}:\mathcal{T}^N\rightarrow\mathbb{R}$.
 
@@
 $G^t_k$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$.
-The estimates $\hat{V}$ is computed by discounting more longer (higher $k$) discouunted sums, $G^k_t$.
+The estimates $\hat{V}$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.
+
+Finally, Figure~\ref{fig:shac} presents an unrolled view of the SHAC algorithm.

### 2025-01-07 paper/icml25/first/sects/introduction.tex
minors + shac/shacpp figs

@@
 Learnign control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields.
-These scenarios are varied, with can consider both single and multi-agent systems, and can have different dynamics and reward functions.
+These scenarios presents a variaty of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions.
 Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings.
-Moreover, reinforcement learning algorithms tend to be sample inefficient due to the high variance of the gradient estimates and the exploration required.
+Moreover, reinforcement learning algorithms tend to be sample inefficient due to the high variance of the gradient estimates and the exploration required \texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}.
 A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.
 
-Despite its effectiveness, the main limitations of SHAC are twofold.
-Firstly, it requires a differentiable simulator and differentiable reward function to work properly.
-Both of which may be not feasible to obtain in practice, especially with inherently sparse rewards.
-Secondly, even when a differentiable simulator is available, SHAC has been shown to be sensitive to complex dynamics such as object collisions~\cite{Georgiev24} or multi-agent interactions, which is currently an unsplore area in the literature.
+Despite its effectiveness, SHAC has two main limitations.
+First, it requires both a differentiable simulator and a differentiable reward function to operate correctly.
+Obtaining these can be impractical, particularly in scenarios with inherently sparse rewards.
+Second, even when a differentiable simulator is available, SHAC has been found to be sensitive to complex dynamics, such as object collisions~\cite{Georgiev24} or interactions in multi-agent systems.
 
-In this work, we aim to address both these limitations by approximating the gradients of the differentiable simulator.
-In particular, we will replace both the simulator and the reward function with a neural network trained in tandem with the policy network and value network.
-Overall, this will provide an algorithm that is more robust to complex dynamics, sparse rewards, and can work in a wider range of environments, from sigle-agent to multi-agent scenarios.
-On the other hand, however, this will also introduce additional complexity and computational cost tied to training additional networks.
+In this work, we aim to address both limitations by approximating the gradients of the differentiable simulator.
+Specifically, we propose replacing both the simulator and the reward function with two neural networks trained alongside the policy and value networks.
+This approach is designed to create an algorithm that is more robust to complex dynamics and sparse rewards, enabling it to function effectively across a broader range of environments, including both single-agent and multi-agent scenarios.
+However, this method also introduces additional complexity and computational overhead due to the need for training these additional networks.
 
-We evaluate our framework---dubbed \fname{}---against PPO~\cite{Schulman17}, a state-of-the-art algorithm for non-differentiable environments (used also in multi-agent scenarios with the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, designed for differentiable environments.
-We compare these algorithms on VMAS---a state-of-the-art simulator for complex physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}---across cooperative tasks with varying numbers of agents, enabling a systematic assessment of how their performance scales as the search space grows.
+We evaluate our framework, referred to as \fname{}, against two baseline algorithms: PPO~\cite{Schulman17}, a state-of-the-art approach for non-differentiable environments (also widely used in multi-agent scenarios through the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, which is tailored for differentiable environments.
+The evaluation is conducted using VMAS, a state-of-the-art simulator designed for complex, physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}.
+We compare these algorithms across cooperative tasks involving varying numbers of agents, providing a systematic analysis of how their performance scales with the increasing size of the search space.
 

### 2025-01-07 paper/icml25/first/sects/shacpp.tex
minors + shac/shacpp figs

@@
+\begin{figure*}[t]
+    \centering
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figs/shac.pdf}
+        \caption{SHAC}
+        \label{fig:shac}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \includegraphics[width=\textwidth]{figs/shacpp.pdf}
+        \caption{\fname{}}
+        \label{fig:shacpp}
+    \end{subfigure}
+    \caption{Comparison between SHAC and \fname{}.
+Nodes with labels (such as $r_i$ or $r'_i$ represents values).
+Nodes without labels (such as $\pi$ or $R$) represents functions.
+Dashed lines are used to represent non differentiability}\label{fig:shac-shacpp}
+\end{figure*}
+
 \section{\fname{}}
@@
 In practice, this is achieved by caching the state-action-reward triplets obtained from the environment and then training the neural network to minimize the MSE loss.
-That is, we minimize the loss $\mathcal{L}_\theta^{R}$:
-$$ \frac{1}{\left|B\right|}\sum_{s,a \in B}\left|\left| R_\theta(s, a) - \hat{R}(s, a) \right|\right|^2 $$
+That is, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+$$ \frac{1}{\left|B\right|}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - \underbrace{r}_{\hat{R}(s, a)} \right|\right|^2 $$
 Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.
 
-Similarly, we aim to train a neural network to approximate the transition function by minimizing.
-However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model\cite{Ma24}.
-This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $N$ actions from $\mathcal{A}^N$, predicts the next $N$ observations from $\mathcal{S}^N$.
+Similarly, we aim to train a neural network to approximate the transition function.
+However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model~\cite{Ma24}.
+This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $N$ actions from $\mathcal{A}^H$, predicts the next $N$ observations from $\mathcal{S}^H$.
 In practice, we aim to minimize:
 
-$$ \mathbb{E}_{\tau}\left[\sum_{i=1}^N \left|\left| F_{\theta}(s^0, a^0, \dots, a^N)^i - \hat{F}(s^i, a^i) \right|\right|^2\right] $$
+$$ \mathbb{E}_{\tau}\left[\sum_{i=1}^H \left|\left| F_{\theta}(s_0, a_0, \dots, a_N)_i - \hat{F}(s_i, a_i) \right|\right|^2\right] $$
 
-Where $\tau$ corresponds to a trajectory $(s^0, a^0, s^1, a^1, \dots, s^N, a^N)$.
+Where $\tau\in\mathcal{T}$.
 In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss.
-That is, we minimize the loss $\mathcal{L}_\theta^{F}$:
-$$ \frac{1}{\left|B\right|}\sum_{\tau \in B}\sum_{i=1}^N\left|\left| F_{\theta}(s^0, a^0, \dots, a^N)^i - \hat{F}(s^i, a^i) \right|\right|^2 $$
+That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+$$ \frac{1}{NH}\sum_{\tau \in B}\sum_{i=1}^H\left|\left| F_{\theta}(s_0, a_0, \dots, a_H)_i - \underbrace{s_{i}}_{\hat{F}(s_{i-1}, a_{i-1})} \right|\right|^2 $$
 Where $B$ represents a batch of trajectories sampled uniformly from the cache.
 
-While SHAC is aims to minimize $\mathcal{L}\theta^\pi$ and $\mathcal{L}_\theta^V$, we also aim to minimize also $\mathcal{L}_\theta^R$ and $\mathcal{L}_\theta^F$.
 Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space.
@@
 In our experiments, we set $\alpha=1$.
-In practice, actions are usually either clipped to the allowed action space or, an activation function is such as hyperbolic tangent is used to ensure that the actions are within the action space.
+Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range.
 This is acceptable for PPO and SHAC.
@@
 To understand why, consider that we are training the policy to optimize two neural networks that approximate the environment.
-Further, these networks are not trained on data generated from action inside the action space.
-As a result, it is very likely that these networks achieve their maximum for actions outside the allowed range.
-Leading to higher than normal pre-activation actions that disrupt the training process.
+Further, these networks are never trained outside the allowed action range.
+As a result, it is likely that these networks will be maximized outside the allowed range thus pushing the policy to generate actions far outside the allowed range.
+This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.
 
@@
 A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.
+Figure~\ref{fig:shacpp} presents an unrolled view of the \fname{} algorithm.
 
 However, consider that, to ensure fast convergence of the reward and transition networks, we employ a cache to store seen trajectories.
-To ensure that the cache is balanced, and not filled completely with $0$-reward trajectories, we divide the cache $K$ equally sized bins.
+To ensure that the cache is balanced, and not filled with $0$-reward trajectories (which are common in case of sparse rewards), we divide the cache in $K$ equally sized bins.
 Each bin is filled with trajectories that have a similar reward.
-For the reward network, this corresponds to the immediate reward.
-For the transition network, this corresponds to the sum of reward accumulated during the trajectory.
 
@@
     \end{algorithmic}
+    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode}
     \label{alg:shacpp}
-    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode}
 \end{algorithm}

### 2025-01-08 paper/icml25/first/figs/shac.tex
revised figs to avoid overlaps

@@
         \node[circle, fill=RoyalBlue,draw=RoyalBlue] (a\i) at (\ax, -1.4) {\color{white}$a_\i$};
-        \node[circle, fill=DarkOrchid,draw=DarkOrchid] (r\i) at (\ax, +1.4) {\color{white}$r_\i$};
+        \node[circle, fill=DarkOrchid,draw=DarkOrchid] (r\i) at (\ax, +3.4) {\color{white}$r_\i$};
 
@@
         \draw[-latex,line width=1,color=DarkOrchid] (a\i) -- (r\i);
-        \draw[-latex,line width=1,color=DarkOrchid] (s\i) -- ($(s\i) + (.6,0)$) to[out=0,in=270]   ($(r\i) - (0,.8)$) -- (r\i);
-        \draw[-latex,line width=1,color=DarkOrchid] (s\j) -- ($(s\j) - (.6,0)$) to[out=180,in=270] ($(r\i) - (0,.8)$) -- (r\i);
+        \draw[-latex,line width=1,color=DarkOrchid,rounded corners=1ex] (s\i) -- ($(s\i) + (+.5,.6)$) -- ($(a\i) + (0,2)$) -- (r\i);
+        \draw[-latex,line width=1,color=DarkOrchid,rounded corners=1ex] (s\j) -- ($(s\j) + (-.5,.6)$) -- ($(a\i) + (0,2)$) -- (r\i);
         \node[circle, fill=DarkOrchid,draw=white, inner sep=1.5pt] (tmp) at ($(r\i) - (0,.8)$) {};
@@
         \node[circle, fill=RoyalBlue,draw=white, inner sep=1.5pt] (tmp) at ($(s\i) - (0,1.2)$) {};
-        \node at ($(tmp) - (.2,0)$) {\color{RoyalBlue}$\pi$};
+        \node at ($(tmp) - (.25,0)$) {\color{RoyalBlue}$\pi_\theta$};
     }
@@
     \foreach \i in {1, ..., \nnodes} {
-        \node[circle, fill=BlueGreen,draw=BlueGreen] (v\i) at ($(s\i) + (0,1.4)$) {\color{white}$v_\i$};
+        \node[circle, fill=BlueGreen,draw=BlueGreen] (v\i) at ($(s\i) + (0,3.4)$) {\color{white}$v_\i$};
         \draw[-latex,line width=1, color=BlueGreen] (s\i) --  (v\i);
@@
         \draw[-latex,line width=1, color=orange] (s\i) -- (s\j) ;
-        \draw[-latex,line width=1, color=orange] (a\i) -- ($(a\i) + (0,1.1)$) to[out=90,in=180] ($(s\j) - (.9,0)$) -- (s\j);
+        \draw[-, line width=1, color=orange, rounded corners=1ex] (a\i) -- ($(a\i) + (.2,.5)$) -- ($(s\j) - (.8,0)$); 
         \node[circle, fill=orange,draw=white, inner sep=1.5pt] (tmp) at ($(s\j) - (.8,0)$) {};
-        \node at ($(tmp) - (0,.3)$) {\color{orange}$\hat{F}$};
+        \node at ($(tmp) + (.1,.3)$) {\color{orange}$\hat{F}$};
     }

### 2025-01-08 paper/icml25/first/figs/shacpp.tex
revised figs to avoid overlaps

@@
         \pgfmathtruncatemacro{\j}{\i+1}
-        \draw[-,line width=1,color=DarkOrchid] (a\i) -- (R\i);
-        \draw[-,line width=1,color=DarkOrchid] (sa\i) -- ($(sa\i) + (.6,0)$) to[out=0,in=270]   ($(R\i) - (0,.2)$) -- (R\i);
-        \draw[-,line width=1,color=DarkOrchid] (sa\j) -- ($(sa\j) - (.6,0)$) to[out=180,in=270] ($(R\i) - (0,.2)$) -- (R\i);
+        \draw[-,line width=1,color=DarkOrchid, rounded corners=1ex] (a\i) -- ($(a\i) + (-.2,.5)$) -- ($(R\i) - (.2,1.25)$) -- ($(R\i) - (0,1)$) -- (R\i);
+        \draw[-,line width=1,color=DarkOrchid, rounded corners=1ex] (sa\i) -- ($(sa\i) + (.6,0)$) -- ($(R\i) - (0,.7)$) -- ($(R\i) - (0,.2)$) -- (R\i);
+        \draw[-,line width=1,color=DarkOrchid, rounded corners=1ex] (sa\j) -- ($(sa\j) - (.6,0)$) -- ($(R\i) - (0,.7)$) -- ($(R\i) - (0,.2)$) -- (R\i);
     }
@@
         \draw[-latex,line width=1, color=orange, dashed] (s\i) -- (s\j) ;
-        \draw[-latex,line width=1, color=orange, dashed] (a\i) -- ($(a\i) + (0,1.1)$) to[out=90,in=180] ($(s\j) - (.9,0)$) -- (s\j);
+        \draw[-, line width=1, color=orange, dashed] (a\i) -- ($(a\i) + (.2,.5)$) -- ($(s\j) - (.8,0)$); 
         \node[circle, fill=orange,draw=white, inner sep=1.5pt] (tmp) at ($(s\j) - (.8,0)$) {};
-        \node at ($(tmp) - (0,.3)$) {\color{orange}$\hat{F}$};
+        \node at ($(tmp) + (0,.3)$) {\color{orange}$\hat{F}$};
     }

### 2025-01-08 paper/icml25/first/sects/ablation.tex
added evaluation (WIP)

\section{Ablation Study}\label{sect:ablation}
While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.

In fact, the transition function is the most complex component to replace as it requires modeling with precision the observations obtained from the environment after a number of $H$ actions from each agent. Avoiding the need of such network lightens the algorithm but it reintroduce the need for a differentiable environment. Nonetheless, the need for a differentiable reward, made by SHAC, is still relaxed.

Studying such setup can provide insights on the importance and difficulty of the transition function in the SHAC algorithm. TODO

### 2025-01-08 paper/icml25/first/sects/discussion.tex
added evaluation (WIP)

\section{Discussion}

Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. The result shown in Sect.~\ref{sect:experiments} seems to support this conclusion. More specifically, let us answer the research questions posed in Sect.~\ref{sect:introduction}.

\begin{center}
\RQ{1} \textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
\end{center}

The propose framework, \fname{}, appears to be successful in emulating the behavior of SHAC in differentiable environments such as transport and sampling. Therefore, we can conclude that it is definitely possible to approximate the gradients of a differentiable environment with a neural network. However, we note that this approach, while obtaining comparable results in most setups, it tends to display higher variance. Further, before ebcoming effective, both neural network representing the reward and transition function need to become accurate which could require several episode of training.

\begin{center}
\RQ{2} \textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
\end{center}
TODO

\begin{center}
\RQ{3} \textbf{How does the performance of these algorithms change as the search space increases?}
\end{center}
TODO

### 2025-01-08 paper/icml25/first/sects/experiments.tex
added evaluation (WIP)

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-1-mlp.pdf}
        \caption{Dispersion, MLP, 1 agent}
        \label{fig:dispersion-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-3-transformer.pdf}
        \caption{Dispersion, Transformer, 3 agents}
        \label{fig:dispersion-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/dispersion-5-transformer.pdf}
        \caption{Dispersion, Transformer, 5 agents}
        \label{fig:dispersion-transformer-5}
    \end{subfigure}

    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-1-mlp.pdf}
        \caption{Transport, MLP, 1 agent}
        \label{fig:transport-mlp-1}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-3-transformer.pdf}
        \caption{Transport, Transformer, 3 agents}
        \label{fig:transport-transformer-3}
    \end{subfigure}
    \begin{subfigure}[b]{0.32\textwidth}
        \includegraphics[width=\textwidth]{figs/transport-5-transformer.pdf}
        \caption{Transport, Transformer, 5 agents}
        \label{fig:transport-transformer-5}
    \end{subfigure}

\end{figure*}

\section{experiments}\label{sect:experiments}

\paragraph{Scenarios}
As mentioned previously, we choose a few multi-angent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO. In particular, we choose the following scenario:
\begin{compactitem}
    \item \textbf{Dispersion}: a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent rechieves a sparse reward of $1$. Agents observe only their own position, velocity and relative position to each goal. Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario.
    \item \textbf{Transport}: a scenario where $n$ agents need to collaborate to push a package into a goal location. Agents observe their own position, velocity, relative position to the package and relative position between package and the goal. The agents receive a reward proportional to the distance between package and the goal. In this scenario, both the reward and transition functions are differentiable.
    \item \textbf{Sampling}: a scenario where $n$ agents need to collect rewards from a grid. The reward are sampled from $k$ gaussians. Agents observe their own position, velocity, lidar values to sence each other, and rewards in a 3x3 grid around themselves.  In this scenario, both the reward and transition functions are differentiable.
    \item \textbf{Discovery}: a scenario where $n$ agents need to collect $k$ goal place randomly. A goal is collected if at least $s$ agents stand in close proximity to it, in this case they receive a reward of $1$. If agents collide with each other, they receive a penalty. Every agent observes its position, velocity, lidar measurements to other agents and goals. In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario.
\end{compactitem}

\paragraph{Architectures}
We test SHAC, PPO, and \fname{} with the both a MLP and a Transformer architectures. In particular, we use the following:
\begin{compactitem}
    \item \textbf{Policy Network}: a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for PPO, SHAC, and \fname{}).
    \item \textbf{Value Network}:  a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for PPO, SHAC, and \fname{}).
    \item \textbf{Reward Network}: a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for \fname{}).
    \item \textbf{Transition Network}: a 3-layer single-head Transformer with hidden size $64$ (for \fname{}).
\end{compactitem}
While the MLP architecture is the most simple, the Transformer baseline benefits from position invariance property of the agents observations. We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.

\paragraph{Hyperparameters}
For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}. For all scenarios, each training episode consists of $512$ environments of $32$ steps each. Each validation episode is composed of $512$ environments of $512$ steps each. We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments. If early stopping is not triggered, we stop training after $20,000$ episodes. The discount factor is set to $0.99$ and lambda factor is set to $0.95$. We report results for increasing number of agents $n\in\{1,3,5\}$. We repeat and report results for $3$ runs with different seeds.

Overall, considering different training algorithms, architectures, scenarios, hyperparameters, we completed a total of $216$ runs lasting between $1$ to $8$ hours each.

\paragraph{Hardware Setup}
\texttt{\color{DarkOrchid}TODO: Qua metterei le specifiche CPU e GPU del vostro server}

\subsection{Results}

\paragraph{Dispersion}
Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion. We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture. In single agent setting, show in Fig.~\ref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes. With $3$ agents, shown in Fig.~\ref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes. With $5$ agents, shown in Fig.~\ref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping. In this last scenario, \fname{} manifest also high reward variance.

\paragraph{Transport}
Now, we compare the performance of these algorithms on a fully differentiable scenario, transport. On the single agent setting, shown in Fig.~\ref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well. While \fname{} appears to converge earlier SHAC appear to be more consistent. With $3$ and $5$ agents, shown in Fig.~\ref{fig:transport-transformer-3} and \ref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.

\paragraph{Sampling}
TODO

\paragraph{Dispersion}
TODO

### 2025-01-08 paper/icml25/first/sects/introduction.tex
added evaluation (WIP)
chore: try to clarify sample inefficient part

@@
-\section{Introduction}
+\section{Introduction}\label{sect:introduction}
 
-Learnign control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields.
-These scenarios presents a variaty of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions.
+Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields.
+These scenarios present a variety of different challenges, such as sparse rewards, complex dynamics, and multi-agent interactions.
 Current approaches often focus on either single- or multi-agent scenarios, lacking a unified solution that works effectively in both settings.
-Moreover, reinforcement learning algorithms tend to be sample inefficient due to the high variance of the gradient estimates and the exploration required \texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}.
+Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
+\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}.
 A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.

### 2025-01-08 paper/icml25/first/sects/shacpp.tex
added evaluation (WIP)
chore: minors on shac++ flow

@@
 
-In practice, this is achieved by caching the state-action-reward triplets obtained from the environment and then training the neural network to minimize the MSE loss.
-That is, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment.
+The neural network is then trained to minimize the MSE loss.
+Specifically, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
 $$ \frac{1}{\left|B\right|}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - \underbrace{r}_{\hat{R}(s, a)} \right|\right|^2 $$
@@
 Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range.
-This is acceptable for PPO and SHAC.
-However, in our scenario, this penalty is crucial.
-To understand why, consider that we are training the policy to optimize two neural networks that approximate the environment.
-Further, these networks are never trained outside the allowed action range.
-As a result, it is likely that these networks will be maximized outside the allowed range thus pushing the policy to generate actions far outside the allowed range.
+While this approach is acceptable for algorithms like PPO and SHAC, the penalty we employ is crucial in our specific scenario.
+This necessity arises from the fact that we are training the policy to optimize two neural networks that serve as environment approximations.
+Because these networks are never trained with data outside the permissible action space, they are likely to be maximized beyond these bounds.
+Consequently, the policy may generate actions far outside the allowed range.
 This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.
@@
 
-And, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.
+Additionally, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.
 
@@
     \begin{algorithmic}[1]
-    \STATE Initialize $\pi_\theta,V_\theta,R_\theta,W_\theta$ networks
-    \STATE Initialize learning rates $\eta_\pi, \eta_R, \eta_V, \eta_W$
+    \STATE Initialize $\pi_\theta,V_\theta,R_\theta,F_\theta$ networks
+    \STATE Initialize learning rates $\eta_\pi, \eta_R, \eta_V, \eta_F$
     \STATE Initialize environment $\mathcal{E}$
@@
         \STATE $R_\theta \gets R_\theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,r)$
-        \STATE $W_\theta \gets W_\theta - \eta_W \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$
+        \STATE $F_\theta \gets F_\theta - \eta_F \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$
 

### 2025-01-08 paper/icml25/first/sects/experiments.tex
minor on experiment section
compact the networks description a bit (@f14-bertolotti check pls)

@@
 
-\section{experiments}\label{sect:experiments}
+\section{Experiments}\label{sect:experiments}
 
 \paragraph{Scenarios}
-As mentioned previously, we choose a few multi-angent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO.
+As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO.
 In particular, we choose the following scenario:
 \begin{compactitem}
-    \item \textbf{Dispersion}: a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent rechieves a sparse reward of $1$.
+    \item \textbf{Dispersion}: a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
 Agents observe only their own position, velocity and relative position to each goal.
@@
 The reward are sampled from $k$ gaussians.
-Agents observe their own position, velocity, lidar values to sence each other, and rewards in a 3x3 grid around themselves.
+Agents observe their own position, velocity, lidar values to since each other, and rewards in a 3x3 grid around themselves.
 In this scenario, both the reward and transition functions are differentiable.
     \item \textbf{Discovery}: a scenario where $n$ agents need to collect $k$ goal place randomly.
-A goal is collected if at least $s$ agents stand in close proximity to it, in this case they receive a reward of $1$.
+A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$.
 If agents collide with each other, they receive a penalty.
@@
 \paragraph{Architectures}
-We test SHAC, PPO, and \fname{} with the both a MLP and a Transformer architectures. In particular, we use the following:
-\begin{compactitem}
-    \item \textbf{Policy Network}: a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for PPO, SHAC, and \fname{}).
-    \item \textbf{Value Network}:  a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for PPO, SHAC, and \fname{}).
-    \item \textbf{Reward Network}: a 1-layer MLP with hidden size $64$ or a 1-layer single-head Transformer with hidden size $64$ (for \fname{}).
-    \item \textbf{Transition Network}: a 3-layer single-head Transformer with hidden size $64$ (for \fname{}).
-\end{compactitem}
-While the MLP architecture is the most simple, the Transformer baseline benefits from position invariance property of the agents observations.
+We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture.
+Specifically, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 for the Policy Network and Value Network (for PPO, SHAC, and \fname{}).
+For the Reward Network, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 (for \fname{}).
+The Transition Network for \fname{} is a 3-layer single-head Transformer with hidden size 64.
+
+While the MLP architecture is the simplest, the Transformer baseline benefits from position invariance property of the agents observations.
 We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.

### 2025-01-08 paper/icml25/first/sects/experiments.tex
hardware setup

@@
 \paragraph{Hardware Setup}
-\texttt{\color{DarkOrchid}TODO: Qua metterei le specifiche CPU e GPU del vostro server}
+Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one Nvidia Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).
 

### 2025-01-08 paper/icml25/first/sects/related.tex
minor + related (WIP)

\section{Related Work}

\paragraph{Differentiable Engines}
Game engines~\cite{Bellemare13,Balla23} and Physics engines \cite{Todorov12,Coumans16} have been wildely used in the field of policy learning. Recently, there has been a trend towards differentiable engines, which allows for the use of gradient-based optimization methods, such as backpropagation. Notable examples are \cite{Freeman21,Howell22,Genesis24}. These framework may be built on top of auto differentiation libraries such \cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients \cite{Carpentier18,Werling21}. Further, several works tackle the challenge of the non-smoothness of the contact dynamics \cite{Degrave19,Moritz20}

\paragraph{Deep Reinforcement Learning}
Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks. Foundational works like DQN \cite{Mnih13} and AlphaZero \cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.

\paragraph{Multi-Agent Reinforcement Learning}

\paragraph{Reinforcement Learning on Differentiable Engines}

### 2025-01-08 paper/icml25/first/sects/shacpp.tex
minor + related (WIP)

@@
 To summarize, our algorithm, aims to minimize the 4 presented losses, $\mathcal{L'}_\theta^{\pi}$, $\mathcal{L}_\theta^V$, $\mathcal{L}_\theta^R$, and $\mathcal{L}_\theta^F$.
+The parameters $\theta$ can be either shared or separated among the different networks, however, we always keep them separated in our experiments.
 A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.
@@
         \STATE $s,a,r,v \sim \mathcal{E}(\pi)$
-        \STATE $\hat{s} = W(s[0], a)$
-        \STATE $\hat{r} = R(\hat{s})$
-        \STATE $\hat{v} = V(\hat{s})$
+        \STATE $\hat{s} = F_\theta(s_{:,0}, a)$
+        \STATE $\hat{r} = R_\theta(\hat{s})$
+        \STATE $\hat{v} = V_\theta(\hat{s})$
 
-        \STATE $\pi_\theta \gets \pi_\theta - \eta_\pi \nabla \mathcal{L'}^\pi_\theta(\hat{r},\hat{v},\gamma,\alpha)$
-        \STATE $V_\theta \gets V_\theta - \eta_V \nabla \mathcal{L}^V_\theta(\hat{v},s,v)$
-        \STATE $R_\theta \gets R_\theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,r)$
-        \STATE $F_\theta \gets F_\theta - \eta_F \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$
+        \STATE $\theta \gets \theta - \eta_\pi \nabla \mathcal{L'}^\pi_\theta(\hat{r},\hat{v},\gamma,\alpha)$
+        \STATE $\theta \gets \theta - \eta_V \nabla \mathcal{L}^V_\theta(\hat{v},s,v)$
+        \STATE $\theta \gets \theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,r)$
+        \STATE $\theta \gets \theta - \eta_F \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$
 
@@
     \end{algorithmic}
-    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode}
+    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode. $s_{:,0}$ denotes the first step of each trajectory in $s$.}
     \label{alg:shacpp}

### 2025-01-09 paper/icml25/first/sects/related.tex
related work on DRL

@@
 \paragraph{Differentiable Engines}
-Game engines~\cite{Bellemare13,Balla23} and Physics engines \cite{Todorov12,Coumans16} have been wildely used in the field of policy learning.
+Game engines~\cite{Bellemare13,Balla23} and Physics engines \cite{Todorov12,Coumans16} have been widely used in the field of policy learning.
 Recently, there has been a trend towards differentiable engines, which allows for the use of gradient-based optimization methods, such as backpropagation.
 Notable examples are \cite{Freeman21,Howell22,Genesis24}.
-These framework may be built on top of auto differentiation libraries such \cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients \cite{Carpentier18,Werling21}.
+These frameworks may be built on top of auto differentiation libraries such \cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients \cite{Carpentier18,Werling21}.
 Further, several works tackle the challenge of the non-smoothness of the contact dynamics \cite{Degrave19,Moritz20}
@@
 Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks. 
-Foundational works like DQN \cite{Mnih13} and AlphaZero \cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.
-
+Foundational works like DQN~\cite{Mnih13} (and its variants)
+and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess. 
+Over the years, several algorithms have been proposed to create more stable and efficient learning methods. 
+Notable examples include A3C for asynchronous learning~\cite{DBLP:conf/icml/MnihBMGLHSK16}, 
+DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17}, 
+and SAC for continuous action spaces with off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}. 
+Following the initial era of model-free algorithms, 
+model-based algorithms have been introduced to enhance sample efficiency. 
+Early works such as Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} have demonstrated promising results in terms of both sample efficiency and performance.
 \paragraph{Multi-Agent Reinforcement Learning}

### 2025-01-09 paper/icml25/first/sects/related.tex
related work on MARL

@@
 Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks. 
-Foundational works like DQN~\cite{Mnih13} (and its variants)
+Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{DBLP:conf/aaai/HesselMHSODHPAS18})
 and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess. 
@@
 Notable examples include A3C for asynchronous learning~\cite{DBLP:conf/icml/MnihBMGLHSK16}, 
-DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17}, 
+DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17} (also used in modern chatbot like ChatGPT), 
 and SAC for continuous action spaces with off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}. 
@@
 Early works such as Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} have demonstrated promising results in terms of both sample efficiency and performance.
+
 \paragraph{Multi-Agent Reinforcement Learning}
+Multi-Agent Reinforcement Learning (MARL) is a subfield of reinforcement learning that focuses on learning policies for multiple agents that interact with each other. 
+This field has a wide range of applications, 
+including robotics, 
+traffic control, 
+and game playing, 
+and involves different types of agents such as cooperative, competitive, and mixed. 
+In this paper, we primarily focus on cooperative multi-agent scenarios, 
+where agents need to collaborate to achieve a common goal. 
+Prominent works in this area include extensions of single-agent algorithms to multi-agent settings, 
+such as MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17} and MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}, 
+as well as novel solutions to unique challenges in multi-agent settings. Examples of these solutions are COMA~\cite{DBLP:conf/aaai/FoersterFANW18} for multi-agent credit assignment, 
+QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination in cooperative multi-agent settings, 
+and Mean Field MARL~\cite{DBLP:conf/icml/YangLLZZW18} for handling a large number of agents.
 

### 2025-01-09 paper/icml25/first/sects/related.tex
more on related work for MARL

@@
 \paragraph{Multi-Agent Reinforcement Learning}
-Multi-Agent Reinforcement Learning (MARL) is a subfield of reinforcement learning that focuses on learning policies for multiple agents that interact with each other. 
+Multi-Agent Reinforcement Learning (MARL) is a subfield of reinforcement learning that focuses on learning policies for multiple agents that interact with each other~\cite{albrecht2024multi}. 
 This field has a wide range of applications, 
-including robotics, 
-traffic control, 
-and game playing, 
-and involves different types of agents such as cooperative, competitive, and mixed. 
+including robotics~\cite{DBLP:journals/air/ChungFYN24}, 
+traffic control~\cite{DBLP:journals/tits/ChuWCL20}, 
+and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20,jaderberg2019human}, 
+and involves different types of agents such as cooperative~\cite{DBLP:journals/corr/abs-2312-05162}, competitive, and mixed. 
 In this paper, we primarily focus on cooperative multi-agent scenarios, 

### 2025-01-09 paper/icml25/first/sects/related.tex
minor on related

@@
 and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20,jaderberg2019human}, 
-and involves different types of agents such as cooperative~\cite{DBLP:journals/corr/abs-2312-05162}, competitive, and mixed. 
+and involves different types of agents such as cooperative, competitive, and mixed. 
 In this paper, we primarily focus on cooperative multi-agent scenarios, 

### 2025-01-13 paper/icml25/first/sects/ablation.tex
minor + figs

@@
+\begin{figure*}[!t]
+    \centering
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion-ablation-1-mlp.pdf}
+        \caption{Dispersion, MLP, 1 agent}
+        \label{fig:dispersion-ablation-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion-ablation-3-transformer.pdf}
+        \caption{Dispersion, Transformer, 3 agents}
+        \label{fig:dispersion-ablation-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion-ablation-5-transformer.pdf}
+        \caption{Dispersion, Transformer, 5 agents}
+        \label{fig:dispersion-ablation-transformer-5}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport-ablation-1-mlp.pdf}
+        \caption{Transport, MLP, 1 agent}
+        \label{fig:transport-ablation-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport-ablation-3-transformer.pdf}
+        \caption{Transport, Transformer, 3 agents}
+        \label{fig:transport-ablation-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport-ablation-5-transformer.pdf}
+        \caption{Transport, Transformer, 5 agents}
+        \label{fig:transport-ablation-transformer-5}
+    \end{subfigure}
+
+    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
+    \label{fig:ablation}
+\end{figure*}
+
 \section{Ablation Study}\label{sect:ablation}
@@
 Avoiding the need of such network lightens the algorithm but it reintroduce the need for a differentiable environment.
-Nonetheless, the need for a differentiable reward, made by SHAC, is still relaxed.
 
-Studying such setup can provide insights on the importance and difficulty of the transition function in the SHAC algorithm. TODO
+Fig.~\ref{fig:ablation} summarizes the results.
+Most notably, using the true transition function does not appear to provide any significant advantage in terms of reward achieved.
+In fact, in some cases, we report better results for \fname{}, namely in Fig.~\ref{fig:transport-ablation-1-mlp}.

### 2025-01-13 paper/icml25/first/sects/abstract.tex
minor + figs

\begin{abstract}
In this work, we introduce a novel framework, \fname{}, inspired by SHAC~\cite{Xu22}, a reinforcement learning algorithm designed for differentiable environments. Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the environment. These neural networks are trained alongside the policy and value networks of SHAC. We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22} and compare its performance against SHAC and PPO. Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to environment gradients.
\end{abstract}

### 2025-01-13 paper/icml25/first/sects/background.tex
minor + figs

@@
-\section{Background}
+\section{Background \& Notation}\label{sect:background}
 
@@
 To provide context for our work, we first introduce the reinforcement learning (RL) framework.
-Each task is modeled as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, where $\mathcal{S}$ represents the state space, $\mathcal{A}$ denotes the action space, $\hat{F}:\mathcal{S} \times \mathcal{A} \rightarrow \mathcal{S}$ is the transition function, $\hat{R}:\mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, $\gamma \in [0, 1]$ is the discount factor, and $T \in \mathbb{N}$ is the time horizon.
+Each task is modeled as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, F, R, T, \gamma, \mu)$, where $\mathcal{S}$ represents the state space, $\mathcal{A}$ denotes the action space, $F:\mathcal{S} \times \mathcal{A} \rightarrow \mathcal{S}$ is the transition function, $R:\mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, $\gamma \in [0, 1]$ is the discount factor, and $T \in \mathbb{N}$ is the time horizon.
+The initial state distribution is denoted by $\mu$.
 
 Further, let $\mathcal{T}$ denote the space of trajectories, defined as $\mathcal{T} = (\mathcal{S} \times \mathcal{A} \times \mathcal{S} \times \mathbb{R})^*$, where each trajectory consists of a sequence of starting states, actions, subsequent states, and rewards.
-We will use the symbols $((s_0, a_0, s_1, r_0), (s_1, a_1, s_2, r_1), \ldots)$ to denote a trajectory in $\mathcal{T}$.
+We will use the symbols $\tau=((s_0, a_0, s_1, r_0), (s_1, a_1, s_2, r_1), \ldots)$ to denote a generic trajectory in $\mathcal{T}$.
+When dealing with multiple trajectories, we will use the notation $s_i^\tau, a_i^\tau$, and $r_i^\tau$.
 
-The goal of the MDP is to find a policy $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maximizes the expected return 
-$$J(\pi) = \mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r(s_t, a_t, s_{t+1})\right]$$
-Where $\tau \in \mathcal{T}$ is a trajectory generated by the policy $\pi$ with the transition function $\hat{F}$.
+A policy is a function $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maps states to actions.
+We can use the policy to sample trajectories from the MDP by iteratively applying the transition and reward function and the policy.
+In this case, we have that the $i$-th element of a trajectory is $(s_i, a_i, s_{i+1}, r_i) = (s_i, \pi(s_i), F(s_i, a_i), R(s_i,a_i,F(s_i, a_i))$.
+The initial state $s_0$ is sampled from the initial state distribution $\mu$.
+
+For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled by following the policy $\pi$.
+
+The goal of the MDP is to find a policy that maximizes the expected return 
+$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r_t^\tau \right]$$
+
+\subsection{Neural Networks}
+A neural network is a function $f_{\theta}:\mathbb{R}^n\rightarrow\mathbb{R}^m$ parameterized by $\theta$ that maps an input vector $x\in\mathbb{R}^n$ to an output vector $y\in\mathbb{R}^m$.
+The parameters $\theta$ of the network are learned by minimizing a loss function $\mathcal{L}_\theta$ with respect to the training data.
+
+In this work, we will often approximate functions, such as transition ($F$) or reward ($R$), with neural network.
+Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$.
+For instance, the transition and its corresponding neural network are denoted as $F$ and $F_{\theta}$, respectively.
+The corresponding loss will be denoted as $\mathcal{L}_\theta^g$ and it will aim to minimize the differences between $g$ and $g_\theta$.
+
+While we use a unique set of parameters $\theta$ for all the networks.
+Each network can either use its own subset of $\theta$ or it can share parameters with other networks.
+In practice, the parameters will never be shared.
 
@@
 
-To solve an MDP $\mathcal{M} = (\mathcal{S}, \mathcal{A}, \hat{F}, \hat{R}, \gamma, T)$, SHAC trains a policy $\pi$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$\mathcal{L}_\theta^{\pi} = \frac{1}{NH}\sum_{i=1}^N \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^i, a_t^i) + \gamma^H V(s^i_{t_0+T}),$$  
-where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^i$ is the state at time $t$ for trajectory $i$, $a_t^i$ is the action taken at time $t$ for trajectory $i$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$.
-Additionally, SHAC assumes that the policy, reward, and transition functions are differentiable, i.e., $\pi, \hat{F}, \hat{R} \in \mathcal{C}^1$.
+To solve an MDP $\mathcal{M}$, SHAC trains a policy $\pi_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$ for a batch of $N$ trajectories $B = \{\tau_1, \ldots, \tau_N\}$ sampled from the MDP.
+The loss is defined as:
+\begin{equation}\label{eq:shac}
+    \mathcal{L}_\theta^{\pi}(B) = \frac{1}{NH}\sum_{\tau\in B} \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^\tau, a_t^\tau) + \gamma^H V_\theta(s^\tau_{t_0+T})
+\end{equation}
+where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$.
+From Equation~\ref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$.
+While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.
 
-The value function is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $\hat{V}$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{V}:\mathcal{T}^N\rightarrow\mathbb{R}$.
+The value function $V_\theta$ is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $V$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{V}:\mathcal{T}^N\rightarrow\mathbb{R}$.
 
-$$ \frac{1}{N}\sum_i^N\left|\left| V(s^i) - \hat{V}(s^i) \right|\right|^2 $$
-$$ \hat{V}(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
+$$ \mathcal{L}_\theta^V(B)=\frac{1}{NT}\sum_{\tau\in B}\sum_{t=0}^T\left|\left| V_\theta(s^\tau_t) - V(s^\tau_t) \right|\right|^2 $$
+$$ V(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
 $$ G_t^k = \left(\sum_{l=0}^{k-1}\gamma^l r_{t+l}\right) + \gamma^k V(s_{t+k})$$
@@
 $G^t_k$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$.
-The estimates $\hat{V}$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.
+The estimates $V$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.
 

### 2025-01-13 paper/icml25/first/sects/experiments.tex
minor + figs

@@
-\begin{figure*}[t]
+\begin{figure*}[!t]
     \centering
@@
 
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-1-mlp.pdf}
+        \caption{Discovery, MLP, 1 agent}
+        \label{fig:discovery-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-3-transformer.pdf}
+        \caption{Discovery, Transformer, 3 agents}
+        \label{fig:discovery-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.32\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-5-transformer.pdf}
+        \caption{Discovery, Transformer, 5 agents}
+        \label{fig:discovery-transformer-5}
+    \end{subfigure}
+
+    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Sampling, and Discovery scenarios.}
+    \label{fig:experiments}
+
 \end{figure*}
@@
 \begin{compactitem}
-    \item \textbf{Dispersion}: a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
+    \item \textbf{Dispersion} (Fig.~\ref{fig:dispersion}): a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
 Agents observe only their own position, velocity and relative position to each goal.
@@
 Therefore, only PPO and \fname{} are applicable to this scenario.
-    \item \textbf{Transport}: a scenario where $n$ agents need to collaborate to push a package into a goal location.
+The maximum reward is $n$.
+    \item \textbf{Transport} (Fig.~\ref{fig:transport}): a scenario where $n$ agents need to collaborate to push a package into a goal location.
 Agents observe their own position, velocity, relative position to the package and relative position between package and the goal.
@@
 In this scenario, both the reward and transition functions are differentiable.
-    \item \textbf{Sampling}: a scenario where $n$ agents need to collect rewards from a grid.
-The reward are sampled from $k$ gaussians.
-Agents observe their own position, velocity, lidar values to since each other, and rewards in a 3x3 grid around themselves.
-In this scenario, both the reward and transition functions are differentiable.
-    \item \textbf{Discovery}: a scenario where $n$ agents need to collect $k$ goal place randomly.
+The maximum reward varies from environement to environment.
+It is the scaled distance between the package and the goal times the number of agents.
+    \item \textbf{Discovery} (Fig.~\ref{fig:discovery}): a scenario where $n$ agents need to collect $k$ goal place randomly.
 A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$.
@@
 Therefore, only PPO and \fname{} are applicable to this scenario.
+When $n=1$ we set $s=1$ and $k=7$.
+When $n>1$ we set $s=2$.
+The maximum rewards depends from the time available, but we set it to $k$.
+    \item \textbf{Sampling} (Fig.~\ref{fig:sampling}): a scenario where $n$ agents need to collect rewards from a grid.
+The reward are sampled from $k$ gaussians.
+Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.
+In this scenario, both the reward and transition functions are differentiable.
+The maximum reward depends on where the gaussians are placed.
 \end{compactitem}
@@
 
+\paragraph{Discovery}
+Similar consideration hold also for the discovery scenario.
+In the single agent setting, shown in Fig.~\ref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO .
+However, with $3$ and $5$ agents, shown in Fig.~\ref{fig:discovery-transformer-3} and \ref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
+
 \paragraph{Sampling}
-TODO
-
-\paragraph{Dispersion}
-TODO

### 2025-01-13 paper/icml25/first/sects/shacpp.tex
minor + figs

@@
 
-\section{\fname{}}
-
-As mentioned previously, we aim to address the limitations of SHAC by approximating the gradients of a differentiable simulator with neural networks.
-In particular, we aim to train a neural network to approximate the reward function by minimizing:
-$$ \mathbb{E}_{s,a}\left[\left|\left| R_{\theta}(s, a) - \hat{R}(s, a) \right|\right|^2\right] $$
-
-In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment.
-The neural network is then trained to minimize the MSE loss.
-Specifically, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$ \frac{1}{\left|B\right|}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - \underbrace{r}_{\hat{R}(s, a)} \right|\right|^2 $$
-Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.
-
-Similarly, we aim to train a neural network to approximate the transition function.
-However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model~\cite{Ma24}.
-This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $N$ actions from $\mathcal{A}^H$, predicts the next $N$ observations from $\mathcal{S}^H$.
-In practice, we aim to minimize:
-
-$$ \mathbb{E}_{\tau}\left[\sum_{i=1}^H \left|\left| F_{\theta}(s_0, a_0, \dots, a_N)_i - \hat{F}(s_i, a_i) \right|\right|^2\right] $$
-
-Where $\tau\in\mathcal{T}$.
-In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss.
-That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$ \frac{1}{NH}\sum_{\tau \in B}\sum_{i=1}^H\left|\left| F_{\theta}(s_0, a_0, \dots, a_H)_i - \underbrace{s_{i}}_{\hat{F}(s_{i-1}, a_{i-1})} \right|\right|^2 $$
-Where $B$ represents a batch of trajectories sampled uniformly from the cache.
-
-Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space.
-That is, suppose the action space is $(a,b)$ (with $a<b$), we minimize the loss $\mathcal{L'}_\theta^\pi:$
-
-$$\mathcal{L}_\theta^\pi + \alpha \frac{1}{NH}\sum_{i=0}^N\sum_{t=t_0}^{t_0+H-1} d(a_t^i,(a,b))$$
-$$d(x,(a,b)) = \begin{cases}(a-x)^2 & \text{if } x < a \\ (x-b)^2 & \text{if } x > b \\ 0 & \text{otherwise} \end{cases}$$
-
-    Where $\alpha$ is a hyper-parameter that controls the strength of the penalty.
-In our experiments, we set $\alpha=1$.
-Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range.
-While this approach is acceptable for algorithms like PPO and SHAC, the penalty we employ is crucial in our specific scenario.
-This necessity arises from the fact that we are training the policy to optimize two neural networks that serve as environment approximations.
-Because these networks are never trained with data outside the permissible action space, they are likely to be maximized beyond these bounds.
-Consequently, the policy may generate actions far outside the allowed range.
-This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.
-
-To summarize, our algorithm, aims to minimize the 4 presented losses, $\mathcal{L'}_\theta^{\pi}$, $\mathcal{L}_\theta^V$, $\mathcal{L}_\theta^R$, and $\mathcal{L}_\theta^F$.
-The parameters $\theta$ can be either shared or separated among the different networks, however, we always keep them separated in our experiments.
-A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.
-Figure~\ref{fig:shacpp} presents an unrolled view of the \fname{} algorithm.
-
-However, consider that, to ensure fast convergence of the reward and transition networks, we employ a cache to store seen trajectories.
-To ensure that the cache is balanced, and not filled with $0$-reward trajectories (which are common in case of sparse rewards), we divide the cache in $K$ equally sized bins.
-Each bin is filled with trajectories that have a similar reward.
-
-Additionally, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.
-
-The full implementation is openly available at 
-\begin{center}
-\url{https://redacted/for/anonymity}
-\end{center}
-
 \begin{algorithm}[t]
@@
         \STATE $\theta \gets \theta - \eta_V \nabla \mathcal{L}^V_\theta(\hat{v},s,v)$
-        \STATE $\theta \gets \theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,r)$
+        \STATE $\theta \gets \theta - \eta_R \nabla \mathcal{L}^R_\theta(\hat{r},s,a,r)$
         \STATE $\theta \gets \theta - \eta_F \nabla \mathcal{L}^F_\theta(\hat{s},s,a)$
@@
 \end{algorithm}
+
+\section{\fname{}}
+
+As mentioned previously, we aim to address the limitations of SHAC by approximating the gradients of a differentiable environment ($R$ and $F$) with neural networks ($R_\theta$ and $F_\theta$).
+In particular, we aim to train a neural network to approximate the reward function by minimizing:
+$$ \mathbb{E}_{\tau\sim\pi}\left[\left|\left| \frac{1}{T}\sum_{i=0}^{T-1} R_{\theta}(s_i^\tau, a_i^\tau, s_{i+1}^\tau) - R(s_i^\tau, a_i^\tau, s_{i+1}^\tau) \right|\right|^2\right] $$
+
+In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment.
+The neural network is then trained to minimize the mean squared error (MSE) loss.
+Specifically, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+$$ \mathcal{L}_\theta^R(B) = \frac{1}{N}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - r \right|\right|^2 $$
+Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.
+
+Similarly, we aim to train a neural network to approximate the transition function.
+However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model~\cite{Ma24}.
+This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $H$ actions from $\mathcal{A}^H$, predicts the next $H$ observations from $\mathcal{S}^H$.
+We aim to minimize:
+
+\begin{equation}
+\begin{aligned}
+\mathbb{E}_{\tau\sim\pi}\left[\frac{1}{NH}\sum_{t=0}^{T-H-1}\sum_{i=0}^{H-1} \left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - \right.\right.\right.
+\\ \left.\left.\left.
+s_{t+i+1}^\tau \right|\right|^2\right]
+\end{aligned}
+\end{equation}
+
+In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss.
+That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
+$$ \frac{1}{NH}\sum_{\tau \in B}\sum_{t=t_0}^{t_0+H-1}\left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - s_{t+1}^\tau \right|\right|^2 $$
+Where $B$ represents a batch of trajectories with horizon $H$ sampled uniformly from the cache.
+
+Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space.
+That is, suppose the action space is $(a,b)$ (with $a<b$), we minimize the loss $\mathcal{L'}_\theta^\pi:$
+\begin{figure*}[!t]
+    \centering
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion.png}
+        \caption{Dispersion}
+        \label{fig:dispersion}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport.png}
+        \caption{Transport}
+        \label{fig:transport}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery.png}
+        \caption{Discovery}
+        \label{fig:discovery}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling.png}
+        \caption{Sampling}
+        \label{fig:sampling}
+    \end{subfigure}
+    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
+\end{figure*}
+
+$$\mathcal{L}_\theta^\pi(B) + \alpha \frac{1}{NH}\sum_{\tau\in B}\sum_{t=t_0}^{t_0+H-1} d(a_t^\tau,(a,b))$$
+$$d(x,(a,b)) = \begin{cases}(a-x)^2 & \text{if } x < a \\ (x-b)^2 & \text{if } x > b \\ 0 & \text{otherwise} \end{cases}$$
+
+    Where $\alpha$ is a hyper-parameter that controls the strength of the penalty.
+In our experiments, we set $\alpha=1$.
+Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range.
+While this approach is acceptable for algorithms like PPO and SHAC, the penalty we employ is crucial in our specific scenario.
+This necessity arises from the fact that we are training the policy to optimize two neural networks that serve as environment approximations.
+Because these networks are never trained with data outside the permissible action space, they are likely to be maximized beyond these bounds.
+Consequently, the policy may generate actions far outside the allowed range.
+This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.
+
+To summarize, our algorithm, aims to minimize the 4 presented losses, $\mathcal{L'}_\theta^{\pi}$, $\mathcal{L}_\theta^V$, $\mathcal{L}_\theta^R$, and $\mathcal{L}_\theta^F$.
+A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.
+Figure~\ref{fig:shacpp} presents an unrolled view of the \fname{} algorithm.
+
+However, consider that, to ensure fast convergence of the reward and transition networks, we employ a cache to store seen trajectories.
+To ensure that the cache is balanced, and not filled with $0$-reward trajectories (which are common in case of sparse rewards), we divide the cache in $K$ equally sized bins.
+Each bin is filled with trajectories that have a similar reward.
+
+Additionally, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.
+
+The full implementation is openly available at 
+\begin{center}
+\url{https://redacted/for/anonymity}
+\end{center}

### 2025-01-14 paper/icml25/first/sects/experiments.tex
added discovery-ablation figures, added hyperparameter tables & script

@@
 
-    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Sampling, and Discovery scenarios.}
+    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
     \label{fig:experiments}
@@
 \begin{compactitem}
-    \item \textbf{Dispersion} (Fig.~\ref{fig:dispersion}): a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
+    \item \textbf{Dispersion} (Fig.~\ref{fig:dispersion}): 
+        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
 Agents observe only their own position, velocity and relative position to each goal.
-Despite the transition function being differentiable, the reward function is not.
-Therefore, only PPO and \fname{} are applicable to this scenario.
-The maximum reward is $n$.
-    \item \textbf{Transport} (Fig.~\ref{fig:transport}): a scenario where $n$ agents need to collaborate to push a package into a goal location.
+        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
+        \sentence{Max Reward} The maximum reward is $n$.
+    \item \textbf{Transport} (Fig.~\ref{fig:transport}): 
+        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location.
 Agents observe their own position, velocity, relative position to the package and relative position between package and the goal.
 The agents receive a reward proportional to the distance between package and the goal.
-In this scenario, both the reward and transition functions are differentiable.
-The maximum reward varies from environement to environment.
-It is the scaled distance between the package and the goal times the number of agents.
-    \item \textbf{Discovery} (Fig.~\ref{fig:discovery}): a scenario where $n$ agents need to collect $k$ goal place randomly.
+        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
+        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
+    \item \textbf{Discovery} (Fig.~\ref{fig:discovery}): 
+        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly.
 A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$.
@@
 Every agent observes its position, velocity, lidar measurements to other agents and goals.
-In this scenario, only the transition function is differentiable.
-Therefore, only PPO and \fname{} are applicable to this scenario.
 When $n=1$ we set $s=1$ and $k=7$.
 When $n>1$ we set $s=2$.
-The maximum rewards depends from the time available, but we set it to $k$.
-    \item \textbf{Sampling} (Fig.~\ref{fig:sampling}): a scenario where $n$ agents need to collect rewards from a grid.
+        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
+        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
+        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents.
+For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
+     \item \textbf{Sampling} (Fig.~\ref{fig:sampling}): 
+        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid.
 The reward are sampled from $k$ gaussians.
 Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.
-In this scenario, both the reward and transition functions are differentiable.
-The maximum reward depends on where the gaussians are placed.
+        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
+        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
+        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
 \end{compactitem}
@@
 We repeat and report results for $3$ runs with different seeds.
+See Table~\ref{apx:tab:ppo} and Table~\ref{apx:tab:shac} for a complete list of hyperparameters.
 
@@
 Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion.
-We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture.
+We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see Fig.~\ref{apx:fig:experiments-mlp}).
 In single agent setting, show in Fig.~\ref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes.
@@
 With $3$ and $5$ agents, shown in Fig.~\ref{fig:transport-transformer-3} and \ref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.
+For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}
 
@@
 However, with $3$ and $5$ agents, shown in Fig.~\ref{fig:discovery-transformer-3} and \ref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
+For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}
 

### 2025-01-14 paper/icml25/first/maxreward_table.py
added relative max reward table and script

@@ -1,0 +1,108 @@
+import click
+import json
+import os
+
+architectures = ["mlp", "transformer"]
+scenarios = ["dispersion", "transport", "discovery", "sampling"]
+algorithms = ["ppo", "shac", "shacrm", "shacwm"]
+agents = [1, 3, 5]
+
+def path2algorithm(path):
+    return path.split("/")[1]
+
+def path2scenario(path):
+    return path.split("/")[2]
+
+def path2agents(path):
+    return int(path.split("/")[3])
+
+def path2architecture(path):
+    return path.split("/")[4]
+
+def path2seed(path):
+    return int(path.split("/")[5])
+
+@click.command()
+@click.option("--input-path"  , "input_path"  , default="data"   , help='Path to the data')
+@click.option("--output-path" , "output_path" , default="output" , help='Path to the output')
+def maxreward_table(input_path, output_path):
+    # get all eval.log
+    path2data = {}
+    for root, dirs, files in os.walk(input_path):
+        for file in files:
+            path = os.path.join(root, file)
+            if file == "eval.log":
+                # read jsonline eval.log
+                with open(path, "r") as f:
+                    data = [json.loads(line) for line in f]
+                    path2data[path] = data
+
+
+    # get max reward
+    path2max = {}
+    for path, data in path2data.items():
+        rewards = [d["message"]["reward"] for d in data]
+        path2max[path] = max(rewards)
+
+    # aggregate rewards by seed
+    path2max_by_seed = {}
+    for path,reward in path2max.items():
+        seed = path.split("/")[-2]
+        path_no_seed = os.path.join(*path.split("/")[:-2],path.split("/")[-1])
+        if path_no_seed not in path2max_by_seed:
+            path2max_by_seed[path_no_seed] = reward
+        else:
+            path2max_by_seed[path_no_seed] = max(path2max_by_seed[path_no_seed], reward)
+
+    # drop all transformer with 1 agent
+    path2max_by_seed = {path:reward if path2agents(path) > 1 or path2architecture(path) == "mlp" else 0 for path,reward in path2max_by_seed.items() }
+
+    # normalize by algorithm max
+    algorithm2max = {}
+    for path in path2max_by_seed:
+        algorithm = path.split("/")[-4]
+        if algorithm not in algorithm2max:
+            algorithm2max[algorithm] = path2max_by_seed[path]
+        else:
+            algorithm2max[algorithm] = max(algorithm2max[algorithm], path2max_by_seed[path])
[…]

### 2025-01-14 paper/icml25/first/maxreward_table.py
1.00 are bold

@@ -70,8 +70,6 @@ def maxreward_table(input_path, output_path):
         algorithm = path.split("/")[-4]
         path2max_by_seed_normalized[path] = path2max_by_seed[path] / algorithm2max[algorithm]
 
-    print(path2max_by_seed_normalized)
-
     # write latex table
     with open(output_path, "w") as f:
         f.write("\\begin{tabular}{ c c c c c c c c c c }\n")
@@ -88,13 +86,11 @@ def maxreward_table(input_path, output_path):
                         path = f"{input_path}/{algorithm}/{scenario}/{agents_}/{architecture}/eval.log"
                         if path not in path2max_by_seed_normalized:
                             f.write(" & -")
-                        elif architecture == "transformer" and agents_ >1 and scenario=="dispersion":
-                            print(path, path2max_by_seed_normalized[path])
-                            f.write(f" & {path2max_by_seed_normalized[path]:.3f}")
                         elif architecture == "transformer" and agents_ == 1:
                             f.write(" & -")
                         elif path in path2max_by_seed_normalized:
-                            f.write(f" & {path2max_by_seed_normalized[path]:.3f}")
+                            value = path2max_by_seed_normalized[path]
+                            f.write(" & \\textbf{"+f"{value:.2f}"+"}" if value == 1 else f" & {value:.2f}")
                         else:
                             raise ValueError("Path not found")
                         

### 2025-01-14 paper/icml25/first/maxreward_table.py
fixed max table

@@ -1,3 +1,4 @@
+import pprint
 import click
 import json
 import os
@@ -44,31 +45,47 @@ def maxreward_table(input_path, output_path):
         rewards = [d["message"]["reward"] for d in data]
         path2max[path] = max(rewards)
 
-    # aggregate rewards by seed
+    # aggregate by seed
     path2max_by_seed = {}
-    for path,reward in path2max.items():
-        seed = path.split("/")[-2]
-        path_no_seed = os.path.join(*path.split("/")[:-2],path.split("/")[-1])
-        if path_no_seed not in path2max_by_seed:
-            path2max_by_seed[path_no_seed] = reward
+    for path, rew in path2max.items():
+        agent = path2agents(path)
+        scenario = path2scenario(path)
+        arch = path2architecture(path)
+        algo = path2algorithm(path)
+        key = (arch, agent, scenario, algo)
+        if key not in path2max_by_seed:
+            path2max_by_seed[key] = rew
         else:
-            path2max_by_seed[path_no_seed] = max(path2max_by_seed[path_no_seed], reward)
+            path2max_by_seed[key] = max(path2max_by_seed[key], rew)
 
-    # drop all transformer with 1 agent
-    path2max_by_seed = {path:reward if path2agents(path) > 1 or path2architecture(path) == "mlp" else 0 for path,reward in path2max_by_seed.items() }
-
-    # normalize by algorithm max
-    algorithm2max = {}
-    for path in path2max_by_seed:
-        algorithm = path.split("/")[-4]
-        if algorithm not in algorithm2max:
-            algorithm2max[algorithm] = path2max_by_seed[path]
+    # get max in scenario-agents
+    path2reference = {}
+    for path, rew in path2max.items():
+        scenario = path2scenario(path)
+        agent = path2agents(path)
+        #architecture = path2architecture(path)
+        #key = (architecture, agent, scenario)
+        key = (agent, scenario)
+        if key not in path2reference:
+            path2reference[key] = rew
         else:
-            algorithm2max[algorithm] = max(algorithm2max[algorithm], path2max_by_seed[path])
-    path2max_by_seed_normalized = {}
-    for path in path2max_by_seed:
-        algorithm = path.split("/")[-4]
-        path2max_by_seed_normalized[path] = path2max_by_seed[path] / algorithm2max[algorithm]
+            path2reference[key] = max(path2reference[(key)], rew)
+
[…]

### 2025-01-14 paper/icml25/first/maxreward_table.py
update figs with max line, updated table

@@ -38,7 +38,6 @@ def maxreward_table(input_path, output_path):
                     data = [json.loads(line) for line in f]
                     path2data[path] = data
 
-
     # get max reward
     path2max = {}
     for path, data in path2data.items():
@@ -58,21 +57,23 @@ def maxreward_table(input_path, output_path):
         else:
             path2max_by_seed[key] = max(path2max_by_seed[key], rew)
 
+    # transformer agent 1 is not available
+    for key in path2max_by_seed:
+        if key[0] == "transformer" and key[1] == 1:
+            path2max_by_seed[key] = -9999
+
+    pprint.pprint(path2max_by_seed)
+    print()
+
     # get max in scenario-agents
     path2reference = {}
-    for path, rew in path2max.items():
-        scenario = path2scenario(path)
-        agent = path2agents(path)
-        #architecture = path2architecture(path)
-        #key = (architecture, agent, scenario)
+    for (arch, agent, scenario, algo), rew in path2max_by_seed.items():
         key = (agent, scenario)
         if key not in path2reference:
             path2reference[key] = rew
         else:
             path2reference[key] = max(path2reference[(key)], rew)
 
-    pprint.pprint(path2max_by_seed)
-    print()
     pprint.pprint(path2reference)
     print()
 
@@ -84,8 +85,6 @@ def maxreward_table(input_path, output_path):
         arch = path2architecture(path)
         algo = path2algorithm(path)
         path2max_by_seed_norm[(arch, agent, scenario, algo)] = path2max_by_seed[(arch, agent, scenario, algo)] / path2reference[(agent, scenario)]
-
-    pprint.pprint(path2max_by_seed_norm)
 
     # write latex table
     with open(output_path, "w") as f:

### 2025-01-15 paper/icml25/first/sects/background.tex
minors + last related para

@@
 
+For simplicity, we formally treat only the case where the policy observe the full state to make the action.
+However, in our experiments, we consider also scenarios where the full state is not available to the policy.
+In such cases, the MDP includes a set of observations, $\mathcal{O}$, an observation function $O:\mathcal{S}\rightarrow\mathcal{O}$, and the policy is a function $\pi:\mathcal{O}\rightarrow\mathcal{A}$.
+In such cases, the MDP takes the name of Partially Observable Markov Decision Process (POMDP).
+
 \subsection{SHAC}

### 2025-01-15 paper/icml25/first/sects/related.tex
minors + last related para

@@
 \paragraph{Reinforcement Learning on Differentiable Engines}
+Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}.
+In \cite{Lv23}, the authors introduce rendering to enhance model-based RL.
+\cite{Zheng24} employes a differentiable symbolic expression search engine.
+Trajectory optimization approach is taken in \cite{Wan24}.
+\cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.
+With respect to this approach, we take a substantially different method, by emulating the differentiable environment with a trained neural network.

### 2025-01-15 src/experiments/configs/shacwm.py
added log grads option

@@ -112,6 +112,9 @@ class Shacwm:
         self.etr                    = 5   # epochs before envs. reset
         self.etv                    = 100 # epochs before evaluation
 
+        # training configurations
+        self.is_deterministic       = True
+
         # rl gamma and lambda coefficients
         self.gamma_factor           = 0.99
         self.lambda_factor          = 0.95

### 2025-01-15 src/trainers/ppo.py

@@ -33,6 +33,7 @@ def ppo(
         compile                : bool                                   ,
         restore_path           : str|None                               ,
         early_stopping         : dict                                   ,
+        log_grads              : bool                                   ,
     ):
 
     ppo_logger    = utils.get_file_logger(os.path.join(dir,"ppo.log"))
@@ -66,15 +67,16 @@ def ppo(
 
         # train policy and value models ##############################
         trainers.routines.ppo_policy_value(
-            policy_model     = policy_model,
-            value_model      = value_model,
-            policy_optimizer = policy_model_optimizer,
-            value_optimizer  = value_model_optimizer,
-            episode_data     = episode_data,
-            batch_size       = batch_size,
-            ppo_logger       = ppo_logger,
-            gamma            = gamma_factor,
-            epochs           = epochs
+            policy_model     = policy_model           ,
+            value_model      = value_model            ,
+            policy_optimizer = policy_model_optimizer ,
+            value_optimizer  = value_model_optimizer  ,
+            episode_data     = episode_data           ,
+            batch_size       = batch_size             ,
+            ppo_logger       = ppo_logger             ,
+            gamma            = gamma_factor           ,
+            epochs           = epochs                 ,
+            log_grads        = log_grads              ,
         )
 
         # checkpoint ##################################################

### 2025-01-15 paper/icml25/first/sects/abstract.tex
minor on abstract

@@
 \begin{abstract}
-In this work, we introduce a novel framework, \fname{}, inspired by SHAC~\cite{Xu22}, a reinforcement learning algorithm designed for differentiable environments.
+In this work, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable environments.
 Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the environment.
 These neural networks are trained alongside the policy and value networks of SHAC.
-We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22} and compare its performance against SHAC and PPO.
+We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO.
 Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to environment gradients.

### 2025-01-16 src/trainers/routines/policy.py
log grads fix; recipes for log targets runs

@@ -24,6 +24,7 @@ def train_policy(
     live_runs = dead_runs.logical_not()
     live_steps = dones[:,live_runs,0].logical_not().sum(0) - 1
 
+
     # compute loss
     optimizer.zero_grad()
     loss = -(
@@ -38,6 +39,7 @@ def train_policy(
 
     # backward pass
     loss.backward()
+    if log_grads: gradnorm = torch.cat([param.grad.detach().flatten() for param in policy_model.parameters()]).norm().item()
     if clip_coefficient is not None: torch.nn.utils.clip_grad_norm_(policy_model.parameters(), clip_coefficient)
     optimizer.step()
 
@@ -47,7 +49,7 @@ def train_policy(
             "done"    : episode_data["dones"][-1,:,0].sum().int().item(),
             "reward"  : episode_data["rewards"].sum(0).mean(0).sum().item()
     } | ({} if not log_grads else {
-            "grads"     : torch.cat([param.grad.detach().flatten() for param in policy_model.parameters()]).norm().item(),
+            "grads"   : gradnorm,
     })))
 
 

### 2025-01-16 paper/icml25/first/sects/experiments.tex
added relative max reward table and script
update figs with max line, updated table
minors + last related para
added norm grads fig

@@
     \label{fig:experiments}
+\end{figure*}
 
-\end{figure*}
+\begin{figure}[t]
+    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
+    \caption{Norm of policy gradients on the Transport environment across 5000 epochs.
+The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent.
+Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurence of generalization (epoch 3200).}
+\end{figure}
 
@@
         \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
+        \sentence{World Model} The world model has enough information to predict the next states accurately.
         \sentence{Max Reward} The maximum reward is $n$.
@@
         \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
+        \sentence{World Model} The world model has enough information to predict the next states accurately.
         \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
@@
 \end{compactitem}
+
+\begin{table*}[!t]
+    \centering
+    \input tables/max.tex
+    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}
+\end{table*}
 

### 2025-01-18 paper/icml25/first/sects/ablation.tex
added discovery-ablation figures, added hyperparameter tables & script
figures are now generated in a grid

@@
     \centering
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-ablation-1-mlp.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-ablation-3-transformer.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-ablation-5-transformer.pdf}
@@
 
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-ablation-1-mlp.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-ablation-3-transformer.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-ablation-5-transformer.pdf}
@@
         \label{fig:transport-ablation-transformer-5}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-ablation-1-mlp.pdf}
+        \caption{Discovery, MLP, 1 agent}
+        \label{fig:discovery-ablation-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-ablation-3-transformer.pdf}
+        \caption{Discovery, Transformer, 3 agents}
+        \label{fig:discovery-ablation-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery-ablation-5-transformer.pdf}
+        \caption{Discovery, Transformer, 5 agents}
+        \label{fig:discovery-ablation-transformer-5}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-ablation-1-mlp.pdf}
+        \caption{Discovery, MLP, 1 agent}
+        \label{fig:sampling-ablation-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-ablation-3-transformer.pdf}
+        \caption{Discovery, Transformer, 3 agents}
+        \label{fig:sampling-ablation-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-ablation-5-transformer.pdf}
+        \caption{Discovery, Transformer, 5 agents}
+        \label{fig:sampling-ablation-transformer-5}
     \end{subfigure}

### 2025-01-18 paper/icml25/first/sects/experiments.tex
figures are now generated in a grid

@@
     \centering
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-1-mlp.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-3-transformer.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion-5-transformer.pdf}
@@
 
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-1-mlp.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-3-transformer.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/transport-5-transformer.pdf}
@@
 
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/discovery-1-mlp.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/discovery-3-transformer.pdf}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.32\textwidth}
+    \begin{subfigure}[b]{0.30\textwidth}
         \includegraphics[width=\textwidth]{figs/discovery-5-transformer.pdf}
@@
         \label{fig:discovery-transformer-5}
+    \end{subfigure}
+
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-1-mlp.pdf}
+        \caption{Discovery, MLP, 1 agent}
+        \label{fig:sampling-mlp-1}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-3-transformer.pdf}
+        \caption{Discovery, Transformer, 3 agents}
+        \label{fig:sampling-transformer-3}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.30\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling-5-transformer.pdf}
+        \caption{Discovery, Transformer, 5 agents}
+        \label{fig:sampling-transformer-5}
     \end{subfigure}
@@
 
-Overall, considering different training algorithms, architectures, scenarios, hyperparameters, we completed a total of $216$ runs lasting between $1$ to $8$ hours each.
+Overall, considering different training algorithms, architectures, scenarios, hyperparameters, we completed a total of $254$ runs lasting between $1$ to $8$ hours each.
 

### 2025-01-18 src/gengif.py

@@ -64,7 +64,7 @@ def run(
         checkpoint = torch.load(input_path)
         print("restoring from:", input_path)
         print("checkpoint ep :", checkpoint["episode"])
-        policy.load_state_dict(checkpoint["policy_state_dict"])
+        policy.load_state_dict({k.replace("_orig_mod.",""):v for k,v in checkpoint["policy_state_dict"].items()})
     
     frames, observation = [], torch.stack(world.reset()).transpose(0,1)
     
@@ -74,14 +74,12 @@ def run(
         total_reward = 0
         for step in range(0, steps):
             actions = policy.act(observation)["actions"]
-            print(actions)
             frames.append(world.render(mode="rgb_array"))
             observation, reward, done, info = world.step(actions.transpose(0,1))
+            print(world.scenario.max_rewards())
             observation = torch.stack(observation).transpose(0,1)
-            print(f"{torch.stack(reward).sum().item():3.2f} ", end=" ")
             total_reward += torch.stack(reward).sum().item()
         
-        print()
         print("total reward", total_reward, world.scenario.max_rewards().sum().item())
     
     

### 2025-01-18 paper/icml25/first/sects/ablation.tex
fixed main-transformer.pdf

--- 2025-01-08 paper/icml25/first/sects/ablation.tex
@@
+\begin{figure*}[!t]
+    \centering
+    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
+    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
+    \label{fig:ablation}
+\end{figure*}
+
 \section{Ablation Study}\label{sect:ablation}

### 2025-01-19 paper/icml25/first/sects/background.tex
prefer Cref on ref

@@
 where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$.
-From Equation~\ref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$.
+From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$.
 While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.
@@
 
-Finally, Figure~\ref{fig:shac} presents an unrolled view of the SHAC algorithm.
+Finally, \Cref{fig:shac} presents an unrolled view of the SHAC algorithm.

### 2025-01-19 paper/icml25/first/sects/discussion.tex
prefer Cref on ref

@@
 Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment.
-The result shown in Sect.~\ref{sect:experiments} seems to support this conclusion.
-More specifically, let us answer the research questions posed in Sect.~\ref{sect:introduction}.
+The result shown in \Cref{sect:experiments} seems to support this conclusion.
+More specifically, let us answer the research questions posed in \Cref{sect:introduction}.
 

### 2025-01-19 paper/icml25/first/sects/experiments.tex
fixed main-transformer.pdf
prefer Cref on ref

@@
 \begin{figure*}[!t]
-    \centering
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/dispersion-1-mlp.pdf}
-        \caption{Dispersion, MLP, 1 agent}
-        \label{fig:dispersion-mlp-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/dispersion-3-transformer.pdf}
-        \caption{Dispersion, Transformer, 3 agents}
-        \label{fig:dispersion-transformer-3}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/dispersion-5-transformer.pdf}
-        \caption{Dispersion, Transformer, 5 agents}
-        \label{fig:dispersion-transformer-5}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/transport-1-mlp.pdf}
-        \caption{Transport, MLP, 1 agent}
-        \label{fig:transport-mlp-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/transport-3-transformer.pdf}
-        \caption{Transport, Transformer, 3 agents}
-        \label{fig:transport-transformer-3}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/transport-5-transformer.pdf}
-        \caption{Transport, Transformer, 5 agents}
-        \label{fig:transport-transformer-5}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/discovery-1-mlp.pdf}
-        \caption{Discovery, MLP, 1 agent}
-        \label{fig:discovery-mlp-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/discovery-3-transformer.pdf}
-        \caption{Discovery, Transformer, 3 agents}
-        \label{fig:discovery-transformer-3}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/discovery-5-transformer.pdf}
-        \caption{Discovery, Transformer, 5 agents}
-        \label{fig:discovery-transformer-5}
-    \end{subfigure}
-
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/sampling-1-mlp.pdf}
-        \caption{Discovery, MLP, 1 agent}
-        \label{fig:sampling-mlp-1}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/sampling-3-transformer.pdf}
-        \caption{Discovery, Transformer, 3 agents}
-        \label{fig:sampling-transformer-3}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.30\textwidth}
-        \includegraphics[width=\textwidth]{figs/sampling-5-transformer.pdf}
-        \caption{Discovery, Transformer, 5 agents}
-        \label{fig:sampling-transformer-5}
-    \end{subfigure}
-
+    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
     \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
@@
 \begin{compactitem}
-    \item \textbf{Dispersion} (Fig.~\ref{fig:dispersion}): 
+    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
         \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
@@
         \sentence{Max Reward} The maximum reward is $n$.
-    \item \textbf{Transport} (Fig.~\ref{fig:transport}): 
+    \item \textbf{Transport} (\Cref{fig:transport}): 
         \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location.
@@
         \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
-    \item \textbf{Discovery} (Fig.~\ref{fig:discovery}): 
+    \item \textbf{Discovery} (\Cref{fig:discovery}): 
         \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly.
@@
 For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
-     \item \textbf{Sampling} (Fig.~\ref{fig:sampling}): 
+     \item \textbf{Sampling} (\Cref{fig:sampling}): 
         \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid.
@@
 We repeat and report results for $3$ runs with different seeds.
-See Table~\ref{apx:tab:ppo} and Table~\ref{apx:tab:shac} for a complete list of hyperparameters.
+See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.
 
@@
 Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion.
-We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see Fig.~\ref{apx:fig:experiments-mlp}).
-In single agent setting, show in Fig.~\ref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes.
-With $3$ agents, shown in Fig.~\ref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes.
-With $5$ agents, shown in Fig.~\ref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping.
+We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see \Cref{apx:fig:experiments-mlp}).
+In single agent setting, show in \Cref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes.
+With $3$ agents, shown in \Cref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes.
+With $5$ agents, shown in \Cref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping.
 In this last scenario, \fname{} manifest also high reward variance.
@@
 Now, we compare the performance of these algorithms on a fully differentiable scenario, transport.
-On the single agent setting, shown in Fig.~\ref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well.
+On the single agent setting, shown in \Cref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well.
 While \fname{} appears to converge earlier SHAC appear to be more consistent.
-With $3$ and $5$ agents, shown in Fig.~\ref{fig:transport-transformer-3} and \ref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.
-For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}
+With $3$ and $5$ agents, shown in \Cref{fig:transport-transformer-3} and \Cref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.
+For the MLP only setting see \Cref{apx:fig:experiments-mlp}
 
@@
 Similar consideration hold also for the discovery scenario.
-In the single agent setting, shown in Fig.~\ref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO .
-However, with $3$ and $5$ agents, shown in Fig.~\ref{fig:discovery-transformer-3} and \ref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
-For the MLP only setting see Fig.~\ref{apx:fig:experiments-mlp}
+In the single agent setting, shown in \Cref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO .
+However, with $3$ and $5$ agents, shown in \Cref{fig:discovery-transformer-3} and \Cref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
+For the MLP only setting see \Cref{apx:fig:experiments-mlp}
 

### 2025-01-19 paper/icml25/first/sects/related.tex
prefer Cref on ref

@@
 \paragraph{Differentiable Engines}
-Game engines~\cite{Bellemare13,Balla23} and Physics engines \cite{Todorov12,Coumans16} have been widely used in the field of policy learning.
+Game engines~\cite{Bellemare13,Balla23} and Physics engines~\cite{Todorov12,Coumans16} have been widely used in the field of policy learning.
 Recently, there has been a trend towards differentiable engines, which allows for the use of gradient-based optimization methods, such as backpropagation.
-Notable examples are \cite{Freeman21,Howell22,Genesis24}.
-These frameworks may be built on top of auto differentiation libraries such \cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients \cite{Carpentier18,Werling21}.
-Further, several works tackle the challenge of the non-smoothness of the contact dynamics \cite{Degrave19,Moritz20}
+Notable examples are~\cite{Freeman21,Howell22,Genesis24}.
+These frameworks may be built on top of auto differentiation libraries such~\cite{Bradbury18,Ansel24} (this is the case of VMAS) or directly with analytical gradients~\cite{Carpentier18,Werling21}.
+Further, several works tackle the challenge of the non-smoothness of the contact dynamics~\cite{Degrave19,Moritz20}
 
@@
 Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}.
-In \cite{Lv23}, the authors introduce rendering to enhance model-based RL.
+In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL.
 \cite{Zheng24} employes a differentiable symbolic expression search engine.
-Trajectory optimization approach is taken in \cite{Wan24}.
+Trajectory optimization approach is taken in~\cite{Wan24}.
 \cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.

### 2025-01-19 paper/icml25/first/sects/abstract.tex
chore: add a little bit of context in the abstract (@f14-bertolotti check)

@@
 \begin{abstract}
-In this work, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable environments.
+Reinforcement learning algorithms have demonstrated success in domains such as robotics and multi-agent systems, 
+yet they often exhibit limited sample efficiency in complex environments, 
+and a unified approach that effectively addresses both single- and multi-agent settings remains elusive.
+To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable environments.
 Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the environment.

### 2025-01-19 paper/icml25/first/sects/shacpp.tex
prefer Cref on ref
typo on shacpp

@@
 Similarly, we aim to train a neural network to approximate the transition function.
-However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar similar to Action World Model~\cite{Ma24}.
+However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}.
 This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $H$ actions from $\mathcal{A}^H$, predicts the next $H$ observations from $\mathcal{S}^H$.
@@
 To summarize, our algorithm, aims to minimize the 4 presented losses, $\mathcal{L'}_\theta^{\pi}$, $\mathcal{L}_\theta^V$, $\mathcal{L}_\theta^R$, and $\mathcal{L}_\theta^F$.
-A minimal pseudocode of our algorithm is presented in \ref{alg:shacpp}.
-Figure~\ref{fig:shacpp} presents an unrolled view of the \fname{} algorithm.
+A minimal pseudocode of our algorithm is presented in \Cref{alg:shacpp}.
+\Cref{fig:shacpp} presents an unrolled view of the \fname{} algorithm.
 

### 2025-01-20 paper/icml25/first/sects/experiments.tex
first draft for results

@@
     \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
-    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
+    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
+The results show the mean and standard deviation of normalized rewards over 3 runs.
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
     \label{fig:experiments}
@@
 The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent.
-Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurence of generalization (epoch 3200).}
+Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
 \end{figure}
@@
     \input tables/max.tex
-    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}
+    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}\label{tab:max-rewards}
 \end{table*}
@@
 \subsection{Results}
+In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
+The results show the mean and standard deviation of normalized rewards over 3 runs.
+An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}.
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.
 
 \paragraph{Dispersion}
-Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion.
-We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see \Cref{apx:fig:experiments-mlp}).
-In single agent setting, show in \Cref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes.
-With $3$ agents, shown in \Cref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes.
-With $5$ agents, shown in \Cref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping.
-In this last scenario, \fname{} manifest also high reward variance.
+In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable).
+In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation.
+In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior.
+In the third column (5 agents), PPO fails to converge, while \fname{} achieves high rewards (with no early stopping) but shows high variance due to increased environment complexity.
+
+\paragraph{Discovery} 
+The second row of \Cref{fig:experiments} shows a scenario with partial observability, namely discovery.
+In the first column (single agent), \fname{} outperforms PPO, though it does not reach the maximum reward, suggesting the world model struggles to capture dynamics.
+In the second and third columns (3 and 5 agents), PPO again fails to converge, whereas \fname{} maintains superior results, indicating cooperative behavior.
 
 \paragraph{Transport}
-Now, we compare the performance of these algorithms on a fully differentiable scenario, transport.
-On the single agent setting, shown in \Cref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well.
-While \fname{} appears to converge earlier SHAC appear to be more consistent.
-With $3$ and $5$ agents, shown in \Cref{fig:transport-transformer-3} and \Cref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.
-For the MLP only setting see \Cref{apx:fig:experiments-mlp}
-
-\paragraph{Discovery}
-Similar consideration hold also for the discovery scenario.
-In the single agent setting, shown in \Cref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO .
-However, with $3$ and $5$ agents, shown in \Cref{fig:discovery-transformer-3} and \Cref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
-For the MLP only setting see \Cref{apx:fig:experiments-mlp}
+The third row of \Cref{fig:experiments} presents a differentiable scenario, transport.
+In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.
+\fname{} converges faster, but SHAC is more consistent due to direct gradient stability.
+In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes, with SHAC again more stable.
+With more agents, convergence is faster because the task becomes easier (more agents can push the package).
+The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package.
+This is a sign of generalization, as the agents learn to avoid collisions and push the package together.
 
 \paragraph{Sampling}
+The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC.
+In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC.
+In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes.
+Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.

### 2025-01-20 paper/icml25/first/sects/discussion.tex
chore: more on discussion and first RQ1

@@
 \section{Discussion}
-
+In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO trying to answer the research questions posed in \Cref{sect:introduction}.
 Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment. 
-The result shown in \Cref{sect:experiments} seems to support this conclusion.
-More specifically, let us answer the research questions posed in \Cref{sect:introduction}.
-
 \begin{center}
@@
 
-The propose framework, \fname{}, appears to be successful in emulating the behavior of SHAC in differentiable environments such as transport and sampling.
+The proposed framework, \fname{}, appears to be successful in emulating the behavior of SHAC in differentiable environments such as transport and sampling. 
 Therefore, we can conclude that it is definitely possible to approximate the gradients of a differentiable environment with a neural network. 
 However, we note that this approach, while obtaining comparable results in most setups, it tends to display higher variance. 
-Further, before ebcoming effective, both neural network representing the reward and transition function need to become accurate which could require several episode of training.
-
+Both neural networks representing the reward and transition functions require sufficient training before becoming effective, which may entail multiple episodes.
+Interestingly, \fname{} outperforms SHAC in the Sampling scenario.
+A possible explanation is that the transition network focuses on the scenario's nuanced dynamics, whereas SHAC relies on broader environmental dynamics, creating additional gradient noise.
+\sentence{GA: francesco verifica che questa discussione può essere allienata con i risultati ottenuti, o magari c'è un'altra spiegazione che mi sono perso?}
 \begin{center}

### 2025-01-20 paper/icml25/first/sects/discussion.tex
on RQ2 on the discussion

@@
 \end{center}
-TODO
-
+The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency.
+In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), 
+it often fails to complete scenarios within the same number of steps. 
+Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}). 
+Despite having to train multiple networks, \fname{} converges much faster to optimal policies, 
+as gradient guidance enables quicker and more accurate convergence to the desired outcome. 
+This pattern becomes even more evident in multi-agent scenarios. 
+In these cases, PPO often struggles to find even marginally good policies. 
+\fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.
+\comment{GA: c'è altro che possiamo discutere rispetto a PPO? Forse dobbiamo dire che anche se converge prima è computazionalmente più costoso?}
 \begin{center}

### 2025-01-20 paper/icml25/first/sects/discussion.tex
on RQ3

@@
 \end{center}
-TODO
+To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents.
+The results demonstrate that performance patterns differ across scenarios.
+In the Dispersion scenario, \fname{} shows increasing difficulty in finding optimal policies as the number of agents grows.
+Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents.
+This counterintuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
+Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.
+
+This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents.
+Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations.
+The main bottleneck lies in the transformer architecture, whose size depends on the agent population in our solution.
+With significantly large populations (~50 agents), network sizes grow considerably, leading to prohibitively long training times.
+
+The Discovery scenario, exhibiting high variance, shows no clear pattern compared to the other three scenarios.
+This might be attributed to partial observability, which significantly complicates convergence and inter-agent cooperation.

### 2025-01-20 paper/icml25/first/sects/ablation.tex
prefer Cref on ref
on ablation

@@
 In fact, the transition function is the most complex component to replace as it requires modeling with precision the observations obtained from the environment after a number of $H$ actions from each agent.
-Avoiding the need of such network lightens the algorithm but it reintroduce the need for a differentiable environment.
+Avoiding the need of such network lightens the algorithm but it reintroduces the need for a differentiable environment.
+\Cref{fig:ablation} presents a comparative analysis between \fname{} and \fnamer{}.
+The results demonstrate that utilizing the true transition function significantly impacts the performance of \fname{}.
+This performance difference is particularly pronounced in the Sampling and Transport scenarios, where \fname{} consistently achieves superior results compared to \fnamer{}, both in terms of maximum reward and sample efficiency.
+This advantage is most evident in the Sampling scenario with one and three agents.
 
-Fig.~\ref{fig:ablation} summarizes the results.
-Most notably, using the true transition function does not appear to provide any significant advantage in terms of reward achieved.
-In fact, in some cases, we report better results for \fname{}, namely in Fig.~\ref{fig:transport-ablation-1-mlp}.
+However, the performance gap narrows in certain scenarios.
+For instance, in the Dispersion scenario, \fnamer{} achieves comparable performance to \fname{}.
+This equivalence might be attributed to the increased complexity of learning the transition function in this particular environment.
+\comment{GA: qui ho cambiato l'ablation perché ora mi sembra che shac++ in realtà vada meglio negli scenari aggiunti, che ne dici?}

### 2025-01-20 paper/icml25/first/sects/conclusion.tex
on conclusion -- first draft!

\section{Conclusion}
In this paper, we proposed \fname{}, an extension of SHAC, 
to provide a reinforcement learning solution that is both sample efficient,
able to work in both single-agent and multi-agent scenarios,
and capable of handling non-differentiable scenarios. 
We demonstrated that this solution maintains alignment with SHAC (\RQ{1}), outperforms the current state-of-the-art PPO in both single-agent and multi-agent scenarios (\RQ{2}), 
and exhibits promising scaling capabilities in multi-agent settings with emergent collaborative behaviors (\RQ{3}).

Despite these encouraging results, several directions for future work remain. 
First, we aim to explore scalability beyond the current limitation of tens of agents. 
Additionally, as evidenced in the discovery scenario, the current approach struggles when agents have highly partial observations of the environment. 
Furthermore, we plan to evaluate our approach in other multi-agent environments, such as those presented from the cooperative AI community~\cite{DBLP:journals/corr/abs-2211-13746} and also in adversarial settings.

### 2025-01-20 paper/icml25/first/sects/abstract.tex
minor on abstract

@@
 These neural networks are trained alongside the policy and value networks of SHAC.
-We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO.
+We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO---a current state-of-the-art algorithm for non-differentiable environments.
 Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to environment gradients.

### 2025-01-20 paper/icml25/first/sects/abstract.tex
on abstract

@@
 and a unified approach that effectively addresses both single- and multi-agent settings remains elusive.
-To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable environments.
-Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the environment.
+To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable simulators.
+Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the simulator.
 These neural networks are trained alongside the policy and value networks of SHAC.

### 2025-01-20 paper/icml25/first/sects/background.tex
revised bg

--- 2025-01-13 paper/icml25/first/sects/background.tex
@@
+\newcommand{\agents}{\mathcal{N}}
+
 \section{Background \& Notation}\label{sect:background}
@@
 To provide context for our work, we first introduce the reinforcement learning (RL) framework.
-Each task is modeled as a finite-horizon Markov Decision Process (MDP) $\mathcal{M} = (\mathcal{S}, \mathcal{A}, F, R, T, \gamma, \mu)$, where $\mathcal{S}$ represents the state space, $\mathcal{A}$ denotes the action space, $F:\mathcal{S} \times \mathcal{A} \rightarrow \mathcal{S}$ is the transition function, $R:\mathcal{S} \times \mathcal{A} \times \mathcal{S} \rightarrow \mathbb{R}$ is the reward function, $\gamma \in [0, 1]$ is the discount factor, and $T \in \mathbb{N}$ is the time horizon.
-The initial state distribution is denoted by $\mu$.
+Each task is modeled as a partially observable, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$:
+$$(\agents,\states,\{\observations{i}\}_{\scriptscriptstyle i\in\agents},\{\actions{i}\}_{\scriptscriptstyle i\in\agents},\{\stateobs{i}\}_{\scriptscriptstyle i\in\agents}, \{\rewards{i}\}_{\scriptscriptstyle i\in\agents}, \transition, \horizon, \discount, \initdist)$$
+Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively.
+$\observations{i}$ and $\actions{i}$ are the observation space and action space for the $i$-th agent ($i\in\agents$).
+$\stateobs{i}:\states\rightarrow\observations{i}$ is the projection to the observation space and $\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}$ is the reward function for the $i$-th agent.
+$\transition:\states\times\{\actions{i}\}_{i\in\agents}$ denotes the transition function.
+$\discount$ and $\horizon$ denote the discount factor and the time horizon respectively.
+Finally, $\initdist$ is the initial state distribution.
 
-Further, let $\mathcal{T}$ denote the space of trajectories, defined as $\mathcal{T} = (\mathcal{S} \times \mathcal{A} \times \mathcal{S} \times \mathbb{R})^*$, where each trajectory consists of a sequence of starting states, actions, subsequent states, and rewards.
-We will use the symbols $\tau=((s_0, a_0, s_1, r_0), (s_1, a_1, s_2, r_1), \ldots)$ to denote a generic trajectory in $\mathcal{T}$.
-When dealing with multiple trajectories, we will use the notation $s_i^\tau, a_i^\tau$, and $r_i^\tau$.
+Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\{\actions{i}\}_{i\in\agents}\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
+Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.
 
-A policy is a function $\pi:\mathcal{S}\rightarrow\mathcal{A}$ that maps states to actions.
-We can use the policy to sample trajectories from the MDP by iteratively applying the transition and reward function and the policy.
-In this case, we have that the $i$-th element of a trajectory is $(s_i, a_i, s_{i+1}, r_i) = (s_i, \pi(s_i), F(s_i, a_i), R(s_i,a_i,F(s_i, a_i))$.
-The initial state $s_0$ is sampled from the initial state distribution $\mu$.
-
-For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled by following the policy $\pi$.
+A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space.
+We will denote the collective policy, determined by the collection of agent-specific policy simply with $\pi$.
+We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions.
+In this case, for trajectory $\tau$, we have that the $i$-th action for the $j$-th agent is $a_{ij}^\tau = \policy{j}(\stateobs{j}(s_i^\tau))$ and the next state is $s_{i+1}^\tau = \transition(s_i^\tau,a_{i1},\dots,a_{i\na})$.
+The $i$-th reward for the $j$-th agent is $r_{ij}^\tau = \rewards{j}(s_i^\tau,a_{ij}^\tau,s_{i+1}^\tau)$.
+Finally, $s_0$ is sampled from the initial state distribution $\initdist$.
+For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.
 
 The goal of the MDP is to find a policy that maximizes the expected return 
-$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{T-1}\gamma^t r_t^\tau \right]$$
+$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]$$
 
@@
 
-In this work, we will often approximate functions, such as transition ($F$) or reward ($R$), with neural network.
+In this work, we will often approximate functions (such as transition $\transition$) with neural network.
 Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$.
@@
 Each network can either use its own subset of $\theta$ or it can share parameters with other networks.
-In practice, the parameters will never be shared.
+In practice, the parameters will be shared only for the policies.
+
+For simplicity, we formally treat only the case where the policy observe the full state to make the action, meaning that $\stateobs{i}$ is the identity function.
+However, in our experiments, we consider also scenarios where the full state is not available to the policies.
 
@@
 
-To solve an MDP $\mathcal{M}$, SHAC trains a policy $\pi_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\mathcal{T}^N\rightarrow\mathbb{R}$ for a batch of $N$ trajectories $B = \{\tau_1, \ldots, \tau_N\}$ sampled from the MDP.
-The loss is defined as:
-\begin{equation}\label{eq:shac}
-    \mathcal{L}_\theta^{\pi}(B) = \frac{1}{NH}\sum_{\tau\in B} \sum_{t=t_0}^{t_0+H-1} \gamma^{t-t_0} R(s_t^\tau, a_t^\tau) + \gamma^H V_\theta(s^\tau_{t_0+T})
-\end{equation}
-where $N$ is the number of sampled trajectories, $H \ll T$ is a time horizon, $s_t^\tau$ is the state at time $t$ for trajectory $\tau$, $a_t^\tau$ is the action taken at time $t$ for trajectory $\tau$, and $V(s): \mathcal{S} \rightarrow \mathbb{R}$ is the value function estimating the expected return starting from state $s$.
-From Equation~\ref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, and transition functions are all differentiable, i.e., $\pi_\theta,V_\theta,F,R \in \mathcal{C}^1$.
+In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$.
+The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:
+\begin{equation*}\label{eq:shac}
+    \frac{1}{\bs\na\mh}\sum_{\tau\in\batch} \sum_{t=t_0}^{t_0+\mh-1} \sum_{i\in\agents} \discount^{t-t_0} R(s_{t}^\tau, a_{ti}^\tau) + \discount^\mh \valuef_\theta(s^\tau_{t_0+\mh})
+\end{equation*}
+where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
+From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.
 While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.
 
-The value function $V_\theta$ is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $V$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{V}:\mathcal{T}^N\rightarrow\mathbb{R}$.
+The value function $\valuef_\theta$ is trained to approximate a temporal difference-$\lambda$ formulation \cite{Sutton98} $\valuef$, which in SHAC translates to minimize the loss $\mathcal{L}_\theta^{\valuef}:\trajectories^\bs\rightarrow\mathbb{R}$.
 
-$$ \mathcal{L}_\theta^V(B)=\frac{1}{NT}\sum_{\tau\in B}\sum_{t=0}^T\left|\left| V_\theta(s^\tau_t) - V(s^\tau_t) \right|\right|^2 $$
-$$ V(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
-$$ G_t^k = \left(\sum_{l=0}^{k-1}\gamma^l r_{t+l}\right) + \gamma^k V(s_{t+k})$$
+$$ \mathcal{L}_\theta^\valuef(\batch)=\frac{1}{\bs\mh}\sum_{\tau\in\batch}\sum_{t=0}^\mh\left|\left| \valuef_\theta(s^\tau_t) - \valuef(s^\tau_t) \right|\right|^2 $$
+$$ \valuef(s_t) = (1-\lambda) \left(\sum_{k=1}^{h-t-1}\lambda^{k-1}G_t^k\right) + \lambda^{h-t-1}G_t^{h-t}$$
+$$ G_t^k = \left(\sum_{l=t}^{k-1+t}\sum_{i\in\agents}\gamma^{l-t} r_{li}\right) + \gamma^k V(s_{t+k})$$
 
@@
 
-Finally, Figure~\ref{fig:shac} presents an unrolled view of the SHAC algorithm.
+Finally, \Cref{fig:shac} presents an unrolled view of the SHAC algorithm for a single-agent scenario.

### 2025-01-20 paper/icml25/first/sects/shacpp.tex
revised bg

@@
 In particular, we aim to train a neural network to approximate the reward function by minimizing:
-$$ \mathbb{E}_{\tau\sim\pi}\left[\left|\left| \frac{1}{T}\sum_{i=0}^{T-1} R_{\theta}(s_i^\tau, a_i^\tau, s_{i+1}^\tau) - R(s_i^\tau, a_i^\tau, s_{i+1}^\tau) \right|\right|^2\right] $$
+\begin{align*}
+    \mathbb{E}_{\tau\sim\pi}\left[\left\vert\left\vert \frac{1}{\horizon\na}\sum_{j=0}^{\horizon-1} \sum_{i\in\agents} {\rewards{i}}_\theta(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right.\right.\right.
+-\\ \left.\left.\left.
+\rewards{i}(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right\vert\right\vert^2\right]
+\end{align*}
 
@@
 The neural network is then trained to minimize the mean squared error (MSE) loss.
-Specifically, we minimize the loss $\mathcal{L}_\theta^{R}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$ \mathcal{L}_\theta^R(B) = \frac{1}{N}\sum_{s,a,r \in B}\left|\left| R_\theta(s, a) - r \right|\right|^2 $$
+Specifically, we minimize the loss $\mathcal{L}_\theta^{\rewards{i}}:(\states\times\actions{i}\times\states)^N\rightarrow\mathbb{R}$:
+$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,r \in \batch}\left|\left| {\rewards{i}}_\theta(s, a) - r \right|\right|^2 $$
 Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.
@@
 However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}.
-This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $H$ actions from $\mathcal{A}^H$, predicts the next $H$ observations from $\mathcal{S}^H$.
+This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $(\{\actions{i}\}_{i\in\agents})^\mh$, predicts the next $\mh$ observations from $\states^\mh$.
 We aim to minimize:
-
-\begin{equation}
-\begin{aligned}
-\mathbb{E}_{\tau\sim\pi}\left[\frac{1}{NH}\sum_{t=0}^{T-H-1}\sum_{i=0}^{H-1} \left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - \right.\right.\right.
-\\ \left.\left.\left.
-s_{t+i+1}^\tau \right|\right|^2\right]
-\end{aligned}
-\end{equation}
+%\sum_{j=0}^{\scriptscriptstyle\mh-1}
+\begin{align*}
+    \mathbb{E}_{\tau\sim\pi}\left[\frac{1}{\bs\mh}\sum_{t=0}^{\scriptscriptstyle\horizon-\mh-1}\left|\left| \transition_{\theta}(s_t^\tau, \{a^\tau_t\}_{i}^{\na},\dots,\{a^\tau_{\scriptscriptstyle t+H-1}\}_{i}^{\na}) \right.\right.\right.-\\\left.\left.\left.
+\left[s_t^\tau,\dots,s_{t+h}^\tau\right] \right|\right|^2\right]
+\end{align*}
 
@@
 That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$ \frac{1}{NH}\sum_{\tau \in B}\sum_{t=t_0}^{t_0+H-1}\left|\left| F_{\theta}(s_t^\tau, a_t^\tau, \dots, a_{t+H-1}^\tau)_i - s_{t+1}^\tau \right|\right|^2 $$
-Where $B$ represents a batch of trajectories with horizon $H$ sampled uniformly from the cache.
+$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, \{a_1^\tau\}_i^\na, \dots, \{a_{\mh}^\tau\}_i^\na) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
+Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.
 
 Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space.
-That is, suppose the action space is $(a,b)$ (with $a<b$), we minimize the loss $\mathcal{L'}_\theta^\pi:$
+Suppose the action space is $(l,r)$ (with $l<r$) for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$
 \begin{figure*}[!t]
@@
 
-$$\mathcal{L}_\theta^\pi(B) + \alpha \frac{1}{NH}\sum_{\tau\in B}\sum_{t=t_0}^{t_0+H-1} d(a_t^\tau,(a,b))$$
-$$d(x,(a,b)) = \begin{cases}(a-x)^2 & \text{if } x < a \\ (x-b)^2 & \text{if } x > b \\ 0 & \text{otherwise} \end{cases}$$
+$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(a,b))$$
+$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > e \\ 0 & \text{otherwise} \end{cases}$$
 

### 2025-01-21 paper/icml25/first/appendices/ablation.tex
refactoring

--- 2025-01-20 paper/icml25/first/sects/ablation.tex
@@
 
-\section{Ablation Study}\label{sect:ablation}
+\section{Ablation Study}\label{apx:ablation}
 While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.
@@
 This equivalence might be attributed to the increased complexity of learning the transition function in this particular environment.
-\comment{GA: qui ho cambiato l'ablation perché ora mi sembra che shac++ in realtà vada meglio negli scenari aggiunti, che ne dici?}

### 2025-01-21 paper/icml25/first/appendices/arch.tex
refactoring

\section{Architectures}\label{apx:arch}

\begin{table}[t]
        \centering
        \input{tables/shacwm.tex}
        \caption{SHAC/\fname{}/\fnamer{} Hyperparameters}
        \label{apx:tab:shac}
\end{table}

\begin{table}[t]
        \centering
        \input{tables/ppo.tex}
        \caption{PPO Hyperparmeters}
        \label{apx:tab:ppo}
\end{table}

### 2025-01-21 paper/icml25/first/appendices/scenarios.tex
refactoring

--- 2025-01-19 paper/icml25/first/sects/experiments.tex
@@
-\begin{figure*}[!t]
-    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
-    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios for the Transformer Architecture.}
-    \label{fig:experiments}
-\end{figure*}
+\section{Environments}\label{apx:scenarios}
 
-\begin{figure}[t]
-    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
-    \caption{Norm of policy gradients on the Transport environment across 5000 epochs.
-The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent.
-Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurence of generalization (epoch 3200).}
-\end{figure}
-
-\section{Experiments}\label{sect:experiments}
-
-\paragraph{Scenarios}
-As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO.
-In particular, we choose the following scenario:
 \begin{compactitem}
@@
 \end{compactitem}
-
-\begin{table*}[!t]
-    \centering
-    \input tables/max.tex
-    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}
-\end{table*}
-
-\paragraph{Architectures}
-We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture.
-Specifically, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 for the Policy Network and Value Network (for PPO, SHAC, and \fname{}).
-For the Reward Network, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 (for \fname{}).
-The Transition Network for \fname{} is a 3-layer single-head Transformer with hidden size 64.
-
-While the MLP architecture is the simplest, the Transformer baseline benefits from position invariance property of the agents observations.
-We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.
-
-\paragraph{Hyperparameters}
-For all networks we use learning rate of $1e\text{-}3$ with Adam Optimizer~\cite{Kingma14}.
-For all scenarios, each training episode consists of $512$ environments of $32$ steps each.
-Each validation episode is composed of $512$ environments of $512$ steps each.
-We employ early stopping when the agents achieve $90\%$ of the maximum reward in $90\%$ of the episode's environments.
-If early stopping is not triggered, we stop training after $20,000$ episodes.
-The discount factor is set to $0.99$ and lambda factor is set to $0.95$.
-We report results for increasing number of agents $n\in\{1,3,5\}$.
-We repeat and report results for $3$ runs with different seeds.
-See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.
-
-Overall, considering different training algorithms, architectures, scenarios, hyperparameters, we completed a total of $254$ runs lasting between $1$ to $8$ hours each.
-
-\paragraph{Hardware Setup}
-Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one Nvidia Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).
-
-\subsection{Results}
-
-\paragraph{Dispersion}
-Firstly, we compare the performance of these algorithms on a non-differentiable scenario (thus SHAC is not aplicable), dispersion.
-We show the comparison between \fname{} and PPO for increasing number of agents for the transformer architecture (see \Cref{apx:fig:experiments-mlp}).
-In single agent setting, show in \Cref{fig:dispersion-mlp-1}, both algorithm perform well, albeit \fname{} tends to converge after fewer episodes.
-With $3$ agents, shown in \Cref{fig:dispersion-transformer-3}, PPO fails to converge, meanwhile \fname{} converges in less than $10,000$ episodes.
-With $5$ agents, shown in \Cref{fig:dispersion-transformer-5}, PPO fails to converge, meanwhile \fname{} achieves high rewards without however triggering the early stopping.
-In this last scenario, \fname{} manifest also high reward variance.
-
-\paragraph{Transport}
-Now, we compare the performance of these algorithms on a fully differentiable scenario, transport.
-On the single agent setting, shown in \Cref{fig:transport-mlp-1}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well.
-While \fname{} appears to converge earlier SHAC appear to be more consistent.
-With $3$ and $5$ agents, shown in \Cref{fig:transport-transformer-3} and \Cref{fig:transport-transformer-5}, PPO fails to converge, meanwhile both SHAC and \fname{} perform similarly well albeit SHAC is more stable.
-For the MLP only setting see \Cref{apx:fig:experiments-mlp}
-
-\paragraph{Discovery}
-Similar consideration hold also for the discovery scenario.
-In the single agent setting, shown in \Cref{fig:discovery-mlp-1}, \fname{} performs better, but comparitavely, to PPO .
-However, with $3$ and $5$ agents, shown in \Cref{fig:discovery-transformer-3} and \Cref{fig:discovery-transformer-5}, PPO fails to converge, meanwhile \fname{} performs better.
-For the MLP only setting see \Cref{apx:fig:experiments-mlp}
-
-\paragraph{Sampling}

### 2025-01-21 paper/icml25/first/sects/background.tex
refactoring

@@
-\newcommand{\agents}{\mathcal{N}}
-
 \section{Background \& Notation}\label{sect:background}
@@
 To provide context for our work, we first introduce the reinforcement learning (RL) framework.
-Each task is modeled as a partially observable, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$:
-$$(\agents,\states,\{\observations{i}\}_{\scriptscriptstyle i\in\agents},\{\actions{i}\}_{\scriptscriptstyle i\in\agents},\{\stateobs{i}\}_{\scriptscriptstyle i\in\agents}, \{\rewards{i}\}_{\scriptscriptstyle i\in\agents}, \transition, \horizon, \discount, \initdist)$$
+Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$.
 Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively.
-$\observations{i}$ and $\actions{i}$ are the observation space and action space for the $i$-th agent ($i\in\agents$).
-$\stateobs{i}:\states\rightarrow\observations{i}$ is the projection to the observation space and $\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}$ is the reward function for the $i$-th agent.
-$\transition:\states\times\{\actions{i}\}_{i\in\agents}$ denotes the transition function.
+$\allo=\{\observations{i}\}_{i\in\agents}$ and $\alla=\{\actions{i}\}_{i\in\agents}$ are the observation space and action space for each agent.
+$\{\stateobs{i}:\states\rightarrow\observations{i}\}_{i\in\agents}$ are the projections from the state to the observation space and $\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}_{i\in\agents}$ are the reward functions for each agent.
+$\transition:\states\times\alla$ denotes the transition function.
 $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively.
@@
 
-Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\{\actions{i}\}_{i\in\agents}\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
+Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
 Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.
@@
 A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space.
-We will denote the collective policy, determined by the collection of agent-specific policy simply with $\pi$.
+We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi$.
 We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions.
@@
 
-In this work, we will often approximate functions (such as transition $\transition$) with neural network.
+In this work, we will often approximate functions (such as transition $\transition$) with neural networks.
 Therefore, for each function $g$, we will denote the corresponding neural network with $g_\theta$.
@@
 
-For simplicity, we formally treat only the case where the policy observe the full state to make the action, meaning that $\stateobs{i}$ is the identity function.
+For simplicity, we formally treat only the case where the policies observe the full state, meaning that $\stateobs{i}$ is the identity function.
 However, in our experiments, we consider also scenarios where the full state is not available to the policies.
@@
 \end{equation*}
-where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
-From \Cref{eq:shac}, it is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.
-While this is an common requirement for policy and value functions, differentiable transition and reward functions are rarely available.
+where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
+It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.
+While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.
 

### 2025-01-21 paper/icml25/first/sects/experiments.tex
minor on table 1
refactoring

@@
     \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
-The results show the mean and standard deviation of normalized rewards over 3 runs.
+The results show the mean and standard deviation of rewards over 3 runs.
 For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
@@
 
+In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures.
+An ablation study when replacing only the transition function (meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.
+
 \paragraph{Scenarios}
-As mentioned previously, we choose a few multi-agent scenarios from the VMAS\footnote{\url{https://github.com/proroklab/VectorizedMultiAgentSimulator}} package to test \fname{} aganist SHAC and PPO.
-In particular, we choose the following scenario:
-\begin{compactitem}
-    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
-        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
-Agents observe only their own position, velocity and relative position to each goal.
-        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
-        \sentence{World Model} The world model has enough information to predict the next states accurately.
-        \sentence{Max Reward} The maximum reward is $n$.
-    \item \textbf{Transport} (\Cref{fig:transport}): 
-        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location.
-Agents observe their own position, velocity, relative position to the package and relative position between package and the goal.
-The agents receive a reward proportional to the distance between package and the goal.
-        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
-        \sentence{World Model} The world model has enough information to predict the next states accurately.
-        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
-    \item \textbf{Discovery} (\Cref{fig:discovery}): 
-        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly.
-A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$.
-If agents collide with each other, they receive a penalty.
-Every agent observes its position, velocity, lidar measurements to other agents and goals.
-When $n=1$ we set $s=1$ and $k=7$.
-When $n>1$ we set $s=2$.
-        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
-        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
-        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents.
-For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
-     \item \textbf{Sampling} (\Cref{fig:sampling}): 
-        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid.
-The reward are sampled from $k$ gaussians.
-Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.
-        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
-        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
-        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
-\end{compactitem}
+As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO.
+In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling.
+For details regarding each scenario, see \Cref{apx:scenarios}.
+Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable.
+Transport and Sampling have differentiable reward functions, thus SHAC is applicable.
+For Transport and Dispersion, the observation space is complete wrt.
+the state.
+For Discovery and Sampling, the observation space is incomplete wrt.
+the state.
+In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC.
 
@@
     \input tables/max.tex
-    \caption{Normalized maximum rewards (relative to the best performing model) for the different scenarios.}\label{tab:max-rewards}
+    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
 \end{table*}
@@
 We test SHAC, PPO, and \fname{} with both an MLP and a Transformer architecture.
-Specifically, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 for the Policy Network and Value Network (for PPO, SHAC, and \fname{}).
-For the Reward Network, we use a 1-layer MLP with hidden size 64 or a 1-layer single-head Transformer with hidden size 64 (for \fname{}).
-The Transition Network for \fname{} is a 3-layer single-head Transformer with hidden size 64.
+Specifically, we use a 1-layer MLP or a 1-layer single-head Transformer for policy, reward, and value network.
+The transition network used by \fname{} is always a 3-layer single-head Transformer
 
-While the MLP architecture is the simplest, the Transformer baseline benefits from position invariance property of the agents observations.
+While the MLP architecture is the simplest, the Transformer baseline benefits from positional invariance property of the agents observations.
 We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.
+For more details on the architectures, see \Cref{apx:arch}.
 
@@
 In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
-The results show the mean and standard deviation of normalized rewards over 3 runs.
+The results show the mean and standard deviation of achieved rewards over 3 runs.
 An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}.
@@
 \fname{} converges faster, but SHAC is more consistent due to direct gradient stability.
-In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes, with SHAC again more stable.
+In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes.
 With more agents, convergence is faster because the task becomes easier (more agents can push the package).

### 2025-01-21 paper/icml25/first/sects/shacpp.tex
refactoring

@@
 \begin{align*}
-    \mathbb{E}_{\tau\sim\pi}\left[\left\vert\left\vert \frac{1}{\horizon\na}\sum_{j=0}^{\horizon-1} \sum_{i\in\agents} {\rewards{i}}_\theta(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right.\right.\right.
--\\ \left.\left.\left.
-\rewards{i}(s_j^\tau, a_j^\tau, s_{j+1}^\tau) \right\vert\right\vert^2\right]
+    \underset{\tau\sim\pi}{E}\left[\hspace{-.7cm}\sum_{\hspace{.7cm}\scriptscriptstyle s_i,a_{ij},s_{i+1}\in\tau} \hspace{-.8cm}\left\vert\left\vert {\rewards{i}}_\theta(s_i, a_{ij}, s_{i+1}) -\rewards{i}(s_i, a_{ij}, s_{i+1}) \right\vert\right\vert^2\right]
 \end{align*}
@@
 Specifically, we minimize the loss $\mathcal{L}_\theta^{\rewards{i}}:(\states\times\actions{i}\times\states)^N\rightarrow\mathbb{R}$:
-$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,r \in \batch}\left|\left| {\rewards{i}}_\theta(s, a) - r \right|\right|^2 $$
-Where $B$ represents a batch of state-action-reward triplets sampled uniformly from the cache.
+$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,s',r \in \batch}\left|\left| {\rewards{i}}_\theta(s, a) - r \right|\right|^2 $$
+Where $B$ represents a batch of state-action-state-reward tuples sampled uniformly from the cache.
 
@@
 However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}.
-This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $(\{\actions{i}\}_{i\in\agents})^\mh$, predicts the next $\mh$ observations from $\states^\mh$.
+This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ observations from $\states^\mh$.
 We aim to minimize:
-%\sum_{j=0}^{\scriptscriptstyle\mh-1}
+
 \begin{align*}
-    \mathbb{E}_{\tau\sim\pi}\left[\frac{1}{\bs\mh}\sum_{t=0}^{\scriptscriptstyle\horizon-\mh-1}\left|\left| \transition_{\theta}(s_t^\tau, \{a^\tau_t\}_{i}^{\na},\dots,\{a^\tau_{\scriptscriptstyle t+H-1}\}_{i}^{\na}) \right.\right.\right.-\\\left.\left.\left.
-\left[s_t^\tau,\dots,s_{t+h}^\tau\right] \right|\right|^2\right]
+    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left|\left| \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right|\right|^2\right]
 \end{align*}
 
+Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$.
 In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss.
 That is, we minimize the loss $\mathcal{L}_\theta^{F}:\mathcal{T}^N\rightarrow\mathbb{R}$:
-$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, \{a_1^\tau\}_i^\na, \dots, \{a_{\mh}^\tau\}_i^\na) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
+$$ \frac{1}{\bs\mh}\sum_{\tau \in B}\left|\left| \transition_{\theta}(s_0^\tau, a_{1*}^\tau, \dots, a_{\mh*}) - \left[ s^\tau_1, \dots, s^\tau_\mh \right] \right|\right|^2 $$
 Where $B$ represents a batch of trajectories of length $\mh$ sampled uniformly from the cache.
@@
 Suppose the action space is $(l,r)$ (with $l<r$) for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$
-\begin{figure*}[!t]
-    \centering
-    \begin{subfigure}[b]{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/dispersion.png}
-        \caption{Dispersion}
-        \label{fig:dispersion}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/transport.png}
-        \caption{Transport}
-        \label{fig:transport}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/discovery.png}
-        \caption{Discovery}
-        \label{fig:discovery}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/sampling.png}
-        \caption{Sampling}
-        \label{fig:sampling}
-    \end{subfigure}
-    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
-\end{figure*}
 
 $$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(a,b))$$
-$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > e \\ 0 & \text{otherwise} \end{cases}$$
+$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$
 

### 2025-01-21 paper/icml25/first/sects/discussion.tex
revised conclusion and discussion

@@
 Interestingly, \fname{} outperforms SHAC in the Sampling scenario.
-A possible explanation is that the transition network focuses on the scenario's nuanced dynamics, whereas SHAC relies on broader environmental dynamics, creating additional gradient noise.
-\sentence{GA: francesco verifica che questa discussione può essere allienata con i risultati ottenuti, o magari c'è un'altra spiegazione che mi sono perso?}
+A possible explanation is that the complex dynamics of the environment may lead to unstable gradients.
+
 \begin{center}
@@
 The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency.
-In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), 
-it often fails to complete scenarios within the same number of steps. 
+In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps.
 Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}).
-Despite having to train multiple networks, \fname{} converges much faster to optimal policies, 
-as gradient guidance enables quicker and more accurate convergence to the desired outcome. 
+Despite having to train multiple networks, \fname{} converges much faster to optimal policies, as gradient guidance enables quicker and more accurate convergence to the desired outcome.
 This pattern becomes even more evident in multi-agent scenarios.
@@
 \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.
-\comment{GA: c'è altro che possiamo discutere rispetto a PPO? Forse dobbiamo dire che anche se converge prima è computazionalmente più costoso?}
+
 \begin{center}
@@
 Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations.
-The main bottleneck lies in the transformer architecture, whose size depends on the agent population in our solution.
-With significantly large populations (~50 agents), network sizes grow considerably, leading to prohibitively long training times.
+The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.
+agents times num.
+train steps).
+With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively long memory requirements.
+In these scenarios, a linear attention mechanism (such as \cite{Beltagy20}) may be employed.
 

### 2025-01-21 paper/icml25/first/sects/experiments.tex
fix

@@
 
-\paragraph{Transport}
+\paragraphTransport}
 The third row of \Cref{fig:experiments} presents a differentiable scenario, transport.
@@
 In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes.
-With more agents, convergence is faster because the task becomes easier (more agents can push the package).
+With more agents, convergence is faster because the task becomes easier (as more agents can push the package).
 The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package.

### 2025-01-21 paper/icml25/first/appendices/additional.tex
minors + appendix (WIP)

\section{Additional Experiments}

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{figs/main-mlp.pdf}
    \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the Transformer architecture see \Cref{fig:experiments}.
    }
    \label{apx:fig:experiments-mlp}
\end{figure}

\begin{figure*}[!t]
    \centering
    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
    \label{fig:ablation}
\end{figure*}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{figs/ablation-mlp.pdf}
    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
    \label{fig:ablation}
\end{figure}

### 2025-01-21 paper/icml25/first/appendices/scenarios.tex
minors + appendix (WIP)

@@
+\begin{figure}[!t]
+    \centering
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion.png}
+        \caption{Dispersion}
+        \label{fig:dispersion}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport.png}
+        \caption{Transport}
+        \label{fig:transport}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery.png}
+        \caption{Discovery}
+        \label{fig:discovery}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling.png}
+        \caption{Sampling}
+        \label{fig:sampling}
+    \end{subfigure}
+    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
+\end{figure}
+
 \section{Environments}\label{apx:scenarios}
 
-\begin{compactitem}
-    \item \textbf{Dispersion} (\Cref{fig:dispersion}): 
-        \sentence{Environment} a scenario where $n$ agents need to reach $n$ goal locations, in such case the agent receives a sparse reward of $1$.
-Agents observe only their own position, velocity and relative position to each goal.
-        \sentence{Differentiability} Despite the transition function being differentiable, the reward function is not. Therefore, only PPO and \fname{} are applicable to this scenario. 
-        \sentence{World Model} The world model has enough information to predict the next states accurately.
-        \sentence{Max Reward} The maximum reward is $n$.
-    \item \textbf{Transport} (\Cref{fig:transport}): 
-        \sentence{Environment} a scenario where $n$ agents need to collaborate to push a package into a goal location.
-Agents observe their own position, velocity, relative position to the package and relative position between package and the goal.
-The agents receive a reward proportional to the distance between package and the goal.
-        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
-        \sentence{World Model} The world model has enough information to predict the next states accurately.
-        \sentence{Max Reward} The maximum reward varies from environement to environment. It is the scaled distance between the package and the goal times the number of agents.
-    \item \textbf{Discovery} (\Cref{fig:discovery}): 
-        \sentence{Environment} a scenario where $n$ agents need to collect $k$ goal place randomly.
-A goal is collected if at least $s$ agents stand in proximity to it, in this case they receive a reward of $1$.
-If agents collide with each other, they receive a penalty.
-Every agent observes its position, velocity, lidar measurements to other agents and goals.
-When $n=1$ we set $s=1$ and $k=7$.
-When $n>1$ we set $s=2$.
-        \sentence{Differentiability} In this scenario, only the transition function is differentiable. Therefore, only PPO and \fname{} are applicable to this scenario. 
-        \sentence{Max Reward} The maximum rewards depends on the number evaluation steps, but for early stopping purposes, we set it to $k$. 
-        \sentence{World Model} Note that, a world model (or an action world model) that uses only the agents observation cannot predict the evolution correctly, as goals may not be visible to the agents.
-For this reason, this scenario is particularly challenging for \fname{} as we do not provide the full world state to the world model.
-     \item \textbf{Sampling} (\Cref{fig:sampling}): 
-        \sentence{Environment} a scenario where $n$ agents need to collect rewards from a grid.
-The reward are sampled from $k$ gaussians.
-Agents observe their own position, velocity, lidar values to each other, and rewards in a 3x3 grid around themselves.
-        \sentence{Differentiability} In this scenario, both the reward and transition functions are differentiable. 
-        \sentence{Max Reward} The maximum reward depends on where the gaussians are placed.
-        \sentence{World Model} Similarly to discovery, this scenario is particularly challenging for \fname{} as we do not provide the full world state, but only the agents observations, to the world model.
-\end{compactitem}
+In this section, we provide a detailed discussion of the environment and scenarios used to test the \fname{} algorithm. Specifically, we utilize four vectorized environments available as part of VMAS.
+
+VMAS is a differentiable simulator designed for various multi-agent tasks.
+These tasks are built on top of a differentiable physics engine, resulting in differentiable transition functions for all tasks.
+However, the reward function is non-differentiable.
+To address this limitation, we created custom reward functions, where possible, to replicate the non-differentiable ones provided by VMAS.
+Consequently, some scenarios are compatible with all tested algorithms (PPO, SHAC, and \fname{}), while others are only applicable to PPO and \fname{}.
+
+Furthermore, the observations produced by VMAS are often only partial representations of the full state.
+While in some cases these observations are sufficient to predict the next observations, in others they are not.
+In such cases, \fname{} is expected to face challenges, as the world model is trained using only partial information.
+In contrast, SHAC computes gradients using access to the full world state.
+We chose to retain this limitation to better reflect the behavior of \fname{} in more realistic environments.
+
+\subsection{Dispersion}
+In the Dispersion task, $n$ agents must reach $n$ randomly placed goals. In \Cref{fig:dispersion}, the agents are represented by blue dots, while the goals are depicted as green dots.
+
+Each agent has a continuous 2-dimensional action space, bounded between $-1$ and $1$, to represent acceleration along the x- and y-axes, respectively.
+Each agent observes its own position, velocity, and relative position to each goal.
+As a consequence, the world model has sufficient information to accurately predict the next states.
+
+Each agent receives a reward of $1$ upon reaching a goal, making the maximum possible reward $n$.
+While the transition function is differentiable, the reward function is not.
+Consequently, only PPO and \fname{} are applicable to this scenario.
+
+\subsection{Transport}
+In the Transport problem, $n$ agents work together to push a package to a randomly placed target location.
+In \Cref{fig:transport}, the agents are represented by blue dots, the package is shown as a red square, and the goal is depicted as a green circle.
+The more agents collaborate to push the package, the faster they can reach the goal and achieve the maximum reward.
+
+Each agent has a continuous 2-dimensional action space, bounded between $-1$ and $1$, to represent acceleration along the x- and y-axes, respectively.
+Each agent observes its own position, velocity, the relative position to the package, and the relative position between the package and the goal.
+As a consequence, the world model has sufficient information to accurately predict the next states.
+
+Each agent receives a reward proportional to the distance between the initial position of the package and the goal.
+Consequently, the maximum reward corresponds to the distance between the package and the goal.
+Since the maximum reward varies across environments, we only plot a reference line representing good performance.
+Although this reward function is differentiable, the implementation provided by VMAS is not.
+Therefore, we created a custom reward function to replicate the original one used by all algorithms (\fname{}, SHAC, and PPO).
+
+\subsection{Dicsovery}
+In the Discovery task, $n$ agents aim to collect as many randomly placed goals as possible.
+When they successfully collect $k$ goals, another $k$ goals are randomly placed.
+To collect a goal, at least $s$ agents must be in proximity to it.
+In \Cref{fig:discovery}, the agents are represented by blue dots, while the goals are depicted as green dots.
+The proximity area is indicated by a dashed circle.
+In our experiments, we set $s=2$ and $k=7$ when $n > 1$.
+When $n = 1$, we set $s=1$ and $k=7$.
+
+Each agent has a continuous 2-dimensional action space, bounded between $-1$ and $1$, to represent acceleration along the x- and y-axes, respectively.
+Each agent observes its own position, velocity, and lidar measurements of other agents and goals.
+As a consequence, the world model lacks sufficient information to accurately predict the next states.
+Specifically, the goal positions are not directly observed but are instead sensed through the lidar measurements.
+We believe this limitation significantly affects the performance of \fname{}.
+
+Each agent receives a reward of $1$ when it collects a goal.
+Thus, the maximum reward depends on the amount of time the agents have to collect the goals.
+For early stopping purposes, we set the maximum reward to $k$.
+Additionally, the reference line is set to $k$.
+While the transition function is differentiable, the reward function is not.
+Therefore, only PPO and \fname{} are applicable to this scenario.
+
+\subsection{Sampling}
+In the Sampling task, $n$ agents must collect rewards from a grid.
+The rewards are sampled from $k$ Gaussian distributions.
+In \Cref{fig:sampling}, the agents are represented by blue dots, while the $k$ reward-generating Gaussians are displayed as a scalar field, where the intensity of the color represents the respective reward value.
+
+Each agent has a continuous 2-dimensional action space, bounded between $-1$ and $1$, to represent acceleration along the x- and y-axes, respectively.
+Each agent observes its own position, velocity, lidar values from other agents, and rewards in a 3x3 grid around itself.
+As a consequence, the world model lacks sufficient information to accurately predict the next states.
+Specifically, it does not have access to the full reward distribution.
+We believe this limitation significantly affects the performance of \fname{}.
+
+Each agent receives a reward equal to the value of the reward in the grid cell where it is located.
+Thus, the maximum reward depends on the placement of the Gaussians.
+Since the maximum reward varies across environments, we only plot a reference line representing good performance.
+While the reward function is differentiable, the one provided by VMAS is not.
+Therefore, we created a custom reward function to match the original one used by all algorithms (\fname{}, SHAC, and PPO).

### 2025-01-21 paper/icml25/first/sects/shacpp.tex
minors + appendix (WIP)

@@
 Nodes without labels (such as $\pi$ or $R$) represents functions.
-Dashed lines are used to represent non differentiability}\label{fig:shac-shacpp}
+Dashed lines are used to represent non differentiability.
+$a_i$ and $r_i$ represent the collection of actions and rewards for all agents}\label{fig:shac-shacpp}
 \end{figure*}
@@
     \end{algorithmic}
-    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode. $s_{:,0}$ denotes the first step of each trajectory in $s$.}
+    \caption{SHAC++ minimal (no cache and no cool-down) pseudocode. $s_{:,0}$ denotes the first step of each trajectory in $s$. $a$ and $r$ denote the actions and rewards for all agents.}
     \label{alg:shacpp}

### 2025-01-22 paper/icml25/first/sects/abstract.tex
minors in the flow

@@
 and a unified approach that effectively addresses both single- and multi-agent settings remains elusive.
-To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, a reinforcement learning algorithm designed for differentiable simulators.
-Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from the simulator.
+To challenge this gap, we introduce a novel framework, \fname{}, inspired by SHAC, which is a reinforcement learning algorithm designed for differentiable simulators.
+Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from simulators.
 These neural networks are trained alongside the policy and value networks of SHAC.
 We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO---a current state-of-the-art algorithm for non-differentiable environments.
-Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to environment gradients.
+Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to simulator gradients.
 \end{abstract}

### 2025-01-22 paper/icml25/first/sects/experiments.tex
minors + appendix (WIP)
minors in the flow

@@
 
-\paragraphTransport}
+\paragraph{Transport}
 The third row of \Cref{fig:experiments} presents a differentiable scenario, transport.
@@
 \fname{} converges faster, but SHAC is more consistent due to direct gradient stability.
-In the second and third columns (3 and 5 agents), PPO fails to converge, while SHAC and \fname{} both complete the task around 10,000 episodes.
+In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes.
 With more agents, convergence is faster because the task becomes easier (as more agents can push the package).

### 2025-01-22 paper/icml25/first/sects/introduction.tex
minors + last related para
prefer Cref on ref
minors in the flow

@@
 Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
-\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}.
+%\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}. 
 A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.
@@
 Second, even when a differentiable simulator is available, SHAC has been found to be sensitive to complex dynamics, such as object collisions~\cite{Georgiev24} or interactions in multi-agent systems.
+Backpropagation through these dynamics has been found, in general, to lead to unstable gradients~\cite{Bengio94,Metz21}.
 
@@
 
-We evaluate our framework, referred to as \fname{}, against two baseline algorithms: PPO~\cite{Schulman17}, a state-of-the-art approach for non-differentiable environments (also widely used in multi-agent scenarios through the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, which is tailored for differentiable environments.
+We evaluate our framework, referred to as \fname{}, against two baseline algorithms: PPO~\cite{Schulman17}, a state-of-the-art approach for non-differentiable environments (also widely used in multi-agent scenarios through the MAPPO variant~\cite{DBLP:conf/nips/YuVVGWBW22}), and SHAC~\cite{Xu22}, which is tailored for differentiable simulators.
 The evaluation is conducted using VMAS, a state-of-the-art simulator designed for complex, physics-based multi-agent scenarios~\cite{DBLP:conf/dars/BettiniKBP22}.

### 2025-01-22 paper/icml25/first/appendices/ablation.tex
appendix wip

--- 2025-01-18 paper/icml25/first/sects/ablation.tex
@@
-\begin{figure*}[!t]
+\begin{figure}[t]
     \centering
     \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
-    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
-    \label{fig:ablation}
-\end{figure*}
+    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.
+Transformer architecture.}
+    \label{fig:ablation-transformer}
+\end{figure}
 
-\section{Ablation Study}\label{sect:ablation}
-While we mainly focus on replacing both the reward and transition function with neural networks, we also investigate the effect of replacing only the latter.
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\columnwidth]{figs/ablation-mlp.pdf}
+    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.
+MLP architecture.}
+    \label{fig:ablation-mlp}
+\end{figure}
 
-In fact, the transition function is the most complex component to replace as it requires modeling with precision the observations obtained from the environment after a number of $H$ actions from each agent.
-Avoiding the need of such network lightens the algorithm but it reintroduce the need for a differentiable environment.
+\section{Ablation Study}\label{apx:ablation}
+In the main paper, we focused on simultaneously replacing both the transition and reward functions with neural networks.
+This approach allows for greater generality and can be applied across a multitude of environments.
+However, when only the transition or reward function is differentiable and available, it is feasible to replace only one of them.
+This simplifies the algorithm, requiring the training of only a single additional network alongside the policy and value functions.
 
-Fig.~\ref{fig:ablation} summarizes the results.
-Most notably, using the true transition function does not appear to provide any significant advantage in terms of reward achieved.
-In fact, in some cases, we report better results for \fname{}, namely in Fig.~\ref{fig:transport-ablation-1-mlp}.
+Specifically, we focus on replacing only the reward function.
+In fact, transition dynamics are often complex and challenging to model with a neural network, especially in multi-agent environments where observations depend not only on the actions of individual agents but also on those of other agents.
+This, in turn, may necessitate the use of larger architectures with greater computational demands, ultimately limiting the algorithm's practicality.
+
+In contrast, the reward function is often simpler and, therefore, easier to model using lightweight neural networks.
+Moreover, a high degree of precision is typically not required in most scenarios.
+As a result, the reward function is an ideal candidate for replacement with a neural network.
+
+This experiment also enables us to compare the final impact of gradients generated by a complex physics engine with those generated by a neural network.
+While the physics engine may produce more unstable gradients, the neural network may generate imprecise gradients, particularly during the early stages of training.
+
+\subsection{Results}
+We refer to the \fname{} algorithm as \fnamer{} when the transition function is differentiable and directly available.
+The results for the transformer architecture are shown in \Cref{fig:ablation-transformer}, while the results for the MLP architecture are presented in \Cref{fig:ablation-mlp}.
+
+For the transformer architecture, \fname{} generally appears to generate better policies, particularly in the Sampling and Transport environments.
+In the Dispersion environment, both algorithms perform comparably.
+Slightly better results are achieved with the \fnamer{} algorithm in the Discovery environment.
+
+For the MLP architecture, the results are more mixed.
+In the Dispersion environment, \fname{} appears to achieve better policies.
+Conversely, in the Transport environment, \fnamer{} leads to substantially better policies.
+In the Discovery environment, both policies struggle to produce meaningful results.
+Finally, in the Sampling environment, \fname{} performs on par with \fnamer{}, except in the single-agent case, where \fnamer{} struggles.
+
+\subsection{Discussion}
+We can easily observe that the transformer architecture leads, generally speaking, to better policies.
+This is also evident by inspecting \Cref{tab:max-rewards}.
+This may be due to the positional invariance property of the transformers leading to more sample efficient networks.
+
+Furthermore, with the transformer architecture, we observe a tendency for \fname{} to produce better results.
+This may be attributed to the generated gradients being more stable, as shown in ~\cref{TODO}.
+
+In contrast, for the MLP architecture, the results are more mixed. This could be attributed to the MLP architecture being less sample-efficient, which ultimately leads to less accurate gradients.

### 2025-01-22 paper/icml25/first/appendices/additional.tex
appendix wip

@@
 \section{Additional Experiments}
+
+Finally, we present the results of the experiments with the MLP architecture in \Cref{apx:fig:experiments-mlp}.
+We observe that the MLP architecture underperforms compared to the Transformer architecture.
+Nonetheless, \fname{} often outperforms PPO and generally performs on par with SHAC.
+
+Interestingly, the Discovery scenario appears to be the most challenging, as none of the algorithms (PPO and \fname{}) are able to make any meaningful progress.
+This may be caused by the sample inefficiency of the MLP architecture, combined with the fact that the world model cannot access the full state.
+
+The only scenario where SHAC outperforms \fname{} is the Transport scenario. This may be due to the MLP's inability to overcome the gradient noise from the world model.
 
@@
     \caption{Comparison between \fname{}, PPO, and SHAC for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios with the MLP architecture.
-    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the Transformer architecture see \Cref{fig:experiments}.
+    We show the mean and standard deviation of the normalized rewards across $3$ runs. For the MLP architecture see \Cref{fig:experiments}.
     }
     \label{apx:fig:experiments-mlp}
+
 \end{figure}
-
-\begin{figure*}[!t]
-    \centering
-    \includegraphics[width=\textwidth]{figs/ablation-transformer.pdf}
-    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
-    \label{fig:ablation}
-\end{figure*}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\columnwidth]{figs/ablation-mlp.pdf}
-    \caption{Comparison between \fname{} with and without transition network, labelled \fnamer{}, for increasing number of agents for Dispersion, Transport, Discovery, and Sampling scenarios.}
-    \label{fig:ablation}
-\end{figure}

### 2025-01-22 paper/icml25/first/appendices/arch.tex
appendix wip

@@
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=.5\textwidth]{figs/attention.pdf} 
+    \caption{Attention matrix for the world model.}
+    \label{fig:apx:attention}
+\end{figure}
+
 \section{Architectures}\label{apx:arch}
+
+As mentioned earlier, we test PPO, SHAC, \fname{}, and \fnamer{} with two different architecture---Transformer and MLP.
+The former is a popular positional invariant architecture for sequence modeling~\cite{Vaswani17}.
+While the latter is a simple feed-forward neural network.
+
+The main advantage of the transformer architecture is its positional invariance property.
+In fact, the order in which the agent's observation are fed into the network should not affect the ultimate outcome, at least, for the tested environments.
+We believe that this leads to better sample efficiency that it is ultimately observed in policies achieving higher rewards (see \Cref{tab:max-rewards}).
+
+In contrast, the MLP is fed with a concatenation of all agent observations.
+This results in a simple and lightweight implementation, leading to faster inference times.
+However, in our experiments, the MLP appears to underperform compared to the transformer architecture.
+
+\subsection{Policy/Value/Reward MLP Network}
+When implementing the policy, value, and reward functions with an MLP, we use a single hidden layer with sizes of $32$, $96$, and $160$, depending on the number of agents.
+This choice was made to account for the increasing complexity that a higher number of agents should introduce.
+The activation function is always ReLU, and dropout is used only for the policy network.
+The learning rate is consistently set to $0.001$.
+The cache for training these networks is set to $30,000$ samples.
+Gradient clipping is applied only to the policy network, with a threshold of $1$.
+For a full list of hyperparameters, see \Cref{apx:tab:shac} and \Cref{apx:tab:ppo}.
+
+\subsection{Policy/Value/Reward Transformer Network}
+When implementing the policy, value, and reward functions with a transformer, we use a single hidden layer with a size of $64$, regardless of the number of agents, and a feed-forward size of $128$.
+We do not scale these sizes with the number of agents, as the transformer is expected to natively scale with the sequence length.
+The remaining hyperparameters are set consistently with those used in the MLP architecture.
+For a full list of hyperparameters, see \Cref{apx:tab:shac} and \Cref{apx:tab:ppo}.
+
+\subsection{World Model}
+When implementing the action world model, we consistently use a transformer architecture.
+Each action is represented as an embedding fed into the network.
+To maintain positional invariance between the agents, we employ a custom attention matrix that allows the action for step $i$ of agent $j$, $a_{ij}$, to attend only to actions from the same step or previous steps of the same agent.
+Formally, $a_{ij}$ can attend to all $a_{kj}$ for $k<i$ and to all $a_{ik}$ for $k \in \agents$.
+This attention pattern is shown in \Cref{apx:fig:attention}.
+As a result, the world models process sequences of length $32 \cdot \na$.
+We use a slightly deeper architecture to account for the increased complexity of the task, with $3$ layers.
+The hidden size is set to $64$ and the feed-forward size to $128$.
+The remaining hyperparameters are set consistently with those used in the MLP architecture.
+For a full list of hyperparameters, see \Cref{apx:tab:shac}.
 

### 2025-01-23 paper/icml25/first/sects/discussion.tex
minors

@@
 train steps).
-With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively long memory requirements.
+With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements.
 In these scenarios, a linear attention mechanism (such as \cite{Beltagy20}) may be employed.

### 2025-01-23 paper/icml25/first/sects/shacpp.tex
minors

@@
 \begin{align*}
-    \underset{\tau\sim\pi}{E}\left[\hspace{-.7cm}\sum_{\hspace{.7cm}\scriptscriptstyle s_i,a_{ij},s_{i+1}\in\tau} \hspace{-.8cm}\left\vert\left\vert {\rewards{i}}_\theta(s_i, a_{ij}, s_{i+1}) -\rewards{i}(s_i, a_{ij}, s_{i+1}) \right\vert\right\vert^2\right]
+    \underset{\tau\sim\pi}{E}\left[\hspace{-.7cm}\sum_{\hspace{.7cm}\scriptscriptstyle s_i,a_{ij},s_{i+1}\in\tau} \hspace{-.8cm}\left( {\rewards{i}}_\theta(s_i, a_{ij}, s_{i+1}) -\rewards{i}(s_i, a_{ij}, s_{i+1}) \right)^2\right]
 \end{align*}
@@
 Specifically, we minimize the loss $\mathcal{L}_\theta^{\rewards{i}}:(\states\times\actions{i}\times\states)^N\rightarrow\mathbb{R}$:
-$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,s',r \in \batch}\left|\left| {\rewards{i}}_\theta(s, a) - r \right|\right|^2 $$
+$$ \mathcal{L}_\theta^{\rewards{i}}(\batch) = \frac{1}{\bs}\sum_{s,a,s',r \in \batch}\left( {\rewards{i}}_\theta(s, a) - r \right)^2 $$
 Where $B$ represents a batch of state-action-state-reward tuples sampled uniformly from the cache.
@@
 We aim to minimize:
-
 \begin{align*}
-    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left|\left| \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right|\right|^2\right]
+    \underset{\tau\sim\pi}{E}\left[\hspace{-1.2cm}\sum_{\scriptscriptstyle \hspace{1cm}\substack{s_t,\dots,s_{t+h+1},\\\hspace{.3cm}a_{t*},\dots,a_{t+h*}\in\tau}}\hspace{-1.2cm} \left\vert\left\vert \transition_{\theta}(s_t, a_{t*},\dots,a_{\scriptscriptstyle i+h*}) - [s_t,\dots,s_{\scriptscriptstyle t+h+1}] \right\vert\right\vert^2\right]
 \end{align*}
-
 Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$.

### 2025-01-24 paper/icml25/first/sects/background.tex
few corrections...

@@
 Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively.
-$\allo=\{\observations{i}\}_{i\in\agents}$ and $\alla=\{\actions{i}\}_{i\in\agents}$ are the observation space and action space for each agent.
-$\{\stateobs{i}:\states\rightarrow\observations{i}\}_{i\in\agents}$ are the projections from the state to the observation space and $\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}_{i\in\agents}$ are the reward functions for each agent.
+$\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent.
+${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent.
 $\transition:\states\times\alla$ denotes the transition function.
@@
 
-Further, let us denote denote the space of trajectories, $\trajectories = \left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
+Further, let us denote denote the space of trajectories, $\trajectories = {\left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)}^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
 Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.
@@
 The goal of the MDP is to find a policy that maximizes the expected return 
-$$\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]$$
+\[\mathbb{E}_{\tau\sim\pi}\left[\sum_{t=0}^{\horizon-1}\sum_{i\in\agents}\gamma^t r_{ti}^\tau \right]\]
 
@@
 \end{equation*}
-where $\mh \ll \horizon$ is the roullout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
+where $\mh \ll \horizon$ is the rollout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
 It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.

### 2025-01-24 paper/icml25/first/sects/discussion.tex
few corrections...

@@
 This pattern becomes even more evident in multi-agent scenarios.
-In these cases, PPO often struggles to find even marginally good policies.
-\fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.
+In these cases, PPO often struggles to find even marginally good policies.\ \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.
 
@@
 Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents.
-This counterintuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
+This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
 Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.
@@
 Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations.
-The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.
-agents times num.
-train steps).
+The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps).
 With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements.
-In these scenarios, a linear attention mechanism (such as \cite{Beltagy20}) may be employed.
+In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.
 

### 2025-01-24 paper/icml25/first/sects/experiments.tex
minors
few corrections...

@@
 The results show the mean and standard deviation of rewards over 3 runs.
-For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
     \label{fig:experiments}
@@
     \caption{Norm of policy gradients on the Transport environment across 5000 epochs.
-The raise in norm for SHAC (epoch 2700) corresponds with occurence of generalization, where collision between agents and the packages become frequent.
+The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent.
 Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
@@
 \paragraph{Scenarios}
-As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO.
+As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO\@.
 In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling.
@@
 Transport and Sampling have differentiable reward functions, thus SHAC is applicable.
-For Transport and Dispersion, the observation space is complete wrt.
-the state.
-For Discovery and Sampling, the observation space is incomplete wrt.
-the state.
-In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC.
+For Transport and Dispersion, the observation space is complete wrt.\ the state.
+For Discovery and Sampling, the observation space is incomplete wrt.\ the state.
+In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC\@.
 
@@
 \paragraph{Hardware Setup}
-Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one Nvidia Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).
+Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one NVIDIA Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).
 
@@
 The third row of \Cref{fig:experiments} presents a differentiable scenario, transport.
-In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.
-\fname{} converges faster, but SHAC is more consistent due to direct gradient stability.
+In the first column (single agent), PPO fails to converge, while SHAC and \fname{} achieve comparable performance.\ \fname{} converges faster, but SHAC is more consistent.
 In the second and third columns (3 and 5 agents), PPO fails as in the single-agent case, while SHAC and \fname{} both complete the task around 10,000 episodes.
@@
 \paragraph{Sampling}
-The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC.
-In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC.
+The fourth row of \Cref{fig:experiments} illustrates the sampling scenario, which is differentiable and thus applicable to SHAC\@.
+In the first column (single agent), PPO fails to converge, while \fname{} outperforms SHAC\@.
 In the second and third columns (3 and 5 agents), PPO again fails to converge, while SHAC and \fname{} both complete training in fewer than 5,000 episodes.

### 2025-01-24 paper/icml25/first/sects/introduction.tex
few corrections...

@@
 \section{Introduction}\label{sect:introduction}
-
 Learning control policies has important applications in robotics~\cite{Singh22}, autonomous driving~\cite{Elallid22}, swarm intelligence~\cite{Tang21} and many other fields.
@@
 Moreover, reinforcement learning algorithms tend to be \emph{sample inefficient}, namely requiring a large number of interactions with the environment to learn a good policy.
-%\texttt{\color{DarkOrchid}{<- FB: non mi è super chiara questa frase}}. 
 A recently introduced method for the latter problem is SHAC~\cite{Xu22} which exploits differentiable simulators to provide precise gradient information, thereby expediting policy optimization and improving learning efficiency.
@@
 \begin{compactitem}
-    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC.
+    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC\@.
     \item We assess the SHAC algorithm in multi-agent scenarios, where it has not been previously evaluated.

### 2025-01-28 paper/icml25/first/sects/background.tex
minors

@@
 ${\{\stateobs{i}:\states\rightarrow\observations{i}\}}_{i\in\agents}$ are the projections from the state to the observation space and ${\{\rewards{i}:\states\times\actions{i}\times\states\rightarrow\mathbb{R}\}}_{i\in\agents}$ are the reward functions for each agent.
-$\transition:\states\times\alla$ denotes the transition function.
+$\transition:\states\times\alla\rightarrow\states$ denotes the transition function.
 $\discount$ and $\horizon$ denote the discount factor and the time horizon respectively.
@@
 A policy for agent $i$ is a function $\policy{i}:\observations{i}\rightarrow\actions{i}$ that maps its observation space to its action space.
-We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi$.
+We will denote the collective policy, determined by the collection of agent-specific policies simply with $\pi=\{\pi_{i}\}_{i\in\agents}$.
 We can use the policy to sample trajectories from the MDP by iteratively applying the transition, reward, projection and policy functions.
@@
 Finally, $s_0$ is sampled from the initial state distribution $\initdist$.
-For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies.
+For brevity, we will simply denote with $\tau\sim\pi$ a trajectory sampled following the policies $\pi$.
 
@@
 
-In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled from $\mdp$.
+In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled usin $\pi$.
 The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:

### 2025-01-28 paper/icml25/first/sects/discussion.tex
minors

@@
 The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.\ agents times num.\ train steps).
-With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements.
+With significantly large populations ($\sim$50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements.
 In these scenarios, a linear attention mechanism (such as~\cite{Beltagy20}) may be employed.

### 2025-01-28 paper/icml25/first/sects/shacpp.tex
few corrections...
minors

@@
         \includegraphics[width=\textwidth]{figs/shac.pdf}
-        \caption{SHAC}
+        \caption{SHAC}%
         \label{fig:shac}
@@
         \includegraphics[width=\textwidth]{figs/shacpp.pdf}
-        \caption{\fname{}}
+        \caption{\fname{}}%
         \label{fig:shacpp}
@@
 However, to avoid vanishing or exploding gradients, we employ a non-recursive approach similar to Action World Model~\cite{Ma24}.
-This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ observations from $\states^\mh$.
+This corresponds to train a neural network that given a starting observation from $\mathcal{S}$ and a sequence of $\mh$ actions from $\alla^\mh$, predicts the next $\mh$ states from $\states^\mh$.
 We aim to minimize:
@@
 
-$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(a,b))$$
-$$d(x,(l,r)) = \begin{cases}(l-x)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$
+$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(l,r))$$
+$$d(x,(l,r)) = \begin{cases}(x-l)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$
 

### 2025-01-28 paper/icml25/first/sects/introduction.tex
add a sentence about inspiration of our work (@f14-bertolotti check)

@@
 
-In this work, we aim to address both limitations by approximating the gradients of the differentiable simulator.
+In this work, we aim to address both limitations by approximating the gradients of the differentiable simulator inspired by model-based reinforcement learning algorithms~\cite{DBLP:conf/icml/HafnerLFVHLD19,DBLP:conf/iclr/HafnerLB020}.
 Specifically, we propose replacing both the simulator and the reward function with two neural networks trained alongside the policy and value networks.

### 2025-01-28 paper/icml25/first/sects/introduction.tex
stronger statement on the contribution (@f14-bertolotti check)

@@
 \begin{compactitem}
-    \item We propose a new reinforcement learning framework that lift the requirement for a differentiable simulator in SHAC\@.
+    \item We propose a new reinforcement learning framework that lifts the requirement for a differentiable simulator in SHAC\@.
     \item We assess the SHAC algorithm in multi-agent scenarios, where it has not been previously evaluated.
     \item We evaluate our algorithm against PPO and SHAC on a set of environments with increasing numbers of agents.
-    \item We show that our algorithm achieve SHAC-like performance in differentiable environments and outperform PPO in non-differentiable environments.
+    \item We demonstrate that our algorithm matches SHAC's performance in differentiable environments while significantly outperforming PPO in terms of sample efficiency, convergence speed, and final reward across both single-agent and complex multi-agent scenarios.
 \end{compactitem}

### 2025-01-28 paper/icml25/first/sects/background.tex
clarify that the formalization is for multi-agent, not the standard RL formalization (@f14-bertolotti check)

@@
 To provide context for our work, we first introduce the reinforcement learning (RL) framework. 
-Each task is modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$.
+Particularly, we focus on multi-agent settings, where multiple agents interact with each other and the environment.
+In doing so, each task is then modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$.
 Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively.

### 2025-01-28 paper/icml25/first/sects/background.tex
clarify some notion

@@
 In doing so, each task is then modeled as a partially observable, decentralized, finite-horizon, Markov Decision Process (MDP)~\cite{Shapley53}, $\mdp$: $(\agents,\states,\allo,\alla,\allp, \allr, \transition, \horizon, \discount, \initdist)$.
-Where $\agents=[\na]$ and $\states$ denote the set of agents and the state space, respectively.
+Where $\agents=\{1, 2, \dots, \na\}=[\na]$ and $\states$ denote the set of agents and the state space, respectively.
 $\allo={\{\observations{i}\}}_{i\in\agents}$ and $\alla={\{\actions{i}\}}_{i\in\agents}$ are the observation space and action space for each agent.
@@
 
-Further, let us denote denote the space of trajectories, $\trajectories = {\left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)}^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
+Further, let us denote the space of trajectories, $\trajectories = {\left(\states\times\alla\times\states\times\mathbb{R}^{n}\right)}^*$, where each trajectory consists of a sequence of: starting state, action per agent, subsequent state, and reward per agent.
 Given a generic trajectory $\tau\in\trajectories$, we denote its $i$-th starting state with $s_i^\tau$, its $i$-th action for agent $j$ with $a_{ij}^\tau$, and its $i$-th reward for agent $j$ with $r_{ij}^\tau$.

### 2025-01-28 paper/icml25/first/sects/background.tex
clarify some notion on 2.2

@@
 \subsection{Neural Networks}
-A neural network is a function $f_{\theta}:\mathbb{R}^n\rightarrow\mathbb{R}^m$ parameterized by $\theta$ that maps an input vector $x\in\mathbb{R}^n$ to an output vector $y\in\mathbb{R}^m$.
-The parameters $\theta$ of the network are learned by minimizing a loss function $\mathcal{L}_\theta$ with respect to the training data.
+A neural network is a function $f_{\theta}:\mathbb{R}^n\rightarrow\mathbb{R}^m$ parameterized by its weights $\theta$, which maps an input vector $x\in\mathbb{R}^n$ to an output vector $y\in\mathbb{R}^m$.
+The network parameters $\theta$ are optimized by minimizing the loss function $\mathcal{L}_\theta$, which quantifies the discrepancy between the network's predictions and the target outputs based on the training data.
 

### 2025-01-28 paper/icml25/first/sects/background.tex
clarify why we do not consider partial observability in the formalization

@@
 \subsection{Neural Networks}
-A neural network is a function $f_{\theta}:\mathbb{R}^n\rightarrow\mathbb{R}^m$ parameterized by its weights $\theta$, which maps an input vector $x\in\mathbb{R}^n$ to an output vector $y\in\mathbb{R}^m$.
+A neural network is a function $f_{\theta}:\mathbb{R}^n\rightarrow\mathbb{R}^m$ parameterized by its weights $\theta \in \mathbb{R}^w$, which maps an input vector $x\in\mathbb{R}^n$ to an output vector $y\in\mathbb{R}^m$.
 The network parameters $\theta$ are optimized by minimizing the loss function $\mathcal{L}_\theta$, which quantifies the discrepancy between the network's predictions and the target outputs based on the training data.
@@
 
-For simplicity, we formally treat only the case where the policies observe the full state, meaning that $\stateobs{i}$ is the identity function.
-However, in our experiments, we consider also scenarios where the full state is not available to the policies.
+For simplicity, we formally consider only the case where policies have access to the full state, meaning that $\stateobs{i}$ is the identity function.
+This allows us to focus on the fundamental aspects of our approach without the added complexity of partial observability.
+Extending our framework to handle partial observability is straightforward; however, we choose not to address it here to maintain clarity and simplicity in our presentation.
+Nonetheless, our experiments do include scenarios where policies do not have access to the complete state information.
 

### 2025-01-28 paper/icml25/first/sects/background.tex
clarify the meaning of rollout (@f14-bertolotti check)

@@
 \end{equation*}
-where $\mh \ll \horizon$ is the rollout size, $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
+where $\mh \ll \horizon$ is the rollout size (namely the number of steps used to estimate the return), $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
 It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.

### 2025-01-28 paper/icml25/first/sects/background.tex
minor on 2.3

@@
 where $\mh \ll \horizon$ is the rollout size (namely the number of steps used to estimate the return), $t_0$ is the initial time step of the rollout, $\valuef_\theta: \states \rightarrow \mathbb{R}$ is the value function estimating the expected return.
-It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta,\transition \in \mathcal{C}^1$.
+It is apparent that SHAC assumes that the policy, reward, value, projection, and transition functions are all differentiable, i.e., $\{\policy{i}\}_{i\in\agents},\{\rewards{i}\}_{i\in\agents},\valuef_\theta, \text{and} \, \transition \in \mathcal{C}^1$.
 While this is a common requirement for policy and value functions, differentiable transition and reward functions are rarely available.

### 2025-01-28 paper/icml25/first/sects/shacpp.tex
provide a anonymous reference to code repository
minor on l' (is it ok like this? @f14-bertolotti
clarify again the goal of the proposed approach in the introduction

@@
     \caption{Comparison between SHAC and \fname{}.
-Nodes with labels (such as $r_i$ or $r'_i$ represents values).
-Nodes without labels (such as $\pi$ or $R$) represents functions.
+Nodes with labels (such as $r_i$ or $r'_i$) values.
+Nodes without labels (such as $\pi$ or $R$) represent functions.
 Dashed lines are used to represent non differentiability.
@@
 
-$$\mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(l,r))$$
+$$\mathcal{L'}_\theta^\pi(B) = \mathcal{L}_\theta^\pi(\batch) + \alpha \frac{1}{\bs\horizon}\sum_{\tau\in\batch}\sum_{a\in\tau} d(a^\tau,(l,r))$$
 $$d(x,(l,r)) = \begin{cases}(x-l)^2 & \text{if } x < l \\ (x-r)^2 & \text{if } x > r \\ 0 & \text{otherwise} \end{cases}$$
@@
 
-The full implementation is openly available at 
-\begin{center}
-\url{https://redacted/for/anonymity}
-\end{center}
+The full implementation is openly available at \url{https://anonymous.4open.science/r/diffagents-8653}\footnote{The link of the original repository will be provided upon acceptance.}.

### 2025-01-28 paper/icml25/first/sects/abstract.tex
few corrections...
minor on abstract and main contribution

@@
 Unlike SHAC, \fname{} eliminates the requirement for differentiable transition and reward functions by utilizing plain neural networks to estimate the gradients that would otherwise be derived directly from simulators.
-These neural networks are trained alongside the policy and value networks of SHAC.
+These neural networks are trained alongside the policy and value networks of SHAC\@.
 We evaluate \fname{} on a set of challenging multi-agent scenarios from VMAS and compare its performance against SHAC and PPO---a current state-of-the-art algorithm for non-differentiable environments.
-Our results demonstrate that \fname{} significantly outperforms PPO and achieves comparable performance to SHAC, despite not having access to simulator gradients.
+Our results demonstrate that \fname{} significantly outperforms PPO in both single- and multi-agent scenarios and achieves comparable performance to SHAC, despite not having access to simulator gradients.
 \end{abstract}

### 2025-01-28 paper/icml25/first/sects/shacpp.tex
minor on fig 1
minor on abstract and main contribution

@@
     \end{subfigure}
-    \caption{Comparison between SHAC and \fname{}.
+    \caption{
+        Comparison between SHAC and \fname{} during a rollout of 4 steps.
         Nodes with labels (such as $r_i$ or $r'_i$) values.
@@
 
-In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the environment.
+In practice, this is achieved by leveraging a \emph{replay buffer} to cache the state-action-reward triplets obtained from the simulator.
 The neural network is then trained to minimize the mean squared error (MSE) loss.

### 2025-03-25 rebuttal.md
added rebuttal

# Reviewer 1

Better discussion on when SHAC++ matches SHAC's performance is needed.

Higher number of agents.

Discussion on partial observability. Not enough support at claim that SHAC++ can handle these scenarios

varying level of observability

Discussing high variance

add training times

# Reviewer 2

revise introduction to remove "methods cannot handle both single-agent and multi-agent scenarios, and also cannot be applied to non-differentiable environments"

Specify that the goal is not to outperform SHAC, but to provide a more general algorithm

MAPPO is already used in the multi-agent setting

Yes modeling the environment may introduce biases degrading performance. 
On the other hand, using the derivative of the environment may lead to complex dynamics and unstable gradients as shown in [TODO].
This can also lead in performance of degradation, while the work of [TODO] takes one road we explore the other.

Yes, we use the local observation to estimate the reward function and state transition. 
This is a deliberate choice to remain faithful to the partial observability of the environment.
While this is possible using PPO or SHAC++, this is not possible for SHAC as the backpropagation needs global information to compute the gradients.

We use a single parameter name, theta, to represent all possible parameter choices for the reward, state, and policy function.
For example, if we wanted separate network, we could set theta=[theta_1, theta_2, theta_3].
if we wanted shared network, we could set theta=[theta_1, theta_1, theta_1].
Ultimately, while this choice may seem confusing, it allows the highest degree of generality.

include observation and reward function accuracy

add experiment with one of shac's experiments

add SAC and explain MAPPO is already added.

Visual representation of cooperation

l,r are the upper and lower bound of the action space

# Reviewer 3

nothing

# Reviewer 4

We will rephrase that sentence.

theoretically characterize the additional smoothness od the network

add experiment with one of shac's experiments

add TD-MPC2 (https://arxiv.org/abs/2310.16828), DreamerV3 (https://arxiv.org/abs/2301.04104)

Yes modeling the environment may introduce biases degrading performance...

______________________

# Reviewer 1

> In some cases, SHAC++ shows slightly lower performance.

**Regarding lower performance** In VMAS all agents experience partial observability, some cases more severe than others. In SHAC++ reward and world model learn to predict the their output from partial observation. Instead SHAC world model and reward function have access to the full state. This can skew the results in favor of SHAC. Despite this, both algorithms perform similarly.

> How well does it scale with larger numbers of agents? Specifically, have the authors conducted any experiments with agent populations larger than those presented?

We executed some experiment up to 7 agents, but discarded the runs as the performance was too heavily impacted to perform the 3 runs with different seeds. Scaling the number of agents has two main effects:
1. It become easier to discover action that lead to rewards, in this sense the enviroment become less sparse.
2. It become more difficult to coordinate between the agents. 
For the dispersion environment 2. seems to be the main factor limiting the performance.
For the transport environement 1. seems to be the main factor pushing the agents to perform better.

> The paper mentions challenges in scenarios with partial observability, such as the Discovery task. Are there any planned modifications or extensions to SHAC++ to better handle these scenarios?

We believe that partial observability is an intrinsic property of some enviroments that RL algorithm need to deal with. SHAC can access the global state through its gradient in order to inform the policy update, however, this feels like cheating. On the other hand, SHAC++ uses reward and world model that access only local observation. While this choice may skew the performance in favor of SHAC, we believe that this is a more realistic scenario. We can include this discussion in the paper.

> Could the authors provide more insights into the factors contributing to the higher variance observed in some scenarios with SHAC++?

In SHAC++ also the reward and world model are intialized randomly. This can lead to this model to learn different solutions that impact the final reward differently, leading to higher variance. While a full analysis of the components that impact this variance would be definetely valuable, we believe that the number of potential factors (NN arch., NN size, environment type, no. agents) would render this analysis too complex for the scope of this paper.

> What are the computational requirements and training times for SHAC++ compared to SHAC and PPO?

We can include the training times in the paper. However, it should be noted that the performance also depends heavily on some parameters such as training epochs for value, policy (and, reward and world model for SHAC++). Therefore, this would not be indicative on what method is the fastest to train. Nonetheless, SHAC++ is definetely more computationally heavy this is because it needs to train the reward and world model in addition to what SHAC does. However, for a comparison consider that for the sampling enviroment (which is the slowest in VMAS) SHAC++ takes 16 hours to complete the 20000 episodes similarly to PPO, while SHAC takes about 10 hours.

### 2025-03-26 rebuttal.md
rebuttal wip

@@
-# Reviewer 1
-
-Better discussion on when SHAC++ matches SHAC's performance is needed.
-
-Higher number of agents.
-
-Discussion on partial observability. Not enough support at claim that SHAC++ can handle these scenarios
-
-varying level of observability
-
-Discussing high variance
-
-add training times
-
-# Reviewer 2
-
-revise introduction to remove "methods cannot handle both single-agent and multi-agent scenarios, and also cannot be applied to non-differentiable environments"
-
-Specify that the goal is not to outperform SHAC, but to provide a more general algorithm
-
-MAPPO is already used in the multi-agent setting
-
-Yes modeling the environment may introduce biases degrading performance. 
-On the other hand, using the derivative of the environment may lead to complex dynamics and unstable gradients as shown in [TODO].
-This can also lead in performance of degradation, while the work of [TODO] takes one road we explore the other.
-
-Yes, we use the local observation to estimate the reward function and state transition. 
-This is a deliberate choice to remain faithful to the partial observability of the environment.
-While this is possible using PPO or SHAC++, this is not possible for SHAC as the backpropagation needs global information to compute the gradients.
-
-We use a single parameter name, theta, to represent all possible parameter choices for the reward, state, and policy function.
-For example, if we wanted separate network, we could set theta=[theta_1, theta_2, theta_3].
-if we wanted shared network, we could set theta=[theta_1, theta_1, theta_1].
-Ultimately, while this choice may seem confusing, it allows the highest degree of generality.
-
-include observation and reward function accuracy
-
-add experiment with one of shac's experiments
-
-add SAC and explain MAPPO is already added.
-
-Visual representation of cooperation
-
-l,r are the upper and lower bound of the action space
-
-# Reviewer 3
-
-nothing
-
-# Reviewer 4
-
-We will rephrase that sentence.
-
-theoretically characterize the additional smoothness od the network
-
-add experiment with one of shac's experiments
-
-add TD-MPC2 (https://arxiv.org/abs/2310.16828), DreamerV3 (https://arxiv.org/abs/2301.04104)
-
-Yes modeling the environment may introduce biases degrading performance...
-
-______________________
-
 # Reviewer 1
@@
 
-**Regarding lower performance** In VMAS all agents experience partial observability, some cases more severe than others.
+**skewed results** In VMAS all agents experience partial observability, some cases more severe than others.
 In SHAC++ reward and world model learn to predict the their output from partial observation.
@@
 
+**Early stopping** We also note that as soon as a run achieves the 90% of the maximum reward, we stop the training, this can lead to some variance in the final performance.
+
 > How well does it scale with larger numbers of agents? Specifically, have the authors conducted any experiments with agent populations larger than those presented?
@@
 Scaling the number of agents has two main effects:
-1. It become easier to discover action that lead to rewards, in this sense the enviroment become less sparse.
+1. It become easier to discover action that lead to rewards, in this sense, the enviroment become less sparse.
 2. It become more difficult to coordinate between the agents. 
@@
 However, for a comparison consider that for the sampling enviroment (which is the slowest in VMAS) SHAC++ takes 16 hours to complete the 20000 episodes similarly to PPO, while SHAC takes about 10 hours.
+
+# Reviewer 2
+
+> problematic claims that are misaligned with each other and with the supporting evidence.
+
+We are sorry for the confusion.
+We will revise the introduction to better reflect the goal of the paper.
+Briefly, we believe that SHAC demostrated excellent perfomance in single-agent and differentiable scenarios.
+In this work, we aim to lift these requirements (single-agent and differentiability) to broaden the applicability of SHAC.
+This led us to develop SHAC++ that can handle multi-agent and non-differentiable environments while mantaining performance comparable to SHAC.
+
+> lower performance in some cases
+
+**Goal**: We want to stress the fact that the goal is not to outperform SHAC, but to lift the requirements of the original framework (single-agent and differentiability) while maintaining comparable performance.
+
+**Skewed results**: In VMAS as in many other enviroments, the agents are able to experience only local observetions and not the whole world state.
+While SHAC++ remains faithful to accessing only this partial information, SHAC access the whole global information through its gradients.
+This puts SHAC in a more favorable position.
+Despite this, SHAC++ and SHAC perform similarly.
+
+**Early stopping** We also note that as soon as a run achieves the 90% of the maximum reward, we stop the training, this can lead to some variance in the final performance.
+
+> The proposed method is not evaluated in such complex environments; the test environments are very simple.
+
+We believe that the chose enviroments offer a good degree of complexity especially when the number of agents is increased.
+This is also reflected by the poor performance achieved by PPO/MAPPO.
+Despite this, we recognize that the statement in the abstract may be misleading.
+We will revise the abstract and introduction to better reflect the complexity of the environments.
+
+> The performance of SHAC++ is questionable in more complex MARL.
+
+We understand that more complex enviroments with more agents may offer more challenging scenarios and more compelling results.
+However, we want to stress that these enviroments are already incapacitating for PPO.
+[RUN experiment with more agents].
+Further, we will consider adding more complex enviroments for future revision, we kindly ask the reviewer to suggest a differentiable enviroment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
+
+> Numerous practical algorithms already exist for such environments, for example MAPPO, HAPPO, and QMIX in multi-agent scenarios.
+
+**PPO & MAPPO confusion**: While we do not evaluate against also HAPPO and QMIX, we always use MAPPO where PPO is evaluated for multi-agent setting.
+Since the MAPPO formulation is fairly natural extension to PPO, we always referred to PPO in the figures and tables, and only briefly mention that MAPPO is used in the case of multi-agent setting.
+We will revise the paper to better reflect this.
+
+**Why MAPPO and not the others**: While HAPPO and QMIX are both valid candidate for baseline.
+In order to offers an ablation study and a degree of statistical significance we had to limit the number of baselines.
+We also note that HAPPO authors designed their algorithm to be non-homogeneous (parameters between agents are not shared).
+This choice renders its application problematic as istantiating and traiing several NN at the same time quickly becomes computationally infeasible.
+Meanwhile MAPPO can be applied with both shared and non-shared parameters.
+In our experiments, in order to be consisted while scaling the number of agents, we always apply parameter sharing.
+
+**Why not SAC**: We appreciate the author suggestion to include SAC and we will consider adding it to the baselines for the next revision.
+However, the original SHAC paper already included a SAC baseline.
+Ultimately, we believe that the inclusion of SAC would not add much to the main discussion.
+
+**Why not the algorithm in related work section**: We did not include PODS, CE-APG, and BPTT as baselines as we see these a precursors of SHAC.
+The authors of SAM-RL introduce a differentiable renderer on top of the differentiable enviroment which is even more difficult to obtain in practice (Although also this component could be replaced with a learned model in our framework).
+Instead, DiffTORI applies a test-time optimization on the agent trajectory.
+This choice render the extension of DiffTORI to the multi-agent setting difficult.
+The authors of AHAC address the issue of stiff dynamic by simply truncating trajectories, while this choice can boost performance of SHAC, it mantains the fundamental limitation of the original framework.
+
+We will provide a more detailed explanation in the paper.
+
+> The authors estimate the reward function and state transition using only local agent states and actions as input. This approach may limit the method's generality and correctness.
+
+Yes, this is a deliberate choice on our part.
+Since in some enviroment their state may be only partially observable then a broad algorithm need to be able to handle this.
+While with SHAC this is not possible, since for gradient computation the global state is needed, SHAC++ can handle this by unsing the local information available to the agents.
+Despite this choice may skew the performance in favor of SHAC, we believe that this is a more realistic scenario.
+We will include this discussion in the paper.
+
+> The parameters for the policy, value, reward, and state transition networks are all denoted by $\theta$.
+
+This is a deliberate choice that allows us to denote a genera framework where the parameters can be arbitrarly shared.
+For example, all parameter may be separate such as in $\theta=[\theta_{\text{policy}}, \theta_{\text{reward}}, \theta_{\text{value}}, \theta_{\text{world}}]$.
+Or, $\theta$ could be designed with some degree of sharing.
+
+> modeling the environment instead of directly utilizing its information might introduce biases and lead to performance degradation.
+
+Modeling the enviroment means that the policy will be only as good as the enviroment model is.
+If the world model is very accurate, the gradient will be informative.
+If the world model is inaccurate or only partially accurate, the gradient will be noisy leading to degradation.
+However, while we train the world model in tandem with the policy, it is possible to bootstrap the world model with already available traces.
+This can lead to a more stable training of the policy.
+We will include this discussion in the paper.
+
+> To what SHAC+ stands for?
+
+The SHAC++ framework allows for replacing only certain components with learned models.
+For example, if a differentialble enviroment is available, but not a differentiable reward.
+Then we can only learn the latter.
+SHAC+ refers to the case where only the reward is learned.
+We will provide a more detailed explanation in the paper.
+
+> Visual analysis of the trained policies' behavior would be more convincing.
+
+We will include snapshot from trajectories from the learned policy.
+
+> Confusion on $l$ and $r$.
+
+$l$ and $r$ denote the lower and upper bound of the action space.
+For example, if the agent is supposed to output action in range of (0,1) then $l=0$ and $r=1$.
+We will provide a better explanation in the paper.
+
+# Reviewer 3
+We thank the reviewer for the positive feedback.
+
+# Reviewer 4
+> a strange way of phrasing the contribution
+
+We understand the reviewer point of view.
+We will revise the introduction to better reflect the goal: to lift the requirements of the original SHAC framework to broaden the applicability of the algorithm.
+
+> the paper conflates two contributions
+
+We believe that SHAC has demonstrated excellent performance in single-agent setting.
+In this work, we aim to broaden its applicability by lifting some of its requirement without impacting the performance.
+Our framework extends SHAC to multi-agent, non-differentiable and partially observable environments.
+We believe this to be a fair contribution to field.
+However, we understand that the introduction may have phrased this in a confusing way.
+We will revise the introduction to better reflect the goal of the paper.
+
+> Not convincing evaluation
+
+We choose VMAS because it offers differentiable and single/multi-agent enviroments.
+Further, these enviroments are able to scale in complexity by increasing the number of agents.
+Ultimately, we believe VMAS to be a good testbed to assess the scaling performance of RL algorithm in general.
+
+> results do not seem significant enough to make a deep statement about SHAC/SHAC++ in multi-agent settings.
+
+We believe that the results are significant in showing that SHAC outperforms heavily PPO/MAPPO even in presence of stiff dynamics.
+Further, SHAC++ is able to match SHAC's performance despite being able to access only partial information.
+For example, consider that when SHAC achieves early stopping threshold, also SHAC++ is able to surpass the same threshold.
+On the other hand, when SHAC++ surpasses the threshold, SHAC is not always able to do so.
+
+> It could be interesting if there are ways to theoretically characterize the additional smoothness and desirable properties that come from the inductive bias of the network.
+
+We agree. While, we do explore the theoretical desirable property, Sect D.2 provides a comparison between the gradients of SHAC++ with an mlp and a transformer.
+
+> why not using the full VMAS suite?
+
+The VMAS suite is fairly extensive offering 21 different enviroments.
+A full evualtion of this enviroments even with a small degree of statistical significance would require thousands of runs.
+Therefore, we limited ourself to the 4 mentioned enviromens which we believe to be representative of the whole suite.
+We will update Sect.A to offer more insight on this choice.
+
+> The most convincing evaluation would be in comparison to exactly the settings from the original SHAC paper
+
+We understand the author point of view.
+In this revision, we mainly focused on a multi-agent evaluation.
+We believe that an evaluation with respect to some original enviroment would be valuable.
+We will work towards this direction for the next revision.
+[Ci penso io a questo]
+
+> Omitted references
+
+We will make sure to include TD-MPC2 and DreamerV3 in the related work section.
+
+> I am also curious about the design space of improving bad simulator gradients.
+
+While we do not perform an extensive analysis on the design choices for improving the enviromen gradients.
+Section D.2 offers a preliminary results comparing the transformer gradients with the mlp gradient.
+Our intuition is that the neural network architecture has definetely an impact.
+However, a more in depth study comparing normalization layer, weigh decay, optimizers and etc.
+would be needed in order to draw meaningful conclusions.

### 2025-03-26 rebuttal.md
minors on rebuttal

@@
 This choice render the extension of DiffTORI to the multi-agent setting difficult.
-The authors of AHAC address the issue of stiff dynamic by simply truncating trajectories, while this choice can boost performance of SHAC, it mantains the fundamental limitation of the original framework.
+The authors of AHAC address the issue of stiff dynamic by simply truncating trajectories, while this choice can boost performance of SHAC, it maintains the fundamental limitation of the original framework.
 
@@
 Yes, this is a deliberate choice on our part.
-Since in some enviroment their state may be only partially observable then a broad algorithm need to be able to handle this.
+Since in some environment their state may be only partially observable then a broad algorithm need to be able to handle this.
 While with SHAC this is not possible, since for gradient computation the global state is needed, SHAC++ can handle this by unsing the local information available to the agents.
@@
 
-This is a deliberate choice that allows us to denote a genera framework where the parameters can be arbitrarly shared.
+This is a deliberate choice that allows us to denote a general framework where the parameters can be arbitrarly shared.
 For example, all parameter may be separate such as in $\theta=[\theta_{\text{policy}}, \theta_{\text{reward}}, \theta_{\text{value}}, \theta_{\text{world}}]$.
@@
 
-Modeling the enviroment means that the policy will be only as good as the enviroment model is.
+Modeling the environment means that the policy will be only as good as the environment model is.
 If the world model is very accurate, the gradient will be informative.
@@
 The SHAC++ framework allows for replacing only certain components with learned models.
-For example, if a differentialble enviroment is available, but not a differentiable reward.
+For example, if a differentiable environment is available, but not a differentiable reward.
 Then we can only learn the latter.
@@
 
-We choose VMAS because it offers differentiable and single/multi-agent enviroments.
-Further, these enviroments are able to scale in complexity by increasing the number of agents.
+We choose VMAS because it offers differentiable and single/multi-agent environments.
+Further, these environments are able to scale in complexity by increasing the number of agents.
 Ultimately, we believe VMAS to be a good testbed to assess the scaling performance of RL algorithm in general.
@@
 
-The VMAS suite is fairly extensive offering 21 different enviroments.
-A full evualtion of this enviroments even with a small degree of statistical significance would require thousands of runs.
-Therefore, we limited ourself to the 4 mentioned enviromens which we believe to be representative of the whole suite.
+The VMAS suite is fairly extensive offering 21 different environments.
+A full evaluation of this environments even with a small degree of statistical significance would require thousands of runs.
+Therefore, we limited ourselves to the 4 mentioned environment which we believe to be representative of the whole suite.
 We will update Sect.A to offer more insight on this choice.
@@
 In this revision, we mainly focused on a multi-agent evaluation.
-We believe that an evaluation with respect to some original enviroment would be valuable.
+We believe that an evaluation with respect to some original environment would be valuable.
 We will work towards this direction for the next revision.
@@
 
-While we do not perform an extensive analysis on the design choices for improving the enviromen gradients.
+While we do not perform an extensive analysis on the design choices for improving the environment gradients.
 Section D.2 offers a preliminary results comparing the transformer gradients with the mlp gradient.
-Our intuition is that the neural network architecture has definetely an impact.
-However, a more in depth study comparing normalization layer, weigh decay, optimizers and etc.
+Our intuition is that the neural network architecture has definitely an impact.
+However, a more in depth study comparing normalization layer, weigh decay, optimizers, etc.
 would be needed in order to draw meaningful conclusions.

### 2025-03-28 rebuttal.md
minors

@@
 
-**skewed results** In VMAS all agents experience partial observability, some cases more severe than others.
-In SHAC++ reward and world model learn to predict the their output from partial observation.
+**skewed results** In VMAS all agents experience partial observability, 
+some cases more severe than others. 
+In SHAC++ reward and world model learn to predict their output from partial observation. 
 Instead SHAC world model and reward function have access to the full state. 
-This can skew the results in favor of SHAC.
-Despite this, both algorithms perform similarly.
+This can skew the results in favor of SHAC. Despite this, both algorithms perform similarly.
 
-**Early stopping** We also note that as soon as a run achieves the 90% of the maximum reward, we stop the training, this can lead to some variance in the final performance.
+**Early stopping** We also note that as soon as a run achieves the 90% of the maximum reward, 
+we stop the training, this can lead to some variance in the final performance.
 
@@
 
-We executed some experiment up to 7 agents, but discarded the runs as the performance was too heavily impacted to perform the 3 runs with different seeds.
-Scaling the number of agents has two main effects:
-1. It become easier to discover action that lead to rewards, in this sense, the enviroment become less sparse.
-2. It become more difficult to coordinate between the agents. 
-For the dispersion environment 2. seems to be the main factor limiting the performance.
-For the transport environement 1. seems to be the main factor pushing the agents to perform better.
+We executed some experiment up to 7 agents, 
+but discarded the runs as the performance was too heavily impacted to perform the 3 runs with different seeds. Scaling the number of agents has two main effects:
+1. It becomes easier to discover action that lead to rewards, in this sense, the environment become less sparse.
+2. It becomes more difficult to coordinate between the agents. 
+For the dispersion environment 2. it seems to be the main factor limiting the performance.
+For the transport environment 1. it seems to be the main factor pushing the agents to perform better.
 
@@
 
-We believe that partial observability is an intrinsic property of some enviroments that RL algorithm need to deal with.
+We believe that partial observability is an intrinsic property of some environments that RL algorithm need to deal with.
 SHAC can access the global state through its gradient in order to inform the policy update, however, this feels like cheating.
@@
 While this choice may skew the performance in favor of SHAC, we believe that this is a more realistic scenario.
-We can include this discussion in the paper.
+We can include this discussion in the paper, and we plan to add this as an extension to the SHAC++ framework in future work.
 
@@
 
-In SHAC++ also the reward and world model are intialized randomly.
+In SHAC++ also the reward and world model are initialized randomly. 
 This can lead to this model to learn different solutions that impact the final reward differently, leading to higher variance. 
-While a full analysis of the components that impact this variance would be definetely valuable, we believe that the number of potential factors (NN arch., NN size, environment type, no.
-agents) would render this analysis too complex for the scope of this paper.
+While a full analysis of the components that impact this variance would be definitely valuable, 
+we believe that the number of potential factors (NN arch., NN size, environment type, no. agents) would render this analysis too complex for the scope of this paper.
 

### 2025-03-28 rebuttal.md
more on rebuttal

@@
 Therefore, this would not be indicative on what method is the fastest to train.
-Nonetheless, SHAC++ is definetely more computationally heavy this is because it needs to train the reward and world model in addition to what SHAC does.
-However, for a comparison consider that for the sampling enviroment (which is the slowest in VMAS) SHAC++ takes 16 hours to complete the 20000 episodes similarly to PPO, while SHAC takes about 10 hours.
+Nonetheless, SHAC++ is definitely more computationally heavy this is because it needs to train the reward and world model in addition to what SHAC does.
+However, for a comparison consider that for the sampling environment (which is the slowest in VMAS) SHAC++ takes 16 hours to complete the 20000 episodes similarly to PPO, while SHAC takes about 10 hours.
 
@@
 We will revise the introduction to better reflect the goal of the paper.
-Briefly, we believe that SHAC demostrated excellent perfomance in single-agent and differentiable scenarios.
+Briefly, we believe that SHAC demonstrated excellent performance in single-agent and differentiable scenarios.
 In this work, we aim to lift these requirements (single-agent and differentiability) to broaden the applicability of SHAC.
-This led us to develop SHAC++ that can handle multi-agent and non-differentiable environments while mantaining performance comparable to SHAC.
+This led us to develop SHAC++ that can handle multi-agent and non-differentiable environments while maintaining performance comparable to SHAC.
 
@@
 
-**Skewed results**: In VMAS as in many other enviroments, the agents are able to experience only local observetions and not the whole world state.
+**Skewed results**: In VMAS as in many other environments, the agents are able to experience only local observations and not the whole world state.
 While SHAC++ remains faithful to accessing only this partial information, SHAC access the whole global information through its gradients.
@@
 
-We believe that the chose enviroments offer a good degree of complexity especially when the number of agents is increased.
+We believe that the chose environments offer a good degree of complexity especially when the number of agents is increased.
 This is also reflected by the poor performance achieved by PPO/MAPPO.
@@
 
-We understand that more complex enviroments with more agents may offer more challenging scenarios and more compelling results.
-However, we want to stress that these enviroments are already incapacitating for PPO.
+We understand that more complex environments with more agents may offer more challenging scenarios and more compelling results.
+However, we want to stress that these environments are already incapacitating for PPO.
 [RUN experiment with more agents].
-Further, we will consider adding more complex enviroments for future revision, we kindly ask the reviewer to suggest a differentiable enviroment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
+Further, we will consider adding more complex environments for future revision, we kindly ask the reviewer to suggest a differentiable environment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
 
@@
 **Why MAPPO and not the others**: While HAPPO and QMIX are both valid candidate for baseline.
-In order to offers an ablation study and a degree of statistical significance we had to limit the number of baselines.
+In order to offer an ablation study and a degree of statistical significance we had to limit the number of baselines.
 We also note that HAPPO authors designed their algorithm to be non-homogeneous (parameters between agents are not shared).
-This choice renders its application problematic as istantiating and traiing several NN at the same time quickly becomes computationally infeasible.
-Meanwhile MAPPO can be applied with both shared and non-shared parameters.
+This choice renders its application problematic as instantiating and training several NNs at the same time quickly becomes computationally infeasible.
+Meanwhile, MAPPO can be applied with both shared and non-shared parameters.
 In our experiments, in order to be consisted while scaling the number of agents, we always apply parameter sharing.
+We will include a more detailed explanation of why we chose MAPPO as a baseline in the paper.
 
@@
 
-**Why not the algorithm in related work section**: We did not include PODS, CE-APG, and BPTT as baselines as we see these a precursors of SHAC.
-The authors of SAM-RL introduce a differentiable renderer on top of the differentiable enviroment which is even more difficult to obtain in practice (Although also this component could be replaced with a learned model in our framework).
+**Why not the algorithm in related work section**: We did not include PODS, CE-APG, and BPTT as baselines as we see these a precursor of SHAC.
+The authors of SAM-RL introduce a differentiable renderer on top of the differentiable environment which is even more difficult to obtain in practice (Although also this component could be replaced with a learned model in our framework).
 Instead, DiffTORI applies a test-time optimization on the agent trajectory.
@@
 Since in some environment their state may be only partially observable then a broad algorithm need to be able to handle this.
-While with SHAC this is not possible, since for gradient computation the global state is needed, SHAC++ can handle this by unsing the local information available to the agents.
+While with SHAC this is not possible, since for gradient computation the global state is needed, SHAC++ can handle this by using the local information available to the agents.
 Despite this choice may skew the performance in favor of SHAC, we believe that this is a more realistic scenario.
@@
 
-This is a deliberate choice that allows us to denote a general framework where the parameters can be arbitrarly shared.
+This is a deliberate choice that allows us to denote a general framework where the parameters can be arbitrary shared.
 For example, all parameter may be separate such as in $\theta=[\theta_{\text{policy}}, \theta_{\text{reward}}, \theta_{\text{value}}, \theta_{\text{world}}]$.
@@
 A full evaluation of this environments even with a small degree of statistical significance would require thousands of runs.
-Therefore, we limited ourselves to the 4 mentioned environment which we believe to be representative of the whole suite.
+Therefore, we limited ourselves to the 4 mentioned environment which we believe to be representative of the whole suite,
+and capture the main challenges we would like to address, like cooperation, partial observability, and stiff dynamics. 
 We will update Sect.A to offer more insight on this choice.

### 2025-03-28 rebuttal.md
on env decision

@@
 In SHAC++ reward and world model learn to predict their output from partial observation. 
-Instead SHAC world model and reward function have access to the full state. 
+Instead, SHAC world model and reward function have access to the full state. 
 This can skew the results in favor of SHAC. Despite this, both algorithms perform similarly.
@@
 but discarded the runs as the performance was too heavily impacted to perform the 3 runs with different seeds. Scaling the number of agents has two main effects:
-1. It becomes easier to discover action that lead to rewards, in this sense, the environment become less sparse.
-2. It becomes more difficult to coordinate between the agents. 
-For the dispersion environment 2. it seems to be the main factor limiting the performance.
-For the transport environment 1. it seems to be the main factor pushing the agents to perform better.
+1) It becomes easier to discover action that lead to rewards, in this sense, the environment become less sparse.
+2) It becomes more difficult to coordinate between the agents to achieve the goal. 
+For the dispersion environment 2) it seems to be the main factor limiting the performance.
+For the transport environment 1) it seems to be the main factor pushing the agents to perform better.
 
@@
 While a full analysis of the components that impact this variance would be definitely valuable, 
-we believe that the number of potential factors (NN arch., NN size, environment type, no. agents) would render this analysis too complex for the scope of this paper.
+we believe that the number of potential factors (NN architecture, NN size, environment type, number of agents) would render this analysis too complex for the scope of this paper.
 
@@
 
-**Skewed results**: In VMAS as in many other environments, the agents are able to experience only local observations and not the whole world state.
-While SHAC++ remains faithful to accessing only this partial information, SHAC access the whole global information through its gradients.
+**Skewed results**: In VMAS as in many other environments, 
+the agents are able to experience only local observations and not the whole world state. While SHAC++ remains faithful to accessing only this partial information, 
+SHAC access the whole global information through its gradients. 
 This puts SHAC in a more favorable position. 
@@
 Despite this, we recognize that the statement in the abstract may be misleading. 
-We will revise the abstract and introduction to better reflect the complexity of the environments.
+We will revise the abstract and introduction to better reflect the complexity of the environments, and we will add more content to the discussion to better explain the choice of the environments.
 
@@
 Ultimately, we believe VMAS to be a good testbed to assess the scaling performance of RL algorithm in general.
+We will add a discussion of the choice of the environments in the paper, comparing them to other environments that could be used for evaluation (e.g., PettingZoo, StarCraft II, etc.).
 

### 2025-03-30 rebuttal.md
minor on why not the full VMAS suite
shorten reply to partial observable env (rev 2)

@@
 We understand that more complex environments with more agents may offer more challenging scenarios and more compelling results.
-However, we want to stress that these environments are already incapacitating for PPO.
-[RUN experiment with more agents].
-Further, we will consider adding more complex environments for future revision, we kindly ask the reviewer to suggest a differentiable environment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
+However, we want to stress that these environments are already incapacitating for PPO/MAPPO.
+Further, we will consider adding more complex environments for future revision (with more agents) to better reflect the performance of SHAC++.
+Finally, we kindly ask the reviewer to suggest a differentiable environment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
 
@@
 
-Yes, this is a deliberate choice on our part.
-Since in some environment their state may be only partially observable then a broad algorithm need to be able to handle this.
-While with SHAC this is not possible, since for gradient computation the global state is needed, SHAC++ can handle this by using the local information available to the agents.
-Despite this choice may skew the performance in favor of SHAC, we believe that this is a more realistic scenario.
-We will include this discussion in the paper.
+Yes, this is intentional for realistic, partially observable environments. SHAC++ can use local information, unlike SHAC (needing global state). We prioritize this realism (details in paper).
 
@@
 The VMAS suite is fairly extensive offering 21 different environments.
-A full evaluation of this environments even with a small degree of statistical significance would require thousands of runs.
+A full evaluation of these environments even with a small degree of statistical significance would require thousands of runs.
 Therefore, we limited ourselves to the 4 mentioned environment which we believe to be representative of the whole suite,

### 2025-03-30 dflex/build/lib/dflex/__init__.py
broken

@@ -1,16 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-from dflex.sim import *
-from dflex.render import *
-from dflex.adjoint import compile
-from dflex import envs
-
-from dflex.util import *
-
-# compiles kernels
-kernel_init()

### 2025-03-30 dflex/build/lib/dflex/adjoint.h

@@ -1,464 +1,0 @@
-#pragma once
-
-#include <cmath>
-#include <stdio.h>
-
-#ifdef CPU
-    #define CUDA_CALLABLE
-    #define __device__
-    #define __host__
-    #define __constant__
-#elif defined(CUDA)
-    #define CUDA_CALLABLE __device__ 
-
-    #include <cuda.h>
-    #include <cuda_runtime_api.h>
-
-    #define check_cuda(code) { check_cuda_impl(code, __FILE__, __LINE__); }
-
-    void check_cuda_impl(cudaError_t code, const char* file, int line)
-    {
-        if (code != cudaSuccess) 
-        {
-            printf("CUDA Error: %s %s %d\n", cudaGetErrorString(code), file, line);
-        }
-    }
-
-    void print_device()
-    {
-        int currentDevice;
-        cudaError_t err = cudaGetDevice(&currentDevice);        
-
-        cudaDeviceProp props;
-        err = cudaGetDeviceProperties(&props, currentDevice);
-        if (err != cudaSuccess)
-            printf("CUDA error: %d\n", err);
-        else
-            printf("%s\n", props.name);
-    }
-
-
-#endif
-
-#ifdef _WIN32
-#define __restrict__ __restrict
-#endif
-
-#define FP_CHECK 0
-
-
-namespace df
-{
-
-template <typename T>
-CUDA_CALLABLE float cast_float(T x) { return (float)(x); }
-
-template <typename T>
-CUDA_CALLABLE int cast_int(T x) { return (int)(x); }
-
-template <typename T>
-CUDA_CALLABLE void adj_cast_float(T x, T& adj_x, float adj_ret) { adj_x += adj_ret; }
-
-template <typename T>
-CUDA_CALLABLE void adj_cast_int(T x, T& adj_x, int adj_ret) { adj_x += adj_ret; }
-
-// avoid namespacing of float type for casting to float type, this is to avoid wp::float(x), which is not valid in C++
-#define float(x) cast_float(x)
-#define adj_float(x, adj_x, adj_ret) adj_cast_float(x, adj_x, adj_ret)
-
-#define int(x) cast_int(x)
-#define adj_int(x, adj_x, adj_ret) adj_cast_int(x, adj_x, adj_ret)
-
-
-#define kEps 0.0f
-
-
-// basic ops for integer types
[…]

### 2025-03-30 dflex/build/lib/dflex/config.py

@@ -1,12 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import os
-
-no_grad = False  # disable adjoint tracking
-check_grad = False  # will perform numeric gradient checking after each launch
-verify_fp = False  # verify inputs and outputs are finite after each launch

### 2025-03-30 dflex/build/lib/dflex/envs/__init__.py

@@ -1,16 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-from .ant import AntEnv
-from .anymal import AnymalEnv
-from .cartpole_swing_up import CartPoleSwingUpEnv
-from .cheetah import CheetahEnv
-from .dflex_env import DFlexEnv
-from .hopper import HopperEnv
-from .humanoid import HumanoidEnv
-from .snu_humanoid import SNUHumanoidEnv
-from .double_pendulum import DoublePendulumEnv

### 2025-03-30 dflex/build/lib/dflex/envs/ant.py

@@ -1,303 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-
-
-class AntEnv(DFlexEnv):
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=4096,
-        episode_length=1000,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=16,
-        early_termination=True,
-        jacobian=False,
-        contact_ke=4.0e4,
-        contact_kd=None,  #  1.0e4,
-        logdir=None,
-        nan_state_fix=False,
-        jacobian_norm=None,
-        termination_height=0.27,
-        action_penalty=0.0,
-        joint_vel_obs_scaling=0.1,
-        up_rew_scale=0.1,
-    ):
-        num_obs = 37
-        num_act = 8
-
-        super(AntEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination = early_termination
-        self.contact_ke = contact_ke
-        self.contact_kd = contact_kd if contact_kd is not None else contact_ke / 4.0
-
-        self.init_sim()
-
-        # other parameters
-        self.termination_height = termination_height
-        self.action_penalty = action_penalty
-        self.joint_vel_obs_scaling = joint_vel_obs_scaling
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/anymal.py

@@ -1,335 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import os
-import sys
-import time
-import urchin
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-
-class AnymalEnv(DFlexEnv):
-    """
-    TODO
-    """
-
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=64,
-        episode_length=1000,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=16,
-        early_termination=True,
-        jacobian=False,
-        logdir=None,
-        nan_state_fix=False,
-        jacobian_norm=None,
-        termination_height=0.25,
-        action_penalty=-0.005,
-        up_rew_scale=0.1,
-        heading_rew_scale=1.0,
-        heigh_rew_scale=1.0,
-    ):
-        num_obs = 49
-        num_act = 12
-
-        super(AnymalEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination = early_termination
-        self.init_sim()
-
-        # MDP parameters
-        self.action_strength = 200.0
-        self.joint_vel_obs_scaling = 0.1
-
-        # reward parameters
-        self.termination_height = termination_height
-        self.action_penalty = action_penalty
-        self.up_rew_scale = up_rew_scale
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/cartpole_swing_up.py

@@ -1,204 +1,0 @@
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-
-class CartPoleSwingUpEnv(DFlexEnv):
-    """ "
-    The real state of the system is [x, x_dot, theta, theta_dot]
-        where x is position on slide and theta is angle of the pendulum.
-        theta=0 is pendulum pointing upwards
-    The observations are [x, x_dot, sin(theta), cos(theta), theta_dot]
-    The actions are [x_ddot]
-    """
-
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=1024,
-        episode_length=240,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=1,
-        jacobian=False,
-        start_state=[0.0, 0.0, 0.0, 0.0],
-        logdir=None,
-        nan_state_fix=False,
-        jacobian_norm=None,
-    ):
-        num_obs = 5
-        num_act = 1
-
-        super(CartPoleSwingUpEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination=False
-        self.start_state = np.array(start_state)
-        assert self.start_state.shape[0] == 4, self.start_state
-
-        self.init_sim()
-
-        # action parameters
-        self.action_strength = 1000.0
-
-        # loss related
-        self.pole_angle_penalty = 1.0
-        self.pole_velocity_penalty = 0.1
-
-        self.cart_position_penalty = 0.05
-        self.cart_velocity_penalty = 0.1
-
-        self.cart_action_penalty = 0.0
-
-        self.setup_visualizer(logdir)
-
-
-    def init_sim(self):
-        self.builder = df.sim.ModelBuilder()
-
-        self.dt = 1.0 / 60.0
-        self.sim_substeps = 4
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/double_pendulum.py

@@ -1,201 +1,0 @@
-# TODO need tom finish this bad boy
-
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-
-class DoublePendulumEnv(DFlexEnv):
-    """ "
-    The real state of the system is [theta1, theta2, theta1_dot, theta2_dot]
-        where the thetas are the angle of the joints. Robot also known as Acrobot
-    The observations are [cos(theta1), sin(theta1), cos(theta2), sin(theta2), theta1_dot, theta2_dot]
-    The actions are [theta2_torque] first joint is not controlled!
-    """
-
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=1024,
-        episode_length=240,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=1,
-        jacobian=False,
-        start_state=[0.0, 0.0, 0.0, 0.0],
-        logdir=None,
-        nan_state_fix=False,
-        jacobian_norm=None,
-    ):
-        num_obs = 6
-        num_act =1
-
-        super(DoublePendulumEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination=False
-        self.start_state = np.array(start_state)
-        assert self.start_state.shape[0] == 4, self.start_state
-
-        self.init_sim()
-
-        # action parameters
-        self.action_strength = 1000.0
-
-        # loss related
-        self.pole_angle_penalty = 1.0
-        self.pole_velocity_penalty = 0.1
-
-        self.setup_visualizer(logdir)
-
-
-    def init_sim(self):
-        self.builder = df.sim.ModelBuilder()
-
-        self.dt = 1.0 / 60.0
-        self.sim_substeps = 4
-        self.sim_dt = self.dt
-
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/hopper.py

@@ -1,267 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-
-class HopperEnv(DFlexEnv):
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=4096,
-        episode_length=1000,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=16,
-        early_termination=True,
-        jacobian=False,
-        contact_ke=2.0e4,
-        contact_kd=None,  #  1.0e3,
-        logdir=None,
-        nan_state_fix=False,
-        jacobian_norm=None,
-        termination_height=-0.45,
-        termination_angle=np.pi / 6.0,
-        termination_height_tolerance=0.15,
-        termination_angle_tolerance=0.05,
-        height_rew_scale=1.0,
-        angle_rew_scale=1.0,
-        action_penalty=-1e-1,
-    ):
-        num_obs = 11
-        num_act = 3
-
-        super(HopperEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination = early_termination
-        self.contact_ke = contact_ke
-        self.contact_kd = contact_kd if contact_kd is not None else contact_ke / 10.0
-
-        self.init_sim()
-
-        # other parameters
-        self.termination_height = termination_height
-        self.termination_angle = termination_angle
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/humanoid.py

@@ -1,388 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-class HumanoidEnv(DFlexEnv):
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=4096,
-        episode_length=1000,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=1,
-        early_termination=True,
-        jacobian=False,
-        contact_ke=2.0e4,
-        contact_kd=None,
-        logdir=None,
-        nan_state_fix=True,  # humanoid env needs this
-        jacobian_norm=None,
-        termination_height=0.74,
-        action_penalty=-0.002,
-        joint_vel_obs_scaling=0.1,
-        termination_tolerance=0.1,
-        height_rew_scale=10.0,
-        up_rew_scale=0.1,
-        heading_rew_scale=1.0,
-        height_rew_type="xu"
-    ):
-        num_obs = 76
-        num_act = 21
-
-        super(HumanoidEnv, self).__init__(
-            num_envs,
-            num_obs,
-            num_act,
-            episode_length,
-            MM_caching_frequency,
-            no_grad,
-            render,
-            nan_state_fix,
-            jacobian_norm,
-            stochastic_init,
-            jacobian,
-            device,
-        )
-
-        self.early_termination = early_termination
-        self.contact_ke = contact_ke
-        self.contact_kd = contact_kd if contact_kd is not None else contact_ke / 10.0
-
-        self.init_sim()
-
-        # other parameters
-        self.termination_height = termination_height
[…]

### 2025-03-30 dflex/build/lib/dflex/envs/torch_utils.py

@@ -1,285 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import gc
-import torch
-from time import sleep
-
-log_output = ""
-
-
-def log(s):
-    print(s)
-    global log_output
-    log_output = log_output + s + "\n"
-
-
-# short hands
-
-
-# torch quat/vector utils
-
-
-def to_torch(x, dtype=torch.float, device="cuda:0", requires_grad=False):
-    return torch.tensor(x, dtype=dtype, device=device, requires_grad=requires_grad)
-
-
-@torch.jit.script
-def quat_mul(a, b):
-    assert a.shape == b.shape
-    shape = a.shape
-    a = a.reshape(-1, 4)
-    b = b.reshape(-1, 4)
-
-    x1, y1, z1, w1 = a[:, 0], a[:, 1], a[:, 2], a[:, 3]
-    x2, y2, z2, w2 = b[:, 0], b[:, 1], b[:, 2], b[:, 3]
-    ww = (z1 + x1) * (x2 + y2)
-    yy = (w1 - y1) * (w2 + z2)
-    zz = (w1 + y1) * (w2 - z2)
-    xx = ww + yy + zz
-    qq = 0.5 * (xx + (z1 - x1) * (x2 - y2))
-    w = qq - ww + (z1 - y1) * (y2 - z2)
-    x = qq - xx + (x1 + w1) * (x2 + w2)
-    y = qq - yy + (w1 - x1) * (y2 + z2)
-    z = qq - zz + (z1 + y1) * (w2 - x2)
-
-    quat = torch.stack([x, y, z, w], dim=-1).view(shape)
-
-    return quat
-
-
-@torch.jit.script
-def normalize(x, eps: float = 1e-9):
-    return x / x.norm(p=2, dim=-1).clamp(min=eps, max=None).unsqueeze(-1)
-
-
-@torch.jit.script
-def quat_apply(a, b):
-    shape = b.shape
-    a = a.reshape(-1, 4)
-    b = b.reshape(-1, 3)
-    xyz = a[:, :3]
[…]

### 2025-03-30 dflex/build/lib/dflex/mat22.h

@@ -1,140 +1,0 @@
-#pragma once
-
-//----------------------------------------------------------
-// mat22
-
-struct mat22
-{
-    inline CUDA_CALLABLE mat22(float m00=0.0f, float m01=0.0f, float m10=0.0f, float m11=0.0f) 
-    {
-        data[0][0] = m00;
-        data[1][0] = m10;
-        data[0][1] = m01;
-        data[1][1] = m11;
-    }
-
-    // row major storage assumed to be compatible with PyTorch
-    float data[2][2];
-};
-
-#ifdef CUDA
-inline __device__ void atomic_add(mat22 * addr, mat22 value) {
-    // *addr += value;
-    atomicAdd(&((addr -> data)[0][0]), value.data[0][0]);
-    atomicAdd(&((addr -> data)[0][1]), value.data[0][1]);
-    atomicAdd(&((addr -> data)[1][0]), value.data[1][0]);
-    atomicAdd(&((addr -> data)[1][1]), value.data[1][1]);
-}
-#endif
-
-inline CUDA_CALLABLE void adj_mat22(float m00, float m01, float m10, float m11, float& adj_m00, float& adj_m01, float& adj_m10, float& adj_m11, const mat22& adj_ret)
-{
-    printf("todo\n");
-}
-
-inline CUDA_CALLABLE float index(const mat22& m, int row, int col)
-{
-    return m.data[row][col];
-}
-
-inline CUDA_CALLABLE mat22 add(const mat22& a, const mat22& b)
-{
-    mat22 t;
-    for (int i=0; i < 2; ++i)
-    {
-        for (int j=0; j < 2; ++j)
-        {
-            t.data[i][j] = a.data[i][j] + b.data[i][j];
-        }
-    }
-
-    return t;
-}
-
-
-inline CUDA_CALLABLE mat22 mul(const mat22& a, float b)
-{
-    mat22 t;
-    for (int i=0; i < 2; ++i)
-    {
-        for (int j=0; j < 2; ++j)
-        {
-            t.data[i][j] = a.data[i][j]*b;
-        }
-    }
-
-    return t;
-}
-
-inline CUDA_CALLABLE mat22 mul(const mat22& a, const mat22& b)
-{
-    mat22 t;
-    for (int i=0; i < 2; ++i)
-    {
-        for (int j=0; j < 2; ++j)
-        {
[…]

### 2025-03-30 dflex/build/lib/dflex/mat33.h

@@ -1,260 +1,0 @@
-#pragma once
-
-//----------------------------------------------------------
-// mat33
-
-struct mat33
-{
-    inline CUDA_CALLABLE mat33(float3 c0, float3 c1, float3 c2)
-    {
-        data[0][0] = c0.x;
-        data[1][0] = c0.y;
-        data[2][0] = c0.z;
-
-        data[0][1] = c1.x;
-        data[1][1] = c1.y;
-        data[2][1] = c1.z;
-
-        data[0][2] = c2.x;
-        data[1][2] = c2.y;
-        data[2][2] = c2.z;
-    }
-
-    inline CUDA_CALLABLE mat33(float m00=0.0f, float m01=0.0f, float m02=0.0f,
-                 float m10=0.0f, float m11=0.0f, float m12=0.0f,
-                 float m20=0.0f, float m21=0.0f, float m22=0.0f) 
-    {
-        data[0][0] = m00;
-        data[1][0] = m10;
-        data[2][0] = m20;
-
-        data[0][1] = m01;
-        data[1][1] = m11;
-        data[2][1] = m21;
-
-        data[0][2] = m02;
-        data[1][2] = m12;
-        data[2][2] = m22;
-    }
-
-    CUDA_CALLABLE float3 get_row(int index) const
-    {
-        return (float3&)data[index]; 
-    }
-
-    CUDA_CALLABLE void set_row(int index, const float3& v)
-    {
-        (float3&)data[index] = v;
-    }
-
-    CUDA_CALLABLE float3 get_col(int index) const
-    {
-        return float3(data[0][index], data[1][index], data[2][index]);
-    }
-
-    CUDA_CALLABLE void set_col(int index, const float3& v)
-    {
-        data[0][index] = v.x;
-        data[1][index] = v.y;
-        data[2][index] = v.z;
-    }
-
-    // row major storage assumed to be compatible with PyTorch
-    float data[3][3];
-};
-
-#ifdef CUDA
-inline __device__ void atomic_add(mat33 * addr, mat33 value) {
-    atomicAdd(&((addr -> data)[0][0]), value.data[0][0]);
-    atomicAdd(&((addr -> data)[1][0]), value.data[1][0]);
-    atomicAdd(&((addr -> data)[2][0]), value.data[2][0]);
[…]

### 2025-03-30 dflex/build/lib/dflex/matnn.h

@@ -1,352 +1,0 @@
-#pragma once
-
-
-CUDA_CALLABLE inline int dense_index(int stride, int i, int j)
-{
-    return i*stride + j;
-}
-
-template <bool transpose>
-CUDA_CALLABLE inline int dense_index(int rows, int cols, int i, int j)
-{
-    if (transpose)
-        return j*rows + i;
-    else
-        return i*cols + j;
-}
-
-
-#ifdef CPU
-
-const int kNumThreadsPerBlock = 1;
-
-template <bool t1, bool t2, bool add>
-CUDA_CALLABLE inline void dense_gemm_impl(int m, int n, int p, const float* __restrict__ A, const float* __restrict__ B, float* __restrict__ C)
-{
-    for (int i=0; i < m; i++)
-    {
-        for (int j=0; j < n; ++j)
-        {
-            float sum = 0.0f;
-
-            for (int k=0; k < p; ++k)
-            {
-                sum += A[dense_index<t1>(m, p, i, k)]*B[dense_index<t2>(p, n, k, j)];
-            }
-            
-            if (add)
-                C[i*n + j] += sum;
-            else
-                C[i*n + j] = sum;
-        }
-    }
-}
-
-#else
-
-const int kNumThreadsPerBlock = 256;
-
-template <bool t1, bool t2, bool add>
-CUDA_CALLABLE inline void dense_gemm_impl(int m, int n, int p, const float* __restrict__ A, const float* __restrict__ B, float* __restrict__ C)
-{
-    // each thread in the block calculates an output (or more if output dim > block dim)
-    for (int e=threadIdx.x; e < m*n; e += blockDim.x)
-    {
-        const int i=e/n;
-        const int j=e%n;
-
-        float sum = 0.0f;
-
-        for (int k=0; k < p; ++k)
-        {
-            sum += A[dense_index<t1>(m, p, i, k)]*B[dense_index<t2>(p, n, k, j)];
-        }
-        
-        if (add)
-            C[i*n + j] += sum;
-        else
-            C[i*n + j] = sum;
-    }
-}
-
-#endif
-
-
-template <bool add=false>
-CUDA_CALLABLE inline void dense_gemm(int m, int n, int p, int t1, int t2, const float* __restrict__ A, const float* __restrict__ B, float* __restrict__ C)
-{
-    if (t1 == 0 && t2 == 0)
-        dense_gemm_impl<false, false, add>(m, n, p, A, B, C);
[…]

### 2025-03-30 dflex/build/lib/dflex/quat.h

@@ -1,321 +1,0 @@
-#pragma once
-
-struct quat
-{
-    // imaginary part
-    float x;
-    float y;
-    float z;
-    
-    // real part
-    float w;
-
-    inline CUDA_CALLABLE quat(float x=0.0f, float y=0.0f, float z=0.0f, float w=0.0) : x(x), y(y), z(z), w(w) {}    
-    explicit inline CUDA_CALLABLE quat(const float3& v, float w=0.0f) : x(v.x), y(v.y), z(v.z), w(w) {}
-};
-
-#ifdef CUDA
-inline __device__ void atomic_add(quat * addr, quat value) {
-    atomicAdd(&(addr -> x), value.x);
-    atomicAdd(&(addr -> y), value.y);
-    atomicAdd(&(addr -> z), value.z);
-    atomicAdd(&(addr -> w), value.w);
-}
-#endif
-
-inline CUDA_CALLABLE void adj_quat(float x, float y, float z, float w, float& adj_x, float& adj_y, float& adj_z, float& adj_w, quat adj_ret)
-{
-    adj_x += adj_ret.x;
-    adj_y += adj_ret.y;
-    adj_z += adj_ret.z;
-    adj_w += adj_ret.w;
-}
-
-inline CUDA_CALLABLE void adj_quat(const float3& v, float w, float3& adj_v, float& adj_w, quat adj_ret)
-{
-    adj_v.x += adj_ret.x;
-    adj_v.y += adj_ret.y;
-    adj_v.z += adj_ret.z;
-    adj_w   += adj_ret.w;
-}
-
-// foward methods
-
-inline CUDA_CALLABLE quat quat_from_axis_angle(const float3& axis, float angle)
-{
-    float half = angle*0.5f;
-    float w = cosf(half);
-
-    float sin_theta_over_two = sinf(half);
-    float3 v = axis*sin_theta_over_two;
-
-    return quat(v.x, v.y, v.z, w);
-}
-
-inline CUDA_CALLABLE quat quat_identity()
-{
-    return quat(0.0f, 0.0f, 0.0f, 1.0f);
-}
-
-inline CUDA_CALLABLE float dot(const quat& a, const quat& b)
-{
[…]

### 2025-03-30 dflex/build/lib/dflex/spatial.h

@@ -1,841 +1,0 @@
-#pragma once
-
-//---------------------------------------------------------------------------------
-// Represents a twist in se(3)
-
-struct spatial_vector
-{
-    float3 w;
-    float3 v;
-
-    CUDA_CALLABLE inline spatial_vector(float a, float b, float c, float d, float e, float f) : w(a, b, c), v(d, e, f) {}
-    CUDA_CALLABLE inline spatial_vector(float3 w=float3(), float3 v=float3()) : w(w), v(v) {}
-    CUDA_CALLABLE inline spatial_vector(float a) : w(a, a, a), v(a, a, a) {}
-
-    CUDA_CALLABLE inline float operator[](int index) const
-    {
-        assert(index < 6);
-
-        return (&w.x)[index];
-    }
-
-    CUDA_CALLABLE inline float& operator[](int index)
-    {
-        assert(index < 6);
-
-        return (&w.x)[index];
-    }
-};
-
-CUDA_CALLABLE inline spatial_vector operator - (spatial_vector a)
-{
-    return spatial_vector(-a.w, -a.v);
-}
-
-
-CUDA_CALLABLE inline spatial_vector add(const spatial_vector& a, const spatial_vector& b)
-{
-    return { a.w + b.w, a.v + b.v };
-}
-
-CUDA_CALLABLE inline spatial_vector sub(const spatial_vector& a, const spatial_vector& b)
-{
-    return { a.w - b.w, a.v - b.v };
-}
-
-CUDA_CALLABLE inline spatial_vector mul(const spatial_vector& a, float s)
-{
-    return { a.w*s, a.v*s };
-}
-
-CUDA_CALLABLE inline float spatial_dot(const spatial_vector& a, const spatial_vector& b)
-{
-    return dot(a.w, b.w) + dot(a.v, b.v);
-}
-
-CUDA_CALLABLE inline spatial_vector spatial_cross(const spatial_vector& a,  const spatial_vector& b)
-{
-    float3 w = cross(a.w, b.w);
-    float3 v = cross(a.v, b.w) + cross(a.w, b.v);
-    
-    return spatial_vector(w, v);
-}
-
-CUDA_CALLABLE inline spatial_vector spatial_cross_dual(const spatial_vector& a,  const spatial_vector& b)
-{
[…]

### 2025-03-30 dflex/build/lib/dflex/util.py

@@ -1,539 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import timeit
-import math
-import numpy as np
-import gc
-import torch
-import cProfile
-
-log_output = ""
-
-
-def log(s):
-    print(s)
-    global log_output
-    log_output = log_output + s + "\n"
-
-
-# short hands
-
-
-def length(a):
-    return np.linalg.norm(a)
-
-
-def length_sq(a):
-    return np.dot(a, a)
-
-
-# NumPy has no normalize() method..
-def normalize(v):
-    norm = np.linalg.norm(v)
-    if norm == 0.0:
-        return v
-    return v / norm
-
-
-def skew(v):
-    return np.array([[0, -v[2], v[1]], [v[2], 0, -v[0]], [-v[1], v[0], 0]])
-
-
-# math utils
-def quat(i, j, k, w):
-    return np.array([i, j, k, w])
-
-
-def quat_identity():
-    return np.array((0.0, 0.0, 0.0, 1.0))
-
-
-def quat_inverse(q):
-    return np.array((-q[0], -q[1], -q[2], q[3]))
-
-
-def quat_from_axis_angle(axis, angle):
-    v = np.array(axis)
-
-    half = angle * 0.5
-    w = math.cos(half)
-
-    sin_theta_over_two = math.sin(half)
-    v *= sin_theta_over_two
-
-    return np.array((v[0], v[1], v[2], w))
-
-
-# rotate a vector
-def quat_rotate(q, x):
-    x = np.array(x)
-    axis = np.array((q[0], q[1], q[2]))
-    return (
-        x * (2.0 * q[3] * q[3] - 1.0)
-        + np.cross(axis, x) * q[3] * 2.0
-        + axis * np.dot(axis, x) * 2.0
-    )
-
-
-# multiply two quats
-def quat_multiply(a, b):
-    return np.array(
-        (
[…]

### 2025-03-30 dflex/dflex/adjoint.m

@@ -1,75 +1,0 @@
-clear all
-
-% quat operations and adjoint
-
-syms ax ay az aw bx by bz bw real
-
-% quat multiplication
-% q = a*b
-
-q = [aw*bx + bw*ax + ay*bz - by*az;
-     aw*by + bw*ay + az*bx - bz*ax;
-     aw*bz + bw*az + ax*by - bx*ay;
-     aw*bw - ax*bx - ay*by - az*bz]
- 
- % Jacobians w.r.t a and b
- dqda = [diff(q, ax), diff(q, ay), diff(q, az), diff(q, aw)]
- dqdb = [diff(q, bx), diff(q, by), diff(q, bz), diff(q, bw)]
- 
- syms rx ry rz rw real
- 
- % J^T*adj_q
- adj_a = dqda'*[rx; ry; rz; rw]
- adj_b = dqdb'*[rx; ry; rz; rw]
- 
- syms px py pz real
- syms qx qy qz qw real
- 
- p = [px; py; pz]
- qi = [qx; qy; qz]
- 
- % vector rotation
- r = p*(2*qw*qw-1) + cross(qi, p)*qw*2.0 + qi*(qi'*p)*2.0
- 
- drdx = [diff(r, px), diff(r, py), diff(r, pz)]
- drdqi = [diff(r, qx), diff(r, qy), diff(r, qz)]
- drdqw = diff(r, qw)
- 
- adj_x = drdx'*[rx; ry; rz]
- adj_qi = drdqi'*[rx; ry; rz]
- adj_qw = drdqw'*[rx; ry; rz]
- 
- ccode(adj_x, 'File', 'adj_x.c')
- ccode([adj_qi; adj_qw], 'File', 'adj_q.c')
- 
- % quat normalize
- q = [qx; qy; qz; qw]
- qhat = q/sqrt(q'*q)
- 
- dqhatdq = [diff(qhat, qx), diff(qhat, qy), diff(qhat, qz), diff(qhat, qw)]
- 
- adj_q = dqhatdq'*q
- 
- % exp map
- syms wx wy wz vx vy vz angle real
- syms adj_tx adj_ty adj_tz real
- 
- v = [vx vy vz]'
- w = [wx wy wz]'
- 
- t = angle*v + cross(w, v)*(1 - cos(angle)) + cross(w, cross(w, v))*(angle - sin(angle))
- 
[…]

### 2025-03-30 dflex/dflex/adjoint.py

@@ -1,2423 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import os
-import imp
-import ast
-import math
-import inspect
-import typing
-import weakref
-import torch
-import torch.utils.cpp_extension
-
-import dflex.config
-
-# Todo
-# -----
-#
-# [ ] Unary ops (e.g.: -)
-# [ ] Inplace ops (e.g.: +=, -=)
-# [ ] Conditionals
-# [ ] Loops (unrolled)
-# [ ] Auto-gen PyTorch operator
-# [ ] CUDA kernel code gen + dynamic compilation
-
-# -----
-
-operators = {}
-functions = {}
-cuda_functions = {}
-kernels = {}
-
-# ----------------------
-# built-in types
-
-
-class float3:
-    def __init__(self):
-        x = 0.0
-        y = 0.0
-        z = 0.0
-
-
-class float4:
-    def __init__(self):
-        x = 0.0
-        y = 0.0
-        z = 0.0
-        w = 0.0
-
-
-class quat:
-    def __init__(self):
-        x = 0.0
-        y = 0.0
-        z = 0.0
-        w = 1.0
-
-
-class mat22:
-    def __init__(self):
-        pass
-
-
-class mat33:
-    def __init__(self):
-        pass
-
-
-class spatial_vector:
-    def __init__(self):
-        pass
-
-
-class spatial_matrix:
-    def __init__(self):
-        pass
-
-
-class spatial_transform:
-    def __init__(self):
-        pass
-
-
-class void:
-    def __init__(self):
-        pass
-
-
-class tensor:
-    def __init__(self, type):
-        self.type = type
-        self.requires_grad = True
-        self.__name__ = "tensor<" + type.__name__ + ">"
-
-
-# ----------------------
-
-
-# register built-in function
-def builtin(key):
-    def insert(func):
-        func.key = key
-        func.prefix = "df::"
-        functions[key] = func
-        return func
-
-    return insert
-
-
-# ---------------------------------
-# built-in operators +,-,*,/
-
-
-@builtin("add")
-class AddFunc:
-    @staticmethod
[…]

### 2025-03-30 dflex/dflex/envs/dflex_env.py

@@ -1,416 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import os
-import sys
-from abc import abstractmethod
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import numpy as np
-import torch
-
-import dflex as df
-
-try:
-    from pxr import Usd
-except ModuleNotFoundError:
-    print("No pxr package")
-
-from gym import spaces
-
-
-def jacobian(output, input, max_out_dim=None):
-    """Computes the jacobian of output tensor with respect to the input"""
-    num_envs, input_dim = input.shape
-    output_dim = output.shape[1]
-    if max_out_dim:
-        output_dim = min(output_dim, max_out_dim)
-    jacobians = torch.zeros((num_envs, output_dim, input_dim), dtype=torch.float32)
-    for out_idx in range(output_dim):
-        select_index = torch.zeros(output.shape[1])
-        select_index[out_idx] = 1.0
-        e = torch.tile(select_index, (num_envs, 1)).cuda()
-        # retain = out_idx != output_dim - 1  # NOTE: experimental
-        try:
-            (grad,) = torch.autograd.grad(
-                outputs=output, inputs=input, grad_outputs=e, retain_graph=True
-            )
-            jacobians[:, out_idx, :] = grad.view(num_envs, input_dim)
-        except RuntimeError as err:
-            print(f"WARN: Couldn't compute jacobian for {out_idx} index")
-            print(err)
-
-
-class DFlexEnv:
-    def __init__(
-        self,
-        num_envs,
-        num_obs,
-        num_act,
-        episode_length,
-        MM_caching_frequency=1,
-        no_grad=True,
-        render=False,
-        nan_state_fix=False,
-        jacobian_norm=None,
-        stochastic_init=False,
-        jacobian=False,
-        device="cuda:0",
-    ):
-        self.no_grad = no_grad
-        df.config.no_grad = self.no_grad
-        self.nan_state_fix = nan_state_fix
-        self.jacobian_norm = jacobian_norm
-        # if true resets all envs on earfly termination
[…]

### 2025-03-30 dflex/dflex/envs/load_utils.py

@@ -1,773 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import urchin
-import math
-import numpy as np
-import os
-import torch
-import random
-
-import xml.etree.ElementTree as ET
-
-import dflex as df
-
-
-def set_np_formatting():
-    np.set_printoptions(
-        edgeitems=30,
-        infstr="inf",
-        linewidth=4000,
-        nanstr="nan",
-        precision=2,
-        suppress=False,
-        threshold=10000,
-        formatter=None,
-    )
-
-
-def set_seed(seed, torch_deterministic=False):
-    if seed == -1 and torch_deterministic:
-        seed = 42
-    elif seed == -1:
-        seed = np.random.randint(0, 10000)
-    print("Setting seed: {}".format(seed))
-
-    random.seed(seed)
-    np.random.seed(seed)
-    torch.manual_seed(seed)
-    os.environ["PYTHONHASHSEED"] = str(seed)
-    torch.cuda.manual_seed(seed)
-    torch.cuda.manual_seed_all(seed)
-
-    if torch_deterministic:
-        # refer to https://docs.nvidia.com/cuda/cublas/index.html#cublasApi_reproducibility
-        os.environ["CUBLAS_WORKSPACE_CONFIG"] = ":4096:8"
-        torch.backends.cudnn.benchmark = False
-        torch.backends.cudnn.deterministic = True
-        torch.use_deterministic_algorithms(True)
-    else:
-        torch.backends.cudnn.benchmark = True
-        torch.backends.cudnn.deterministic = False
-
-    return seed
-
-
-def urdf_add_collision(
-    builder, link, collisions, shape_ke, shape_kd, shape_kf, shape_mu
-):
-    # add geometry
-    for collision in collisions:
-        origin = urchin.matrix_to_xyz_rpy(collision.origin)
-
-        pos = origin[0:3]
-        rot = df.rpy2quat(*origin[3:6])
-
-        geo = collision.geometry
-
-        if geo.box:
-            builder.add_shape_box(
-                link,
-                pos,
-                rot,
-                geo.box.size[0] * 0.5,
-                geo.box.size[1] * 0.5,
[…]

### 2025-03-30 dflex/dflex/envs/snu_humanoid.py

@@ -1,462 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import os
-import sys
-
-import torch
-
-from .dflex_env import DFlexEnv
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-import dflex.envs.load_utils as lu
-import dflex.envs.torch_utils as tu
-
-# Used for rendering. If not installed, the render function will not work
-try:
-    from pxr import Usd, UsdGeom, Gf
-except ModuleNotFoundError:
-    print("No pxr package. Rendering wil not work!")
-
-
-class SNUHumanoidEnv(DFlexEnv):
-    def __init__(
-        self,
-        render=False,
-        device="cuda:0",
-        num_envs=4096,
-        episode_length=1000,
-        no_grad=True,
-        stochastic_init=False,
-        MM_caching_frequency=1,
-        early_termination=True,
-        jacobian=False,
-        contact_ke=2.0e4,
-        contact_kd=None,
-        logdir=None,
-        nan_state_fix=True,  # humanoid env needs this
-        jacobian_norm=None,
-        termination_height=0.46,
-        termination_tolerance=0.05,
-        height_rew_scale=4.0,
-        up_rew_scale=0.1,
-        heading_rew_scale=1.0,
-        action_penalty=-1e-3,
-        joint_vel_obs_scaling=0.1,
-    ):
-        self.filter = {
-            "Pelvis",
-            "FemurR",
-            "TibiaR",
-            "TalusR",
-            "FootThumbR",
-            "FootPinkyR",
-            "FemurL",
-            "TibiaL",
-            "TalusL",
-            "FootThumbL",
-            "FootPinkyL",
-        }
-
-        self.skeletons = []
-        self.muscle_strengths = []
-
-        self.mtu_actuations = True
-
-        self.inv_control_freq = 1
-
-        # "humanoid_snu_lower"
-        self.num_joint_q = 29
-        self.num_joint_qd = 24
-
[…]

### 2025-03-30 dflex/dflex/model.py

@@ -1,2217 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-"""A module for building simulation models and state.
-"""
-
-import math
-import torch
-import numpy as np
-
-from typing import Tuple
-from typing import List
-
-Vec3 = List[float]
-Vec4 = List[float]
-Quat = List[float]
-Mat33 = List[float]
-Transform = Tuple[Vec3, Quat]
-
-from dflex.util import *
-
-# shape geometry types
-GEO_SPHERE = 0
-GEO_BOX = 1
-GEO_CAPSULE = 2
-GEO_MESH = 3
-GEO_SDF = 4
-GEO_PLANE = 5
-GEO_NONE = 6
-
-# body joint types
-JOINT_PRISMATIC = 0
-JOINT_REVOLUTE = 1
-JOINT_BALL = 2
-JOINT_FIXED = 3
-JOINT_FREE = 4
-
-
-class Mesh:
-    """Describes a triangle collision mesh for simulation
-
-    Attributes:
-
-        vertices (List[Vec3]): Mesh vertices
-        indices (List[int]): Mesh indices
-        I (Mat33): Inertia tensor of the mesh assuming density of 1.0 (around the center of mass)
-        mass (float): The total mass of the body assuming density of 1.0
-        com (Vec3): The center of mass of the body
-    """
-
-    def __init__(self, vertices: List[Vec3], indices: List[int]):
-        """Construct a Mesh object from a triangle mesh
-
-        The mesh center of mass and inertia tensor will automatically be
-        calculated using a density of 1.0. This computation is only valid
-        if the mesh is closed (two-manifold).
-
-        Args:
-            vertices: List of vertices in the mesh
-            indices: List of triangle indices, 3 per-element
-        """
-
-        self.vertices = vertices
-        self.indices = indices
-
-        # compute com and inertia (using density=1.0)
-        com = np.mean(vertices, 0)
-
-        num_tris = int(len(indices) / 3)
-
-        # compute signed inertia for each tetrahedron
-        # formed with the interior point, using an order-2
[…]

### 2025-03-30 dflex/dflex/render.py

@@ -1,554 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-"""This optional module contains a built-in renderer for the USD data
-format that can be used to visualize time-sampled simulation data.
-
-Users should create a simulation model and integrator and periodically
-call :func:`UsdRenderer.update()` to write time-sampled simulation data to the USD stage.
-
-Example:
-
-    >>> # construct a new USD stage
-    >>> stage = Usd.Stage.CreateNew("my_stage.usda")
-    >>> renderer = df.render.UsdRenderer(model, stage)
-    >>>
-    >>> time = 0.0
-    >>>
-    >>> for i in range(100):
-    >>>
-    >>>     # update simulation here
-    >>>     # ....
-    >>>
-    >>>     # update renderer
-    >>>     stage.update(state, time)
-    >>>     time += dt
-    >>>
-    >>> # write stage to file
-    >>> stage.Save()
-
-Note:
-    You must have the Pixar USD bindings installed to use this module
-    please see https://developer.nvidia.com/usd to obtain precompiled
-    USD binaries and installation instructions.
-    
-"""
-
-try:
-    from pxr import Usd, UsdGeom, Gf, Sdf
-except ModuleNotFoundError:
-    print("No pxr package")
-
-
-import dflex.sim
-import dflex.util
-
-import math
-
-
-def _usd_add_xform(prim):
-    prim.ClearXformOpOrder()
-
-    t = prim.AddTranslateOp()
-    r = prim.AddOrientOp()
-    s = prim.AddScaleOp()
-
-
-def _usd_set_xform(xform, transform, scale, time):
-    xform_ops = xform.GetOrderedXformOps()
-
-    pos = tuple(transform[0])
-    rot = tuple(transform[1])
-
-    xform_ops[0].Set(Gf.Vec3d(pos), time)
-    xform_ops[1].Set(Gf.Quatf(rot[3], rot[0], rot[1], rot[2]), time)
-    xform_ops[2].Set(Gf.Vec3d(scale), time)
-
-
[…]

### 2025-03-30 dflex/dflex/sim.py

@@ -1,3356 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-"""This module contains time-integration objects for simulating
-models + state forward in time.
-
-"""
-
-import math
-import torch
-import numpy as np
-
-import dflex.util
-import dflex.adjoint as df
-import dflex.config
-
-from dflex.model import *
-import time
-from copy import deepcopy
-
-# Todo
-# -----
-#
-# [x] Spring model
-# [x] 2D FEM model
-# [x] 3D FEM model
-# [x] Cloth
-#     [x] Wind/Drag model
-#     [x] Bending model
-#     [x] Triangle collision
-# [x] Rigid body model
-# [x] Rigid shape contact
-#     [x] Sphere
-#     [x] Capsule
-#     [x] Box
-#     [ ] Convex
-#     [ ] SDF
-# [ ] Implicit solver
-# [x] USD import
-# [x] USD export
-# -----
-
-# externally compiled kernels module (C++/CUDA code with PyBind entry points)
-kernels = None
-
-
-@df.func
-def test(c: float):
-    x = 1.0
-    y = float(2)
-    z = int(3.0)
-
-    print(y)
-    print(z)
-
-    if c < 3.0:
-        x = 2.0
-
-    return x * 6.0
-
-
-def kernel_init():
-    global kernels
-    kernels = df.compile()
-
-
-@df.kernel
-def integrate_particles(
-    x: df.tensor(df.float3),
-    v: df.tensor(df.float3),
-    f: df.tensor(df.float3),
-    w: df.tensor(float),
-    gravity: df.tensor(df.float3),
-    dt: float,
-    x_new: df.tensor(df.float3),
-    v_new: df.tensor(df.float3),
-):
-    tid = df.tid()
-
-    x0 = df.load(x, tid)
[…]

### 2025-03-30 dflex/dflex/vec2.h

@@ -1,7 +1,0 @@
-#pragma once
-
-struct float2
-{
-    float x;
-    float y;
-};

### 2025-03-30 dflex/dflex/vec3.h

@@ -1,223 +1,0 @@
-#pragma once
-
-struct float3
-{
-    float x;
-    float y;
-    float z;
-
-    inline CUDA_CALLABLE float3(float x=0.0f, float y=0.0f, float z=0.0f) : x(x), y(y), z(z) {}
-    explicit inline CUDA_CALLABLE float3(const float* p) : x(p[0]), y(p[1]), z(p[2]) {}
-};
-
-
-//--------------
-// float3 methods
-
-inline CUDA_CALLABLE float3 operator - (float3 a)
-{
-    return { -a.x, -a.y, -a.z };
-}
-
-
-inline CUDA_CALLABLE float3 mul(float3 a, float s)
-{
-    return { a.x*s, a.y*s, a.z*s };
-}
-
-inline CUDA_CALLABLE float3 div(float3 a, float s)
-{
-    return { a.x/s, a.y/s, a.z/s };
-}
-
-inline CUDA_CALLABLE float3 add(float3 a, float3 b)
-{
-    return { a.x+b.x, a.y+b.y, a.z+b.z };
-}
-
-inline CUDA_CALLABLE float3 add(float3 a, float s)
-{
-    return { a.x + s, a.y + s, a.z + s };
-}
-
-
-inline CUDA_CALLABLE float3 sub(float3 a, float3 b)
-{
-    return { a.x-b.x, a.y-b.y, a.z-b.z };
-}
-
-inline CUDA_CALLABLE float dot(float3 a, float3 b)
-{
-    return a.x*b.x + a.y*b.y + a.z*b.z;
-}
-
-inline CUDA_CALLABLE float3 cross(float3 a, float3 b)
-{
-    float3 c;
-    c.x = a.y*b.z - a.z*b.y;
-    c.y = a.z*b.x - a.x*b.z;
-    c.z = a.x*b.y - a.y*b.x;
-
-    return c;
-}
-
-inline CUDA_CALLABLE float index(const float3 & a, int idx)
-{
-#if FP_CHECK
-    if (idx < 0 || idx > 2)
-    {
-        printf("float3 index %d out of bounds at %s %d\n", idx, __FILE__, __LINE__);
-        exit(1);
-    }
-#endif
-
-    return (&a.x)[idx];
-        
-}
-
-inline CUDA_CALLABLE void adj_index(const float3 & a, int idx, float3 & adj_a, int & adj_idx, float & adj_ret)
-{
-#if FP_CHECK
[…]

### 2025-03-30 dflex/docs/conf.py

@@ -1,75 +1,0 @@
-# Configuration file for the Sphinx documentation builder.
-#
-# This file only contains a selection of the most common options. For a full
-# list see the documentation:
-# https://www.sphinx-doc.org/en/master/usage/configuration.html
-
-# -- Path setup --------------------------------------------------------------
-
-# If extensions (or modules to document with autodoc) are in another directory,
-# add these directories to sys.path here. If the directory is relative to the
-# documentation root, use os.path.abspath to make it absolute, like shown here.
-#
-import os
-import sys
-sys.path.insert(0, os.path.abspath('../dflex'))
-
-
-# -- Project information -----------------------------------------------------
-
-project = 'dFlex'
-copyright = '2020, NVIDIA'
-author = 'NVIDIA'
-
-
-# -- General configuration ---------------------------------------------------
-
-# Add any Sphinx extension module names here, as strings. They can be
-# extensions coming with Sphinx (named 'sphinx.ext.*') or your custom
-# ones.
-extensions = [
-    'sphinx.ext.autodoc',
-    'sphinx.ext.napoleon',
-    'sphinx.ext.intersphinx',
- #   'sphinx.ext.autosummary',
-    'sphinx.ext.todo',
-    'autodocsumm'
-]
-
-# put type hints inside the description instead of the signature (easier to read)
-autodoc_typehints = 'description'
-# document class *and* __init__ methods
-autoclass_content = 'both' # 
-
-todo_include_todos = True
-
-intersphinx_mapping = {
-    'python': ("https://docs.python.org/3", None),
-    'numpy': ('http://docs.scipy.org/doc/numpy/', None),
-    'PyTorch': ('http://pytorch.org/docs/master/', None),
-}
-
-# Add any paths that contain templates here, relative to this directory.
-templates_path = ['_templates']
-
-# List of patterns, relative to source directory, that match files and
-# directories to ignore when looking for source files.
-# This pattern also affects html_static_path and html_extra_path.
-exclude_patterns = ['_build', 'Thumbs.db', '.DS_Store']
-
-# -- Options for HTML output -------------------------------------------------
-
-# The theme to use for HTML and HTML Help pages.  See the documentation for
-# a list of builtin themes.
-#
-import sphinx_rtd_theme
-
-html_theme_path = [sphinx_rtd_theme.get_html_theme_path()]
-html_theme = "sphinx_rtd_theme"
-
-# html_theme = 'alabaster'
-
[…]

### 2025-03-30 dflex/extension/dflex.py

@@ -1,460 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-"""dFlex Kit extension
-
-Allows setting up, training, and running inference on dFlex optimization environments.
-
-"""
-
-import os
-import subprocess
-import carb
-import carb.input
-import math
-import numpy as np
-import omni.kit.ui
-import omni.appwindow
-import omni.kit.editor
-import omni.timeline
-import omni.usd
-import omni.ui as ui
-
-from pathlib import Path
-
-ICON_PATH = Path(__file__).parent.parent.joinpath("icons")
-
-
-from pxr import Usd, UsdGeom, Sdf, Gf
-
-import torch
-
-from omni.kit.settings import create_setting_widget, create_setting_widget_combo, SettingType, get_settings_interface
-
-KIT_GREEN = 0xFF8A8777
-LABEL_PADDING = 120
-
-DARK_WINDOW_STYLE = {
-    "Button": {"background_color": 0xFF292929, "margin": 3, "padding": 3, "border_radius": 2},
-    "Button.Label": {"color": 0xFFCCCCCC},
-    "Button:hovered": {"background_color": 0xFF9E9E9E},
-    "Button:pressed": {"background_color": 0xC22A8778},
-    "VStack::main_v_stack": {"secondary_color": 0x0, "margin_width": 10, "margin_height": 0},
-    "VStack::frame_v_stack": {"margin_width": 15, "margin_height": 10},
-    "Rectangle::frame_background": {"background_color": 0xFF343432, "border_radius": 5},
-    "Field::models": {"background_color": 0xFF23211F, "font_size": 14, "color": 0xFFAAAAAA, "border_radius": 4.0},
-    "Frame": {"background_color": 0xFFAAAAAA},
-    "Label": {"font_size": 14, "color": 0xFF8A8777},
-    "Label::status": {"font_size": 14, "color": 0xFF8AFF77}
-}
-
-CollapsableFrame_style = {
-    "CollapsableFrame": {
-        "background_color": 0xFF343432,
-        "secondary_color": 0xFF343432,
[…]

### 2025-03-30 dflex/format.sh

@@ -1,1 +1,0 @@
-yapf --recursive --in-place .

### 2025-03-30 dflex/setup.py

@@ -1,26 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import setuptools
-
-setuptools.setup(
-    name="dflex",
-    version="0.0.1",
-    author="NVIDIA",
-    author_email="mmacklin@nvidia.com",
-    description="Differentiable Multiphysics for Python",
-    long_description="",
-    long_description_content_type="text/markdown",
-    packages=setuptools.find_packages(),
-    include_package_data=True,
-    classifiers=[
-        "Programming Language :: Python :: 3",  
-        "Operating System :: OS Independent",
-    ],
-    python_requires=">=3.9",
-    install_requires=["ninja", "torch", "numpy", "urchin", "usd-core"],
-)

### 2025-03-30 dflex/tests/kit_walker.py

@@ -1,365 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-use_omni = False
-if use_omni:
-    import omni.usd
-
-
-class Experiment:
-    name = "kit_walker"
-
-    network_file = None
-    record = True
-
-    render_time = 0.0
-    render_enabled = True
-
-    def __init__(self):
-        pass
-
-    def reset(self, adapter="cuda"):
-        self.episode_duration = 5.0  # seconds
-
-        self.frame_dt = 1.0 / 60.0
-        self.frame_count = int(self.episode_duration / self.frame_dt)
-
-        self.sim_substeps = 64
-        self.sim_dt = self.frame_dt / self.sim_substeps
-        self.sim_time = 0.0
-
-        self.train_max_iters = 10000
-        self.train_iter = 0
-        self.train_rate = 0.025
-        self.train_loss = []
-        self.train_loss_best = math.inf
-
-        self.phase_count = 8
-        self.phase_step = math.pi / self.phase_count * 2.0
-        self.phase_freq = 5.0
-
-        self.render_time = 0.0
-
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.train_loss = []
-        self.optimizer = None
-
-        # mesh = Usd.Stage.Open("assets/prop.usda")
-        if use_omni == False:
-            stage = Usd.Stage.Open("kit_walker.usda")
-        else:
-            stage = omni.usd.get_context().get_stage()
-
-        # ostrich
-        # geom = UsdGeom.Mesh(stage.GetPrimAtPath("/ostrich"))
-        # points = geom.GetPointsAttr().Get()
-
-        # builder.add_soft_mesh(pos=(0.0, 0.0, 0.0),
-        #                       rot=df.quat_identity(),
-        #                       scale=2.0,
-        #                       vel=(0.0, 0.0, 0.0),
[…]

### 2025-03-30 dflex/tests/rl_swing_up.py

@@ -1,397 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 4
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = True
-
-    name = "cartpole"
-
-    regularization = 1.0e-3
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        link_width = 0.5
-
-        max_depth = depth
-
-        # if (True):
-
-        #     # create a branched tree
-        #     builder.add_articulation()
-        #     test_util.build_tree(builder, angle=0.0, width=link_width, max_depth=max_depth)
-
-        #     self.ground = False
-
-        #     # add weight
-        #     if (False):
-
-        #         radius = 0.5
-
-        #         X_pj = df.transform((link_width * 2.0, 0.0, 0.0), df.quat_from_axis_angle( (0.0, 0.0, 1.0), 0.0))
-        #         X_cm = df.transform((radius, 0.0, 0.0), df.quat_identity())
-
-        #         parent = len(builder.body_mass)-1
-        #         link = builder.add_link(parent, X_pj, (0.0, 0.0, 1.0), df.JOINT_REVOLUTE)
[…]

### 2025-03-30 dflex/tests/test_adjoint.py

@@ -1,20 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-import cProfile
-
-import numpy as np
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df

### 2025-03-30 dflex/tests/test_allegro.py

@@ -1,363 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 64
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = False
-
-    name = "allegro"
-
-    regularization = 1.0e-3
-
-    env_count = 1
-    env_dofs = 2
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        # allegro
-        for i in range(self.env_count):
-            test_util.urdf_load(
-                builder,
-                # "assets/franka_description/robots/franka_panda.urdf",
-                "assets/allegro_hand_description/allegro_hand_description_right.urdf",
-                df.transform(
-                    (0.0, 0.0, 0.0),
-                    df.quat_from_axis_angle((0.0, 0.0, 1.0), math.pi * 0.5),
-                ),
-                floating=False,
-                limit_ke=0.0,  # 1.e+3,
-                limit_kd=0.0,
-            )  # 1.e+2)
-
-        # set fingers to mid-range of their limits
-        for i in range(len(builder.joint_q_start)):
-            if builder.joint_type[i] == df.JOINT_REVOLUTE:
-                dof = builder.joint_q_start[i]
-                mid = (
[…]

### 2025-03-30 dflex/tests/test_ant.py

@@ -1,535 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-import xml.etree.ElementTree as ET
-
-
-class Robot:
-    frame_dt = 1.0 / 60.0
-
-    episode_duration = 2.0  # seconds
-    episode_frames = int(episode_duration / frame_dt)
-
-    sim_substeps = 32
-    sim_dt = frame_dt / sim_substeps
-    sim_steps = int(episode_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 1024
-    train_rate = 0.001
-
-    phase_count = 8
-    phase_step = math.pi / phase_count * 2.0
-    phase_freq = 6.0
-
-    ground = True
-
-    name = "humanoid"
-
-    regularization = 1.0e-3
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        self.parse_mjcf(
-            "./assets/" + self.name + ".xml",
-            builder,
-            stiffness=0.0,
-            damping=0.0,
-            contact_ke=1.0e3,
-            contact_kd=1.0e3,
-            contact_kf=1.0e2,
-            contact_mu=0.75,
-            limit_ke=1.0e2,
-            limit_kd=1.0e1,
-        )
-
-        # base transform
-
-        # set joint targets to rest pose in mjcf
-        if self.name == "ant":
-            builder.joint_q[0:3] = [0.0, 0.70, 0.0]
[…]

### 2025-03-30 dflex/tests/test_articulation_fk.py

@@ -1,284 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-import tinyobjloader
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 4
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 64
-    train_rate = 0.05  # 1.0/(sim_dt*sim_dt)
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        x = 0.0
-        w = 0.5
-
-        max_depth = 3
-
-        # create a branched tree
-        builder.add_articulation()
-        test_util.build_tree(builder, angle=0.0, width=w, max_depth=max_depth)
-
-        # add weight
-        if True:
-            radius = 0.1
-
-            X_pj = df.transform(
-                (w * 2.0, 0.0, 0.0), df.quat_from_axis_angle((0.0, 0.0, 1.0), 0.0)
-            )
-            X_cm = df.transform((radius, 0.0, 0.0), df.quat_identity())
-
-            parent = len(builder.body_mass) - 1
-            link = builder.add_link(parent, X_pj, (0.0, 0.0, 1.0), df.JOINT_REVOLUTE)
-            shape = builder.add_shape_sphere(link, pos=(0.0, 0.0, 0.0), radius=radius)
-
-        self.model = builder.finalize(adapter)
-        self.model.contact_ke = 1.0e4
-        self.model.contact_kd = 1000.0
-        self.model.contact_kf = 100.0
-        self.model.contact_mu = 0.75
-        self.model.ground = False
[…]

### 2025-03-30 dflex/tests/test_ball_joint.py

@@ -1,280 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class JointTest:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = False
-
-    name = "ball_joint"
-
-    regularization = 1.0e-3
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        width = 0.1
-        radius = 0.05
-
-        builder.add_articulation()
-
-        body = -1
-        for i in range(8):
-            body = builder.add_link(
-                body,
-                df.transform((2.0 * width, 0.0, 0.0), df.quat_identity()),
-                axis=(0.0, 0.0, 1.0),
-                damping=1.0,
-                stiffness=500.0,
-                type=df.JOINT_BALL,
-            )
-            shape = builder.add_shape_capsule(
-                body, pos=(width, 0.0, 0.0), half_width=width, radius=radius
-            )
-
-        builder.joint_qd[1] = 1.0
-
-        self.pole_angle_penalty = 10.0
-        self.pole_velocity_penalty = 0.5
-
-        self.cart_action_penalty = 1.0e-7
-        self.cart_velocity_penalty = 1.0
-        self.cart_position_penalty = 2.0
-
-        # finalize model
[…]

### 2025-03-30 dflex/tests/test_ballistic.py

@@ -1,132 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import torch
-import time
-import math
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Ballistic:
-    sim_duration = 2.0  # seconds
-    sim_substeps = 10
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 5
-    train_rate = 0.1  # 1.0/(sim_dt*sim_dt)
-
-    def __init__(self, adapter="cpu"):
-        builder = df.sim.ModelBuilder()
-
-        builder.add_particle((0, 1.0, 0.0), (0.1, 0.0, 0.0), 1.0)
-
-        self.model = builder.finalize(adapter)
-
-        self.target = torch.tensor((2.0, 1.0, 0.0), device=adapter)
-
-        # -----------------------
-        # set up Usd renderer
-
-        self.stage = Usd.Stage.CreateNew("outputs/ballistic.usda")
-
-        self.renderer = df.render.UsdRenderer(self.model, self.stage)
-        self.renderer.draw_points = True
-        self.renderer.draw_springs = True
-        self.renderer.draw_shapes = True
-
-        self.renderer.add_sphere(self.target.tolist(), 0.1, "target")
-
-        self.integrator = df.sim.SemiImplicitIntegrator()
-
-    def loss(self):
-        # -----------------------
-        # run simulation
-
-        self.state = self.model.state()
-
-        for i in range(0, self.sim_steps):
-            self.state = self.integrator.forward(self.model, self.state, self.sim_dt)
-
-            if (i % self.sim_substeps) == 0:
-                self.renderer.update(self.state, self.sim_time)
-
-            self.sim_time += self.sim_dt
-
-        loss = torch.norm(self.state.particle_q[0] - self.target)
-        return loss
-
[…]

### 2025-03-30 dflex/tests/test_beam.py

@@ -1,228 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Beam:
-    sim_duration = 3.0  # seconds
-    sim_substeps = 32
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 64
-    train_rate = 1.0
-
-    def __init__(self, device="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-        builder.add_soft_grid(
-            pos=(0.0, 0.0, 0.0),
-            rot=df.quat_identity(),
-            vel=(0.0, 0.0, 0.0),
-            dim_x=20,
-            dim_y=2,
-            dim_z=2,
-            cell_x=0.1,
-            cell_y=0.1,
-            cell_z=0.1,
-            density=10.0,
-            k_mu=1000.0,
-            k_lambda=1000.0,
-            k_damp=5.0,
-            fix_left=True,
-            fix_right=True,
-        )
-
-        self.model = builder.finalize(device)
-
-        # disable triangle dynamics (just used for rendering)
-        self.model.tri_ke = 0.0
-        self.model.tri_ka = 0.0
-        self.model.tri_kd = 0.0
-        self.model.tri_kb = 0.0
-
-        self.model.particle_radius = 0.05
-        self.model.ground = False
-
-        self.target = torch.tensor((-0.5)).to(device)
-        self.material = torch.tensor(
-            (100.0, 50.0, 5.0), requires_grad=True, device=device
-        )
-
-        # -----------------------
-        # set up Usd renderer
-
-        self.stage = Usd.Stage.CreateNew("outputs/beam.usd")
-
-        if self.stage:
-            self.renderer = df.render.UsdRenderer(self.model, self.stage)
-            self.renderer.draw_points = True
[…]

### 2025-03-30 dflex/tests/test_bending.py

@@ -1,249 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-import cProfile
-
-import numpy as np
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Bending:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 32
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 200
-    train_rate = 0.01
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        if True:
-            mesh = Usd.Stage.Open("assets/icosphere_open.usda")
-            geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/Shell/Mesh"))
-
-            # mesh = Usd.Stage.Open("assets/cylinder_long_open.usda")
-            # geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/CylinderLong/CylinderLong"))
-
-            points = geom.GetPointsAttr().Get()
-            indices = geom.GetFaceVertexIndicesAttr().Get()
-            counts = geom.GetFaceVertexCountsAttr().Get()
-
-            linear_vel = np.array((1.0, 0.0, 0.0))
-            angular_vel = np.array((0.0, 0.0, 0.0))
-            center = np.array((0.0, 1.6, 0.0))
-            radius = 0.5
-
-            r = df.quat_multiply(
-                df.quat_from_axis_angle((0.0, 0.0, 1.0), math.pi * 0.0),
-                df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi * 0.5),
-            )
-
-            builder.add_cloth_mesh(
-                pos=center,
-                rot=(0.0, 0.0, 0.0, 1.0),
-                scale=radius,
-                vel=(0.0, 0.0, 0.0),
[…]

### 2025-03-30 dflex/tests/test_cage.py

@@ -1,217 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Cage:
-    sim_duration = 2.0  # seconds
-    sim_substeps = 8
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 20
-    train_rate = 0.1  # 1.0/(sim_dt*sim_dt)
-
-    def __init__(self, mode="quad", adapter="cpu"):
-        builder = df.sim.ModelBuilder()
-
-        if mode == "quad":
-            # anchors
-            builder.add_particle((-1.0, 1.0, 0.0), (0.0, 0.0, 0.0), 0.0)
-            builder.add_particle((1.0, 1.0, 0.0), (0.0, 0.0, 0.0), 0.0)
-            builder.add_particle((1.0, -1.0, 0.0), (0.0, 0.0, 0.0), 0.0)
-            builder.add_particle((-1.0, -1.0, 0.0), (0.0, 0.0, 0.0), 0.0)
-
-            # ball
-            builder.add_particle((0.0, 0.0, 0.0), (0.0, 0.0, 0.0), 1.0)
-
-            ke = 1.0e2
-            kd = 10.0
-
-            # springs
-            builder.add_spring(0, 4, ke, kd, 0)
-            builder.add_spring(1, 4, ke, kd, 0)
-            builder.add_spring(2, 4, ke, kd, 0)
-            builder.add_spring(3, 4, ke, kd, 0)
-
-            self.target_pos = torch.tensor((0.85, 0.5, 0.0), device=adapter)
-            self.target_index = 4
-
-        if mode == "box":
-            # anchors
[…]

### 2025-03-30 dflex/tests/test_cartpole.py

@@ -1,373 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 4
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = True
-
-    name = "cartpole"
-
-    regularization = 1.0e-3
-
-    env_count = 16
-    env_dofs = 2
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        link_width = 0.5
-
-        max_depth = depth
-
-        # cartpole
-        for i in range(self.env_count):
-            test_util.urdf_load(
-                builder,
-                "assets/" + self.name + ".urdf",
-                df.transform(
-                    (0.0, 2.5, -2.0 + i * 2.0),
-                    df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi * 0.5),
-                ),
-                floating=False,
-            )
-
-            builder.joint_q[i * 2 + 0] = 0
-            builder.joint_q[i * 2 + 1] = -math.pi * 0.5  # + i*0.25
-
-        self.pole_angle_penalty = 10.0
-        self.pole_velocity_penalty = 0.5
-
-        self.cart_action_penalty = 1.0e-7
-        self.cart_velocity_penalty = 1.0
-        self.cart_position_penalty = 2.0
-
[…]

### 2025-03-30 dflex/tests/test_chain.py

@@ -1,143 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Chain:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 2
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 20
-    train_rate = 0.1  # 1.0/(sim_dt*sim_dt)
-
-    def __init__(self, adapter="cpu"):
-        builder = df.sim.ModelBuilder()
-
-        # anchor
-        builder.add_particle((0.0, 1.0, 0.0), (0.0, 0.0, 0.0), 0.0)
-
-        for i in range(1, 10):
-            builder.add_particle((i, 1.0, 0.0), (0.0, 0.0, 0.0), 1.0)
-            builder.add_spring(i - 1, i, 1.0e6, 0.0, 0)
-
-        self.model = builder.finalize(adapter)
-        self.model.ground = False
-
-        self.impulse = torch.tensor((0.0, 0.0, 0.0), requires_grad=True, device=adapter)
-
-        # -----------------------
-        # set up Usd renderer
-
-        self.stage = Usd.Stage.CreateNew("outputs/chain.usda")
-
-        self.renderer = df.render.UsdRenderer(self.model, self.stage)
-        self.renderer.draw_points = True
-        self.renderer.draw_springs = True
-        self.renderer.draw_shapes = True
-
-        # self.integrator = df.sim.SemiImplicitIntegrator()
-        self.integrator = df.sim.XPBDIntegrator()
-
-    def loss(self):
-        # -----------------------
-        # run simulation
-
-        self.state = self.model.state()
-        self.state.particle_qd[1] = self.impulse
-
-        for i in range(0, self.sim_steps):
[…]

### 2025-03-30 dflex/tests/test_cloth.py

@@ -1,235 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-# uncomment to output timers
-df.ScopedTimer.enabled = True
-
-
-class Cloth:
-    sim_duration = 5.0  # seconds
-    sim_substeps = 16
-    sim_dt = (
-        1.0 / 60.0
-    ) / sim_substeps  # 60 frames per second, 16 updates between frames,
-    # sim_steps = int(sim_duration/sim_dt)
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    render_time = 0.0
-
-    train_iters = 100
-    train_rate = 0.01
-
-    phase_count = 4
-
-    def __init__(self, dim=20, mode="cloth", adapter="cpu"):
-        torch.manual_seed(42)
-
-        height = 2.5
-
-        builder = df.sim.ModelBuilder()
-        builder.add_cloth_grid(
-            pos=(0.0, height, 0.0),
-            rot=df.quat_from_axis_angle((1.0, 0.0, 0.0), math.pi * 1.04),
-            vel=(0.0, 0.0, 0.0),
-            dim_x=dim,
-            dim_y=dim,
-            cell_x=0.2,
-            cell_y=0.2,
-            mass=400 / (dim**2),
-        )  # , fix_left=True, fix_right=True, fix_top=True, fix_bottom=True)
-
-        attach0 = 0
-        attach1 = 20
-
-        anchor0 = builder.add_particle(
-            pos=builder.particle_q[attach0] - (1.0, 0.0, 0.0),
-            vel=(0.0, 0.0, 0.0),
-            mass=0.0,
-        )
-        anchor1 = builder.add_particle(
-            pos=builder.particle_q[attach1] + (1.0, 0.0, 0.0),
[…]

### 2025-03-30 dflex/tests/test_cloth_collisions.py

@@ -1,271 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-# uncomment to output timers
-df.ScopedTimer.enabled = True
-
-
-class Cloth:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 16
-    sim_dt = (
-        1.0 / 60.0
-    ) / sim_substeps  # 60 frames per second, 16 updates between frames,
-    # sim_steps = int(sim_duration/sim_dt)
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    render_time = 0.0
-
-    train_iters = 100
-    train_rate = 0.01
-
-    phase_count = 4
-
-    def __init__(self, dim=20, mode="cloth", adapter="cpu"):
-        torch.manual_seed(42)
-
-        height = 4.0
-
-        builder = df.sim.ModelBuilder()
-
-        # builder.add_particle(pos=(2.5, 3.0, 2.5), vel=(0.0, 0.0, 0.0), mass=1.0)
-        # builder.add_particle(pos=(2.5, 4.0, 2.5), vel=(0.0, 0.0, 0.0), mass=1.0)
-        # builder.add_particle(pos=(2.5, 5.0, 2.5), vel=(0.0, 0.0, 0.0), mass=1.0)
-
-        builder.add_cloth_grid(
-            pos=(0.0, height, 0.0),
-            rot=df.quat_from_axis_angle((1.0, 0.0, 0.0), math.pi / 2),
-            vel=(0, 5.0, 0.0),
-            dim_x=dim,
-            dim_y=dim,
-            cell_x=0.2,
-            cell_y=0.2,
-            mass=400 / (dim**2),
-        )  # , fix_left=True, fix_right=True, fix_top=True, fix_bottom=True)
-
[…]

### 2025-03-30 dflex/tests/test_cloth_sphere_collisions.py

@@ -1,250 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-# uncomment to output timers
-df.ScopedTimer.enabled = True
-
-
-class Cloth:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 16
-    sim_dt = (
-        1.0 / 60.0
-    ) / sim_substeps  # 60 frames per second, 16 updates between frames,
-    # sim_steps = int(sim_duration/sim_dt)
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    render_time = 0.0
-
-    train_iters = 100
-    train_rate = 0.01
-
-    phase_count = 4
-
-    def __init__(self, dim=20, mode="cloth", adapter="cpu"):
-        torch.manual_seed(42)
-
-        height = 4.0
-
-        builder = df.sim.ModelBuilder()
-
-        builder.add_cloth_grid(
-            pos=(0.0, height, 0.0),
-            rot=df.quat_from_axis_angle((1.0, 0.0, 0.0), math.pi / 2),
-            vel=(0, 5.0, 0.0),
-            dim_x=dim,
-            dim_y=dim,
-            cell_x=0.2,
-            cell_y=0.2,
-            mass=400 / (dim**2),
-        )  # , fix_left=True, fix_right=True, fix_top=True, fix_bottom=True)
-
-        usd = Usd.Stage.Open("assets/sphere.usda")
-        geom = UsdGeom.Mesh(usd.GetPrimAtPath("/Sphere/Sphere"))
-
-        points = geom.GetPointsAttr().Get()
-        indices = geom.GetFaceVertexIndicesAttr().Get()
-        counts = geom.GetFaceVertexCountsAttr().Get()
-
[…]

### 2025-03-30 dflex/tests/test_contact.py

@@ -1,147 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Contact:
-    sim_duration = 2.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 16
-    train_rate = 0.1  # 1.0/(sim_dt*sim_dt)
-
-    def __init__(self, adapter="cpu"):
-        builder = df.sim.ModelBuilder()
-
-        builder.add_particle((0.0, 1.5, 0.0), (0.0, 0.0, 0.0), 0.25)
-
-        self.target_pos = torch.tensor((3.0, 0.0, 0.0), device=adapter)
-        self.target_index = 0
-
-        self.model = builder.finalize(adapter)
-        self.model.contact_ke = 1.0e3
-        self.model.contact_kf = 10.0
-        self.model.contact_kd = 10.0
-        self.model.contact_mu = 0.25
-
-        self.model.particle_qd.requires_grad = True
-
-        # -----------------------
-        # set up Usd renderer
-
-        self.stage = Usd.Stage.CreateNew("outputs/contact.usda")
-
-        self.renderer = df.render.UsdRenderer(self.model, self.stage)
-        self.renderer.draw_points = True
-        self.renderer.draw_springs = True
-        self.renderer.draw_shapes = True
-
-        self.renderer.add_sphere(self.target_pos.tolist(), 0.1, "target")
-
-        self.integrator = df.sim.SemiImplicitIntegrator()
-
-    def loss(self):
-        # -----------------------
-        # run simulation
-
-        self.state = self.model.state()
-
-        for i in range(0, self.sim_steps):
-            self.state = self.integrator.forward(self.model, self.state, self.sim_dt)
-
[…]

### 2025-03-30 dflex/tests/test_fem.py

@@ -1,285 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import cProfile
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class FEM:
-    sim_duration = 5.0  # seconds
-    sim_substeps = 32
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 16
-    train_rate = 0.01  # 1.0/(sim_dt*sim_dt)
-
-    phase_count = 8
-    phase_step = math.pi / phase_count * 2.0
-    phase_freq = 2.5
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        mesh = Usd.Stage.Open("assets/prop.usda")
-        geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/mesh"))
-
-        points = geom.GetPointsAttr().Get()
-
-        tet_indices = geom.GetPrim().GetAttribute("tetraIndices").Get()
-        tri_indices = geom.GetFaceVertexIndicesAttr().Get()
-        tri_counts = geom.GetFaceVertexCountsAttr().Get()
-
-        r = df.quat_multiply(
-            df.quat_from_axis_angle((0.0, 0.0, 1.0), math.pi * 0.0),
-            df.quat_from_axis_angle((1.0, 0.0, 0.0), math.pi * 0.0),
-        )
-
-        builder.add_soft_mesh(
-            pos=(0.0, 2.0, 0.0),
-            rot=r,
-            scale=1.0,
-            vel=(1.5, 0.0, 0.0),
-            vertices=points,
-            indices=tet_indices,
[…]

### 2025-03-30 dflex/tests/test_fem_contact.py

@@ -1,330 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import cProfile
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-import test_util
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class FEMContact:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 64
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 16
-    train_rate = 0.01  # 1.0/(sim_dt*sim_dt)
-
-    phase_count = 8
-    phase_step = math.pi / phase_count * 2.0
-    phase_freq = 2.5
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        solid = True
-
-        if solid:
-            builder.add_soft_grid(
-                pos=(0.5, 1.0, 0.0),
-                rot=(0.0, 0.0, 0.0, 1.0),
-                vel=(0.0, 0.0, 0.0),
-                dim_x=3,
-                dim_y=10,
-                dim_z=3,
-                cell_x=0.1,
-                cell_y=0.1,
-                cell_z=0.1,
-                density=1000.0,
-                k_mu=10000.0,
-                k_lambda=10000.0,
-                k_damp=1.0,
-            )
-
-        else:
-            builder.add_cloth_grid(
-                pos=(-0.7, 1.0, -0.7),
-                rot=df.quat_from_axis_angle((1.0, 0.0, 0.0), math.pi * 0.5),
-                vel=(0.0, 0.0, 0.0),
-                dim_x=20,
-                dim_y=20,
-                cell_x=0.075,
-                cell_y=0.075,
-                mass=0.2,
-            )
-
-        # art = builder.add_articulation()
-
-        # link = builder.add_link(
[…]

### 2025-03-30 dflex/tests/test_franka.py

@@ -1,299 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 4
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = False
-
-    name = "franka"
-
-    regularization = 1.0e-3
-
-    env_count = 1
-    env_dofs = 2
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        # cartpole
-        for i in range(self.env_count):
-            test_util.urdf_load(
-                builder,
-                "assets/franka_description/robots/franka_panda.urdf",
-                df.transform(
-                    (0.0, 0.0, 0.0),
-                    df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi * 0.5),
-                ),
-                floating=False,
-                limit_ke=1.0e3,
-                limit_kd=1.0e2,
-            )
-
-        for i in range(len(builder.joint_target_kd)):
-            builder.joint_target_kd[i] = 1.0
-
-        # finalize model
-        self.model = builder.finalize(adapter)
-        self.model.ground = self.ground
-        self.model.gravity = torch.tensor((0.0, -9.81, 0.0), device=adapter)
-        # self.model.gravity = torch.tensor((0.0, 0.0, 0.0), device=adapter)
-
[…]

### 2025-03-30 dflex/tests/test_hopper.py

@@ -1,316 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    frame_dt = 1.0 / 60.0
-
-    episode_duration = 2.0  # seconds
-    episode_frames = int(episode_duration / frame_dt)
-
-    sim_substeps = 16
-    sim_dt = frame_dt / sim_substeps
-    sim_steps = int(episode_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 1024
-    train_rate = 0.001
-
-    ground = True
-
-    name = "hopper"
-
-    regularization = 1.0e-3
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        # humanoid
-        test_util.urdf_load(
-            builder,
-            "assets/hopper.urdf",
-            # df.transform((0.0, 1.35, 0.0), df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi*0.5)),
-            df.transform((0.0, 0.65, 0.0), df.quat_identity()),
-            floating=True,
-            shape_ke=1.0e3 * 2.0,
-            shape_kd=1.0e2,
-            shape_kf=1.0e2,
-            shape_mu=0.5,
-        )
-
-        # set pd-stiffness
-        for i in range(len(builder.joint_target_ke)):
-            builder.joint_target_ke[i] = 10.0
-            builder.joint_target_kd[i] = 1.0
-
-        # finalize model
-        self.model = builder.finalize(adapter)
-        self.model.ground = self.ground
[…]

### 2025-03-30 dflex/tests/test_humanoid.py

@@ -1,328 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class Robot:
-    frame_dt = 1.0 / 60.0
-
-    episode_duration = 2.0  # seconds
-    episode_frames = int(episode_duration / frame_dt)
-
-    sim_substeps = 16
-    sim_dt = frame_dt / sim_substeps
-    sim_steps = int(episode_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 1024
-    train_rate = 0.05
-
-    ground = True
-
-    name = "humanoid"
-
-    regularization = 1.0e-3
-
-    phase_count = 8
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        # humanoid
-        test_util.urdf_load(
-            builder,
-            "assets/humanoid.urdf",
-            df.transform(
-                (0.0, 1.35, 0.0),
-                df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi * 0.5),
-            ),
-            # df.transform((0.0, 0.65, 0.0), df.quat_identity()),
-            floating=True,
-            shape_ke=1.0e3 * 5.0,
-            shape_kd=1.0e2 * 2.0,
-            shape_kf=1.0e2,
-            shape_mu=0.5,
-        )
-
-        # set pd-stiffness
-        for i in range(len(builder.joint_target_ke)):
-            builder.joint_target_ke[i] = 10.0
-            builder.joint_target_kd[i] = 1.0
-
-        # finalize model
[…]

### 2025-03-30 dflex/tests/test_jelly.py

@@ -1,282 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-import timeit
-import cProfile
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Bending:
-    sim_duration = 5.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 200
-    train_rate = 0.01
-
-    phase_count = 8
-    phase_step = math.pi / phase_count * 2.0
-    phase_freq = 2.5
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        r = df.quat_multiply(
-            df.quat_from_axis_angle((0.0, 0.0, 1.0), math.pi * 0.0),
-            df.quat_from_axis_angle((1.0, 0.0, 0.0), -math.pi * 0.5),
-        )
-
-        builder = df.sim.ModelBuilder()
-
-        mesh = Usd.Stage.Open("assets/jellyfish.usda")
-        geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/Icosphere/Icosphere"))
-
-        points = geom.GetPointsAttr().Get()
-        indices = geom.GetFaceVertexIndicesAttr().Get()
-        counts = geom.GetFaceVertexCountsAttr().Get()
-
-        face_materials = [-1] * len(counts)
-        face_subsets = UsdGeom.Subset.GetAllGeomSubsets(geom)
-
-        for i, s in enumerate(face_subsets):
-            face_subset_indices = s.GetIndicesAttr().Get()
-
-            for f in face_subset_indices:
-                face_materials[f] = i
-
-        active_material = 0
-        active_scale = []
-
-        def add_edge(f0, f1):
-            if (
-                face_materials[f0] == active_material
-                and face_materials[f1] == active_material
-            ):
[…]

### 2025-03-30 dflex/tests/test_lift_drag.py

@@ -1,214 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-from torch.utils.tensorboard import SummaryWriter
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-# uncomment to output timers
-df.ScopedTimer.enabled = False
-
-
-class Cloth:
-    sim_duration = 2.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    render_time = 0.0
-
-    train_iters = 4
-    train_rate = 0.01 / sim_substeps
-
-    phase_count = 4
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        height = 2.5
-
-        builder = df.sim.ModelBuilder()
-        builder.add_cloth_grid(
-            pos=(0.0, height, 0.0),
-            rot=df.quat_from_axis_angle((1.0, 0.5, 0.0), math.pi * 0.5),
-            vel=(0.0, 0.0, 0.0),
-            dim_x=16,
-            dim_y=16,
-            cell_x=0.125,
-            cell_y=0.125,
-            mass=1.0,
-        )  # , fix_left=True, fix_right=True, fix_top=True, fix_bottom=True)
-
-        self.model = builder.finalize(adapter)
-        self.model.tri_ke = 10000.0
-        self.model.tri_ka = 10000.0
-        self.model.tri_kd = 100.0
-        self.model.tri_lift = 10.0
-        self.model.tri_drag = 5.0
-
-        self.model.contact_ke = 1.0e4
-        self.model.contact_kd = 1000.0
-        self.model.contact_kf = 1000.0
-        self.model.contact_mu = 0.5
-
-        self.model.particle_radius = 0.01
-        self.model.ground = False
-
[…]

### 2025-03-30 dflex/tests/test_muscle.py

@@ -1,355 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class MuscleTest:
-    sim_duration = 4.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 10.0
-
-    ground = False
-
-    name = "muscle"
-
-    regularization = 1.0e-3
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        length = 0.5
-        width = 0.1
-        radius = 0.05
-
-        builder.add_articulation()
-
-        body = -1
-        for i in range(2):
-            if i == 0:
-                body = builder.add_link(
-                    body,
-                    df.transform((2.0 * length, 0.0, 0.0), df.quat_identity()),
-                    axis=(0.0, 0.0, 1.0),
-                    damping=1.0,
-                    stiffness=0.0,
-                    type=df.JOINT_FIXED,
-                )
-            else:
-                body = builder.add_link(
-                    body,
-                    df.transform((2.0 * length, 0.0, 0.0), df.quat_identity()),
-                    axis=(0.0, 0.0, 1.0),
-                    damping=1.0,
-                    stiffness=0.0,
-                    type=df.JOINT_BALL,
-                )
-
-            shape = builder.add_shape_box(
[…]

### 2025-03-30 dflex/tests/test_paper.py

@@ -1,214 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-import cProfile
-
-import numpy as np
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Paper:
-    sim_duration = 10.0  # seconds
-    sim_substeps = 64
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 200
-    train_rate = 0.01
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-        np.random.seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        mesh = Usd.Stage.Open("assets/paper.usda")
-        geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/Grid/Grid"))
-
-        # mesh = Usd.Stage.Open("assets/dart.usda")
-        # geom = UsdGeom.Mesh(mesh.GetPrimAtPath("/planes_001/planes_001"))
-
-        points = geom.GetPointsAttr().Get()
-        indices = geom.GetFaceVertexIndicesAttr().Get()
-        counts = geom.GetFaceVertexCountsAttr().Get()
-
-        center = np.array((0.0, 10.0, 0.0))
-        radius = 5.0
-
-        for i in range(1):
-            center = np.array([0.0, 5.0, 0.0]) + np.random.ranf((3,)) * 10.0
-            axis = df.normalize(np.random.ranf((3,)))
-            angle = (
-                np.random.ranf(
-                    1,
-                )
-                * math.pi
-            )
-
-            builder.add_cloth_mesh(
-                pos=center,
-                rot=df.quat_from_axis_angle(axis, angle),
-                scale=radius,
-                vel=(0.0, 0.0, 0.0),
-                vertices=points,
-                indices=indices,
[…]

### 2025-03-30 dflex/tests/test_rigid_bounce.py

@@ -1,318 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class RigidBounce:
-    frame_dt = 1.0 / 60.0
-
-    episode_duration = 2.0  # seconds
-    episode_frames = int(episode_duration / frame_dt)
-
-    sim_substeps = 16
-    sim_dt = frame_dt / sim_substeps
-    sim_steps = int(episode_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 0.01
-
-    ground = True
-
-    name = "rigid_bounce"
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        builder.add_articulation()
-
-        # add sphere
-        link = builder.add_link(
-            -1,
-            df.transform((0.0, 0.0, 0.0), df.quat_identity()),
-            (0, 0, 0),
-            df.JOINT_FREE,
-        )
-        shape = builder.add_shape_sphere(
-            link,
-            (0.0, 0.0, 0.0),
-            df.quat_identity(),
-            radius=0.1,
-            ke=1.0e4,
-            kd=10.0,
-            kf=1.0e2,
-            mu=0.25,
-        )
-
-        builder.joint_q[1] = 1.0
-
-        # v_s = df.get_body_twist((0.0, 0.0, 0.0), (1.0, -1.0, 0.0), builder.joint_q[0:3])
-
-        w_m = (0.0, 0.0, 3.0)  # angular velocity (expressed in world space)
-        v_m = (
-            0.0,
-            0.0,
-            0.0,
[…]

### 2025-03-30 dflex/tests/test_rigid_slide.py

@@ -1,260 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-import tinyobjloader
-
-import numpy as np
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class RigidSlide:
-    sim_duration = 3.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    train_iters = 64
-    train_rate = 0.1
-
-    discount_scale = 1.0
-    discount_factor = 0.5
-
-    def __init__(self, adapter="cpu"):
-        torch.manual_seed(42)
-
-        # load mesh
-        usd = Usd.Stage.Open("assets/suzanne.usda")
-        geom = UsdGeom.Mesh(usd.GetPrimAtPath("/Suzanne/Suzanne"))
-
-        points = geom.GetPointsAttr().Get()
-        indices = geom.GetFaceVertexIndicesAttr().Get()
-        counts = geom.GetFaceVertexCountsAttr().Get()
-
-        builder = df.sim.ModelBuilder()
-
-        mesh = df.sim.Mesh(points, indices)
-
-        articulation = builder.add_articulation()
-
-        rigid = builder.add_link(
-            parent=-1,
-            X_pj=df.transform((0.0, 0.0, 0.0), df.quat_identity()),
-            axis=(0.0, 0.0, 0.0),
-            type=df.JOINT_FREE,
-        )
-
-        ke = 1.0e4
-        kd = 1.0e3
-        kf = 1.0e3
-        mu = 0.5
-
-        # shape = builder.add_shape_mesh(
-        #     rigid,
-        #     mesh=mesh,
-        #     scale=(0.2, 0.2, 0.2),
-        #     density=1000.0,
-        #     ke=1.e+4,
[…]

### 2025-03-30 dflex/tests/test_snu.py

@@ -1,429 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class HumanoidSNU:
-    sim_duration = 1.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-
-    sim_time = 0.0
-
-    train_iters = 128
-    train_rate = 0.01
-
-    ground = False
-
-    name = "humanoid_snu_neck"
-
-    regularization = 1.0e-3
-
-    env_count = 16
-    env_dofs = 2
-
-    def __init__(self, depth=1, mode="numpy", render=True, adapter="cpu"):
-        torch.manual_seed(41)
-
-        builder = df.sim.ModelBuilder()
-
-        self.adapter = adapter
-        self.mode = mode
-        self.render = render
-
-        self.filter = {}
-
-        if self.name == "humanoid_snu_arm":
-            self.filter = {"ShoulderR", "ArmR", "ForeArmR", "HandR", "Torso", "Neck"}
-            self.ground = False
-
-        if self.name == "humanoid_snu_neck":
-            self.filter = {"Torso", "Neck", "Head", "ShoulderR", "ShoulderL"}
-            self.ground = False
-
-        self.node_map, self.xform_map, self.mesh_map = test_util.parse_skeleton(
-            "assets/snu/arm.xml", builder, self.filter
-        )
-        self.muscles = test_util.parse_muscles(
-            "assets/snu/muscle284.xml", builder, self.node_map, self.xform_map
-        )
-
-        # set initial position 1m off the ground
-        if self.name == "humanoid_snu":
-            builder.joint_q[1] = 1.0
-
[…]

### 2025-03-30 dflex/tests/test_snu_mlp.py

@@ -1,570 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import os
-import sys
-
-import torch.nn as nn
-import torch.nn.functional as F
-from torch.utils.tensorboard import SummaryWriter
-
-# to allow tests to import the module they belong to
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-import numpy as np
-
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-from pxr import Usd, UsdGeom, Gf
-
-import test_util
-
-
-class MultiLayerPerceptron(nn.Module):
-    def __init__(self, n_in, n_out, n_hd, adapter, inference=False):
-        super(MultiLayerPerceptron, self).__init__()
-
-        self.n_in = n_in
-        self.n_out = n_out
-        self.n_hd = n_hd
-
-        # self.ll = nn.Linear(n_in, n_out)
-        self.fc1 = nn.Linear(n_in, n_hd).to(adapter)
-        self.fc2 = nn.Linear(n_hd, n_hd).to(adapter)
-        self.fc3 = nn.Linear(n_hd, n_out).to(adapter)
-        self.bn1 = nn.LayerNorm(n_in, elementwise_affine=False).to(adapter)
-        self.bn2 = nn.LayerNorm(n_hd, elementwise_affine=False).to(adapter)
-        self.bn3 = nn.LayerNorm(n_out, elementwise_affine=False).to(adapter)
-
-    def forward(self, x: torch.Tensor):
-        x = F.leaky_relu(self.bn2(self.fc1(x)))
-        x = F.leaky_relu(self.bn2(self.fc2(x)))
-
-        x = torch.tanh(self.bn3(self.fc3(x)) - 2.0)
-
-        return x
-
-
-class HumanoidSNU:
-    train_iters = 100000000
-    train_rate = 0.001
-    train_size = 128
-    train_batch_size = 4
-    train_batch_iters = 128
-    train_batch_count = int(train_size / train_batch_size)
-    train_data = None
-
[…]

### 2025-03-30 dflex/tests/test_urdf.py

@@ -1,35 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import torch
-
-t = torch.tensor((1.0), requires_grad=True)
-
-i = t.item()
-
-print(i)
-
-from urchin import URDF
-
-
-# robot = URDF.load("assets/trifinger/urdf/trifinger_with_stage.urdf")
-# robot = URDF.load("assets/franka_description/robots/franka_panda.urdf")
-# robot = URDF.load("assets/anymal_b_simple_description/urdf/anymal.urdf")
-# robot = URDF.load("assets/kinova_description/urdf/kinova.urdf")
-# robot = URDF.load("assets/ur5/urdf/ur5_robot.urdf")
-# robot = URDF.load("assets/kuka_allegro_description/allegro.urdf")
-robot = URDF.load("assets/allegro_hand_description/allegro_hand_description_left.urdf")
-
-
-for link in robot.links:
-    dir(link)
-    print(link)
-
-for joint in robot.joints:
-    print(joint)
-
-robot.show()

### 2025-03-30 dflex/tests/test_util.py

@@ -1,493 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import urchin
-import math
-import numpy as np
-import os
-
-import xml.etree.ElementTree as ET
-
-import dflex as df
-
-
-def urdf_add_collision(
-    builder, link, collisions, shape_ke, shape_kd, shape_kf, shape_mu
-):
-    # add geometry
-    for collision in collisions:
-        origin = urchin.matrix_to_xyz_rpy(collision.origin)
-
-        pos = origin[0:3]
-        rot = df.rpy2quat(*origin[3:6])
-
-        geo = collision.geometry
-
-        if geo.box:
-            builder.add_shape_box(
-                link,
-                pos,
-                rot,
-                geo.box.size[0] * 0.5,
-                geo.box.size[1] * 0.5,
-                geo.box.size[2] * 0.5,
-                ke=shape_ke,
-                kd=shape_kd,
-                kf=shape_kf,
-                mu=shape_mu,
-            )
-
-        if geo.sphere:
-            builder.add_shape_sphere(
-                link,
-                pos,
-                rot,
-                geo.sphere.radius,
-                ke=shape_ke,
-                kd=shape_kd,
-                kf=shape_kf,
-                mu=shape_mu,
-            )
-
-        if geo.cylinder:
-            # cylinders in URDF are aligned with z-axis, while dFlex uses x-axis
-            r = df.quat_from_axis_angle((0.0, 1.0, 0.0), math.pi * 0.5)
-
-            builder.add_shape_capsule(
-                link,
-                pos,
-                df.quat_multiply(rot, r),
-                geo.cylinder.radius,
-                geo.cylinder.length * 0.5,
-                ke=shape_ke,
-                kd=shape_kd,
-                kf=shape_kf,
-                mu=shape_mu,
-            )
-
-        if geo.mesh:
-            for m in geo.mesh.meshes:
-                faces = []
-                vertices = []
-
-                for v in m.vertices:
-                    vertices.append(np.array(v))
-
[…]

### 2025-03-30 dflex/tests/test_walker.py

@@ -1,227 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import math
-import torch
-import time
-
-# include parent path
-import os
-import sys
-
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), "..")))
-
-import dflex as df
-
-from pxr import Usd, UsdGeom, Gf
-
-
-class Walker:
-    sim_duration = 5.0  # seconds
-    sim_substeps = 16
-    sim_dt = (1.0 / 60.0) / sim_substeps
-    sim_steps = int(sim_duration / sim_dt)
-    sim_time = 0.0
-
-    render_time = 0.0
-
-    train_iters = 50
-    train_rate = 0.0001
-
-    def __init__(self, mode="walker", adapter="cpu"):
-        self.phase_count = 8
-        self.phase_step = math.pi / self.phase_count * 2.0
-        self.phase_freq = 20.0
-
-        torch.manual_seed(42)
-
-        builder = df.sim.ModelBuilder()
-
-        walker = Usd.Stage.Open("assets/walker.usda")
-        mesh = UsdGeom.Mesh(walker.GetPrimAtPath("/Grid/Grid"))
-
-        points = mesh.GetPointsAttr().Get()
-        indices = mesh.GetFaceVertexIndicesAttr().Get()
-
-        for p in points:
-            builder.add_particle(tuple(p), (0.0, 0.0, 0.0), 1.0)
-
-        for t in range(0, len(indices), 3):
-            i = indices[t + 0]
-            j = indices[t + 1]
-            k = indices[t + 2]
-
-            builder.add_triangle(i, j, k)
-
-        self.model = builder.finalize(adapter)
-        self.model.tri_ke = 10000.0
-        self.model.tri_ka = 10000.0
-        self.model.tri_kd = 100.0
-        self.model.tri_lift = 0.0
-        self.model.tri_drag = 0.0
-
-        self.edge_ke = 0.0
-        self.edge_kd = 0.0
-
-        self.model.contact_ke = 1.0e4
-        self.model.contact_kd = 1000.0
[…]

### 2025-03-30 src/environments/scenarios/Ant.py

@@ -23,19 +23,7 @@ except ModuleNotFoundError:
 
 class AntEnv(DFlexEnv):
 
-    def __init__(
-            self, 
-            render               = False    ,
-            device               = 'cuda:0' ,
-            num_envs             = 4096     ,
-            seed                 = 0        ,
-            episode_length       = 1000     ,
-            no_grad              = True     ,
-            stochastic_init      = False    ,
-            MM_caching_frequency = 1        ,
-            early_termination    = True
-        ):
-
+    def __init__(self, render=False, device='cuda:0', num_envs=4096, seed=0, episode_length=1000, no_grad=True, stochastic_init=False, MM_caching_frequency = 16, early_termination = True):
         num_obs = 37
         num_act = 8
     
@@ -233,7 +221,7 @@ class AntEnv(DFlexEnv):
     '''
     cut off the gradient from the current state to previous states
     '''
-    def zero_grad(self, checkpoint = None):
+    def clear_grad(self, checkpoint = None):
         with torch.no_grad():
             if checkpoint is None:
                 checkpoint = {}
@@ -250,12 +238,15 @@ class AntEnv(DFlexEnv):
             self.actions = checkpoint['actions'].clone()
             self.progress_buf = checkpoint['progress_buf'].clone()
 
+    def zero_grad(self):
+        self.clear_grad()
+
     '''
     This function starts collecting a new trajectory from the current states but cuts off the computation graph to the previous states.
     It has to be called every time the algorithm starts an episode and it returns the observation vectors
     '''
     def initialize_trajectory(self):
-        self.zero_grad()
+        self.clear_grad()
         self.calculateObservations()
 
         return self.obs_buf
@@ -349,4 +340,4 @@ class Ant:
 
     def step(self, actions):
         obs, rew, done, _ = self.world.step(actions[0])
-        return [obs], [rew], done, {}
+        return [obs.clone()], [rew.clone()], done.clone(), {}

### 2025-03-30 src/gengif.py

@@ -79,6 +79,7 @@ def run(
             print(world.scenario.max_rewards())
             observation = torch.stack(observation).transpose(0,1)
             total_reward += torch.stack(reward).sum().item()
+            
         
         print("total reward", total_reward, world.scenario.max_rewards().sum().item())
     

### 2025-03-30 src/trainers/routines/policy.py

@@ -24,12 +24,11 @@ def train_policy(
     live_runs = dead_runs.logical_not()
     live_steps = dones[:,live_runs,0].logical_not().sum(0) - 1
 
-
     # compute loss
     optimizer.zero_grad()
     loss = -(
-        (episode_data["proxy_rewards"] * gammas * dones.logical_not()).sum() + 
-        ((gammas * episode_data["values"])[live_steps,live_runs]).sum()
+        (episode_data["proxy_rewards"] * gammas * dones.logical_not()).sum() 
+        # + ((gammas * episode_data["values"])[live_steps,live_runs]).sum()
     ) / (steps * envs)
 
     # add action space loss

### 2025-03-30 src/trainers/shac.py

@@ -76,7 +76,7 @@ def shac(
             world         = train_world,
             unroll_steps  = train_steps,
             policy_model  = policy_model.sample,
-            use_diffreward= use_diffreward,
+            use_diffreward= False,
         )
 
         episode_data["proxy_rewards"] = episode_data["rewards"]

### 2025-03-30 src/trainers/shacwm.py

@@ -178,25 +178,25 @@ def shacwm(
             clip_coefficient = reward_clip_coefficient ,
         )
 
-        # train value model ###########################################
-        trainers.routines.train_value(
-            episode          = episode               ,
-            model            = value_model           ,
-            optimizer        = value_model_optimizer ,
-            episode_data     = episode_data          ,
-            cached_data      = value_cache           ,
-            training_epochs  = value_epochs          ,
-            batch_size       = value_batch_size      ,
-            cache_size       = value_cache_size      ,
-            slam             = lambda_factor         ,
-            gamma            = gamma_factor          ,
-            bins             = value_bins            ,
-            logger           = value_logger          ,
-            ett              = value_ett             ,
-            stop_threshold   = value_stop_threshold  ,
-            tolerance        = value_tolerance       ,
-            clip_coefficient = value_clip_coefficient
-        )
+        # # train value model ###########################################
+        # trainers.routines.train_value(
+        #     episode          = episode               ,
+        #     model            = value_model           ,
+        #     optimizer        = value_model_optimizer ,
+        #     episode_data     = episode_data          ,
+        #     cached_data      = value_cache           ,
+        #     training_epochs  = value_epochs          ,
+        #     batch_size       = value_batch_size      ,
+        #     cache_size       = value_cache_size      ,
+        #     slam             = lambda_factor         ,
+        #     gamma            = gamma_factor          ,
+        #     bins             = value_bins            ,
+        #     logger           = value_logger          ,
+        #     ett              = value_ett             ,
+        #     stop_threshold   = value_stop_threshold  ,
+        #     tolerance        = value_tolerance       ,
+        #     clip_coefficient = value_clip_coefficient
+        # )
 
         # checkpoint ##################################################
         if episode % etv == 0:

### 2025-03-30 src/unroll.py

@@ -17,13 +17,16 @@ def unroll(
         for i in dones[:,0].nonzero():
             observations[i] = torch.stack(world.reset_at(i)).transpose(0,1)[i]
 
+    if observations is not None and dones is not None:
+        observations = world.world.reset(env_ids=torch.arange(world.batch_dim, device=world.device)[dones[:,0]]).unsqueeze(1)
+
     if observations is None and dones is None: 
         observations = torch.stack(world.reset()).transpose(0,1)
         dones        = torch.zeros(observations.size(0), observations.size(1), device=observations.device).bool()
 
     max_reward   = world.scenario.max_rewards() if max_reward else torch.zeros(world.batch_dim, device=world.device)
 
-    observations = observations.detach()
+    observations = world.world.initialize_trajectory().unsqueeze(1)
 
     observation_cache = []
     action_cache      = []
@@ -37,7 +40,7 @@ def unroll(
         observation_cache.append(observations)
         done_cache       .append(dones)
         policy_result = policy_model(observations)
-        actions  = policy_result["actions"]
+        actions  = torch.nn.functional.tanh( policy_result["actions"] )
         logprobs = policy_result.get("logprobs",torch.empty(0))
         entropy  = policy_result.get("entropy" ,torch.empty(0))
         logits   = policy_result.get("logits"  ,torch.empty(0))
@@ -61,6 +64,7 @@ def unroll(
     entropy_cache     = torch.stack(entropy_cache)
     logits_cache      = torch.stack(logits_cache)
 
+    print(world.world.state.joint_q.view(world.batch_dim, -1)[0].mean())
 
     return { 
             "logprobs"      : logprobs_cache    ,

### 2025-03-30 rebuttal.md
shrink it down to 5k

@@
 
-We are sorry for the confusion. 
-We will revise the introduction to better reflect the goal of the paper. 
-Briefly, we believe that SHAC demonstrated excellent performance in single-agent and differentiable scenarios. 
-In this work, we aim to lift these requirements (single-agent and differentiability) to broaden the applicability of SHAC. 
-This led us to develop SHAC++ that can handle multi-agent and non-differentiable environments while maintaining performance comparable to SHAC.
+We apologize for the confusion.
+We'll revise the introduction to clarify our goals.
+SHAC demonstrated excellent performance in single-agent, differentiable settings.
+Our work aims to lift these constraints by developing SHAC++, which handles multi-agent and non-differentiable environments while maintaining comparable performance to SHAC.
 
@@
 
-**Goal**: We want to stress the fact that the goal is not to outperform SHAC, but to lift the requirements of the original framework (single-agent and differentiability) while maintaining comparable performance.
+**Goal**: Lift SHAC's requirements to broaden its applicability---see previous point.
 
-**Skewed results**: In VMAS as in many other environments, 
-the agents are able to experience only local observations and not the whole world state. While SHAC++ remains faithful to accessing only this partial information, 
-SHAC access the whole global information through its gradients. 
-This puts SHAC in a more favorable position. 
-Despite this, SHAC++ and SHAC perform similarly.
+**Skewed results**: In VMAS, agents only access local observations.
+SHAC++ respects this limitation while SHAC accesses global state through its gradients, giving SHAC an advantage.
+Despite this, both algorithms perform similarly.
 
@@
 
-> The proposed method is not evaluated in such complex environments; the test environments are very simple.
+> The proposed method is not evaluated in such complex environments; the test environments are very simple
 
-We believe that the chose environments offer a good degree of complexity especially when the number of agents is increased. 
-This is also reflected by the poor performance achieved by PPO/MAPPO. 
-Despite this, we recognize that the statement in the abstract may be misleading. 
-We will revise the abstract and introduction to better reflect the complexity of the environments, and we will add more content to the discussion to better explain the choice of the environments.
+While these environments present significant challenges for PPO/MAPPO (as shown by their poor performance), we acknowledge they could be more complex.
+We'll revise our paper to better reflect this and explain our environmental choices in the discussion section.
 
-> The performance of SHAC++ is questionable in more complex MARL.
+> The performance of SHAC++ is questionable in more complex MARL
 
-We understand that more complex environments with more agents may offer more challenging scenarios and more compelling results.
-However, we want to stress that these environments are already incapacitating for PPO/MAPPO.
-Further, we will consider adding more complex environments for future revision (with more agents) to better reflect the performance of SHAC++.
-Finally, we kindly ask the reviewer to suggest a differentiable environment that can scale in the number of agents which he would believe to be adequate for the evaluation of SHAC++.
+We understand that more complex environments may offer more challenging scenarios. However, we want to stress that these environments are already incapacitating for PPO/MAPPO. 
+We plan to evaluate SHAC++ in more complex environments with larger agent populations in the next revision.
+We would appreciate the reviewer's suggestions for suitable differentiable, scalable multi-agent environments that could provide more comprehensive benchmarks.
 
-> Numerous practical algorithms already exist for such environments, for example MAPPO, HAPPO, and QMIX in multi-agent scenarios.
+> Numerous practical algorithms already exist for such environments, for example MAPPO, HAPPO, and QMIX in multi-agent scenarios
 
-**PPO & MAPPO confusion**: While we do not evaluate against also HAPPO and QMIX, we always use MAPPO where PPO is evaluated for multi-agent setting.
-Since the MAPPO formulation is fairly natural extension to PPO, we always referred to PPO in the figures and tables, and only briefly mention that MAPPO is used in the case of multi-agent setting.
-We will revise the paper to better reflect this.
+**PPO & MAPPO confusion**: We always use MAPPO where PPO is evaluated for multi-agent setting (we just briefly mention in the paper). We will fix this in the paper.
 
-**Why MAPPO and not the others**: While HAPPO and QMIX are both valid candidate for baseline.
-In order to offer an ablation study and a degree of statistical significance we had to limit the number of baselines.
-We also note that HAPPO authors designed their algorithm to be non-homogeneous (parameters between agents are not shared).
-This choice renders its application problematic as instantiating and training several NNs at the same time quickly becomes computationally infeasible.
-Meanwhile, MAPPO can be applied with both shared and non-shared parameters.
-In our experiments, in order to be consisted while scaling the number of agents, we always apply parameter sharing.
-We will include a more detailed explanation of why we chose MAPPO as a baseline in the paper.
+**Why MAPPO and not others**: While HAPPO and QMIX are valid candidates, we limited baselines to maintain statistical significance.
+HAPPO's non-homogeneous design (non-shared parameters between agents) makes it computationally expensive and creates an unfair comparison.
+In our experiments, we consistently use parameter sharing for scalability.
+MAPPO supports both shared and non-shared parameters, making it appropriate for our evaluation.
 
@@
 
-**Why not the algorithm in related work section**: We did not include PODS, CE-APG, and BPTT as baselines as we see these a precursor of SHAC.
-The authors of SAM-RL introduce a differentiable renderer on top of the differentiable environment which is even more difficult to obtain in practice (Although also this component could be replaced with a learned model in our framework).
-Instead, DiffTORI applies a test-time optimization on the agent trajectory.
-This choice render the extension of DiffTORI to the multi-agent setting difficult.
-The authors of AHAC address the issue of stiff dynamic by simply truncating trajectories, while this choice can boost performance of SHAC, it maintains the fundamental limitation of the original framework.
+**Why not other algorithms**: We excluded PODS, CE-APG, and BPTT as they are precursors to SHAC.
+SAM-RL requires a differentiable renderer beyond the differentiable environment, making it even less practical (though our framework could learn this component too).
+DiffTORI's test-time optimization approach is difficult to extend to multi-agent settings.
+AHAC addresses stiff dynamics by truncating trajectories, which could improve SHAC's performance but doesn't address its fundamental limitations.
 
-We will provide a more detailed explanation in the paper.
-
-> The authors estimate the reward function and state transition using only local agent states and actions as input. This approach may limit the method's generality and correctness.
+> The authors estimate the reward function and state transition using only local agent states and actions as input. This approach may limit the method's generality and correctness
 
@@
 
-> The parameters for the policy, value, reward, and state transition networks are all denoted by $\theta$.
+> The parameters for the policy, value, reward, and state transition networks are all denoted by $\theta$
 
-This is a deliberate choice that allows us to denote a general framework where the parameters can be arbitrary shared.
-For example, all parameter may be separate such as in $\theta=[\theta_{\text{policy}}, \theta_{\text{reward}}, \theta_{\text{value}}, \theta_{\text{world}}]$.
-Or, $\theta$ could be designed with some degree of sharing.
+This is intentional, providing flexibility.
+Parameters can be fully separate ($\theta=[\theta_{\text{policy}}, \theta_{\text{reward}}, \theta_{\text{value}}, \theta_{\text{world}}]$) or have varying degrees of sharing.
 
-> modeling the environment instead of directly utilizing its information might introduce biases and lead to performance degradation.
+> modeling the environment instead of directly utilizing its information might introduce biases and lead to performance degradation
 
-Modeling the environment means that the policy will be only as good as the environment model is.
-If the world model is very accurate, the gradient will be informative.
-If the world model is inaccurate or only partially accurate, the gradient will be noisy leading to degradation.
-However, while we train the world model in tandem with the policy, it is possible to bootstrap the world model with already available traces.
-This can lead to a more stable training of the policy.
-We will include this discussion in the paper.
+We agree that learned models introduce potential for inaccuracy.
+Policy performance depends on model quality - accurate models provide informative gradients, while inaccurate ones introduce noise.
+However, we can mitigate this by bootstrapping the world model using existing trajectories before policy training, improving stability.
 
@@
 
-The SHAC++ framework allows for replacing only certain components with learned models.
-For example, if a differentiable environment is available, but not a differentiable reward.
-Then we can only learn the latter.
-SHAC+ refers to the case where only the reward is learned.
-We will provide a more detailed explanation in the paper.
+SHAC+ refers to a variant where only the reward function is learned while using the true environment dynamics.
 
-> Visual analysis of the trained policies' behavior would be more convincing.
+> Visual analysis of the trained policies' behavior would be more convincing
 
@@
 
-> Confusion on $l$ and $r$.
+> Confusion on $l$ and $r$
 
-$l$ and $r$ denote the lower and upper bound of the action space.
-For example, if the agent is supposed to output action in range of (0,1) then $l=0$ and $r=1$.
-We will provide a better explanation in the paper.
+$l$ and $r$ are the lower and upper bounds of the action space (e.g., for actions in range (0,1), $l=0$ and $r=1$).
 

### 2025-03-30 src/environments/DFlexEnv.py
reverting dflex

@@ -1,110 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import os
-import sys
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), '..')))
-
-import numpy as np
-import torch
-
-import dflex as df
-import xml.etree.ElementTree as ET
-
-from gym import spaces
-
-
-class DFlexEnv:
-    
-    def __init__(self, num_envs, num_obs, num_act, episode_length, MM_caching_frequency = 1, seed=0, no_grad=True, render=False, device='cuda:0'):
-        self.seed = seed
-
-        self.no_grad = no_grad
-        df.config.no_grad = self.no_grad
-
-        self.episode_length = episode_length
-
-        self.device = device
-
-        self.visualize = render
-
-        self.sim_time = 0.0
-
-        self.num_frames = 0 # record the number of frames for rendering
-
-        self.num_environments = num_envs
-        self.num_agents = 1
-
-        self.MM_caching_frequency = MM_caching_frequency
-        
-        # initialize observation and action space
-        self.num_observations = num_obs
-        self.num_actions = num_act
-
-        self.obs_space = spaces.Box(np.ones(self.num_observations) * -np.inf, np.ones(self.num_observations) * np.inf)
-        self.act_space = spaces.Box(np.ones(self.num_actions) * -1., np.ones(self.num_actions) * 1.)
-
-        # allocate buffers
-        self.obs_buf = torch.zeros(
-            (self.num_envs, self.num_observations), device=self.device, dtype=torch.float, requires_grad=False)
-        self.rew_buf = torch.zeros(
-            self.num_envs, device=self.device, dtype=torch.float, requires_grad=False)
-        self.reset_buf = torch.ones(
-            self.num_envs, device=self.device, dtype=torch.long, requires_grad=False)
-        # end of the episode
-        self.termination_buf = torch.zeros(
-            self.num_envs, device=self.device, dtype=torch.long, requires_grad=False)
-        self.progress_buf = torch.zeros(
-            self.num_envs, device=self.device, dtype=torch.long, requires_grad=False)
[…]

### 2025-03-30 src/environments/scenarios/Ant.py

@@ -1,343 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-from environments.DFlexEnv import DFlexEnv
-import math
-import torch
-import utils
-import os
-import sys
-sys.path.insert(0, os.path.abspath(os.path.join(os.path.dirname(__file__), '..')))
-import dflex as df
-import numpy as np
-np.set_printoptions(precision=5, linewidth=256, suppress=True)
-
-try:
-    from pxr import Usd
-except ModuleNotFoundError:
-    print("No pxr package")
-
-class AntEnv(DFlexEnv):
-
-    def __init__(self, render=False, device='cuda:0', num_envs=4096, seed=0, episode_length=1000, no_grad=True, stochastic_init=False, MM_caching_frequency = 16, early_termination = True):
-        num_obs = 37
-        num_act = 8
-    
-        super(AntEnv, self).__init__(num_envs, num_obs, num_act, episode_length, MM_caching_frequency, seed, no_grad, render, device)
-
-        self.stochastic_init = stochastic_init
-        self.early_termination = early_termination
-
-        self.init_sim()
-
-        # other parameters
-        self.termination_height = 0.27
-        self.action_strength = 200.0
-        self.action_penalty = 0.0
-        self.joint_vel_obs_scaling = 0.1
-
-        #-----------------------
-        # set up Usd renderer
-        if (self.visualize):
-            self.stage = Usd.Stage.CreateNew("outputs/" + "Ant_" + str(self.num_envs) + ".usd")
-
-            self.renderer = df.render.UsdRenderer(self.model, self.stage)
-            self.renderer.draw_points = True
-            self.renderer.draw_springs = True
-            self.renderer.draw_shapes = True
-            self.render_time = 0.0
-
-    def init_sim(self):
-        self.builder = df.sim.ModelBuilder()
-
-        self.dt = 1.0/60.0
-        self.sim_substeps = 16
-        self.sim_dt = self.dt
-
-        self.ground = True
-
-        self.num_joint_q = 15
-        self.num_joint_qd = 14
-
[…]

### 2025-03-30 src/environments/scenarios/__init__.py

@@ -3,4 +3,3 @@ from environments.scenarios.Transport  import Transport
 from environments.scenarios.Sampling   import Sampling
 from environments.scenarios.Flocking   import Flocking
 from environments.scenarios.Discovery  import Discovery
-from environments.scenarios.Ant        import Ant

### 2025-03-30 src/environments/utils.py

@@ -83,12 +83,5 @@ def get_environment(name:str, envs:int, agents:int, device:str, grad_enabled:boo
                 dict_spaces        = False        ,
                 seed               = seed         ,
             )
-        case "ant" :
-            return environments.scenarios.Ant(
-                device       = device       ,
-                num_envs     = envs         ,
-                seed         = seed         ,
-                grad_enabled = grad_enabled ,
-            )
         case _:
             raise ValueError(f"Unknown environment {name}.")

### 2025-03-30 src/experiments.py

@@ -6,7 +6,6 @@ ENVNAMES = [
     "transport" ,
     "sampling"  ,
     "discovery" ,
-    "ant"       ,
 ]
 
 NNS = [

### 2025-03-30 src/trainers/routines/policy.py

@@ -27,8 +27,8 @@ def train_policy(
     # compute loss
     optimizer.zero_grad()
     loss = -(
-        (episode_data["proxy_rewards"] * gammas * dones.logical_not()).sum() 
-        # + ((gammas * episode_data["values"])[live_steps,live_runs]).sum()
+        (episode_data["proxy_rewards"] * gammas * dones.logical_not()).sum() + 
+        ((gammas * episode_data["values"])[live_steps,live_runs]).sum()
     ) / (steps * envs)
 
     # add action space loss

### 2025-03-30 src/trainers/shacwm.py

@@ -178,25 +178,25 @@ def shacwm(
             clip_coefficient = reward_clip_coefficient ,
         )
 
-        # # train value model ###########################################
-        # trainers.routines.train_value(
-        #     episode          = episode               ,
-        #     model            = value_model           ,
-        #     optimizer        = value_model_optimizer ,
-        #     episode_data     = episode_data          ,
-        #     cached_data      = value_cache           ,
-        #     training_epochs  = value_epochs          ,
-        #     batch_size       = value_batch_size      ,
-        #     cache_size       = value_cache_size      ,
-        #     slam             = lambda_factor         ,
-        #     gamma            = gamma_factor          ,
-        #     bins             = value_bins            ,
-        #     logger           = value_logger          ,
-        #     ett              = value_ett             ,
-        #     stop_threshold   = value_stop_threshold  ,
-        #     tolerance        = value_tolerance       ,
-        #     clip_coefficient = value_clip_coefficient
-        # )
+        # train value model ###########################################
+        trainers.routines.train_value(
+            episode          = episode               ,
+            model            = value_model           ,
+            optimizer        = value_model_optimizer ,
+            episode_data     = episode_data          ,
+            cached_data      = value_cache           ,
+            training_epochs  = value_epochs          ,
+            batch_size       = value_batch_size      ,
+            cache_size       = value_cache_size      ,
+            slam             = lambda_factor         ,
+            gamma            = gamma_factor          ,
+            bins             = value_bins            ,
+            logger           = value_logger          ,
+            ett              = value_ett             ,
+            stop_threshold   = value_stop_threshold  ,
+            tolerance        = value_tolerance       ,
+            clip_coefficient = value_clip_coefficient
+        )
 
         # checkpoint ##################################################
         if episode % etv == 0:

### 2025-03-30 src/utils/load_utils.py

@@ -1,718 +1,0 @@
-# Copyright (c) 2022 NVIDIA CORPORATION.  All rights reserved.
-# NVIDIA CORPORATION and its licensors retain all intellectual property
-# and proprietary rights in and to this software, related documentation
-# and any modifications thereto.  Any use, reproduction, disclosure or
-# distribution of this software and related documentation without an express
-# license agreement from NVIDIA CORPORATION is strictly prohibited.
-
-import urdfpy
-import math
-import numpy as np
-import os
-import torch
-import random
-
-import xml.etree.ElementTree as ET
-
-import dflex as df
-
-
-def set_np_formatting():
-    np.set_printoptions(edgeitems=30, infstr='inf',
-                        linewidth=4000, nanstr='nan', precision=2,
-                        suppress=False, threshold=10000, formatter=None)
-
-def set_seed(seed, torch_deterministic=False):
-    if seed == -1 and torch_deterministic:
-        seed = 42
-    elif seed == -1:
-        seed = np.random.randint(0, 10000)
-    print("Setting seed: {}".format(seed))
-
-    random.seed(seed)
-    np.random.seed(seed)
-    torch.manual_seed(seed)
-    os.environ['PYTHONHASHSEED'] = str(seed)
-    torch.cuda.manual_seed(seed)
-    torch.cuda.manual_seed_all(seed)
-
-    if torch_deterministic:
-        # refer to https://docs.nvidia.com/cuda/cublas/index.html#cublasApi_reproducibility
-        os.environ['CUBLAS_WORKSPACE_CONFIG'] = ':4096:8'
-        torch.backends.cudnn.benchmark = False
-        torch.backends.cudnn.deterministic = True
-        torch.use_deterministic_algorithms(True)
-    else:
-        torch.backends.cudnn.benchmark = True
-        torch.backends.cudnn.deterministic = False
-
-    return seed
-
-def urdf_add_collision(builder, link, collisions, shape_ke, shape_kd, shape_kf, shape_mu):
-        
-    # add geometry
-    for collision in collisions:
-        
-        origin = urdfpy.matrix_to_xyz_rpy(collision.origin)
-
-        pos = origin[0:3]
-        rot = df.rpy2quat(*origin[3:6])
-
-        geo = collision.geometry
-
-        if (geo.box):
-            builder.add_shape_box(
-                link,
-                pos, 
-                rot, 
-                geo.box.size[0]*0.5, 
-                geo.box.size[1]*0.5, 
-                geo.box.size[2]*0.5,
[…]

### 2025-03-30 src/unroll.py
reverting dflex

@@ -17,16 +17,13 @@ def unroll(
         for i in dones[:,0].nonzero():
             observations[i] = torch.stack(world.reset_at(i)).transpose(0,1)[i]
 
-    if observations is not None and dones is not None:
-        observations = world.world.reset(env_ids=torch.arange(world.batch_dim, device=world.device)[dones[:,0]]).unsqueeze(1)
-
     if observations is None and dones is None: 
         observations = torch.stack(world.reset()).transpose(0,1)
         dones        = torch.zeros(observations.size(0), observations.size(1), device=observations.device).bool()
 
     max_reward   = world.scenario.max_rewards() if max_reward else torch.zeros(world.batch_dim, device=world.device)
 
-    observations = world.world.initialize_trajectory().unsqueeze(1)
+    observations = observations.detach()
 
     observation_cache = []
     action_cache      = []
@@ -40,7 +37,7 @@ def unroll(
         observation_cache.append(observations)
         done_cache       .append(dones)
         policy_result = policy_model(observations)
-        actions  = torch.nn.functional.tanh( policy_result["actions"] )
+        actions  = policy_result["actions"]
         logprobs = policy_result.get("logprobs",torch.empty(0))
         entropy  = policy_result.get("entropy" ,torch.empty(0))
         logits   = policy_result.get("logits"  ,torch.empty(0))
@@ -63,8 +60,6 @@ def unroll(
     logprobs_cache    = torch.stack(logprobs_cache)
     entropy_cache     = torch.stack(entropy_cache)
     logits_cache      = torch.stack(logits_cache)
-
-    print(world.world.state.joint_q.view(world.batch_dim, -1)[0].mean())
 
     return { 
             "logprobs"      : logprobs_cache    ,

### 2025-03-30 src/utils/__init__.py

@@ -19,6 +19,3 @@ from utils.random_dispatch    import random_dispatch
 from utils.save_config        import save_config
 from utils.seed_everything    import seed_everything
 from utils.value_loss         import value_loss
-
-import utils.torch_utils
-import utils.load_utils

### 2025-04-22 paper/ecai2025/sects/conclusion.tex
update to ecai 2025 template

--- 2025-01-20 paper/icml25/first/sects/conclusion.tex
@@
-\section{Conclusion}
-In this paper, we proposed \fname{}, an extension of SHAC, 
-to provide a reinforcement learning solution that is both sample efficient,
-able to work in both single-agent and multi-agent scenarios,
-and capable of handling non-differentiable scenarios. 
-We demonstrated that this solution maintains alignment with SHAC (\RQ{1}), outperforms the current state-of-the-art PPO in both single-agent and multi-agent scenarios (\RQ{2}), 
-and exhibits promising scaling capabilities in multi-agent settings with emergent collaborative behaviors (\RQ{3}).
+\section{Conclusions \& Future Works}\label{sect:conclusion}
+In this paper, we proposed \fname{}, an extension of SHAC, to provide a reinforcement learning solution that is both sample efficient, able to work in both single-agent and multi-agent scenarios, and capable of handling non-differentiable scenarios.
+We demonstrated that this solution maintains alignment with SHAC (\RQ{1}), outperforms the current state-of-the-art PPO in both single-agent and multi-agent scenarios (\RQ{2}), and exhibits promising scaling capabilities in multi-agent settings with emergent collaborative behaviors (\RQ{3}).
 

### 2025-04-22 paper/ecai2025/sects/related.tex
update to ecai 2025 template

--- 2025-01-19 paper/icml25/first/sects/related.tex
@@
 Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}.
-In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL.
-\cite{Zheng24} employes a differentiable symbolic expression search engine.
-Trajectory optimization approach is taken in~\cite{Wan24}.
-\cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.
+In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL\@.\ \citet{Zheng24} employ a differentiable symbolic expression search engine.
+Trajectory optimization approach is taken in~\cite{Wan24}.\ \citet{Georgiev24} tackles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.
 With respect to this approach, we take a substantially different method, by emulating the differentiable environment with a trained neural network.

### 2025-04-22 paper/ecai2025/appendices/arch.tex
first pass on spacing

--- 2025-01-22 paper/icml25/first/appendices/arch.tex
@@
+\section{Architectures}\label{apx:arch}
+
 \begin{figure}[t]
     \centering
-    \includegraphics[width=.5\textwidth]{figs/attention.pdf} 
-    \caption{Attention matrix for the world model.}
-    \label{fig:apx:attention}
+    \includegraphics[width=.45\textwidth]{figs/attention.pdf} 
+    \caption{Attention matrix for the world model.}\vspace{0.5cm}
+    \label{apx:fig:attention}
 \end{figure}
 
-\section{Architectures}\label{apx:arch}
-
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=.45\textwidth]{figs/grads-mlp-transport.pdf}
+    \caption{Gradient norm for a transformer architecture wrt. a MLP architecture.}\vspace{0.5cm}
+    \label{fig:grads-mlp}
+\end{figure}
 As mentioned earlier, we test PPO, SHAC, \fname{}, and \fnamer{} with two different architecture---Transformer and MLP.
@@
 For a full list of hyperparameters, see \Cref{apx:tab:shac}.
-
-\begin{table}[t]
-        \centering
-        \input{tables/shacwm.tex}
-        \caption{SHAC/\fname{}/\fnamer{} Hyperparameters}
-        \label{apx:tab:shac}
-\end{table}
-
-\begin{table}[t]
-        \centering
-        \input{tables/ppo.tex}
-        \caption{PPO Hyperparmeters}
-        \label{apx:tab:ppo}
-\end{table}

### 2025-04-22 paper/ecai2025/appendices/scenarios.tex
first pass on spacing

--- 2025-01-21 paper/icml25/first/appendices/scenarios.tex
@@
-\begin{figure}[!t]
+\section{Environments}\label{apx:scenarios}
+\begin{figure}
     \centering
-    \begin{subfigure}[b]{0.24\textwidth}
+    \begin{subfigure}{0.24\textwidth}
         \includegraphics[width=\textwidth]{figs/dispersion.png}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
+    \begin{subfigure}{0.24\textwidth}
         \includegraphics[width=\textwidth]{figs/transport.png}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
+    \begin{subfigure}{0.24\textwidth}
         \includegraphics[width=\textwidth]{figs/discovery.png}
@@
     \end{subfigure}
-    \begin{subfigure}[b]{0.24\textwidth}
+    \begin{subfigure}{0.24\textwidth}
         \includegraphics[width=\textwidth]{figs/sampling.png}
@@
     \end{subfigure}
-    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}
+    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
 \end{figure}
-
-\section{Environments}\label{apx:scenarios}
 
@@
 
-\subsection{Dicsovery}
+\subsection{Discovery}
 In the Discovery task, $n$ agents aim to collect as many randomly placed goals as possible.

### 2025-04-26 paper/ecai2025/appendices/scenarios.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

@@
 \section{Environments}\label{apx:scenarios}
+\begin{table}[htbp]
+  \centering
+  \caption{Comparison of Multi-Agent Reinforcement Learning Environments/Frameworks}
+  \label{tab:marl_comparison}
+  % Optional: resizebox to fit width if needed
+  \resizebox{\textwidth}{!}{% 
+  \begin{tabular}{@{}l l l l l l l l@{}}
+    \toprule % Use booktabs rule, or \hline\hline
+    \textbf{Feature} & \textbf{VMAS} & \textbf{PettingZoo} & \textbf{SMAC} & \textbf{MA-MuJoCo} & \textbf{PyBullet} & \textbf{Neural MMO} & \textbf{Melting Pot} \\
+    \midrule % Use booktabs rule, or \hline
+    Key Focus & Fast, Vectorized & Standardized & StarCraft II & Multi-Agent & General 3D & Large-Scale Agent & Social Dilemmas \\
+     & 2D Physics & MARL API & Micromanagement & Robotics (MuJoCo & Physics & Survival/Econ. & / Cooperation \\
+     & (PyTorch) & (Collection) & & tasks) & Engine & & \\
+    \midrule % Separator line
+    Single/Multi & Primarily Multi & Primarily Multi & Multi Only & Primarily Multi & Both & Multi Only & Multi Only \\
+     & (can do Single) & (API standard) & & (adaptations) & & (Massively) & \\
+    \midrule % Separator line
+    Differentiable & Yes & No & No & No & No (standard & No & No \\
+     & & & & & use) & & \\
+    \midrule % Separator line
+    Generality & Medium & Very High (via & Low (Specific game & Medium (MuJoCo & High & Medium (Large- & Medium (Social \\
+     & (Physics-based, & wrapped envs) & benchmark) & robotics) & (General & scale agent focus) & dilemmas focus) \\
+     & PyTorch) & & & & physics) & & \\
+    \bottomrule % Use booktabs rule, or \hline
+  \end{tabular}
+  } % Closing bracket for resizebox, if used
+\end{table}
+
 \begin{figure}
@@
 
-In this section, we provide a detailed discussion of the environment and scenarios used to test the \fname{} algorithm. Specifically, we utilize four vectorized environments available as part of VMAS.
+In this section, we provide a detailed overview of the environments used to evaluate the \fname{} algorithm.
+For our experiments, we selected four vectorized scenarios from VMAS (Vectorized Multi-Agent Simulator), which offers an ideal balance of features for our research objectives.
 
-VMAS is a differentiable simulator designed for various multi-agent tasks.
-These tasks are built on top of a differentiable physics engine, resulting in differentiable transition functions for all tasks.
-However, the reward function is non-differentiable.
-To address this limitation, we created custom reward functions, where possible, to replicate the non-differentiable ones provided by VMAS.
-Consequently, some scenarios are compatible with all tested algorithms (PPO, SHAC, and \fname{}), while others are only applicable to PPO and \fname{}.
+VMAS stands out among alternative multi-agent reinforcement learning frameworks such as PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, PyBullet~\cite{DBLP:journals/corr/abs-1710-06347}, Neural MMO~\cite{DBLP:conf/aaai/DiehlB21}, and Melting Pot~\cite{DBLP:conf/icml/GuptaK0G20}.
+As summarized in Table~\ref{tab:marl_comparison}, VMAS provides several key advantages for our work:
 
-Furthermore, the observations produced by VMAS are often only partial representations of the full state.
-While in some cases these observations are sufficient to predict the next observations, in others they are not.
-In such cases, \fname{} is expected to face challenges, as the world model is trained using only partial information.
-In contrast, SHAC computes gradients using access to the full world state.
-We chose to retain this limitation to better reflect the behavior of \fname{} in more realistic environments.
+\begin{compactitem}
+    \item \textbf{Multi-Agent Framework}: VMAS is specifically designed for multi-agent scenarios, making it appropriate for our experimental setup.
+    \item \textbf{Differentiable Physics}: Unlike most alternatives, VMAS offers a differentiable physics engine that enables gradient-based optimization techniques.
+    \item \textbf{Vectorized Implementation}: The framework supports parallel execution of multiple environments, significantly improving training efficiency.
+    \item \textbf{Physics-based Simulation}: VMAS provides realistic modeling of agent interactions and dynamics through its physics engine.
+    \item \textbf{Open Source Accessibility}: The codebase is fully accessible, allowing necessary modifications for our experiments.
+\end{compactitem}
+
+While VMAS provides differentiable transition functions across all tasks, its reward functions are generally non-differentiable.
+To address this limitation, we implemented custom differentiable reward functions where possible to replicate the original VMAS behavior.
+Consequently, some scenarios are compatible with all tested algorithms (PPO, SHAC, and \fname{}), while others can only be used with PPO and \fname{}.
+
+An additional consideration is that VMAS often provides agents with partial observations rather than complete state information.
+In some scenarios, these observations contain sufficient information to predict subsequent states, while in others they do not.
+This partial observability presents a particular challenge for \fname{}, as its world model must operate with incomplete information.
+In contrast, SHAC computes gradients with access to the full world state.
+We deliberately maintained these observability constraints to assess \fname{}'s performance under more realistic conditions.
+
+Figure~\ref{fig:scenarios} illustrates the four VMAS environments used in our experiments, which are detailed in the following subsections.
 

### 2025-04-26 paper/ecai2025/sects/abstract.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

\begin{abstract}
Reinforcement learning (RL) algorithms show promise in robotics and multi-agent systems but often suffer from low sample efficiency. While methods like SHAC leverage differentiable simulators to improve efficiency, they are limited to specific settings: they require fully differentiable environments (including transition and reward functions) and have primarily been demonstrated in single-agent scenarios. To overcome these limitations, we introduce \fname{}, a novel framework inspired by SHAC\@. \fname{} removes the need for differentiable simulator components by using neural networks to approximate the required gradients, training these networks alongside the standard policy and value networks. This allows the core SHAC approach to be applied in both non-differentiable and multi-agent environments. We evaluate \fname{} on challenging multi-agent tasks from the VMAS suite, comparing it against SHAC (where applicable) and PPO, a standard algorithm for non-differentiable settings. Our results demonstrate that \fname{} significantly outperforms PPO in both single- and multi-agent scenarios. Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems
\end{abstract}

### 2025-04-26 paper/ecai2025/sects/experiments.tex
first pass on spacing
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

--- 2025-01-24 paper/icml25/first/sects/experiments.tex
@@
-\begin{figure*}[!t]
+\begin{figure*}[t]
     \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
@@
 \begin{figure}[t]
+    \centering
     \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
@@
 Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
+    \vspace{0.5cm}
 \end{figure}
-
 \section{Experiments}\label{sect:experiments}
@@
 In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures.
-An ablation study when replacing only the transition function (meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.
+An ablation study when replacing only the transition function (SHAC+,meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.
 
 \paragraph{Scenarios}
-As mentioned previously, we choose a few multi-agent scenarios from the VMAS package to test \fname{}, SHAC and PPO\@.
-In particular, we choose the scenarios Dispersion, Discovery, Transport, and Sampling.
-For details regarding each scenario, see \Cref{apx:scenarios}.
-Dispersion and Discovery have non-differentiable reward functions, thus SHAC is not applicable.
-Transport and Sampling have differentiable reward functions, thus SHAC is applicable.
-For Transport and Dispersion, the observation space is complete wrt.\ the state.
-For Discovery and Sampling, the observation space is incomplete wrt.\ the state.
-In these scenario, \fname{} is slightly disadvantaged as it does not have access to the full state of the environment compared to SHAC\@.
+\begin{figure*}
+    \centering
+    \begin{subfigure}{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/dispersion.png}
+        \caption{Dispersion}
+        \label{fig:dispersion}
+    \end{subfigure}
+    \begin{subfigure}{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/transport.png}
+        \caption{Transport}
+        \label{fig:transport}
+    \end{subfigure}
+    \begin{subfigure}{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/discovery.png}
+        \caption{Discovery}
+        \label{fig:discovery}
+    \end{subfigure}
+    \begin{subfigure}{0.24\textwidth}
+        \includegraphics[width=\textwidth]{figs/sampling.png}
+        \caption{Sampling}
+        \label{fig:sampling}
+    \end{subfigure}
+    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
+\end{figure*}
+We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
+For more details about the choice of environments, see \Cref{apx:scenarios}.
+We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details).
+These scenarios can be categorized along two dimensions:
+\begin{compactitem}
+    \item \textbf{Reward differentiability:} Dispersion and Discovery feature non-differentiable reward functions (making SHAC inapplicable), while Transport and Sampling have differentiable rewards (allowing for SHAC implementation).
+    \item \textbf{Observation completeness:} Transport and Dispersion provide agents with complete observation spaces relative to the environment state.
+In contrast, Discovery and Sampling have incomplete observation spaces, which presents an additional challenge for \fname{} compared to SHAC, as \fname{} lacks access to the complete environmental state.
+\end{compactitem}
 
-\begin{table*}[!t]
+This selection of scenarios enables a thorough evaluation of how each algorithm handles different types of multi-agent cooperation challenges.
+
+\begin{table*}[t]
     \centering

### 2025-04-26 paper/ecai2025/sects/introduction.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro

\section{Introduction}\label{sect:introduction}
Learning control policies is crucial for applications like robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, and swarm intelligence \cite{Tang21}. These domains often present challenges such as sparse rewards, complex dynamics, and multi-agent interactions. Many current RL approaches focus specifically on either single- or multi-agent settings, lacking a unified solution. Furthermore, RL algorithms are often \emph{sample inefficient}, requiring extensive interaction with the environment to learn effective policies.

A promising direction to improve sample efficiency involves utilizing \emph{differentiable simulators}, namely simulators that allow for backpropagation through their components. SHAC \cite{Xu22} exemplifies this approach, using gradients obtained directly from the simulator to accelerate policy optimization. However, SHAC's applicability is constrained by two key limitations. First, it necessitates both a differentiable transition function \emph{and} a differentiable reward function, which is often impractical, especially with sparse or complex reward structures. Second, while powerful, backpropagation through complex dynamics, like object collisions~\cite{Georgiev24} or intricate multi-agent interactions, can lead to unstable gradients~\cite{Bengio94, Metz21}, potentially hindering learning. The original SHAC work also focused primarily on single-agent tasks.

In this work, we propose \fname{} to address these limitations and extend gradient-based policy optimization beyond SHAC's constraints. Our goal is to develop a sample-efficient reinforcement learning framework that maintains the benefits of gradient-based optimization without requiring differentiable simulators. Drawing inspiration from model-based RL \cite{DBLP:conf/icml/HafnerLFVHLD19, DBLP:conf/iclr/HafnerLB020}, \fname{} substitutes direct simulator differentiation with neural approximations of the gradient functions. By training networks that approximate transition and reward gradients concurrently with policy optimization, our approach preserves the sample efficiency benefits of gradient information while eliminating strict differentiability requirements. This design enhances robustness to complex dynamics and enables application in both non-differentiable environments and multi-agent scenarios, albeit with additional computational overhead compared to SHAC.

We evaluate \fname{} against two established baselines: PPO \cite{Schulman17} (including its multi-agent variant MAPPO \cite{DBLP:conf/nips/YuVVGWBW22}) as the standard for non-differentiable environments, and the original SHAC \cite{Xu22} where applicable. Our experiments utilize the VMAS simulator \cite{DBLP:conf/dars/BettiniKBP22}, which provides physics-based multi-agent environments with varying levels of complexity. Through this comparative analysis on cooperative tasks with different agent populations, we investigate the following research questions:

\begin{compactitem}
    \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
    \item \RQ{2} How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?
    \item \RQ{3} How does the performance of these algorithms change as the search space increases?
\end{compactitem}

Our contributions include:
\begin{compactitem}
    \item A novel RL framework (\fname{}) that employs learned gradient approximations, eliminating the need for differentiable simulators
    \item The first empirical evaluation of SHAC in multi-agent environments, establishing a baseline for differentiable MARL
    \item A comprehensive comparison across environments with varying agent counts and differentiability properties
    \item Experimental evidence demonstrating that \fname{} matches SHAC's performance in differentiable settings while substantially outperforming PPO in sample efficiency and final reward across both single-agent and complex multi-agent scenarios
\end{compactitem}

### 2025-04-26 paper/ecai2025/sects/related.tex
chore: minors on related

--- 2025-01-19 paper/icml25/first/sects/related.tex
@@
 Precursors of SHAC with similar intuition are PODS~\cite{Mora21} and CE-APG~\cite{Gillen22}.
-In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL.
-\cite{Zheng24} employes a differentiable symbolic expression search engine.
+In~\cite{Lv23}, the authors introduce rendering to enhance model-based RL\@.\ In~\cite{Zheng24} authors employ a differentiable symbolic expression search engine.
 Trajectory optimization approach is taken in~\cite{Wan24}.
-\cite{Georgiev24} takles unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.
+In~\cite{Georgiev24} authors tackle unstable gradient due to stiff dynamics by adaptively truncating trajectories when these occur.
 With respect to this approach, we take a substantially different method, by emulating the differentiable environment with a trained neural network.

### 2025-04-26 paper/ecai2025/sects/experiments.tex
minors

@@
 
-In this section, we test \fname{} with PPO and SHAC for a set of multi-agent scenarios and architectures.
-An ablation study when replacing only the transition function (SHAC+,meaning that the reward function is assumed differentiable) is also conducted in \Cref{apx:ablation}.
+In this section, we evaluate \fname{} against both PPO and SHAC across various multi-agent scenarios and architectures.
+An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
+We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC \cite{Haarnoja2018SoftAA}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}).
+This makes PPO a representative benchmark for state-of-the-art performance in non-differentiable environments.
 
 \paragraph{Scenarios}
-\begin{figure*}
-    \centering
-    \begin{subfigure}{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/dispersion.png}
-        \caption{Dispersion}
-        \label{fig:dispersion}
-    \end{subfigure}
-    \begin{subfigure}{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/transport.png}
-        \caption{Transport}
-        \label{fig:transport}
-    \end{subfigure}
-    \begin{subfigure}{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/discovery.png}
-        \caption{Discovery}
-        \label{fig:discovery}
-    \end{subfigure}
-    \begin{subfigure}{0.24\textwidth}
-        \includegraphics[width=\textwidth]{figs/sampling.png}
-        \caption{Sampling}
-        \label{fig:sampling}
-    \end{subfigure}
-    \caption{VMAS environments used in the experiments.}\label{fig:scenarios}\vspace{0.5cm}
-\end{figure*}
 We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
@@
 \subsection{Results}
+\begin{figure}
+    \centering
+    \includegraphics[width=0.32\columnwidth]{figs/step-1.png}
+    \includegraphics[width=0.32\columnwidth]{figs/step-2.png}
+    \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
+    \caption{Snapshots of the result of the Transport scenario.
+The agents are represented as colored circles, the package as a red square, and the goal as a green circle.
+The agents are trained to push the package to the goal.}\label{fig:transport-trained}
+    \vspace{0.5cm}
+\end{figure}
 In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
@@
 With more agents, convergence is faster because the task becomes easier (as more agents can push the package).
-The emergent of cooperation can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package.
+The emergent of cooperation (see \Cref{fig:transport-trained}) can be also display but looking to the gradient norm in \Cref{fig:grads-transformer-transport} we can see that the norm of the gradients for SHAC and \fname{} increases when the agents start to collide with each other and the package.
 This is a sign of generalization, as the agents learn to avoid collisions and push the package together.

### 2025-05-02 paper/ecai2025/sects/limitations.tex
added limitations

\section{Limitations}

Environments often exhibit vastly different characteristics and challenges, and the proposed method is not a universal solution suitable for all scenarios. One of the most difficult components to learn is the world model, which relies on meaningful interactions between agents and the environment—interactions that do not necessarily yield immediate rewards. This creates a cold-start problem: the world model cannot be learned until agents are able to interact effectively with the environment. To address this, as in prior works, we use stochastic action sampling instead of a deterministic policy. In the environments considered, this approach eventually leads to the successful learning of a world model, which in turn allows agents to develop a viable policy. However, this strategy may not be effective in more complex environments where random actions fail to produce meaningful progress. In such cases, the proposed method may not succeed in learning a policy.

This limitation can be mitigated by introducing a cold-start dataset—either manually curated or generated using alternative algorithms such as PPO or MPPI~\cite{Alvarez24}. Such data can help initialize the world model and provide a useful foundation for the learning process.

### 2025-05-04 paper/ecai2025/appendices/arch.tex
add more experiments, minor on discussion

@@
     \centering
-    \includegraphics[width=.45\textwidth]{figs/attention.pdf} 
-    \caption{Attention matrix for the world model.}\vspace{0.5cm}
+    \begin{subfigure}[b]{0.45\textwidth}
+        \centering
+        \includegraphics[width=\textwidth]{figs/attention.pdf}
+        \caption{Attention matrix for the world model.}
         \label{apx:fig:attention}
-\end{figure}
-
-\begin{figure}[t]
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.45\textwidth}
         \centering
-    \includegraphics[width=.45\textwidth]{figs/grads-mlp-transport.pdf}
-    \caption{Gradient norm for a transformer architecture wrt. a MLP architecture.}\vspace{0.5cm}
+        \includegraphics[width=\textwidth]{figs/grads-mlp-transport.pdf}
+        \caption{Gradient norm for a transformer architecture wrt. a MLP architecture.}
         \label{fig:grads-mlp}
+    \end{subfigure}
+    \caption{Network architecture visualizations: (left) attention pattern in the world model; (right) gradient norm comparison between transformer and MLP architectures.}
+    \label{fig:combined-arch}
+    \vspace{0.5cm}
 \end{figure}

### 2025-05-04 paper/ecai2025/sects/discussion.tex
start handling some reviewer point: why we choose vmas and the scenarion, clearer intro
add more experiments, minor on discussion

--- 2025-01-23 paper/icml25/first/sects/discussion.tex
@@
 Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment.
-\begin{center}
-\RQ{1} \textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
-\end{center}
+\begin{rqbox}
+\textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
+\end{rqbox}
 
@@
 
-\begin{center}
-\RQ{2} \textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
-\end{center}
+\begin{replybox}
+Yes, \fname{} approximates simulator gradients, achieving comparable performance to SHAC in differentiable settings, occasionally outperforming it despite initially higher variance.
+\end{replybox}
+
+\begin{rqbox}
+\textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
+\end{rqbox}
+
 The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency.
-In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps.
+
+\noindent\paragraph{Single-agent:} In single-agent cases, while PPO manages to find near-good policies (e.g., in the Dispersion scenario), it often fails to complete scenarios within the same number of steps.
 Indeed, \fname{} achieves better results in all environments (see \Cref{tab:max-rewards}).
-Despite having to train multiple networks, \fname{} converges much faster to optimal policies, as gradient guidance enables quicker and more accurate convergence to the desired outcome.
-This pattern becomes even more evident in multi-agent scenarios.
+Compared to SHAC (where applicable), \fname{} performs comparably.
+
+\noindent\paragraph{Multi-agent:} This pattern becomes even more evident in multi-agent scenarios.
 In these cases, PPO often struggles to find even marginally good policies.
 \fname{}, on the other hand, eventually succeeds in understanding the environment and subsequently improves the policy to enable cooperation---see both \Cref{tab:max-rewards} and \Cref{fig:experiments}.
+Its performance remains comparable to SHAC where SHAC can be applied.
+Overall, despite having to train multiple networks, \fname{} converges much faster to optimal policies than PPO, as gradient guidance enables quicker and more accurate convergence to the desired outcome.
 
-\begin{center}
-\RQ{3} \textbf{How does the performance of these algorithms change as the search space increases?}
-\end{center}
-To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents.
+\begin{replybox}
+\fname{} outperforms PPO across all scenarios, with PPO struggling especially in multi-agent settings.
+When compared to SHAC (where applicable), \fname{} achieves comparable performance while enabling emergent cooperative behaviors.
+\end{replybox}
+
+\begin{rqbox}
+\textbf{How does the performance of these algorithms change as the search space increases?}
+\end{rqbox}
+
+To analyze the impact of increasing search spaces, we primarily focused on scenarios with varying numbers of agents (up to 5 in the main experiments).
 The results demonstrate that performance patterns differ across scenarios.
@@
 Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents.
-This counterintuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
+This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
 Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.
 
-This advantage does not hold in the Dispersion scenario, where complexity increases substantially with additional agents.
-Nevertheless, we observe the emergence of cooperation across all scenarios, suggesting potential scalability to larger agent populations.
-The main bottleneck lies in the transformer architecture for the action world model, where the sequence length scales with the number of agents (sequence length = num.
-agents times num.
-train steps).
-With significantly large populations (~50 agents), the attention mechanism becomes computationally expensive, leading to prohibitively large memory requirements.
-In these scenarios, a linear attention mechanism (such as \cite{Beltagy20}) may be employed.
+This scaling advantage is not observed in the Dispersion scenario, where complexity increases quadratically with agent count.
+Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (quantitative analysis with up to 20 agents is provided in \Cref{apx:experiments}).
+The primary computational constraint lies in the transformer architecture employed for the action world model, where sequence length scales linearly with the number of agents (specifically, sequence length = $n \times s$, where $n$ is the number of agents and $s$ is the number of training steps).
+With agent populations exceeding approximately 50 agents, the self-attention mechanism's $O(n^2)$ complexity leads to prohibitive memory requirements.
+For such large-scale multi-agent systems, linear attention mechanisms (e.g.,~\cite{Beltagy20}) represent a viable architectural alternative.
 
-The Discovery scenario, exhibiting high variance, shows no clear pattern compared to the other three scenarios.
-This might be attributed to partial observability, which significantly complicates convergence and inter-agent cooperation.
+\begin{replybox}
+Performance varies with scale: Transport/Sampling converge faster with more agents, while Dispersion becomes harder.
+Cooperation emerges across scenarios.
+Main scalability limitation is transformer attention complexity, becoming prohibitive at large agent counts ($~$50), though moderate populations remain feasible.
+\end{replybox}

### 2025-05-04 paper/ecai2025/sects/experiments.tex
add more experiments, minor on discussion

@@
 \begin{figure*}[t]
-    \includegraphics[width=\textwidth]{figs/main-transformer.pdf}
+    \centering
+    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
     \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
@@
     \centering
-    \includegraphics[width=\columnwidth]{figs/grads-transformer-transport.pdf}
-    \caption{Norm of policy gradients on the Transport environment across 5000 epochs.
-The raise in norm for SHAC (epoch 2700) corresponds with occurrence of generalization, where collision between agents and the packages become frequent.
-Also \fname{} experience a raise in norm, albeit more behaved, corresponds with the occurrence of generalization (epoch 3200).}\label{fig:grads-transformer-transport}
+    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
+    \caption{Policy gradient norms in the Transport environment over 5000 epochs.
+Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
     \vspace{0.5cm}
@@
     \includegraphics[width=0.32\columnwidth]{figs/step-3.png}
-    \caption{Snapshots of the result of the Transport scenario.
-The agents are represented as colored circles, the package as a red square, and the goal as a green circle.
-The agents are trained to push the package to the goal.}\label{fig:transport-trained}
+    \caption{Snapshots of trained agents with \fname{} in the Transport scenario. Colored circles represent agents pushing a red package (square) toward the green goal.}\label{fig:transport-trained}
     \vspace{0.5cm}

### 2025-05-05 paper/ecai2025/sects/abstract.tex
few typos

@@
 Our results demonstrate that \fname{} significantly outperforms PPO in both single- and multi-agent scenarios.
-Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems
+Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems.
 \end{abstract}

### 2025-05-05 paper/ecai2025/sects/background.tex
few typos

--- 2025-01-28 paper/icml25/first/sects/background.tex
@@
 
-In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled usin $\pi$.
+In its trivial multi-agent extension, SHAC trains policies ${\pi_1}_\theta,\dots,{\pi_\na}_\theta$ by minimizing the loss function $\mathcal{L}_\theta^{\pi}:\trajectories^\bs\rightarrow\mathbb{R}$ for a batch of $\bs$ trajectories $\batch = \{\tau_1, \ldots, \tau_\bs\}$ sampled using $\pi$.
 The loss $\mathcal{L}_\theta^{\pi}(\batch)$ is defined as:

### 2025-05-05 paper/ecai2025/sects/introduction.tex
few typos

@@
 
-Our contributions include:
+\noindent Our contributions include:
 \begin{compactitem}
-    \item A novel RL framework (\fname{}) that employs learned gradient approximations, eliminating the need for differentiable simulators
-    \item The first empirical evaluation of SHAC in multi-agent environments, establishing a baseline for differentiable MARL
-    \item A comprehensive comparison across environments with varying agent counts and differentiability properties
-    \item Experimental evidence demonstrating that \fname{} matches SHAC's performance in differentiable settings while substantially outperforming PPO in sample efficiency and final reward across both single-agent and complex multi-agent scenarios
+    \item a novel RL framework (\fname{}) that employs learned gradient approximations, eliminating the need for differentiable simulators;
+    \item the first empirical evaluation of SHAC in multi-agent environments, establishing a baseline for differentiable MARL;
+    \item a comprehensive comparison across environments with varying agent counts and differentiability properties;
+    \item experimental evidence demonstrating that \fname{} matches SHAC's performance in differentiable settings while substantially outperforming PPO in sample efficiency and final reward across both single-agent and complex multi-agent scenarios.
 \end{compactitem}

### 2025-05-05 paper/ecai2025/sects/discussion.tex
minor on fig 1

@@
-\section{Discussion}
-In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO trying to answer the research questions posed in \Cref{sect:introduction}.
-Note that \fname{} does not aim to outperform SHAC, it merely aims to perform comparably with SHAC while being applicable even in non-differentiable environment.
+In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO, addressing the research questions posed in \Cref{sect:introduction}.
+Note that \fname{} does not aim to outperform SHAC, but rather to perform comparably while being applicable even in non-differentiable environments.
+
+\subsection{Gradient Approximation with Neural Networks}
 \begin{rqbox}
@@
 Therefore, we can conclude that it is definitely possible to approximate the gradients of a differentiable environment with a neural network.
-However, we note that this approach, while obtaining comparable results in most setups, it tends to display higher variance.
+However, we note that this approach, while obtaining comparable results in most setups, tends to display higher variance.
 Both neural networks representing the reward and transition functions require sufficient training before becoming effective, which may entail multiple episodes.
@@
 
+\subsection{Comparison to PPO and SHAC}
 \begin{rqbox}
@@
 
-The proposed algorithm appears capable of outperforming PPO across all scenarios showing also a better sample efficiency.
+The proposed algorithm appears capable of outperforming PPO across all scenarios, also showing better sample efficiency.
 
@@
 
+\subsection{Scaling with Search Space}
 \begin{rqbox}
@@
 Conversely, in Transport and Sampling scenarios, the system appears to converge more quickly with more agents.
-This counter intuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
+This counterintuitive behavior likely stems from the fact that, despite larger search spaces, the fundamental tasks in Sampling and Transport become relatively easier to solve with multiple agents, making policy discovery more straightforward.
 Even with approximated world dynamics (transition and reward functions), the guidance provided is sufficient to lead policies toward optimal behavior.

### 2025-05-05 paper/ecai2025/sects/shacpp.tex
minor on fig 1

--- 2025-01-28 paper/icml25/first/sects/shacpp.tex
@@
         Comparison between SHAC and \fname{} during a rollout of 4 steps.
-        Nodes with labels (such as $r_i$ or $r'_i$) values.
-Nodes without labels (such as $\pi$ or $R$) represent functions.
-Dashed lines are used to represent non differentiability.
-$a_i$ and $r_i$ represent the collection of actions and rewards for all agents}\label{fig:shac-shacpp}
+        Nodes with labels (such as $r_i$ or $r'_i$) indicate values, while nodes without labels (such as $\pi$ or $R$) represent functions.
+Dashed lines denote non-differentiability.
+$a_i$ and $r_i$ refer to the collection of actions and rewards for all agents.
+Note that the main advantage of \fname{} is that the transition function $F$ on the right can be non-differentiable, unlike the left $F$ in SHAC, which must be differentiable.
+        }\label{fig:shac-shacpp}
 \end{figure*}

### 2025-05-06 paper/ecai2025/sects/discussion.tex
minor on discussion

@@
+\section{Discussion}
 In the following, we discuss the results obtained in \Cref{sect:experiments} and provide some insights into the performance of \fname{} compared to SHAC and PPO, addressing the research questions posed in \Cref{sect:introduction}.

### 2025-07-26 paper/camera/appendices/additional.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-01-22 paper/icml25/first/appendices/additional.tex
@@
-\section{Additional Experiments}
+\section{Additional Experiments}\label{apx:experiments}
 
@@
 
+To check also the scalability of \fname{} we run the experiments with up to $20$ agents in the Transport scenario.
+The results are shown in \Cref{apx:fig:experiments-mlp}.
+We observe that \fname{} is able to scale well with the number of agents, achieving good results even with $20$ agents, also in this case similarly to SHAC.
 \begin{figure}[t]
@@
     We show the mean and standard deviation of the normalized rewards across $3$ runs. For the MLP architecture see \Cref{fig:experiments}.
-    }
+    }\vspace{0.5cm}
     \label{apx:fig:experiments-mlp}
+\end{figure}
 
+\begin{figure}[t]
+    \centering
+    \includegraphics[width=\columnwidth]{figs/transport-high-agents.pdf}
+    \caption{Comparison between \fname{} and SHAC for increasing number of agents (up to 20) in the Transport scenario with the Transformer architecture.
+Results demonstrate the scalability of both approaches as the number of agents increases.}\label{apx:fig:experiments-mlp-20}
+
+    \vspace{0.5cm}
 \end{figure}

### 2025-07-26 paper/camera/figs/attention.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\documentclass[crop,dvipsnames,tikz]{standalone}

\usetikzlibrary{calc}

\definecolor{darkorchid}{rgb}{0.6, 0.2, 0.8}

\begin{document}
\begin{tikzpicture}
    \foreach \a in {1,...,3} {
        %\foreach \i in {0,...,5} {
        %    \node at (\i+0.5, 6.5) {$a_{\a\i}$};
            \foreach \j in {1,...,5} {
                \node at (\a*5+\j-5,16) {$a_{\a\j}$};
                \node at (0,16-\a*5-\j+5) {$a_{\a\j}$};
                \foreach \aa in {1,...,3} {
                    \foreach \jj in {1,...,5} {

                        %% when j < jj and a == aa
                        \ifnum \j<\jj \ifnum \a=\aa
                            \node[fill=darkorchid, draw=darkorchid, minimum size=.95cm, inner sep=0pt, outer sep=0pt, rounded corners=1ex] at (\a*5+\j-5,16-\aa*5-\jj+5) {};
                        \fi\fi
                        \if \j\jj
                            \node[fill=darkorchid, draw=darkorchid, minimum size=.95cm, inner sep=0pt, outer sep=0pt, rounded corners=1ex] at (\a*5+\j-5,16-\aa*5-\jj+5) {};
                        \fi

                    }
                }
            }
        %}
    }
\end{tikzpicture}
\end{document}

### 2025-07-26 paper/camera/sects/abstract.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-05-05 paper/ecai2025/sects/abstract.tex
@@
 Reinforcement learning (RL) algorithms show promise in robotics and multi-agent systems but often suffer from low sample efficiency.
-While methods like SHAC leverage differentiable simulators to improve efficiency, they are limited to specific settings: they require fully differentiable environments (including transition and reward functions) and have primarily been demonstrated in single-agent scenarios.
-To overcome these limitations, we introduce \fname{}, a novel framework inspired by SHAC\@.
+While methods like SHAC leverage differentiable simulators to improve efficiency, they are limited to specific settings: they require fully differentiable environments, including transition and reward functions, and have primarily been demonstrated in single-agent scenarios.
+To overcome these limitations, we introduce \fname{}, a novel framework inspired by SHAC.
 \fname{} removes the need for differentiable simulator components by using neural networks to approximate the required gradients, training these networks alongside the standard policy and value networks.
-This allows the core SHAC approach to be applied in both non-differentiable and multi-agent environments.
+This enables the core SHAC approach to be applied in both non-differentiable and multi-agent environments.
 We evaluate \fname{} on challenging multi-agent tasks from the VMAS suite, comparing it against SHAC (where applicable) and PPO, a standard algorithm for non-differentiable settings.
@@
 Furthermore, in differentiable environments where SHAC operates, \fname{} achieves comparable performance despite lacking direct access to simulator gradients, thus successfully extending SHAC's benefits to a broader class of problems.
+The full implementation is openly available at
+\begin{center}
+    \url{https://github.com/f14-bertolotti/shacpp}.
+\end{center}
 \end{abstract}

### 2025-07-26 paper/camera/sects/acknowledgements.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\section*{Acknowledgements}

\begin{itemize}
    \item Computational resources provided by INDACO Platform, which is a project of High Performance Computing at the University of MILAN http://www.unimi.it
    \item This work was partly supported by the MUR project “T-LADIES” (PRIN2020TL3X8X).
    \item TODO (unibo)
\end{itemize}

### 2025-07-26 paper/camera/sects/background.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-05-05 paper/ecai2025/sects/background.tex
@@
 \subsection{Markov Decision Processes}
-To provide context for our work, we first introduce the reinforcement learning (RL) framework. 
+To provide context for our work, we first introduce the reinforcement learning framework. 
 Particularly, we focus on multi-agent settings, where multiple agents interact with each other and the environment.
@@
 
-While we use a unique set of parameters $\theta$ for all the networks.
-Each network can either use its own subset of $\theta$ or it can share parameters with other networks.
-In practice, the parameters will be shared only for the policies.
+Throughout this notation, we use a single global set of parameters $\theta$ to describe all networks.
+Each network can either use its own distinct subset of $\theta$ or share parameters with other networks.
+In practice, parameter sharing occurs only among the policy networks.
 
-For simplicity, we formally consider only the case where policies have access to the full state, meaning that $\stateobs{i}$ is the identity function.
-This allows us to focus on the fundamental aspects of our approach without the added complexity of partial observability.
-Extending our framework to handle partial observability is straightforward; however, we choose not to address it here to maintain clarity and simplicity in our presentation.
+To keep the notation simple, we formally consider only the case where policies have access to the full state, meaning that $\stateobs{i}$ is the identity function.
+Extending our framework to handle partial observability is straightforward.
 Nonetheless, our experiments do include scenarios where policies do not have access to the complete state information.
@@
 \subsection{SHAC}
+
+Short Horizon Actor-Critic (SHAC) is a model-based RL algorithm designed to leverage differentiable simulators for efficient policy learning.
+It alternates between rolling out short trajectories (a fixed small horizon h) in parallel and updating an actor–critic pair using gradients that flow directly through the simulator’s dynamics and reward model.
 
@@
 
-$G^t_k$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$.
+$G^k_t$ represents the discounted sum of rewards up to step $k$, plus the bootstrapped value of the state at step $t+k$.
 The estimates $V$ is computed by discounting longer (higher $k$) discounted sums, $G^k_t$.

### 2025-07-26 paper/camera/sects/conclusion.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\section{Conclusions \& Future Work}\label{sect:conclusion}

In this paper, we introduced \fname{}, an extension of SHAC, designed to provide a reinforcement learning solution that is sample-efficient, applicable in both single-agent and multi-agent settings, and capable of handling non-differentiable environments. We demonstrated that \fname{} preserves the core alignment with SHAC (\RQ{1}), consistently outperforms the state-of-the-art PPO algorithm in both single-agent and multi-agent tasks (\RQ{2}), and exhibits promising scalability in multi-agent environments, including the emergence of collaborative behaviors (\RQ{3}).

Despite these encouraging results, several avenues remain for future exploration. One key direction is scaling beyond the current limit of tens of agents. Another challenge lies in environments with extreme partial observability, such as the discovery scenario, where limited sensory input hinders learning. Expanding the evaluation to a broader set of environments—particularly those from the cooperative AI community~\cite{DBLP:journals/corr/abs-2211-13746} and adversarial settings—would further validate the robustness and versatility of \fname{}.

Addressing the cold-start problem more effectively could also enhance performance, particularly in environments with sparse or delayed rewards. Several promising techniques for mitigating this issue include:

\begin{itemize}
    \item \textbf{World model pre-training:} Leveraging datasets generated by alternative algorithms or human demonstrations to initialize the world model before policy training begins.
    \item \textbf{Model Predictive Path Integral (MPPI) control~\cite{Alvarez24}:} Using MPPI to generate high-quality initial trajectories that can seed the world model and guide early exploration.
\end{itemize}

The latter approach may also be integrated into \fname{} during training to periodically introduce diverse trajectories, potentially improving robustness and convergence speed.

We also note that linear attention mechanisms, such as Performer~\cite{Choromanski20} and Mamba~\cite{Gu23}, could be leveraged to improve the scalability of \fname{} in environments with a large number of agents, without incurring the high computational cost associated with training classical Transformer architectures~\cite{Vaswani17}.

### 2025-07-26 paper/camera/sects/experiments.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-05-04 paper/ecai2025/sects/experiments.tex
@@
-\begin{figure*}[t]
-    \centering
-    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
-    \caption{Performance comparison of \fname{}, PPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
-The results show the mean and standard deviation of rewards over 3 runs.
-For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
-    \label{fig:experiments}
-\end{figure*}
-
-\begin{figure}[t]
-    \centering
-    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
-    \caption{Policy gradient norms in the Transport environment over 5000 epochs.
-Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
-    \vspace{0.5cm}
-\end{figure}
 \section{Experiments}\label{sect:experiments}
 
-In this section, we evaluate \fname{} against both PPO and SHAC across various multi-agent scenarios and architectures.
+In this section, we evaluate \fname{} against both PPO (for single agent settings), MAPPO (for multi-agent settings), and SHAC across various multi-agent scenarios and architectures.
 An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
@@
 
-This selection of scenarios enables a thorough evaluation of how each algorithm handles different types of multi-agent cooperation challenges.
-
-\begin{table*}[t]
-    \centering
-    \input tables/max.tex
-    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
-\end{table*}
+\noindent This selection of scenarios enables a thorough evaluation of how each algorithm handles different types of multi-agent cooperation challenges.
 
@@
 See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.
+Hyperparameter choices are the result of a small grid search on the transport scenario.
 
@@
 \paragraph{Hardware Setup}
-Our experiments were conducted on a cluster of machines composed of 2 computational nodes, each with one NVIDIA Tesla V100 GPU (with 32GB of memory), 200GB of RAM, and two Intel Xeon Gold 6226R processors (32 cores each).
+All experiments were conducted on a cluster with two nodes, each equipped with a V100 GPU (32GB of memory) and 200GB of RAM, and one node with four A100 GPUs (40GB of memory) and 512GB of RAM.
 
-\subsection{Results}
 \begin{figure}
@@
     \caption{Snapshots of trained agents with \fname{} in the Transport scenario. Colored circles represent agents pushing a red package (square) toward the green goal.}\label{fig:transport-trained}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
+    \caption{Policy gradient norms in the Transport environment over 5000 epochs.
+Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
     \vspace{0.5cm}
 \end{figure}
-In this section, we compare the performance of \fname{}, PPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
+
+\subsection{Results}
+
+In this section, we compare the performance of \fname{}, PPO/MAPPO, and SHAC across different scenarios using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
 The results show the mean and standard deviation of achieved rewards over 3 runs.
@@
 Notably, SHAC does not exhibit the same stability advantage it displayed in other scenarios, possibly because gradients derived directly from the environment are noisier compared to those learned by the transition network.
+
+\begin{figure*}
+    \centering
+    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
+    \caption{Performance comparison of \fname{}, PPO/MAPPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
+The results show the mean and standard deviation of rewards over 3 runs.
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
+    \label{fig:experiments}
+\end{figure*}
+
+\begin{table*}
+    \centering
+    \input tables/max.tex
+    \vspace{0.2cm}
+    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
+\end{table*}

### 2025-07-26 paper/camera/sects/introduction.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-05-05 paper/ecai2025/sects/introduction.tex
@@
 \section{Introduction}\label{sect:introduction}
-Learning control policies is crucial for applications like robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, and swarm intelligence \cite{Tang21}.
-These domains often present challenges such as sparse rewards, complex dynamics, and multi-agent interactions.
+Learning control policies is crucial for applications such as robotics \cite{Singh22}, autonomous driving \cite{Elallid22}, and swarm intelligence \cite{Tang21}.
+These domains often present challenges including sparse rewards, complex dynamics, and multi-agent interactions.
 Many current RL approaches focus specifically on either single- or multi-agent settings, lacking a unified solution.
@@
 
-A promising direction to improve sample efficiency involves utilizing \emph{differentiable simulators}, namely simulators that allow for backpropagation through their components.
-SHAC \cite{Xu22} exemplifies this approach, using gradients obtained directly from the simulator to accelerate policy optimization.
-However, SHAC's applicability is constrained by two key limitations.
-First, it necessitates both a differentiable transition function \emph{and} a differentiable reward function, which is often impractical, especially with sparse or complex reward structures.
-Second, while powerful, backpropagation through complex dynamics, like object collisions~\cite{Georgiev24} or intricate multi-agent interactions, can lead to unstable gradients~\cite{Bengio94, Metz21}, potentially hindering learning.
+A promising direction to improve sample efficiency involves utilizing \emph{differentiable simulators}, namely simulators that allow backpropagation through their components.
+SHAC \cite{Xu22} exemplifies this approach by using gradients obtained directly from the simulator to accelerate policy optimization.
+However, SHAC’s applicability is constrained by two key limitations.
+First, it necessitates both a differentiable transition function \emph{and} a differentiable reward function, which is often impractical, especially in the presence of sparse or complex reward structures.
+Second, while powerful, backpropagation through complex dynamics, such as object collisions~\cite{Georgiev24} or intricate multi-agent interactions, can lead to unstable gradients~\cite{Bengio94, Metz21}, potentially hindering learning.
 The original SHAC work also focused primarily on single-agent tasks.
@@
 Our goal is to develop a sample-efficient reinforcement learning framework that maintains the benefits of gradient-based optimization without requiring differentiable simulators.
-Drawing inspiration from model-based RL \cite{DBLP:conf/icml/HafnerLFVHLD19, DBLP:conf/iclr/HafnerLB020}, \fname{} substitutes direct simulator differentiation with neural approximations of the gradient functions.
-By training networks that approximate transition and reward gradients concurrently with policy optimization, our approach preserves the sample efficiency benefits of gradient information while eliminating strict differentiability requirements.
-This design enhances robustness to complex dynamics and enables application in both non-differentiable environments and multi-agent scenarios, albeit with additional computational overhead compared to SHAC.
+Drawing inspiration from model-based RL \cite{DBLP:conf/icml/HafnerLFVHLD19, DBLP:conf/iclr/HafnerLB020}, \fname{} substitutes direct simulator differentiation with neural approximations by training networks that approximate the transition and reward functions concurrently with the policy function.
+This approach preserves the benefits of gradient information while eliminating strict differentiability requirements.
+It also enhances robustness to complex dynamics and enables application in both non-differentiable environments and multi-agent scenarios, albeit with additional computational overhead compared to SHAC.
 
-We evaluate \fname{} against two established baselines: PPO \cite{Schulman17} (including its multi-agent variant MAPPO \cite{DBLP:conf/nips/YuVVGWBW22}) as the standard for non-differentiable environments, and the original SHAC \cite{Xu22} where applicable.
+We evaluate \fname{} against two established baselines: PPO \cite{Schulman17} (including its multi-agent variant MAPPO \cite{DBLP:conf/nips/YuVVGWBW22}), as the standard for non-differentiable environments, and the original SHAC \cite{Xu22}, where applicable.
 Our experiments utilize the VMAS simulator \cite{DBLP:conf/dars/BettiniKBP22}, which provides physics-based multi-agent environments with varying levels of complexity.
@@
     \item \RQ{1} Can we train a neural network to approximate the gradients of a differentiable simulator?
-    \item \RQ{2} How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?
+    \item \RQ{2} How does our algorithm compare to PPO/MAPPO and SHAC in both single-agent and multi-agent settings?
     \item \RQ{3} How does the performance of these algorithms change as the search space increases?

### 2025-07-26 paper/camera/sects/limitations.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

\section{Limitations}

While the proposed method shows promise, it is not a one-size-fits-all solution. Environments vary significantly in complexity and dynamics, and several limitations must be acknowledged.

A central challenge lies in learning the world model, which depends on rich, meaningful interactions between agents and the environment. These interactions often yield sparse or delayed rewards, leading to a cold-start problem: until the world model is sufficiently trained, agents may struggle to explore or act effectively, even when guided by stochastic policies. This issue can be partially addressed by seeding the learning process with a cold-start dataset—either manually designed or generated using alternative algorithms such as PPO or MPPI~\cite{Alvarez24}. Such data helps bootstrap the world model, providing a more stable foundation for training.

Despite careful attention to implementation, including reference to best practices (e.g., the well-known "37 Implementation Details of Proximal Policy Optimization" blog~\cite{Shengyi22}), we observed that PPO and MAPPO performed suboptimally in our experiments. To ensure the validity of these findings, we cross-verified results using a third-party implementation~\cite{Bettini24}, which yielded consistent outcomes.

Although early stopping is primarily used to conserve computational resources, it can give the false impression that the algorithm is inherently robust to policy collapse. In reality, we observed multiple instances of partial policy collapse—some of which recovered—and a few cases of complete collapse without recovery. One example of partial collapse appears in the five-agent variant of the dispersion scenario shown in \Cref{apx:fig:experiments-mlp}.

The discovery scenario itself poses considerable difficulty for all tested algorithms due to its inherent partial observability. Agents rely solely on a simulated LiDAR system for perception, which provides a narrow, local view of the environment. This makes it particularly difficult for any method—including \fname{}—to construct a complete and accurate world model, often resulting in suboptimal performance. Ultimately, this may indicate that partial observability may be challenging to overcome with the current approach.

\fname{} addresses the limitations of SHAC by introducing additional training components, specifically the transition and reward networks. However, this improvement comes at the cost of increased computational overhead. The training burden is especially pronounced in environments with a large number of agents, where the transition network must capture complex inter-agent dynamics. While this trade-off is acceptable in many scenarios, it significantly slows down the training process and demands computational resources that may not be readily available in all settings.

Regarding interpretability, \fname{} does not offer significant advantages over its predecessors. As the policy is governed by a neural network, understanding or explaining its behavior remains a challenging task. This limitation is inherent to most deep reinforcement learning methods and highlights the ongoing need for techniques that enhance policy transparency and interpretability.

### 2025-07-26 paper/camera/sects/related.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-04-26 paper/ecai2025/sects/related.tex
@@
 Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks.
-Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{DBLP:conf/aaai/HesselMHSODHPAS18})
+Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{Hessel18})
 and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.
 Over the years, several algorithms have been proposed to create more stable and efficient learning methods.
-Notable examples include A3C for asynchronous learning~\cite{DBLP:conf/icml/MnihBMGLHSK16},
+Notable examples include A3C for asynchronous learning~\cite{Mnih16},
 DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17} (also used in modern chatbot like ChatGPT),
@@
 QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination in cooperative multi-agent settings,
-and Mean Field MARL~\cite{DBLP:conf/icml/YangLLZZW18} for handling a large number of agents.
+and Mean Field MARL~\cite{Yang18} for handling a large number of agents.
 

### 2025-07-26 paper/camera/sects/shacpp.tex
camera
Co-authored-by: Francesc bertolotti <f14.bertolotti@gmail.com>

--- 2025-05-05 paper/ecai2025/sects/shacpp.tex
@@
 \end{align*}
-Where $a_{t*}=\{a_{ti}\}_{i\in\agents}$.
+Where $a_{t*}$ denotes the set of actions taken from all agents at time $t$, $\{a_{ti}\}_{i\in\agents}$.
 In practice, this is achieved by caching trajectories and then training the neural network to minimize the MSE loss.
@@
 Further, we modify the SHAC policy loss, $\mathcal{L}_\theta^{\pi}$, to include a term that penalizes actions outside the action space.
-Suppose the action space is $(l,r)$ (with $l<r$) for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$
+Suppose the action space is the open interval $(l,r)$ for all agents, we minimize the loss $\mathcal{L'}_\theta^\pi:$
 
@@
 
-    Where $\alpha$ is a hyper-parameter that controls the strength of the penalty.
-In our experiments, we set $\alpha=1$.
-Usually, actions are either clipped to the allowed action space or, an activation function (such as hyperbolic tangent or sigmoid) is used to ensure that the actions are within the allowed range.
-While this approach is acceptable for algorithms like PPO and SHAC, the penalty we employ is crucial in our specific scenario.
-This necessity arises from the fact that we are training the policy to optimize two neural networks that serve as environment approximations.
-Because these networks are never trained with data outside the permissible action space, they are likely to be maximized beyond these bounds.
-Consequently, the policy may generate actions far outside the allowed range.
-This can ultimately result in 0-gradients when action clipping is applied or high pre-activation actions when activation function is used.
+    Here, $\alpha$ is a hyperparameter that controls the strength of the penalty, and in our experiments we set $\alpha=1$.
+Typically, actions are either clipped to the allowed action space or constrained via activation functions such as hyperbolic tangent or sigmoid to ensure they remain within valid bounds.
+While these approaches are sufficient for algorithms like PPO and SHAC, in our case the penalty is crucial.
+This is because our policy is trained to optimize two neural networks that approximate the environment dynamics and reward.
+Since these networks are only trained on data within the valid action space, their outputs are unreliable for out-of-bound actions.
+As a result, the policy may exploit these approximations by generating actions far outside the allowed range.
+Action clipping may lead to zero gradients, while activation functions may results large pre-activation values.
 
@@
 Additionally, to minimize the time spent between swapping among the different training procedures (policy, value, reward, transition), we apply a cool-down of $K^V$/$K^R$/$K^F$ episodes between running training procedure for the value/reward/transition networks.
-
-The full implementation is openly available at \url{https://anonymous.4open.science/r/diffagents-8653}\footnote{The link of the original repository will be provided upon acceptance.}.

### 2025-07-26 paper/camera/locals2table.py

@@ -1,0 +1,123 @@
+import click
+import json
+import os
+
+def walk_dict(d):
+    for key, value in d.items():
+        if type(value) == dict:
+            for subkey, subvalue in walk_dict(value):
+                yield key + "." + subkey, subvalue
+        else:
+            yield key, value
+transformer_rename = {
+    "train_envs"                         : "number of training environments"      ,
+    "train_steps"                        : "training horizon"                     ,
+    "eval_envs"                          : "number of evaluation environments"              ,
+    "eval_steps"                         : "evaluation horizon"                   ,
+    "policy_layers"                      : "policy layers"                        ,
+    "policy_hidden_size"                 : "policy hidden size"                   ,
+    "policy_feedforward"                 : "policy feedforward size"              ,
+    "policy_heads"                       : "policy heads"                         ,
+    "policy_dropout"                     : "policy dropout"                       ,
+    "policy_activation"                  : "policy activation"                    ,
+    "policy_var"                         : "policy variance"                      ,
+    "value_layers"                       : "value layers"                         ,
+    "value_hidden_size"                  : "value hidden size"                    ,
+    "value_feedforward"                  : "value feedforward size"               ,
+    "value_dropout"                      : "value dropout"                        ,
+    "value_activation"                   : "value activation"                     ,
+    "reward_layers"                      : "reward layers"                        ,
+    "reward_hidden_size"                 : "reward hidden size"                   ,
+    "reward_feedforward"                 : "reward feedforward size"              ,
+    "reward_dropout"                     : "reward dropout"                       ,
+    "reward_activation"                  : "reward activation"                    ,
+    "world_layers"                       : "world layers"                         ,
+    "world_hidden_size"                  : "world hidden size"                    ,
+    "world_feedforward"                  : "world feedforward size"               ,
+    "world_dropout"                      : "world dropout"                        ,
+    "world_activation"                   : "world activation"                     ,
+    "policy_learning_rate"               : "policy learning rate"                 ,
+    "value_learning_rate"                : "value learning rate"                  ,
+    "reward_learning_rate"               : "reward learning rate"                 ,
+    "policy_clip_coefficient"            : "policy clip coefficient"              ,
[…]

### 2025-08-04 paper/camera/sects/conclusion.tex
to 7 pages for contribution

@@
 
-Despite these encouraging results, several avenues remain for future exploration.
-One key direction is scaling beyond the current limit of tens of agents.
-Another challenge lies in environments with extreme partial observability, such as the discovery scenario, where limited sensory input hinders learning.
-Expanding the evaluation to a broader set of environments—particularly those from the cooperative AI community~\cite{DBLP:journals/corr/abs-2211-13746} and adversarial settings—would further validate the robustness and versatility of \fname{}.
+Despite these encouraging results, several avenues remain for future exploration: scaling beyond tens of agents, handling extreme partial observability (e.g., discovery scenarios with limited sensory input), and expanding evaluation to broader environments---particularly from the cooperative AI community~\cite{DBLP:journals/corr/abs-2211-13746} and adversarial settings—to further validate \fname{}'s robustness and versatility.
 
@@
 Several promising techniques for mitigating this issue include:
-
-\begin{itemize}
-    \item \textbf{World model pre-training:} Leveraging datasets generated by alternative algorithms or human demonstrations to initialize the world model before policy training begins.
-    \item \textbf{Model Predictive Path Integral (MPPI) control~\cite{Alvarez24}:} Using MPPI to generate high-quality initial trajectories that can seed the world model and guide early exploration.
-\end{itemize}
-
+\begin{enumerate*}[label=\emph{\roman*})]
+    \item leveraging datasets generated by alternative algorithms or human demonstrations to initialize the world model before policy training begins (world-model pre-training).
+    \item using Model Predictive Path Integral (MPPI)~\cite{Alvarez24}  to generate high-quality initial trajectories that can seed the world model and guide early exploration.
+\end{enumerate*}
+%
 The latter approach may also be integrated into \fname{} during training to periodically introduce diverse trajectories, potentially improving robustness and convergence speed.

### 2025-08-04 paper/camera/sects/experiments.tex
to 7 pages for contribution

@@
     \centering
-    \includegraphics[width=0.9\columnwidth]{figs/grads-transformer-transport.pdf}
+    \includegraphics[width=0.7\columnwidth]{figs/grads-transformer-transport.pdf}
     \caption{Policy gradient norms in the Transport environment over 5000 epochs.
@@
 
-\begin{figure*}
+\begin{figure*}[t]
     \centering
-    \includegraphics[width=0.9\textwidth]{figs/main-transformer.pdf}
+    \includegraphics[width=0.75\textwidth]{figs/main-transformer.pdf}
     \caption{Performance comparison of \fname{}, PPO/MAPPO, and SHAC across different scenarios (Dispersion, Transport, Discovery, and Sampling) using the Transformer architecture for multi-agent settings and the MLP architecture for single-agent settings.
@@
 
-\begin{table*}
+\begin{table}[t]
     \centering
@@
     \vspace{0.2cm}
-    \caption{Normalized rewards (relative to the best performing model) for the different scenarios. Best results are in bold.}\label{tab:max-rewards}
-\end{table*}
+    \caption{Normalized rewards (relative to the best performing model) for the different scenarios.
+Best results are in bold.
+\textsc{D} stands for Dispersion, \textsc{T} for Transport, \textsc{Di} for Discovery, and \textsc{S} for Sampling.}\label{tab:max-rewards}.
+\end{table}

### 2025-08-04 paper/camera/sects/limitations.tex
to 7 pages for contribution

@@
 
-A central challenge lies in learning the world model, which depends on rich, meaningful interactions between agents and the environment.
-These interactions often yield sparse or delayed rewards, leading to a cold-start problem: until the world model is sufficiently trained, agents may struggle to explore or act effectively, even when guided by stochastic policies.
-This issue can be partially addressed by seeding the learning process with a cold-start dataset—either manually designed or generated using alternative algorithms such as PPO or MPPI~\cite{Alvarez24}.
-Such data helps bootstrap the world model, providing a more stable foundation for training.
+\paragraph{World Model Learning and Cold-Start Problem}
+A central challenge lies in learning the world model, which depends on rich agent-environment interactions.
+Sparse or delayed rewards create a cold-start problem: until the world model is sufficiently trained, agents struggle to explore effectively.
+This can be addressed by seeding with a cold-start dataset---manually designed or generated using algorithms like PPO or MPPI~\cite{Alvarez24}---to bootstrap the world model and provide a stable training foundation.
 
-Despite careful attention to implementation, including reference to best practices (e.g., the well-known "37 Implementation Details of Proximal Policy Optimization" blog~\cite{Shengyi22}), we observed that PPO and MAPPO performed suboptimally in our experiments.
+\paragraph{Baseline Performance and Implementation}
+Despite careful attention to implementation, including reference to best practices (e.g., the well-known ``37 Implementation Details of Proximal Policy Optimization'' blog~\cite{Shengyi22}), we observed that PPO and MAPPO performed suboptimally in our experiments.
 To ensure the validity of these findings, we cross-verified results using a third-party implementation~\cite{Bettini24}, which yielded consistent outcomes.
 
+\paragraph{Policy Collapse and Robustness}
 Although early stopping is primarily used to conserve computational resources, it can give the false impression that the algorithm is inherently robust to policy collapse.
-In reality, we observed multiple instances of partial policy collapse—some of which recovered—and a few cases of complete collapse without recovery.
+In reality, we observed multiple instances of partial policy collapse---some of which recovered---and a few cases of complete collapse without recovery.
 One example of partial collapse appears in the five-agent variant of the dispersion scenario shown in \Cref{apx:fig:experiments-mlp}.
 
-The discovery scenario itself poses considerable difficulty for all tested algorithms due to its inherent partial observability.
-Agents rely solely on a simulated LiDAR system for perception, which provides a narrow, local view of the environment.
-This makes it particularly difficult for any method—including \fname{}—to construct a complete and accurate world model, often resulting in suboptimal performance.
-Ultimately, this may indicate that partial observability may be challenging to overcome with the current approach.
+\paragraph{Partial Observability Challenges}
+The discovery scenario poses considerable difficulty for all algorithms due to inherent partial observability.
+Agents rely solely on simulated LiDAR for perception, providing only a narrow, local view.
+This makes it difficult for any method---including \fname{}---to construct accurate world models, often resulting in suboptimal performance and suggesting that partial observability remains challenging for the current approach.
 
-\fname{} addresses the limitations of SHAC by introducing additional training components, specifically the transition and reward networks.
-However, this improvement comes at the cost of increased computational overhead.
-The training burden is especially pronounced in environments with a large number of agents, where the transition network must capture complex inter-agent dynamics.
-While this trade-off is acceptable in many scenarios, it significantly slows down the training process and demands computational resources that may not be readily available in all settings.
+\paragraph{Computational Overhead}
+\fname{} addresses SHAC's limitations by adding transition and reward networks, but this increases computational overhead.
+Training burden is especially pronounced with many agents, where the transition network must capture complex inter-agent dynamics.
+While this trade-off is often acceptable, it significantly slows training and demands substantial computational resources.
 
+\paragraph{Interpretability}
 Regarding interpretability, \fname{} does not offer significant advantages over its predecessors.

### 2025-08-04 paper/camera/sects/conclusion.tex
compact more

@@
 
-We also note that linear attention mechanisms, such as Performer~\cite{Choromanski20} and Mamba~\cite{Gu23}, could be leveraged to improve the scalability of \fname{} in environments with a large number of agents, without incurring the high computational cost associated with training classical Transformer architectures~\cite{Vaswani17}.
+Finally, linear attention mechanisms like Performer~\cite{Choromanski20} and Mamba~\cite{Gu23} could improve \fname{}'s scalability in large multi-agent environments while avoiding the computational overhead of classical Transformers~\cite{Vaswani17}.

### 2025-08-04 paper/camera/sects/discussion.tex
compact more

--- 2025-05-06 paper/ecai2025/sects/discussion.tex
@@
 \subsection{Gradient Approximation with Neural Networks}
-\begin{rqbox}
-\textbf{Can we train a neural network to approximate the gradients of a differentiable simulator?}
-\end{rqbox}
+%\begin{rqbox}
+\textbf{RQ:} \emph{Can we train a neural network to approximate the gradients of a differentiable simulator?}
+%\end{rqbox}
 
@@
 
-\begin{replybox}
-Yes, \fname{} approximates simulator gradients, achieving comparable performance to SHAC in differentiable settings, occasionally outperforming it despite initially higher variance.
-\end{replybox}
+%\begin{replybox}
+%Yes, \fname{} approximates simulator gradients, achieving comparable performance to SHAC in differentiable settings, occasionally outperforming it despite initially higher variance.
+%\end{replybox}
 
 \subsection{Comparison to PPO and SHAC}
-\begin{rqbox}
-\textbf{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
-\end{rqbox}
+%\begin{rqbox}
+\textbf{RQ:} \emph{How does our algorithm compare to PPO and SHAC in both single-agent and multi-agent settings?}
+%\end{rqbox}
 
@@
 
-\begin{replybox}
-\fname{} outperforms PPO across all scenarios, with PPO struggling especially in multi-agent settings.
+%\begin{replybox}
+%\fname{} outperforms PPO across all scenarios, with PPO struggling especially in multi-agent settings.
 When compared to SHAC (where applicable), \fname{} achieves comparable performance while enabling emergent cooperative behaviors.
-\end{replybox}
+%\end{replybox}
 
 \subsection{Scaling with Search Space}
-\begin{rqbox}
-\textbf{How does the performance of these algorithms change as the search space increases?}
-\end{rqbox}
+%\begin{rqbox}
+\textbf{RQ:} \emph{How does the performance of these algorithms change as the search space increases?}
+%\end{rqbox}
 
@@
 
-\begin{replybox}
-Performance varies with scale: Transport/Sampling converge faster with more agents, while Dispersion becomes harder.
-Cooperation emerges across scenarios.
-Main scalability limitation is transformer attention complexity, becoming prohibitive at large agent counts ($~$50), though moderate populations remain feasible.
-\end{replybox}
+%\begin{replybox}
+%Performance varies with scale: Transport/Sampling improve with more agents, while Dispersion degrades. Cooperation emerges consistently. Transformer attention limits scalability beyond ~50 agents.
+%\end{replybox}

### 2025-08-04 paper/camera/sects/related.tex
compact more

@@
 \paragraph{Deep Reinforcement Learning}
-Deep reinforcement learning has revolutionized policy learning tasks by combining reinforcement learning with deep neural networks.
-Foundational works like DQN~\cite{Mnih13} (and its variants~\cite{Hessel18})
-and AlphaZero~\cite{Silver17} demonstrated human-level performance in complex environments such as Atari games and chess.
-Over the years, several algorithms have been proposed to create more stable and efficient learning methods.
-Notable examples include A3C for asynchronous learning~\cite{Mnih16},
-DDPG and PPO for continuous action spaces~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17} (also used in modern chatbot like ChatGPT),
-and SAC for continuous action spaces with off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}.
-Following the initial era of model-free algorithms,
-model-based algorithms have been introduced to enhance sample efficiency.
-Early works such as Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} have demonstrated promising results in terms of both sample efficiency and performance.
+Reinforcement learning has achieved significant advancements through the integration of deep neural networks.
+Foundational works such as DQN~\cite{Mnih13} and its variants~\cite{Hessel18}
+and AlphaZero~\cite{Silver17} achieved human-level performance in complex environments.
+Subsequent algorithms improved efficiency, including A3C~\cite{Mnih16}, 
+DDPG and PPO for continuous control~\cite{DBLP:journals/corr/LillicrapHPHETS15,Schulman17}, 
+and SAC for off-policy learning~\cite{DBLP:conf/icml/HaarnojaZAL18}.
+Model-based approaches like Dreamer~\cite{DBLP:conf/iclr/HafnerLB020} and PlaNet~\cite{DBLP:conf/icml/HafnerLFVHLD19} 
+later enhanced sample efficiency.
 
@@
 traffic control~\cite{DBLP:journals/tits/ChuWCL20},
-and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20,jaderberg2019human},
+and game playing~\cite{DBLP:conf/iclr/BakerKMWPMM20},
 and involves different types of agents such as cooperative, competitive, and mixed.
-In this paper, we primarily focus on cooperative multi-agent scenarios,
-where agents need to collaborate to achieve a common goal.
-Prominent works in this area include extensions of single-agent algorithms to multi-agent settings,
-such as MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17} and MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22},
-as well as novel solutions to unique challenges in multi-agent settings. Examples of these solutions are COMA~\cite{DBLP:conf/aaai/FoersterFANW18} for multi-agent credit assignment,
-QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination in cooperative multi-agent settings,
-and Mean Field MARL~\cite{Yang18} for handling a large number of agents.
+In this paper, we primarily focus on cooperative multi-agent scenarios where agents collaborate to achieve common goals.
+Prominent works include extensions of single-agent algorithms like MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17} and MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22},
+as well as novel solutions such as COMA~\cite{DBLP:conf/aaai/FoersterFANW18} for credit assignment,
+QMIX~\cite{DBLP:conf/icml/RashidSWFFW18} for coordination,
+and Mean Field MARL~\cite{Yang18} for large-scale scenarios.
 

### 2025-08-05 paper/camera/sects/experiments.tex
compact more
ready to submit

@@
 An ablation study replacing only the transition function (SHAC+, where the reward function remains differentiable) is provided in \Cref{apx:ablation}.
-We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC \cite{Haarnoja2018SoftAA}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}).
+We select PPO as our baseline because, despite the emergence of more recent algorithms HAPPO \cite{Kuba2021TrustRM}, and SAC~\cite{DBLP:conf/icml/HaarnojaZAL18}, it remains one of the few approaches that scales effectively to multi-agent settings without substantial modifications (e.g., MAPPO~\cite{DBLP:conf/nips/YuVVGWBW22}).
 This makes PPO a representative benchmark for state-of-the-art performance in non-differentiable environments.
@@
     \centering
-    \includegraphics[width=0.7\columnwidth]{figs/grads-transformer-transport.pdf}
+    \includegraphics[width=0.65\columnwidth]{figs/grads-transformer-transport.pdf}
     \caption{Policy gradient norms in the Transport environment over 5000 epochs.
 Spikes in gradient norm for SHAC (epoch 2700) and \fname{} (epoch 3200) correspond to generalization events when agents begin frequent interactions with the package and each other.}\label{fig:grads-transformer-transport}
-    \vspace{0.5cm}
+    \vspace{0.1cm}
 \end{figure}
@@
     \label{fig:experiments}
+    \vspace{0.1cm}
 \end{figure*}
@@
     \input tables/max.tex
-    \vspace{0.2cm}
+    \vspace{0.1cm}
     \caption{Normalized rewards (relative to the best performing model) for the different scenarios.

### 2025-08-05 paper/camera/sects/limitations.tex
ready to submit

@@
 \paragraph{World Model Learning and Cold-Start Problem}
-A central challenge lies in learning the world model, which depends on rich agent-environment interactions.
-Sparse or delayed rewards create a cold-start problem: until the world model is sufficiently trained, agents struggle to explore effectively.
-This can be addressed by seeding with a cold-start dataset---manually designed or generated using algorithms like PPO or MPPI~\cite{Alvarez24}---to bootstrap the world model and provide a stable training foundation.
+A central challenge lies in learning the world model from agent-environment interactions.
+Sparse or delayed rewards create a cold-start problem: until sufficiently trained, agents struggle to explore effectively.
+This can be addressed by seeding with a cold-start dataset---manually designed or generated using algorithms like PPO or MPPI~\cite{Alvarez24}---to bootstrap the world model and provide stable training foundation.
 
@@
 \paragraph{Computational Overhead}
-\fname{} addresses SHAC's limitations by adding transition and reward networks, but this increases computational overhead.
+\fname{} addresses SHAC's limitations by adding transition and reward networks, increasing computational overhead.
 Training burden is especially pronounced with many agents, where the transition network must capture complex inter-agent dynamics.
-While this trade-off is often acceptable, it significantly slows training and demands substantial computational resources.
+While often acceptable, this significantly slows training and demands substantial resources.
 

### 2025-08-13 paper/camera/sects/discussion.tex
prepare for camera ready

@@
 This scaling advantage is not observed in the Dispersion scenario, where complexity increases quadratically with agent count.
-Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (quantitative analysis with up to 20 agents is provided in \Cref{apx:experiments}).
+Nevertheless, we observe consistent emergence of cooperative behavior patterns across all experimental scenarios, suggesting potential scalability to larger agent populations (see \Cref{apx:experiments} for quantitative analysis with up to 20 agents in appendix~\cite{appendix}).
 The primary computational constraint lies in the transformer architecture employed for the action world model, where sequence length scales linearly with the number of agents (specifically, sequence length = $n \times s$, where $n$ is the number of agents and $s$ is the number of training steps).

### 2025-08-13 paper/camera/sects/experiments.tex
prepare for camera ready

@@
 We evaluate our algorithms on several multi-agent scenarios from VMAS~\cite{DBLP:conf/dars/BettiniKBP22}, which we selected over alternatives like PettingZoo~\cite{DBLP:conf/icml/PettingZoo}, SMAC~\cite{DBLP:conf/icml/ZhangZLZ20}, MA-MuJoCo~\cite{DBLP:conf/aaai/GuptaK0G20}, and others due to its key advantages: differentiable physics (enabling gradient-based optimization), multi-agent design, vectorized implementation (for efficient parallel training), realistic physics-based simulation, and open-source accessibility.
-For more details about the choice of environments, see \Cref{apx:scenarios}.
-We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} for details).
+For more details about the choice of environments, see \Cref{apx:scenarios} in appendix~\cite{appendix}.
+We selected four distinct scenarios: Dispersion, Discovery, Transport, and Sampling, specifically for their cooperative nature and varying levels of required coordination (see \Cref{fig:scenarios} and \Cref{apx:scenarios} in appendix~\cite{appendix} for details).
 These scenarios can be categorized along two dimensions:
@@
 We apply the transformer architecture only if the number of agents is greater than $1$, otherwise we use only the MLP architecture.
-For more details on the architectures, see \Cref{apx:arch}.
+For more details on the architectures, see \Cref{apx:arch} in appendix~\cite{appendix}.
 
@@
 We repeat and report results for $3$ runs with different seeds.
-See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} for a complete list of hyperparameters.
+See \Cref{apx:tab:ppo} and \Cref{apx:tab:shac} in appendix~\cite{appendix} for a complete list of hyperparameters.
 Hyperparameter choices are the result of a small grid search on the transport scenario.
@@
 An overview of the results is shown in \Cref{fig:experiments} and \Cref{tab:max-rewards}.
-For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}.
 
@@
 In \Cref{fig:experiments}, the first row displays results for the dispersion scenario (non-differentiable, thus SHAC is not applicable).
-In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation.
+In the first column (single agent, see also \Cref{tab:max-rewards} and \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}), both PPO and \fname{} perform well, though \fname{} converges faster thanks to gradient approximation.
 In the second column (3 agents), PPO fails to converge, whereas \fname{} converges in fewer than 10,000 episodes, suggesting emergent cooperative behavior.
@@
 The results show the mean and standard deviation of rewards over 3 runs.
-For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp}.}%
+For results using the MLP architecture in all scenarios, refer to \Cref{apx:fig:experiments-mlp} in appendix~\cite{appendix}.}%
     \label{fig:experiments}

### 2025-08-17 paper/camera/sects/acknowledgements.tex
to 7 pages for contribution
ready to submit
separated appendix from the main paper

@@
     \item Computational resources provided by INDACO Platform, which is a project of High Performance Computing at the University of MILAN http://www.unimi.it
-    \item This work was partly supported by the MUR project “T-LADIES” (PRIN2020TL3X8X).
-    \item TODO (unibo)
+    \item This work was partly supported by the MUR project ``T-LADIES'' (PRIN2020TL3X8X).
+    \item This work was partly supported by the Italian PRIN project ``CommonWears'' (2020 HCWWLP).
+    \item Computational cloud resources were provided by the Gemma 2 Google Research Award 2024.
+    %\item TODO (unibo)
 \end{itemize}
